---
title: "Bent Fall 2016"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Fall 2016


## Problem 1

    Solve: ONE / FIVE = .TWO with TWO being even.
    Different letters are different digits and
    there are no leading zeros.

    -- Journal of Recreational Mathematics

Solution: solved in the cryptic arithmetic notebook

    Cryptic fraction:

      ONE / FIVE = . TWO

      with TWO being even.
      Note the decimal point in the equation!

    a=make_crypt_df('onefivetwo')
    k=(make_num(a,'one')*1000 == make_num(a,'five')*make_num(a,'two'))
    k=k&(a$o!=0)&(a$f!=0)&(a$t!=0)
    sum(k)
    [1] 2
    make_num(a,'two')[k,]
    [1] 264 375
    a[k,]
           o n e f i v t w
    897834 4 9 5 1 8 7 2 6
    963122 5 2 8 1 4 0 3 7

    Note the first solution has TWO = 264 (even) as requested.


## Problem 2

    John is shopping for a used car and has found
    one being offered by Alice for $12,000, based
    on $2,000 down and a $10,000 balance to be
    paid to Alice over 36 months at 6% (annual)
    simple interest paid monthly on the unpaid
    balance (John's monthly payments would be
    $10,000/36 + unpaid balance x 0.06/12).
    However, if he pays cash, he can get a
    reduction of D dollars so the cash payment
    would be $12,000 - D. Although Alice will
    finance the car at 6%, she believes she can
    better invest her money at 10% elsewhere. If
    the 36 monthly payments are discounted to the
    present at a rate of 10%, how big a reduction
    in the cash price (to the nearest dollar) can
    Alice afford to give John so that buying the
    car for a single cash payment and the net
    present value of financing over 36 months
    (plus $2,000 added to account for the down
    payment), are equivalent propositions? Assume
    all months have 30 days, and a year is 360
    days.

    -- John W. Langhaar, PA A'33

TODO

    Some kind of sinking fund at the difference in
    interest rates.
    
    They made a spreadsheet with 36 rows.  The start
    of the spreadsheet is shown here:
    
    Month Loan       Int.  Princ.  Payment Factor       DCF
    --------------------------------------------------------
    0     10000.00  6.0%  2000.00  2000.00 10.0%     2000.00
    1      9722.22 50.00   277.78   327.78 1.0083     325.07
    2      9444.44 48.61   277.78   326.39 1.0167     321.02
    ...
    35      277.70  2.78   277.78   280.56 1.3370    209.84
    36        0.00  1.39   277.70   279.09 1.3482    207.01
    Totals:       925.00 12000.00 12925.00         11443.45

    The first value in the Int. column is just the interest rate.
    The first value in the Factor column is the investment return rate.
    
    The loan is 10000 with 6% interest compounded annually so the
    interest for the next month is the product of the principal*(1+0.06/12).
    The principal payment is just 10000/36.  The total payment is the sum
    of the principal and interest payments.  The loan column is then reduced
    by the amount of the principal payment.  The "factor" column is the
    net present value assuming she can get 10% on her investment.
    Specifically, the Factor column = (1+0.1/12)^month.
    
    The final column (DCF) is the discounted value of the payment in
    terms of today's money (e.g. the NPV of the payment).
    
    The final conclusion is that a fair reduction in price is the difference
    between the original purchase price (12000) and the NPV of the cash
    flow that she is anticipating(11443.35).  This difference is
    just:
    
        12000-11443.35 = 556.55
        
    I am not sure I agree that this is "fair" but it is the answer they
    were looking for.
    

## Problem 3

    Algernon, Basil and Clarence are either
    English or Irish, either Conservative or
    Liberal, and either Protestant or Catholic.
    In each case there is, as they all know, at
    least one of each.  None of them know which
    the others are, but Algernon and Basil have
    both been told that if Clarence is an Irish
    Catholic he cannot be a Liberal.  Basil asks
    Algernon whether he is a Conservative or a
    Liberal, and whether he is English or Irish.
    Algernon tells him.  After a pause for
    reflection, Basil is able to announce
    correctly the full particulars (English or
    Irish, Protestant or Catholic, Conservative
    or Liberal) of both Algernon and Clarence.
    What are the particulars for the three men?

    -- Brain Puzzler's Delight by E.R. Emmet

TODO

    meh


## Problem 4

    Our library has several copies of
    Boremaster's commentary on Hegel. It is not
    exactly a jolly read, as you well know if you
    have ever waded through its 36 chapters, but
    is much in demand on the grounds that it is
    less painful than Hegel himself. Even so, I
    was surprised to see my friend Jones leaving
    the library with three copies under his arm.
    "Steady on, old bean," I exclaimed, "there
    are other readers to think of." "The other
    copies are all on the shelf," he replied
    airily, "but I had to take three to get a
    complete text. Some rotter has snipped whole
    chapters out of every copy." "Well, surely
    two copies would have done." "No, no two
    copies would yield a full text." "Do you mean
    that I shall have to check every copy if I
    want to be sure of a full text?" "Oh no. Just
    take any three at random, as I did. You are
    bound to get a full text, even though no
    chapter is present in all copies. For each
    pair of chapters there is at least one copy
    with only one of them." For this to be true,
    what is the minimum number of copies that the
    library have in total? Also, provide an N row
    by 36 column matrix, where each row
    represents a copy of the book, each column
    represents a chapter, and each element of the
    matrix is either 1 if the chapter is present
    in the indicated copy or 0 if the chapter is
    absent.

    -- a Tantalizer by Martin Hollis in New
    Scientist

TODO

    This reminds me of the locks at the bank.
    Like that problem -- I don't care.


## Problem 5

    A hexagon inscribed in a circle has side
    lengths of 2, 2, 2, 11, 11, and 11. What is
    the exact radius of the circle?
    
    -- Puzzle Corner in Technology Review
    
Solution:

    I am fairly certain I can solve this numerically.
    Assume a circle of radius=1 and have two unknown
    for the sides (or central angles).  The contraints
    are:
    
        r = 1
        a/b = 11/2
        a/2 = r*sin(alpha) --> alpha = asin(a/(2*r))
        b/2 = r*sin(beta)  --> beta  = asin(b/(2*r))
        
        3*2*alpha+3*2*beta = 2*pi

    I don't know that this will be exact since it is numeric
    but I will try it anyhow.
    
```{r}

foo = function(x) {
    b=x
    a=b*11/2
    alpha=asin(a/2)   # Half-angle of the wide pie slice
    beta =asin(b/2)   # Half-angle of the narrow pie slice
    return(abs(6*(alpha+beta)-2*pi))
}

optimize(foo,c(0.01,4/11))

```

The numerical solution yields:

    b = 0.2857004

In the original problem statement, however, b=2 so the circle must
be scaled up by the same factor:

    r = 2/0.2857004 = 7.00034
    
I suspect the "exact" radius is 7.

The Bent did this analytically but they get the same answer.
My solution was actually much easier, albeit less exact.


## Bonus

    Sally's Spa has installed a security system
    consisting of a row of N switches wired so
    that, unless the following rules are
    followed, an alarm will go off: (1) the
    switch at the far right can be turned on or
    off at any time and (2) any other switch may
    be turned on or off only if the switch to its
    immediate right is on and all other switches
    (if any) to its right are off. If all N
    switches are initially on, what is the least
    number of moves required to deactivate (all
    switches off) the alarm system? A move
    consists of turning a switch on or off.
    Express your answer without using recursion
    or summation. You can use (N mod 2) or (-1)^N
    as a way to find even versus odd.

    -- Problem Solving Through Recreational
    Mathematics by Bonnie Averbach and Orin Chein

TODO

    meh


## Computer Bonus

    Let f(N) be the smallest positive integer
    whose prime factorization contains the N
    digits 1 to N each exactly once. So,
    f(3)=26=2x13. Find f(N) and its prime
    factorization, for N equal to 4 through 9,
    inclusive.

    -- Adapted from Puzzle Corner in Technology
       Review

TODO

    Let's try exhaustive search and see how we do.
    Up to N=8 this works very well.  Not so good for
    the case N=9 but it eventually gets to the solution.

```{r}
library(gmp)

needs=c(4,5,6,7,8,9)

n=10
while (T) {
    m=sort(strsplit(paste0(as.numeric(factorize(n)),collapse=""),"")[[1]])
    if (length(m) == length(unique(m))) {
        m=as.numeric(m)
        if (m[1]==1 & all(diff(m)==1)) {
            if (length(m) %in% needs) {
                print(paste0(sprintf("%d  ",c(length(m),n,as.numeric(factorize(n)))),collapse=""))
                k=which(length(m) %in% needs)
                needs=needs[-k]
                if (length(needs)==0) {
                    break
                }
            }
        }
    }
    n=n+1
}
```

Solution:

    N    f(N)         factors
    -----------------------------------------------------
    4    246          2   3           41
    5    1230         2   3   5       41
    6    13830        2   3   5            461
    7    60710        2       5    13        467
    8    446010       2   3   5                    14867
    9    2992890      2   3   5         67      1489

I don't really see any obvious patterns other than the preponderance
of 2, 3, and 5.  After that the solutions seem scattered.

The Bent solution does not offer any insight.  It does match this
solution.
