---
title: "Bent Fall 2002"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Fall 2002


## Problem 1

    What is the total number of pips
    on the 136 dominoes making up a
    double-fifteen set?
    
    -- Madachy's Mathematical
    Recreations by Joseph S. Madachy

Solution:

    I had no idea what a double-15 set of dominoes was so I looked
    on the Web and saw the answer in an add for a company that sells them.
    The Wikipedia article on dominoes describes this in detail.
    
    A set of double-n dominoes contains every possible combination of
    0-n pips on each half domino.  For a double-n set we have:
    
    0 paired with 0-n --> n+1
    1 paired with 1-n --> n
    2 paired with 2-n --> n-1
    ...
    n paired with n   --> 1
    
    Total number of dominoes is (n+1)*(n+2)/2
    Total number of pips is n*(n+1)*(n+2)/2

    The number of dominoes is obvious for a summation of 1 to n+1.
    The number of pips just multiplies by the average number of pips per domino.

    For double-15 set: Number of pips = 15*16*17/2 = 2040

## Problem 2

    Solve the following cryptic
    multiplication: PEN x INK =
    LETTER with different letters
    being different digits, the same
    letter being the same digit throughout,
    and no leading zeros.
    
    -- Howard G. McIlvried, PA Gamma'53

Solution: solved in the cryptic arithmetic notebook.

    PEN = 615
    INK = 354
    LETTER = 217710


## Problem 3

    In a game of bridge, South is
    declarer at a contract of three no
    trump and holds A, K, 10, x, x of
    spades, while dummy holds J, x, x of
    spades. (Declarer has no knowledge
    of how the rest of the spade suit is
    distributed.) After winning the first
    trick, on which no spades were
    played, declarer leads the A of
    spades. Both opponents follow suit,
    but the Q does not drop. What is the
    probability that the Q will fall if
    declarer now leads the K of spades?
    
    -- John W. Langhaar, PA A'33

Solution:

    The solution in the Bent is WAY more complex.
    Need to consider the split of the cards before the
    first round of spades a well as after.
    
    End results is P(Q) = 0.3 instead of 0.25.
    
    My original solution:
    
    Prior to leading the A there are 5 spades
    in the hands of the opposition (E-W).
    
    After leading the A there are three remaining.
    
    These are Q, x, x and they can either be located
    in E or W.  The only way the Q drops is if it is
    a singleton.  These are the possible distributions
    of the three spades:
    
    Q   x   x   result
    ------------------
    E   E   E
    E   E   W
    E   W   E
    E   W   W   Q falls
    W   E   E   Q falls
    W   E   W
    W   W   E
    W   W   W

    So, Q falls in 2/8 of the distributions = 1/4 of the time.
    

## Problem 4

Problem is missing -- there is no problem 4!
The solutions in next month explained that there was a typo
in the problem and there was no unique solution so the
problem was not counted for determining perfect
solution submissions.


## Problem 5

    What is the maximum number of
    knights that can be placed on a
    standard 8x8 chess board, so that no
    knight threatens another knight,
    i. e., no knight can move to a square
    occupied by another knight? Also,
    what is the placement of the knights
    on the board? A knight moves two
    squares in one direction and one
    square in another direction to end up
    on the diagonally opposite corner of a
    2 x 3 grid. The move can occur even if
    intervening squares are occupied.
    
    -- Daryl Cooper

TODO

I think I solved this once before - but I'm not sure if that was
the case of no knight attacking another knight or no knight that
was not under attack.  I remember some blocks of knights in the
four corners.

I peeked at their solution -- not the problem I recall.  They note
that you can just put all the knights on the same color and they
can't attack one another -- e.g. 32 knights.


## Bonus

    Most people are familiar
    with how water sloshes back and
    forth in a bathtub, rising and falling
    at the ends of the tub, similar to
    ocean tides, but on a much higher
    frequency. A similar phenomenon
    can occur in a lake, usually initiated
    due to the action of the wind piling
    up water at one end. Assume the
    Lake of Geneva behaves like a tank
    of water, 70 km long, 8 km wide, and
    150 m deep. What is the period of
    the unimodal oscillations in the lake?
    
    -- Vibrations and Waves
    by A.P. French

TODO

I don't have a clue.

They cite some obscure formula -- not worth the effort!!!???


## Double Bonus

    Consider the equation: x^7 + y^3 = z^2.
    Two solutions to this equation are: 
    
        x = 1, y = 2, z = 3; and 
        x = 2, y = 17, z = 71.
        
    Find another solution in which x, y, and z
    are relatively prime positive integers.
    
    -- Mathematical Mountaintops
    by John L. Casti

Solution:  let's try exhaustive search.  Keep x small to prevent
numerical overflow.

```{r}
source('rgc_lib.R')

for (x in 1:20) {
    for (y in 1:100000) {
        m=x^7+y^3
        z=floor(round(sqrt(m)))
        if (z*z == m) {
            s1=prime_factor(x)
            s2=prime_factor(y)
            s3=prime_factor(z)
            if (any(s1 %in% s2) | any(s1 %in% s3) | any(s2 %in% s3) ) {
                # Not a solution
            } else {
                print(paste0("x = ",x,"   y = ",y,"   z = ",z))
                # print(c("   x factors = ",s1))
                # print(c("   y factors = ",s2))
                # print(c("   z factors = ",s3))
            }
        }
    }
}
```

Most of the solutions have common prime factors so we really need to
filter out the extraneous solutions.  Feasible solutions for the
range of x=[0-20], y=[0-100e3]:

    x         y        z
    ---------------------
    1         2        3
    2        17       71
    17    76271 21063928

Exhaustive search with sympy:

    x = 17         = 17
    y = 76271      = 13*5867
    z = 210639280  = 2^4 * 5 * 79 * 33329

FYI: the author's note that this is the only other known solution
to this problem.
