---
title: "Bent Summer 2017"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Summer 2017


## Problem 1

    At lunch one day, math Prof. Lohwater
    mentioned to economics Prof. Baird that he
    had to go pick up a birthday present for his
    daughter, Tatiana. Baird knew that Lohwater
    had three children of different ages, but did
    not recall which was which and asked her age.
    Lohwater, being in a puzzling mood, wrote on
    a paper napkin a polynomial f(x) with integer
    coefficients and said, "Tatiana's age is the
    only positive integer which is a zero of this
    polynomial." Baird was not much impressed and
    not in the mood for algebra, but decided to
    give it the old college try. More or less at
    random, Baird picked 7, but was told f(7) =
    77. Then he tried a larger integer n, which
    gave f(n) = 85. Lohwater exclaimed that his
    daughter was older than n. How old is
    Tatiana?

    -- The Surprise Attack in Math- ematical
    Problems by L.A. Graham

Solution:

    The polynomial can be written as:
    
    f(x) = (x-t)*f2(x)
    
    We know:
    
        f(t) = 0
        f(7) = 77 = (7-t)*f2(7)
        f(n) = 85 = (n-t)*f2(n)
        t>n>7

    I think there are an infinite number of solutions?
    Do we make some assumption about the degree of the
    polynomial?  What does the constraint that t is the
    only positive integer zero of the polynomial mean?

    I gave up and peeked at the solution.
    
    They argued that f(n)-f(7) must be divisible by (n-7).
    (I need to confirm this -- it isn't blatantly obvious to me)
    
    Consider an arbitrary poly: f(x) = sum(a_i * x^i)
    
    Then: f(n1) = sum(a_i*n1^i)
          f(n2) = sum(a_i*n2^i)
          
    f(n1)-f(n2) = sum(a_i*(n1^i-n2^i))
    
    I can't prove this is a multiple of n1-n2 analytically for any i
    but I did test it for several different values of i and it generates
    integer quotients -- the coefficients are the binomial coefficients.
    
    Let's just accept that the diff of like powers is dvisible by the
    difference.
    
    If so - then:
    
    (85-77) / (n-7) = 8 /(n-7) --> n-7 = 1, 2, 4, 8 --> n=8,9,11,15
    
    Also -- since f(t) = 0 we know:
    
    (77-0) / (7-t) = 77/(7-t) --> 7-t = 11, 7, 77 --> t=-4,0,-70 ???
    
    I don't understand them -- because they use t-7 instead.  Apparently
    the order of the terms are interchangeable.  This leads to:
    
        t-7 = 11, 7, 77 --> t = 14, 18, 84
        
    Finally, use f(t) and f(n) then:
    
    f(n)/(t-n) = integer --> 85/(t-n) = 1, 5, 17, 85
    
    or
    
    t-n = 85,17, 5, 1
    
    select one from each set of parens:
    
        (14, 18, 84) - (8, 9, 11, 15) = (1, 5, 17, 85)
    
    The only combination that works is: 14-9 = 5, so
    
        t = 14
        n =  9


## Problem 2


    Inside a hemispherical building is a
    concentric hemispherical atrium. Each floor
    of the building is an annulus which extends
    from the outside of the atrium to the inside
    of the outside wall. On each floor, a regular
    geometric figure has been painted. On the top
    floor, the outside wall circumscribes an
    equilateral triangle which circumscribes the
    atrium. Similarly, on the floor immediately
    below the top floor, a painted square
    inscribes the outer wall and circumscribes
    the atrium. One floor down, the figure is a
    regular pentagon, and the figure is a regular
    hexagon on the level below that. The top
    floor is 5 m higher above ground level than
    the floor below it, and the painted square is
    60 m on a side. What is the area of the floor
    with the painted hexagon, what is its outside
    perimeter and its height above ground?

    -- Cecil T. Ho, NY Beta'68

Solution:

    I floundered around on this for a while but I probably
    would have figured it out eventually.  I did peek at
    their solution but they didn't have anything I didn't
    already know.  Basically I wasn't sure if the distance
    between all floors was 5m or just the square-triangle.
    Turns out it was just the square-triangle.
    
    Basic solution:
    
    Each floor is in the form of an annulus with inner
    radius, r1, and outer radius, r2.  The radii are related
    by the inscribed/circumscribed relationship with a polyon
    of n sides.  Consider a polygon of n sides:
    
    theta = angle of each pie slice = 2*pi/n
    alpha = theta/2 = pi/n
    L = length of a side of the polygon
    
    For inscribed circle:
    
        (L/2)/r1 = tan(alpha) = sin(alpha)/cos(alpha)
        
    For circumscribed circle:
    
        (L/2)/r2 = sin(alpha)

    Taking the raio:
    
        r2/r1 = 1/cos(alpha) --> r1=r2*cos(alpha)
        
    The area of each floor:
    
    Consider a vertical cross section of the building.
    Let R1 = radius of inside hemisphere (atrium), R2 = radius of
    outside hemisphere.  Let y = height of a particular floor.
    
        y^2 + r1^2 = R1^2
        
        y^2 + r2^2 = R2^2
        
    Equating the two expressions for y^2 gives:
    
        r2^2 - r1^2 = R2^2 - R1^2
        
    Since R1,R2 are constants (independent of floor) this is a constant.
    The area of each floor is just:
    
        Area = pi*r2^2 - pi*r1^2 = pi*(r2^2 - r1^2)
    
    We know something about the floor with the square.  The side of the
    square is 60m and there are 4 sides to the polygon.  The inscribed
    circle has a radius of 30m, the circumscribed is 30*sqrt(2) m.
    
        Area = pi*(30^2 * 2 - 30^2) = 900*pi m^2
        
    For the floor with the hexagon we have:
    
        r1 = r2*cos(pi/6) = r2*sqrt(3)/2
        
        pi*(r2^2 - r2^2 * 3/4) = 900*pi
        
        r2^2 = 4*900 = 3600  --> r2 = 60m, Perimeter = 2*pi*r2

    The height above the ground requires use of the triangle-square
    floor distance of 5m.
    
    For the triangle floor:
    
        r1 = r2*cos(pi/3) = r2/2
        
        pi*(r2^2 - r2^2 * 1/4) = 900*pi
        
        r2^2 = 4/3*900 = 1200 --> r2 = 20*sqrt(3), r1=10*sqrt(3)
        
    For the square floor:
    
        r1 = 30, r2=30*sqrt(2)
        
    We know:
    
        y^2 + r1^2 = R1^2
        
        y^2 + r2^2 = R2^2
        
    Let the height of the square floor be y=h.
    
        h^2 + 30^2 = R1^2
        
        h^2 + (30*sqrt(2))^2 = R2^2
        
    The height of the triangle floor is y=h+5.
    
        (h+5)^2 + (10*sqrt(3))^2 = R1^2
        
        (h+5)^2 + (20*sqrt(3))^2 = R2^2

    Solving either one yields:
    
        h = 115/2 = 57.5m for height of square floor
        
    Substituting back into the given equations gives:
    
        R1 = 5/2*sqrt(673), R2 = 5/2*sqrt(817)
        
    For the hexagon floor:
    
        h^2 + (30*sqrt(3))^2 = (5/2*sqrt(673))^2
        
        h = 5/2*sqrt(241) ~ 38.81m

    In summary:
    
        Area of each floor = 900*pi m^2         (~ 2827   m^2)
        Perimeter of hexagon floor = 120*pi m   (~  377   m)
        Height of hexagon floor = 5/2*sqrt(241) (~  38.81 m)

    Just out of curiosity, the height of some other floors:
    
    Triangle:  h^2 + (10*sqrt(3))^2 = (5/2*sqrt(673))^2
               h = 125/2 = 62.5
    
    Square:    h = 57.5 as noted above
    
    Pentagon:  r1=r2*cos(pi/5) --> r2^2*(1-cos(pi/5)^2) = 900
        Solution is ugly:
        r2 = 51.03904850112239
        r1 = 41.29145761413522
        h = 50.01265367984459
        
    Hexagon:   h = 5/2*sqrt(241) ~ 38.81

    Next floor: r1=r2*cos(pi/7) --> r2^2*(1-cos(pi/7)^2) = 900
        r2 = 69.14294612887468
        r1 = 62.29564189717018
        h = 18.04170171074603
        
    The next one is below ground so I quit.


## Problem 3

    Give me an integer that, when divided by the
    product of its digits yields a quotient of 3,
    and if you add 18 to the integer, its digits
    are reversed.

    -- Little Man Tate, 1991

Solution: Brute force?

```{r}

n=1
while (T) {
    digits=as.numeric(strsplit(sprintf("%d",n),"")[[1]])
    p=prod(digits)
    if (3*p == n) {
        digits2=as.numeric(strsplit(as.character(n+18),"")[[1]])
        if (all(digits2 == rev(digits))) {
            print(n)
            break
        }
    }
    n=n+1
}

```

Didn't take much brute or force.

    Solution is n=24
    Product of digits = 8, 24/8 = 3
    
    For n+18 we get 42 == rev(24)

    They assert that it is obvious the solution has two digits but
    that wasn't obvious to me.  If I knew it was only two digits
    I would have done it algebraically.
    
    They argue that since adding 18 reverses all the digits the
    original number can only have 2 digits.  Really?


## Problem 4

    A data entry service has assigned a different
    code number in the inclusive range of 1
    through 10 to each of the days of the week:
    SUN, MON, TUE, WED, THU, FRI, and SAT. It
    turns out that the code numbers for a pair of
    days have a common factor greater than 1
    if -- and only if -- the abbreviations for the days
    have at least one letter in common. In
    addition, if one substitutes code numbers for
    days, MON + TUE < WED + THU < FRI + SAT. In
    the order SUN through SAT, what are the code
    numbers of the days of the week?

    -- Susan Denham in New Scientist

Solution: exhaustive search

    Condition is for gcd > 1 for lots of cases.
    Build the permutation matrix and then do the
    testing.
    
```{r}
library(gtools)
library(gmp)

dow=c('sun','mon','tue','wed','thu','fri','sat')

a=permutations(10,7)
a=data.frame(a)
colnames(a)=dow
k=rep(T,nrow(a))
for (n1 in 1:6) {
    for (n2 in (n1+1):7) {
        i=dow[n1]
        j=dow[n2]
        print(sprintf("Checking %s vs %s",i,j))
        if (any(strsplit(i,"")[[1]] %in% strsplit(j,"")[[1]])) {
            k=k&(gcd(a[,i],a[,j])>1)
        } else {
            k=k&(gcd(a[,i],a[,j])==1)
        }
    }
}
n1=a[,'mon']+a[,'tue']
n2=a[,'wed']+a[,'thu']
n3=a[,'fri']+a[,'sat']
k=k&(n1<n2)&(n2<n3)

print(a[k,])

```
 
 Solution:
 
            sun mon tue wed thu fri sat
    575356  10   5   6   9   4   7   8


## Problem 5

    Five Tau Bates, who understand not a word of
    Chinese, enter a restaurant in China, where
    none of the staff speak any English.
    Undaunted, they sit down and look at the
    menu, which has nine items listed, each
    identified by a Chinese symbol, but which,
    for the sake of simplicity, we will call A,
    B, C, D, E, F, G, H, and I. They each order
    one item from the menu by pointing to it.
    When the server brings the five dishes, he
    sets them in the middle of the table so that
    the diners can share. The same thing happens
    for the next two nights. However, by the time
    they return on the fourth day, they have
    deduced what culinary item each letter
    represents, and each is able to order his
    favorite meal with total confidence. What
    dishes did they order each night? Present
    your answer as three strings, each with five
    letters.

    -- Hard-to-Solve Brainteasers by Jaime and
       Lea Poniachik

TODO

    meh


## Bonus

    The ancient Egyptians had a unique way of
    depicting rational numbers when the numerator
    was greater than 1; they expressed such
    fractions as the sum of a series of unit
    fractions, that is fractions with a numerator
    of 1. Furthermore, in a so-called Egyptian
    fraction, none of these unit fractions could
    be repeated. For example, 5/6 could be
    written as 1/2 + 1/3, and 7/12 as 1/2 + 1/12
    or 1/3 + 1/4 or 1/4 + 1/6 + 1/8 + 1/24, but
    not as 1/4 + 1/4 + 1/12.  In ancient Egypt,
    five women went into a bakery to buy bread,
    but only three identical loaves were left.
    The baker said not to worry as she and her
    two apprentices would each take a loaf and
    divide it into pieces which were unit
    fractions of a whole loaf. Each loaf would be
    cut into the same number of pieces, but the
    pieces from each loaf would form a different
    Egyptian fraction. That is, no two loaves
    were cut into exactly the same set of pieces,
    although the Egyptian fractions for the three
    loaves may have had some sizes in common.
    The baker was more skilled with a knife than
    the apprentices, as she was capable of
    cutting off a slice as small as 1/100 of a
    loaf, but the smallest piece the apprentices
    were competent to slice off was 1/50 of a
    loaf. The pieces would be distributed to the
    five customers in such a way that each would
    get the same number of pieces, the volume of
    the pieces for each customer would total 3/5
    of a loaf, and each buyer's pieces would form
    a different Egyptian fraction. What size
    pieces did each woman get?  We want the
    answer with the fewest pieces. Present your
    answer as five different series of unit
    fractions. For simplicity, the answers need
    only list the denominators of the unit
    fractions.

    -- Ancient Puzzles-The Greatest Puzzles Ever
       Solved, by Tim Dedopulos


Solution:

    If each loaf is cut into n pieces then we have
    a total of 3*n pieces to be distributed to the
    five women.  This must be an integer so:
    
        m = 3*n/5 = integer
        
        n = 5/3*m --> m = 3, 6, 9, ... multiples of 3
        
    We can do an exhaustive search for m=3 and then
    look for the fractions that sum to 1 with only
    one denominator > 50.
    
        Need 1/a + 1/b + 1/c = 3/5 with a<b<c
        
    If a solution can be found for m=3 then there is
    no need to look at m=6, ... since it will result
    in more pieces and the problem requests minimum pieces.
    
```{r}
library(gmp)

solution=c()
for (a in 1:98) {
    for (b in (a+1):99) {
        for (c in (b+1):100) {
            if(as.bigq(1,a)+as.bigq(1,b)+as.bigq(1,c) == as.bigq(3,5)) {
                solution=append(solution,c(a,b,c))
            }
        }
    }
}
solution=matrix(solution,ncol=3,byrow=T)
print(solution)

```

    There are six fraction combinations that add to 3/5.
    Two of the fractions include 1/60 which can only be
    done by the baker.  We will arbitrarily pick the first
    one and see if we can rearrange the pieces to form
    three whole loaves using proper Egyptian fractions:

             [,1] [,2] [,3]
        [1,]    2   12   60  <-- pick this
        [2,]    2   14   35
        [3,]    2   15   30
        [4,]    3    4   60  <-- omit this
        [5,]    3    5   15
        [6,]    3    6   10
    
    So the challenge is to rearrange the 15 values into
    three sets of 5 that sum to 1 without any duplicated
    fractions.
    
        2   12   60
        2   14   35
        2   15   30
        3    5   15
        3    6   10
    
    Since there are three fractions of 1/2 we need one for each
    loaf.  There are two fractions of 1/3 that must be in separate
    loaves and two fractions of 1/15 that must be in separate loaves.
    
    The known denominators must be: 2, 2, 2, 60
    
        Baker: 2, 60
        Asst1: 2
        Asst2: 2
    
    Remaining pieces:
    
        3, 3, 5, 6, 10, 12, 14, 15, 15, 30, 35
    
    We know the baker must have 1/2 and 1/60 and either 1/3 or 1/15 or both.
    We can try various combinations until we find a solution that sums to 1.
    
    We need to find a combination of 3 of these that yields a total of 1.

```{r}

baker=as.bigq(1,2)+as.bigq(1,60)

a=combinations(9,3,c(3,5,6,10,12,14,15,30,35))
k=apply(a,1,function(x) {as.numeric(baker+as.bigq(1,x[1])+as.bigq(1,x[2])+as.bigq(1,x[3]))})

print(a[k==1,])
```

    The only combination that works is:

        Baker: 2, 3, 12, 15, 60
    
    This leaves the following for the assistants:

        Asst1: 2, 4 more pieces
        Asst2: 2, 4 more pieces
    
    Remaining pieces: 3, 5, 6, 10, 14, 15, 30, 35

    These need to be partitioned into two groups where each
    one sums to 1/2.

```{r}
asst=as.bigq(1,2)

a=combinations(8,4,c(3,5,6,10,14,15,30,35))
k=apply(a,1,function(x) {as.numeric(asst+as.bigq(1,x[1])+as.bigq(1,x[2])+as.bigq(1,x[3])+as.bigq(1,x[4]))})

print(a[k==1,])
```

    The two assistants and the baker are:

    Asst1: 2, 3, 14, 15, 35
    Asst2: 2, 5,  6, 10, 30
    Baker: 2, 3, 12, 15, 60

    At the start of this problem we arbitrarily picked
    one of the two solutions that required the skill of
    the baker.  What if we had picked the other one?
    Would a solution still be possible?
    
    e.g. what if our original sets were:
    
        2   14   35
        2   15   30
        3    4   60  <-- must be baker for 1/60 loaf
        3    5   15
        3    6   10
    
    All three must use a piece of 1/3 and then choose from the
    remaining pieces:
    
    Then the baker starting point is:
    
        baker: 3, 60,  three more pieces

    We search for a feasible solution of three pieces from the remainders:
    
```{r}

baker=as.bigq(1,3)+as.bigq(1,60)

a=combinations(9,3,c(2,4,5,6,10,14,15,30,35))
k=apply(a,1,function(x) {as.numeric(baker+as.bigq(1,x[1])+as.bigq(1,x[2])+as.bigq(1,x[3]))})

print(a[k==1,])
``` 

    There is no feasible solution with this choice for the baker.
    
    Conclude: the ony solution is the one found above.
    The solution is repeated here for convenience.

    The two assistants and the baker are:

    Asst1: 2, 3, 14, 15, 35
    Asst2: 2, 5,  6, 10, 30
    Baker: 2, 3, 12, 15, 60


## Computer Bonus

    Many states have license plates with a format
    consisting of three letters followed by four
    digits. Assume each such license plate
    represents an integer in the lowest possible
    base greater than 10, with a = 10, b = 11, c
    = 12, . . ., z = 35. For example, the license
    plate TBP1885 would represent an integer in
    base 30, the lowest base where T is a valid
    digit, and its base 10 equivalent would be
    21,428,584,445 (base-10). Exactly four of these
    license plates, when expressed in their base
    10 equivalent, can be expressed as x^y, where
    x and y are positive integers (y > 3). One of
    these plates is BBG1802 which is in base 17;
    its base 10 equivalent is:
    
        282,475,249 (base 10) = 
        49^5 =
        7^10. 
        
    In license plate format, what are the other
    three integers?

    -- Dakota Ebersold, son of member

Solution:  Can probably enumerate all the solutions.  Warning that this runs very
slow.  A much more efficient solution follows.

```{r, eval=FALSE}

is_solution = function(x) {
    b=max(x)+1
    n=0
    for (i in x) { n=b*n+i }
    # n = license plate number in base 10
    # How do I check if n = x^y ???  with y>3
    xmax=round(n^(1/3.8))
    y=log(n)/log(2:xmax)
    k=which.min(abs(y-trunc(y)))
    x=(2:xmax)[k]
    y=round(log(n)/log(x))
    if (y>3) {
        return(x^y == n)
    }
    return(FALSE)
}

t1=Sys.time()
for (n1 in 10:35) {
    print(sprintf("Checking n1 = %d",n1))
    print(Sys.time()-t1)
    t1=Sys.time()
    for (n2 in 10:35) {
        for (n3 in 10:35) {
            for (d1 in 0:9) {
                for (d2 in 0:9) {
                    for (d3 in 0:9) {
                        for (d4 in 0:9) {
                            plate=c(n1,n2,n3,d1,d2,d3,d4)
                            if (is_solution(plate)) {
                                print("Solution set:")
                                print(plate)
                            }
                        }
                    }
                }
            }
        }
    }
}

```

Solutions:

    11 11 16  1  8  0  2 --> BBG1802
    11 13 10  8  7  7  9 --> BDA8779
    12 14 14  2  7  4  4 --> CEE2744
    18 29 18  0  8  0  1 --> ITI0801

Run time ~ 78 minutes

The Bent calls this the naive brute force approach.  They suggest that it
is more efficient to look for the combinations of x^y with y>=4 that are
feasible.  In particular they note that the largest value is ZZZ9999 (base 36).
This value is 78362916381 which is just slightly larger than 529^4.  This means
that any solution to this problem can be expressed as x^y with x in the
range of 2:529 and y>4.  Increase the value of y until you reach the max value.

They don't really say what to do with the x,y pairs once you get them but it
seems like you need to convert this value with each different number base and
see if it meets the template for a licence plate (three letters + four numbers).

We can try that:

```{r}

int_to_digits = function(n,b=10,num_digits=7) {
    a=c()
    for (i in 1:num_digits) {
        m=n%%b
        n=trunc(n/b)
        a=append(a,m)
    }
    if (n != 0) {
        a=rep(b-1,num_digits)
    } else {
        a=rev(a)
    }
    return(a)
}

digits_to_int = function(a,b=10) {
    n=0
    for (i in a) { n=b*n+i }
    return(n)
}

t1=Sys.time()

# The largest possible licence plate value
# No need to search for x^y > pmax
pmax=digits_to_int(c(35,35,35,9,9,9,9),36)

# Create a vector of all the feasible x^y quantities
numlist=c()
for (x in 2:529) {
    y=4
    while (T) {
        m=x^y
        if (m<pmax) {
            numlist=append(numlist,m)
            y=y+1
        } else {
            break
        }
    }
}
# Remove redundant entries
numlist=unique(numlist)

# Check each value against the 26 possible bases.
# If the pattern matches the letters+digits of a
# licence plate then add it to the solutions.
solutions=c()
for (n in numlist) {
    for (b in 11:36) {
        a=int_to_digits(n,b)
        if (all(a[1:3]>9) & all(a[4:7]<10) & (b==(max(a)+1))) {
            print(sprintf("base = %d   n = %.0f",b,n))
            print(a)
            solutions=append(solutions,a)
        }
    }
}
solutions=matrix(solutions,ncol=7,byrow=T)
t2=Sys.time()

print(solutions)
print(t2-t1)
```

    OMG -- the solution time is about 1 second!!!

    I did the factorization manually and added it to the table below:


         [,1] [,2] [,3] [,4] [,5] [,6] [,7]
    [1,]   11   11   16    1    8    0    2   282475249    7^10 (base 17)
    [2,]   18   29   18    0    8    0    1   13841287201  7^12 (base 30)
    [3,]   12   14   14    2    7    4    4   148035889    23^6 (base 15)
    [4,]   11   13   10    8    7    7    9   90224199     39^5 (base 14)
    > print(t2-t1)
    Time difference of 1.025259 secs

    The original brute force solution took over an hour.
    
    I guess algorithms are important.


