---
title: "Bent Fall 2017"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Fall 2017


## Problem 1

    Aunt Alice is a bit hard of hearing, so that
    when her nephew Ned asked various neighbors
    some questions about their house numbers, she
    failed to hear the answers, though she heard
    the questions fine. Ned has lived on
    Christmas Crescent for some time, and his
    aunt knows Ned's number, but Alice has only
    just bought a vacant house there, but Ned
    doesn't know that. The Crescent has houses
    numbered rather curiously from 5 to 105,
    inclusive.  Ned asked the same three
    questions to three people, first to X, then
    to Y, then to Z, who live in separate houses
    on the Crescent. The questions were: (i) Is
    the number of your house a multiple of 4?
    (ii) Is it a perfect square?  (iii) Is it a
    multiple of 9?  No two sets of three answers
    were exactly the same. After hearing X's
    answers, Ned says to him: "If I knew whether
    your house number is greater than 83, I would
    know what it is." (Alice heard her nephew say
    this and was able to write down X's number
    correctly.) After hearing Y's answers, Ned
    says to him: "If I knew whether your number
    is greater than 50, I could tell you what it
    is." (Alice heard her nephew say this, and as
    she happened to know that Y's number is
    greater than her own, she was able to write
    it down correctly.  After hearing Z's
    answers, Ned says to him: "If I knew whether
    your house number is greater than 30, I could
    tell you what it is." (Alice heard this and
    notes with interest that her own number is
    certainly less than Z's. But she too has no
    way of deciding whether Z's number is greater
    than 30. However, being anxious to continue
    building her reputation for logical deduction
    and intuition, guessed that it was greater
    than 30 and wrote down the number, which,
    fortunately was quite correct.) What are the
    house numbers of Ned, Alice, X, Y, and Z?

    -- Brain Puzzler's Delight by E.R. Emmet

TODO

    meh


## Problem 2

    The Rev. Obadiah Slope does not actually care
    about soccer but, since becoming a Rural
    Dean, he likes to seem as if he does. So he
    takes a dutiful interest in the local league
    (each team plays each other team once in a
    season). "How is it going this year, my son?"
    he recently asked the retired tax inspector
    who acts as secretary.  "Well, it's not
    finished yet you know," was the answer, "even
    though only four sides are competing this
    time. Parminster United have so far played
    two games and lost them both, with three
    goals for and six against in total. Quondam
    Athletic have also played two, winning one
    and drawing the other, with a total of four
    goals for and three against. Real Episcopi
    have drawn one and won two, totaling two
    goals for and none against. And Salem
    Dynamo...." "Splendid! Splendid!" Rev. Slope
    broke in, conscious that the lunch hour was
    nigh and suddenly remembering that, "I must
    see a man about a misericord." What was the
    score in each of Salem's matches?

    -- A Tantalizer by Martin Hollis in New
    Scientist

TODO

    meh


## Problem 3

    On a 12-hour analog clock, at what time near
    8:18 are the hands symmetric about 6, that
    is, 6 splits the angle between the minute and
    hour hands? At what time near 8:18 is the
    clockwise angle from the hour hand to the
    minute hand the Golden Mean fraction of the
    whole circle, that is, the angle between the
    minute and the hour hands is 360 degrees
    (sqrt(5) - 1)/2?
    Express your two answers (to the nearest
    millisecond) in the form hh:mm:ss.ttt.

    -- Allan Gottlieb's Puzzle Corner in
       Technology Review

Solution:

    Let t = current time (in seconds)
    
    h = angle for the hour hand
    m = angle for the minute hand
    
    h=t/(86400/2)*360
    m=t/3600*360 with a repeat every 3600 seconds

    Want: h-180 = 180-m
    
    So: t/(86400/2)*360 - 180 = 180 - (t/3600-n)*360
    
    where n = 8 = number of hours
    
    t = 29907.69230769231 seconds
    
    Solving for t = 8:18:27.692

    As far as the Golden Mean fraction - I don't know what
    that really means.  I guess it means the minute hand
    leads the hour hand by 360*(sqrt(5)-1)/2 degrees.
    
    so: (t/3600-n)*360 - t/(86400/2)*360 = 360*(sqrt(5)-1)/2
    
    Where n = 7 or 8 (might need to allow minute hand to be in
    the next hour).
    
    t = 08:18:38.097119
    
    The solution in the Bent appears different but is the same.
    They have M=(360*sqrt(5)-600)/11 after 8 which is the same
    time as above.


## Problem 4

    In the game of tennis, two players hit a ball
    back and forth across a net into a defined
    court. A player loses a point by either
    hitting the ball outside the court or failing
    to return a ball before it has bounced twice.
    The first player to win four points, provided
    he is ahead by at least two points, wins the
    game. Thus, the final score can be 4-0, 4-1,
    4-2, 5-3, 6-4, etc. The same player serves
    for a complete game, and on each point he has
    two chances to make a successful serve.
    Assume a player has a probability, p1, of a
    successful first serve, a probability, p2, of
    a successful second serve if the first serve
    is bad, a probability q1 of winning a point
    if his first serve is good, and a probability
    q2 of winning a point if his second serve is
    good. If he is serving, what is his
    probability w of winning a point (in terms of
    p's and q's)? What is his probability g of
    winning the game (in terms of w)? If p1=0.8,
    p2=0.7, q1=0.6, and q2=0.5, what are w and g?

    -- H.G. McIlvried III, PA G'53

Solution:

    The probability of winning a point should be:
    
    w = p1*q1 + (1-p1)*p2*q2

    To win a game:
    
        win 4 points in a row:                 w^4
        win 4 out of 5 (lose 1 of first 4):    C(4,1)*w^4*(1-w)
        win 4 out of 6 (lose 2 of first 5):    C(5,2)*w^4*(1-w)^2
        win 3 out of 6 --> deuce:              C(6,3)*w^3*(1-w)^3*W(deuce)
        
        Once you get to deuce there are two alternatives or go back
        to deuce:
        
            ww
            wL -> Deuce
            Lw -> Deuce
            LL
            
        W(deuce) = w^2/(w^2 + (1-w)^2)
        
    Summing all the possibilities:
    
        P = w^4 + 
            C(4,1)*w^4*(1-w) + 
            C(5,2)*w^4*(1-w)^2 + 
            C(6,3)*w^3*(1-w)^3*w^2/(w^2+(1-w)^2)
        
        P = w^4*(15 - 34*w + 28*w^2 - 8*w^3)/(2*w^2 - 2*w + 1)
        
    For the specific case:
    
        p1=0.8, p2=0.7, q1=0.6, q2=0.5
        
        w = 0.8*0.6 + (1-0.8)*0.7*0.5 = 0.55 (55/100)
        
        P ~ 0.6231485024752476
        
          = 50350399/80800000 (precisely)


## Problem 5

    Solve this cryptic addition with the usual
    rules: different letters are different
    digits, same letter is same digit, no leading
    zeros, base 10. 

    FIVE + FIVE + TEN + TEN + TEN + TEN + THIRTY = EIGHTY.

    -- Journal of Recreational Mathematics

Solution: from the cryptic arithmetic notebook

    Cryptic addition:

      FIVE + FIVE + TEN + TEN + TEN + TEN + THIRTY = EIGHTY

    a=make_crypt_df('fivetenthirtyeighty')
    x1=make_num(a,'five')*2+make_num(a,'ten')*4+make_num(a,'thirty')
    x2=make_num(a,'eighty')
    k=(x1==x2)&(a$f!=0)&(a$t!=0)&(a$e!=0)
    sum(k)
    [1] 1
    a[k,]
            f i v e t n h r y g
    2923964 8 0 5 2 1 4 9 3 7 6


## Bonus

    Four square blocks of sizes from 1^2 up to 4^2
    can be fitted into a 7 x 5 rectangle (as
    shown), but not into a smaller rectangle. As
    you take a larger series of blocks, 1^2, 2^2,
    3^2, ..., n^2, it is not always easy to choose
    the rectangle which will accommodate them
    with the least "waste". With n=4, the minimum
    waste is 5. What size rectangles give the
    minimum waste for n=11 and for n=12? Consider
    the rectangle as a pxq grid, and draw in the
    n squares. Next, fill in each of these
    squares with digits; 1 for the 1^2 square, 2
    for the 2^2 square, 3 for the 3^2 square, ...
    Use a for 10^2 , b for 11^2 , etc. Use a dash for
    wasted space. Present your answers as q rows
    of p digits, as indicated in the example
    below.

        3334444
        3334444
        3334444
        2214444
        22-----

    -- Stephen Ainley in New Scientist

TODO

    meh


## Computer Bonus

    Find the (base 10) count of and the sum of
    all palindromic numbers between 1 and
    100,000,000 (base), inclusive, for bases 2
    through 10, inclusive. A palindromic number
    is a number which reads the same forward and
    backward.

    -- Rolf B. Karlsson, MI Z'96

Solution: Try brute force?

Preserved for posterity but the run time is very long.

```{r, eval=FALSE}
library(gmp)

int_to_digits=function(n,b=10) {
    a=c()
    while (n>0) {
        m=n%%b
        n=trunc(n/b)
        a=append(a,m)
    }
    a=rev(a)
    return(a)
}

for (b in 2:10) {
    np=0
    total=as.bigz(0)
    nmax=b^8
    for (i in 1:nmax) {
        a=int_to_digits(i,b)
        if (all(a==rev(a))) { np=np+1; total=total+i }
    }
    print(sprintf("%d   %d   %s",b,np,as.character(total)))
}

```

    While this was running forever I went and solved it in Python:
    
    Python Solution:
    
    Base   Num     Sum
    ---------------------------
    2         30           2632
    3        160         327039
    4        510        9986436
    5       1248      141759070
    6       2590     1240259700
    7       4800     7770547197
    8       8190    38125189144
    9      13120   155181063996
    10     19998   545045045040


    R Solution:
    
    Base   Num     Sum
    ---------------------------
    2         30           2632
    3        160         327039
    4        510        9986436
    5       1248      141759070
    6       2590     1240259700
    7       4800     7770547197
    8       8190    38125189144
    9      13120   155181063996
    10     19998   545045045040

    I repeated the final base in R using bigz values to avoid
    overflow and loss of precision.

Need something to make pandoc happy.

```{r}
1
```
