---
title: "Bent Winter 2019"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Winter 2019


## Problem 1

    My elder son has just entered as a
    freshman at Cornell University (NY
    Delta). Being a California native, I was
    naturally concerned how he would
    adjust to his first chilly winter there
    in Ithaca. Find a unique solution
    to the following cryptarithm to
    alleviate my fears:

        CORNELL + STUDENT = ENDURES.

    Standard rules apply: each different
    letter stands for a different digit
    and each different digit is always
    represented by the same letter;
    leading zeros are not allowed.

    -- Jeffrey R. Stribling, CA A'92

Solution: from the cryptic arithmetic notebook

    Cryptic addition:

      CORNELL + STUDENT = ENDURES

    a=make_crypt_df('cornellstudentendures')
    k=(make_num(a,'cornell')+make_num(a,'student') == make_num(a,'endures'))
    k=k&(a$c!=0)&(a$s!=0)&(a$e!=0)
    sum(k)
    [1] 1
    a[k,]
            c o r n e l s t u d
    2007354 5 4 8 1 9 7 3 6 2 0


## Problem 2

    In the World of Bonkers, the ditty for
    remembering the numbers of days in their
    months runs as follows:

    'Nineteen days hath Cucumber,
    Strawberry too, and the number
    of days there are in Tiddleywinks
    is just the same as 'tis in Jinks,
    which is seven whole days more
    than the ten there are in Pinafore.
    With the month of Collywobble,
    diary makers have some trouble;
    thirteen days if year is even,
    forty-four if odd. And Stephen
    quite the least of Bonkers' months,
    in fact it's sometimes called the
    runt, as you can easily understand,
    since it hath but four days
    and another two in years of Heaven
    (that's if the year's date ends in
    seven).'

    There are five days in each week:
    Joyday, Funday, Laughday,
    Blissday, and Workday, in that
    order.

    In the year 17 a.b. (after bliss) the King of
    Bonkers went off for his holiday on Funday,
    Strawberry 17th and returned, just over 7
    weeks later, on Blissday, Tiddleywinks 8th.
    In the year 19 a.b., there were 41st to
    Pinafore 7th. Note that Collywobble is the
    first and most important month of the seven
    which comprise the year.

    What is the order of the seven months of the
    Bonkers' year? Given Strawberry 13th was a
    Workday in the year 21 a.b., what day of the
    week was Collywobble 2nd?

    -- Brain Puzzler's Delight
    by E.R. Emmet

TODO

    meh


## Problem 3

    Find the smallest rational x for which x-5, x
    and x+5 are all squares of rational numbers
    (that is, each number can be written as some
    p^2/q^2).

    -- Mathematical Tournament from 1225

TODO

    Well I was flailing all over and was off on a
    Wikipedia hunt of Diophantine equations and
    Pythagorean triples and finally gave up.
    Clearly I was going off the deep end.
    
    I looked at the Bent solution.
    
    It uses Diophantine equations and Pythagorean triples!

    Basic idea:
    
        x = p^2/q^2 --> x +/- 5 = (p^2 +/- 5*q^2)/q^2
        
        Consider just the numerators:
        
        a^2 = p^2 - 5*q^2
        b^2 = p^2
        c^2 = p^2 + 5*q^2
        
        D = c^2-b^2 = b^2-a^2 = 5*q^2, or
        
        c^2 = b^2+D
        a^2 = b^2-D
        
        Multiplying: (c*a)^2 + D^2 = (b^2)^2
        
        Which is the Pythagorean theorem (a Diophantine equation).
        
        As an aside -- for arbitrary m,n this generates Pythagorean triples:
        
            c*a = m^2-n^2
            D   = 2*m*n
            b^2 = m^2+n^2
            
        Let:
        
            b   = r^2 + s^2
            m   = 2*r*s
            n   = r^2 - s^2
            
        Then: for integer r, s:
        
            D   = 4*r*s*(r^2-s^2)  = 5*q^2
            b^2 = (r^2+s^2)^2      = p^2

        Need to find r,s to make smallest integer of
        the form D = 5*q^2
        
        They assert either r=5 or s=5 (why?).
        
        If r=5, s=4 then D = 5*(16)(9) = 5*144 = 5*12^2
        b^2 = (r^2 + s^2)^2 = 41^2 = p^2
        
        p/q = 41/12 --> x   = (p/q)^2  = (41/12)^2 = 1681/144
                        x-5 = 961/144  = (31/12)^2
                        x+5 = 2401/144 = (49/12)^2

    Supposedly this solution was due to Fibonacci.


## Problem 4

    Primeministan County has a cluster of small
    towns interconnected by a series of roads,
    such that there is only a single route
    between any two towns. Interestingly, when
    sorted by distance, the ten routes in the
    county (measured in km) are unique
    consecutive primes. What is the minimal total
    road length in Primeministan? How many towns
    are there and how are the roads connected in
    this topology?

    -- Hunter W. Hagadorn, PA E'59

TODO

    meh


## Problem 5

    Start with a 10 cm square piece of paper and
    color in a 3 mm wide red border on both sides
    along and including all four edges. Fold the
    paper so that no red shows, not even a corner
    or an edge. Paper must either fully wrap
    around an edge or the edge must be recessed
    at least 3 mm from the outer perimeter of the
    folded shape (and be covered) to be
    considered hidden. If the paper has been
    folded and a new fold makes more than one
    crease, the new fold is considered as only
    one fold. What is the minimum number of folds
    required?

    -- The Man in the Milk Carton by Stephen Barr

TODO

    meh -- I have trouble even visualizing this.
    I would probably need to make a prop.


## Bonus

    Frank has a tablecloth covered with 1 cm
    diameter black polka dots. The polka dots
    form a square grid with 2 cm between
    neighboring centers. He takes a flat 4 cm by
    4 cm piece of clear plastic and paints four 1
    cm diameter circles on it in the same pattern
    as on the tablecloth. After laying the
    tablecloth on a flat surface, he randomly
    tosses the piece of plastic onto it. To three
    significant figures, what is the probability
    that none of the polka dots on the plastic
    will overlap any portion of any of the polka
    dots on the tablecloth?

    --Don A. Dechman, TX A'57

Solution:

    I am positive I solved this before and wrote
    code in many languages as a Monte-Carlo benchmark.
    
    Need to dig that out.

    See files: c:\rgc\test\play\bent\2019\0_winter\old_pc
    
    A LOT of work done on this.


## Computer Bonus

    In the range from 1 to 1,000,000,000, what is
    the longest string of consecutive integers
    without a semiprime? A semiprime is an
    integer that is the product of exactly two
    (not necessarily different) primes. As an
    example, 96 to 105 is a string of 10 integers
    without a semiprime.

    -- Howard G. McIlvried III, PA G'53

Solution:

    Brute force

    Final solution:
    
    Maximum string without semiprime = 94 integers
    First value = 981270903
    Last value  = 981270996
    
    Through n=400e6 the longest string is 86.
    The run time is very, very slow in R.
    Not sure it would be any faster in a different language.
    Still have to find the semiprimes!!!
    
    According to the Bent the solution should be 94 with the solution
    near the end of the range (~981e6).

    I let it run overnight and my results match the Bent.
    
    Preserved for posterity but will not run by default.

```{r, eval=FALSE}
library(gmp)

m=0
mmax=0
lastval=0

for (n in 1:1000000000) {
    if (length(factorize(n))==2) {
        m=0
    } else {
        m=m+1
        if (m>mmax) { mmax=m; lastval=n }
    }
    if (n%%10000000 == 0) {
        print(sprintf("Working for n = %f, mmax= %d",n/1e6,mmax))
    }
}

print(mmax)
print(lastval)

```

    Final solution:
    
    Maximum string without semiprime = 94 integers
    First value = 981270903
    Last value  = 981270996
    
