---
title: "Bent Winter 2008"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Winter 2008


## Problem 1

    Ann, Beth, and Clara have been
    challenged to find a solution to the
    following cryptic:

    FOUR is a perfect square;
    FIVE is a Fibonacci number; &
    SIX is a semi-prime (product of
    two primes).

    Each girl found a different solution,
    and the values of FOUR, FIVE, and
    SIX in Clara's solution were not used
    in Ann's or Beth's answer. What was
    Clara's solution? The usual rules
    apply. No leading zeroes, and each
    letter represents a different digit.

    -- Richard England in New Scientist

Solution: solved in the cryptic arithmetic notebook

      FOUR is a perfect square
      FIVE is a Fibonacci number
      SIX is a semi-prime (product of two prime numbers)
      
      Ann, Beth, and Clara found a different solution
      
      FOUR, FIVE, and SIX in Clara's solution not in Ann's or Beth's.
      
      What is Clara's solution?


    1. There are only two 4 digit Fibonacci numbers with 4 distinct digits:

        1597 and 2584

    2. There are 36 4 digit perfect squares with 4 distinct digits.  The
       only ones that start with the same digit as the Fibonacci numbers
       are of the form 1*** or 2***.  Furthermore, the perfect squares
       can not have any duplicate digits with the Fibonacci numbers other
       than the leading digit.  This limits the solution sets to:

        FOUR   FIVE
        -------------------------------
        1597   1024
        2584   2601 2916

    3. Three digit semi-primes: there are lots of them (265).  The middle
       digit must match the second digit of FIVE and the other digits
       can not occur in FOUR or FIVE.  This reduces the solution set
       to the following:


        FOUR   FIVE     SIX
        -------------------
        1597   1024     803  <-- Clara's solution
        2584   2601     763
        2584   2916     793

        Clara's solution must be the unique row (the first one).
        
        Can't say which one is Ann or Beth.


## Problem 2

    Consider a regular pentagram
    (five-pointed star) with a circumscribing
    circle of unit radius. Fold
    the five points of the star so that
    their tips meet in a point to form a
    pyramid with a regular pentagon as
    its base. What is the volume of this
    pyramid?

    -- Slicing Pizzas, Racing Turtles,
    and Further Adventures in Applied
    Mathematics by Robert B. Banks

Solution:

    Let's look at a picture first.
    We will draw the circle and the star and then
    figure things out:
    
```{r}
r=1
theta=seq(0,2*pi,length=200)
xo=r*cos(theta)
yo=r*sin(theta)
r2=0.382
xi=r2*cos(theta)
yi=r2*sin(theta)
plot(xo,yo,type='n',asp=1)
polygon(xo,yo)
polygon(xi,yi,border='green')
dtheta=2*pi/5
theta=seq(5)*dtheta
x=r*cos(theta)
y=r*sin(theta)
#
lines(x[c(1,3)],y[c(1,3)],col='red')
lines(x[c(1,4)],y[c(1,4)],col='red')
#
lines(x[c(2,4)],y[c(2,4)],col='red')
lines(x[c(2,5)],y[c(2,5)],col='red')
#
lines(x[c(3,5)],y[c(3,5)],col='red')

polygon(x,y,lty='dashed',border='blue')

```

    We need to work out the size of the internal pentagon and the
    length of the sides of the "points" of the star.  Once we have
    that we should be able to find the volume.
    
    A pentagon divides the circle into 5 parts so each angle is 360/5 = 72 degrees.
    The internal angle (between sides) of the pentagon are just 180-72=108 degrees.
    
    The sides of the outside pentagon connect points on the unit circle so we
    can calculate the length of a side as:
    
        L = 2*sin(36 degrees) = 1.175571
        
    The isosceles triangle with this as a base has two 36 degree base angles with
    a height of h/(L/2) = tan(36) so
    
        h = (L/2)*tan(36) = sin(36)*tan(36)
        
    So, the distance from the center of the circle to the point(s) of the internal
    pentagon are:
    
        r2 = 1*cos(36)-sin(36)*tan(36)
        
    Now that we know the radius of the interior circle we can calculate the length
    of each side of the smaller pentagon.  This is just:
    
        L2 = 2*r2*sin(36)
        
    The points of the stars have base angles of 180-108=72 degrees.  The length of
    a side of the point is then:
    
        (L2/2)/d = cos(72) --> d = (L2/2)/cos(72)
        
    The height of each star point is:
    
        hpoint = d*sin(72)
        
    The distance from the center of the circle to the middle of a side of the small
    pentagon is:
    
        d2 = r2*cos(36) = (1-sin(36)*tan(36))*cos(36)
        
    Based on d,hpoint we can calculate the height of the 5-sided pyramid when the
    points are flipped up:
    
        hp = sqrt(hpoint^2 - d2^2)
        
    So - at this point we know the base dimensions and the height of the pyramid.
    This is sufficient to calculate the volume.
    
```{r}
theta=36*pi/180
L=2*sin(theta)
h=(L/2)*tan(theta)
r2=1*cos(theta)-sin(theta)*tan(theta)
L2=2*r2*sin(theta)
d=(L2/2)/cos(2*theta)
hpoint=d*sin(2*theta)
d2=r2*cos(theta)
hp=sqrt(hpoint^2-d2^2)
# Area of the central pentagon
area=1/2*L2*d2*5
# Volume of the pyramid
V=5*L2*d2*hp/6


print(paste0("Side of outside pentagon = ",L,collapse=""))
print(paste0("Height of outer triangle = ",h,collapse=""))
print(paste0("Radius of inside circle  = ",r2,collapse=""))
print(paste0("Side of inside  pentagon = ",L2,collapse=""))
print(paste0("Height of star point     = ",hpoint,collapse=""))
print(paste0("Height = ",hp,collapse=""))
print(paste0("Area of base pentagon    = ",area,collapse=""))
print(paste0("Volume of pyramid        = ",V,collapse=""))
```

Might be able to calculate this another way.  Consider a circle
of radius=r that circumscribes a pentagon.  The ratio of the
pentagon area to the circle area is:

    area_p = 1/2*L*d*5
       d=cos(theta)
       L/2=sin(theta)
       theta = pi/5
       
    area_p = sin(theta)*cos(theta)*5
           = 2.377641
           
    area_c = pi*r^2 = pi
    
    ratio = area_p/area_c = 0.7568267

At each slice the area ratio holds so if we calculate the volume
of a cone of height hp the volume of the pentagonal pyramid should
be this ratio times that volume.  For the central circle/pentagon
given above:

    V_Cone = 1/3*pi*r2**2*hp
           = 1/3*pi*(0.38196601125010^2)*0.618033988749895
           = 0.09442574

so:

    V_pyramid = V_cone * 0.7568267
              = 0.07146392

Which is the same as above.


## Problem 3

    Find the positive root of the following
    equation:

        (x + 9)**(1/3) - (x - 9)**(1/3) = 3.

    Express your answer exactly in
    terms of surds.

    -- Mathematics Teacher

Solution:

    x = 4*sqrt(5) is the positive root.

    Visualize this for some range of x to get an idea of what
    is going on:
    
```{r}
x=seq(8,10,length=100)
y=rep(0,length(x))
for (i in seq_along(x)) {
    if (x[i]<9) {
        y[i]=(x[i]+9)^(1/3)+(9-x[i])^(1/3) - 3
    } else {
        y[i]=(x[i]+9)^(1/3)-(x[i]-9)^(1/3) - 3
    }
}
plot(x,y,type='b')
abline(h=0,col='red',lwd=3)
```

So - we expect a solution somewhere near x=9.  This is confusing due to the sign
issue for cube root of (x-9) for x<9.  We can eliminate this if we rearrange the
equation and cube it:

    (x+9)^(1/3) - (x-9)^(1/3) = 3
    
    (x+9)^(1/3) - 3 = (x-9)^(1/3)
    
    ((x+9)^(1/3) - 3)^3 = x-9
    
    (x+9) - 9 * (x+9)^(2/3) + 27 * (x+9)^(1/3) - 27 = x-9
    
    (18-27) - 9*(x+9)^(2/3) + 27*(x+9)^(1/3) = 0

    -9 - 9*(x+9)^(2/3) + 27*(x+9)^(1/3) = 0
    
    1 + (x+9)^(2/3) - 3*(x+9)^(1/3) = 0
    
    Let u = (x+9)^(1/3) then x = u^3 - 9
    
    u**2 - 3*u + 1 = 0
    
    u = (3 +/- sqrt(9-4))/2
    
    Choosing the positive root:
    
    u = (3+sqrt(5))/2
    
    So
    
    x = u^3 - 9 = ((3+sqrt(5))/2)^3 - 9
      = 4*sqrt(5)
      ~ 8.944272


## Problem 4

    In a garden is a circular pond
    with a horizontal bottom and vertical
    sides. The owner of the garden
    decides to erect a statue in the
    middle of the pond. In preparation,
    three identical cubical (solid and
    impermeable) concrete blocks are
    placed side by side in the pond to
    form a base for the statue. When the
    first block is placed in the pond, the
    water level rises three inches. When
    the second block is placed, the water
    level rises another four inches. And,
    when the third block is added, the
    water level again rises four inches.
    What is the length of the edges of
    the blocks?

    -- Almost Impossible Brain Bafflers
    by Tim Sole and Rod Marshall

Solution:

    When the cubes are placed in the water they will
    raise the water level based on their displacement.
    If the cubes were fully submerged then adding each
    cube would raise the water level by the same amount.
    Since this does not occur, we know the first cube
    is not fully submerged.  Whether or not the second
    cube is full submerged is difficult to know.  We will
    assume so and look for contradictions.
    
    A = cross sectional area of pond
    h0 = initial water height
    
    V = A*h0 -> volume of water in pond
    
    Add first block (partially submerged):
    
        A*(h0+3)-S*S*(h0+3) = A*h0   [1]
        
    Add second block (two blocks fully submerged):
    
        A*(h0+7)-2*S*S*S = A*h0   [2]
        
    Add third block (three blocks fully submerged):
    
        A*(h0+11)-3*S*S*S = A*h0   [3]
        
    From equation [2]: 7*A-2*S*S*S = 0  --> A = (2/7)*S*S*S
    From equation [3]: 11*A-3*S*S*S = 0 --> A = (3/11)*S*S*S
    
    CONTRADICTION!!!
    
    Second scenario:
    
    Add first block (partially submerged):
    
        A*(h0+3)-S*S*(h0+3) = A*h0   [1]
        
    Add second block (partially submerged):
    
        A*(h0+7)-2*S*S*(h0+7) = A*h0   [2]
        
    Add third block (three blocks fully submerged):
    
        A*(h0+11)-3*S*S*S = A*h0   [3]

    From equation [1]: 3*A-S*S*(h0+3) = 0
    From equation [2]: 7*A-2*S*S*(h0+7) = 0
    
    Combining: 1/3*(h0+3) = 2/7*(h0+7)
       7*(h0+3) = 6*(h0+7)
       h0 = 21 in

    Substituting into equations [1-3]:
    
        [1] 3*A - 24*S*S = 0  -> A = 8*S*S
        [2] 7*A - 56*S*S = 0  -> useless
        [3] 11*A - 3*S*S*S = 0 -> 88*S*S - 3*S*S*S = 0
        
        S = 88/3 in
          = 29.3333 in
        
        A = 8*S*S = 6883.556 in^2
          = 47.802 ft^2


## Problem 5

    Using a standard set of 28 dominos,
    Carl laid out in a row, in typical
    domino style with ends matching,
    e.g., 6-3, 3-5, 5-5, 5-1. He started with
    0-0 and then added dominos, building
    on either end of the string, so
    that the total number of pips in the
    row equaled 1, then 4, then 9, working
    through each perfect square and
    continuing as far as possible. His row
    could not have been shorter at any
    stage. When he reached the highest
    square possible, he was left with five
    unused dominos, four of which were
    doubles. What were those five
    dominos? List the complete order of the domino row of
    your solution.

    -- Adrian Somerfield in New Scientist

TODO

    Don't care
    

## Bonus

    In the game of Thirty Five, each player is dealt
    a hand of nine cards from a standard deck of 52 cards. To
    win a pot, one must be able to select cards in a single suit
    to obtain a score of 35, where face cards count 10, aces
    count 1, and all other cards have their designated value.
    What is the probability of getting a hand that can win the
    pot?

    -- Howard G. McIlvried III, PA Gamma'53

Solution: try Monte-Carlo

```{r}
library(gtools)

is_winner = function(h) {
    for (r in 4:length(h)) {
        a=combinations(length(h),r,h,set=F)
        if (any(apply(a,1,sum)==35)) {
            return(TRUE)
        }
    }
    return(FALSE)
}

pipval=c(1,2,3,4,5,6,7,8,9,10,10,10,10)
cards=0:51

ntrials=1000000

set.seed(2020)
win=0
for (trial in 1:ntrials) {
    s=sample(cards,9)
    winner=0
    for (i in 0:3) {
        h=pipval[s[trunc(s/13)==i]%%13+1]
        if (sum(h)>=35) { 
            if (is_winner(h)) {
                winner=1
            }
        }
    }
    win=win+winner
}
print(win)
print(win/ntrials)
```

Probability ~ 3.9113% (1e6 trials)

    Pr         Trials    Seed
    0.038966   1e6       42
    0.039335   1e6       123
    0.039265   1e6       2020
    
According to the closed form counting in the Bent:

    Pr = 0.0391
    
    (143784024/3679075400)

After looking at the solution I don't think counting is worth it.
Random sampling is good enough for me.


## Double Bonus

    In 2006, the age of a famous composer
    would be, if he were still alive, equal to 10 times 10 less
    than his age when he died. His age at death equals 2006
    minus the reverse of the year he died. Who is this composer?

    -- Dr. Aziz S. Inan, CA Epsilon'78

Solution:

    b = year of birth
    d = year of death = wxyz
    a = age at death
    
    (2006-b) = 10*(d-b-10)
    
    (d-b) = 2006-zyxw
    
    Solve first equation for b: 2006-10*d+100 = -9*b -> b = (10*d-2106)/9
    substituting in the second equation:
    
    (d-(10*d-2106)/9) = 2006-zyxw
    
    (9*d-10*d+2106)/9 = 2006-zyxw
    
    2106-d = 9*(2006-zyxw)
    
    2106-wxyz = 9*(2006-zyxw)
    
    By exhaustive elimination:
    
    d = wxyz = 1791 (year of death)
    b = 1756
    age at death = 35
    
    Famous Composer = Mozart
