---
title: "Bent Fall 2010"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Fall 2010


## Problem 1

    Ann, Barb, and Cleo are three
    remarkable women: two are remarkably
    intelligent, two are remarkably
    beautiful, two are remarkably artistic,
    and two are remarkably rich. Each
    has no more than three remarkable
    characteristics. If Ann is remarkably
    intelligent, she is remarkably rich. If
    Barb or Cleo is remarkably beautiful,
    she is remarkably artistic. If Ann or
    Cleo is remarkably rich, she is remarkably
    artistic. Which woman is not
    remarkably rich?

    -- Test Your Logic by George J. Summers

TODO

    meh - logic
    
    
## Problem 2

    A researcher is interested in studying
    cheating at his alma mater, but he feels
    that if he asks students directly if they
    cheat, he will not get an honest answer.
    Therefore, he decides to use the following
    procedure. He places three cards
    in an opaque bag. The first card reads,
    "Have you cheated on a test in the past
    year?"; the second card reads, "Answer
    Yes," and the third card reads, "Answer
    No." He hands the bag to a student, asks
    him to select a card at random from the
    bag, answer Yes or No, and replace the
    card in the bag, all without letting the
    researcher see the card. The researcher
    interviewed 2,352 students and received
    928 yeses and 1,424 nos. Assuming that
    all of the students either answer truthfully
    or answer yes or no when directed,
    what is the estimated percentage of
    students who cheat?

    -- Why Do Buses Come in Threes?
    by Rob Eastaway & Jeremy Wyndham

Solution:

    If we get an even sample then 1/3 will answer
    yes as instructed, 1/3 will answer no as instructed,
    and 1/3 will answer the real question.
    
    So:  the forced responses are 784=yes, 784=no.
    The remainder are 784=honest
    
    honest responses are: 928-784 = 144 yes, 1424-784=640 no
    
    Estimated percentage of cheaters = 144/784 = 9/49 ~ 0.184 = 18.4%


## Problem 3

    A banker agrees to lend you a hundred
    million dollars at an interest rate
    of 100%. If the compounding period for
    the interest is yearly, then you will owe
    the banker two hundred million dollars
    after one year (assuming that you make
    no other payments). However, if the
    interest rate is compounded monthly
    at an interest rate of 1/12 of 100%, then
    you will owe $261,303,529.02 after one
    year; if the interest is compounded
    weekly at a rate of 1/52 of 100%, you
    will owe $269,259,695.44. How much
    will you owe, to the nearest penny, after
    one year, if the interest is compounded
    continuously?

    -- John L. Bradshaw, PA A'82

Solution:

    This is a relatively well-known result.  Continuous
    compounding is exponential:
    
    total = loan * exp(i)
          ~ 2.718281828459045 * loan

    This is the limiting case of (1+i/n)^n as n-->infinity

    I will leave the rounding to the nearest penny as an exercise
    but it should yield 271828182.85


## Problem 4

    Three spheres of radii 2, 3, and 4
    cm lie on a horizontal plane, tangent
    to each other. A fourth sphere of radius
    1 cm is placed in the pocket formed by
    the other three. What is the distance of
    the center of the smallest sphere from
    the plane surface?

    -- Technology Review

TODO

    This seems like a devilishly simple problem
    that makes my head hurt.  I should definitely
    work on this.
    

## Problem 5

    A young girl is riding in a swing with
    chains 8 ft in length and a seat 2 ft above
    the ground. She wants to jump off and land
    as far from the swing's rest position as
    possible. At what angle of the swing (the
    angle is 0 degrees  when the chains are
    vertical) should she release from the seat,
    and how far from the point directly under
    the swing's rest position can she land?
    Treat the girl as a point mass, and assume
    that the highest she can pump the swing is
    45 degrees with respect to vertical.

    -- Towing Icebergs, Falling Dominoes, and
    Other Adventures in Applied Mathematics by
    Robert B.  Banks

Solution:

    This seems like a fun problem.  Need to trade
    potential and kinetic energy along with ballistic
    motion.


## Bonus

    A football punter, standing at the middle of
    the 50-yard line, wishes to make a "coffin
    corner" kick. That is, to kick the ball out
    of bounds as close to the goal line as
    possible, but not into the end zone. If
    successful, the opposing team will get the
    ball at the yard line where it goes out of
    bounds; if unsuccessful, it will get the
    ball at the 20-yard line. Assume the ball
    travels in a straight line, has enough
    velocity to cross the sideline or goal line,
    and there is no run-back. If the kicker's
    launch angle has a normal distribution with
    a standard deviation of 7.5 degrees from
    where he aims, at what sideline yard marker
    should the punter aim so that the expected
    value of the opposing team's starting
    position is minimized? What is that expected
    value (in yards from their goal line)? A
    football field, not counting the end zones,
    is 300 feet by 160 feet.

    -- Steven E. Bradley, AZ A'77

Solution:

    Let's use Monte Carlo
    
    A measure the angle of the kick as the angle relative to the center
    of the field.  For a very shallow angle the ball will go into the
    end zone and come out to the 20.  If the angle is "steep enough" it
    will go out of bounds.
    
```{r}
num_kicks=1000000       # Number of kicks at each target yardline
kasd=7.5*pi/180         # Kick angle standard deviation in radians
width=80/3              # Width of half-field: kicker to the sideline

acrit=atan(width/50)    # Critical kick angle.  If the kick angle is less than
                        # this value it will go into the end-zone.

set.seed(123)           # reproducibility

yardlist=seq(0,15,0.2)  # Kick target yardlines
x=c()                   # Mean response

for (target in yardlist) {
    theta = atan(width/(50-target))
    ka=rnorm(num_kicks,theta,kasd)
    y=ifelse(ka<acrit,20,50-width/tan(ka))
    x=c(x,mean(y))
}
plot(yardlist,x,type='b',xlab="Target",ylab="Mean Start",main="Coffin Corner Kick")
grid()
```

```{r}
print(paste0("Optimal starting yardline = ",min(x)))
print(paste0("Optimal target yardline   = ",yardlist[which.min(x)]))
```

The optimal target is:

    Aim for ~6 yardline
    Mean starting result is 14.4 yardline

Of course we could also do this analytically but this was more fun.


## Computer Bonus

    The smallest palindromic square with an even
    number of digits is 698,896 (the square of
    836).  What are the next two smallest
    palindromic squares with an even number of
    digits?

    -- The Colossal Book of Mathematics by
    Martin Gardner

Solution:

    We should be able to solve this with exhaustive search.
    We will simply start counting up from 837, calculate
    the square and check to see if it is a palindrome.
    To ensure this is valid for large numbers we can use
    gmp for wide integers or use reticulate/python.
    
    Let's try gmp first:

```{r}
library(gmp)

is_palindrome = function(n) {
    s1=as.character(as.bigz(n))
    s2=paste0(rev(strsplit(s1,"")[[1]]),collapse="")
    if (s1==s2) {
        return(TRUE)
    }
    return(FALSE)
}

# Comment this out unless you really want to run it.
# It takes about 40 minutes to run!
if (FALSE) {
    t1=Sys.time()
    curr_target=0
    n=1
    num_palindromes=0
    while(1) {
        m=as.bigz(n)^2
        if (nchar(as.character(m))%%2 == 0) {
            if (nchar(as.character(m)) != curr_target) {
                curr_target=nchar(as.character(m))
                print(paste0("Searching num digits = ",curr_target))
            }
            if (is_palindrome(m)) {
                print(n)
                print(paste0("Time = ",Sys.time()-t1))
                num_palindromes=num_palindromes+1
                if (num_palindromes==3) {
                    break
                }
            }
        }
        n=n+1
    }
} else {
    print("No output produced since run-time is ~ 40 minutes")
}

```

The first two are easy:

    [1] 836     ( 0.54 seconds)
    [1] 798644  (28.36 seconds)

but then it gets more difficult.  No solution as of n=59908778.
Finally found it:

    [1] 64030648  (2337.56 seconds)


Repeated the analysis with python and it was much faster:

    836         836**2      =           698896  ( 6 digits), t= 0.00 seconds
    798644      798644**2   =     637832238736  (12 digits), t= 0.83 seconds
    64030648    64030648**2 = 4099923883299904  (16 digits), t=70.42 seconds
    
This is about 33X faster than R with gmp.

