---
title: "Bent Spring 2012"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Spring 2012


## Problem 1

    Solve the following cryptarithm: 
    
    O/NE + T/WO + S/IX = NI/NE, where / indicates division.

    As expected, TWO is divisible by 2, SIX is
    divisible by 6, and NINE is divisible by 9,
    but oddly TEN is divisible by 7. Each
    different letter represents a different
    digit, the same letter represents the same
    digit, and there are no leading zeros.

    -- H. G. McIlvried, III, PA Gamma '53

Solution: solved in the crypti arithmetic notebook

  Cyptic fractions:

      O/NE + T/WO + S/IX = NI/NE

      TWO is divisible by 2
      SIX is divisible by 6
      NINE is divisible by 9
      TEN is divisible by 7

  Putting over a common demoninator:

      O * WO*IX + T * NE*IX + S * NE*WO = NI * WO*IX

    a=make_crypt_df('onetwosixnine')
    x_o=make_num(a,'o')
    x_ne=make_num(a,'ne')
    x_t=make_num(a,'t')
    x_wo=make_num(a,'wo')
    x_s=make_num(a,'s')
    x_ix=make_num(a,'ix')
    x_ni=make_num(a,'ni')
    k=(x_o*x_wo*x_ix + x_t*x_ne*x_ix + x_s*x_ne*x_wo == x_ni*x_wo*x_ix)
    k=k & (a$o != 0)
        & (a$n != 0)
        & (a$t != 0)
        & (a$w != 0)
        & (a$s != 0)
        & (a$i != 0)
    k=k & (make_num(a,'two') %% 2 == 0)
    k=k & (make_num(a,'six') %% 6 == 0)
    k=k & (make_num(a,'nine') %% 9 == 0)
    k=k & (make_num(a,'ten') %% 7 == 0)
    sum(k)
    [1] 1
    > a[k,]
            o n e t w s i x
    1138194 6 2 4 7 5 9 1 8

    So:

      O/NE + T/WO + S/IX = NI/NE
   
      6/24 + 7/56 + 9/18 = 21/24

      2/8  + 1/8  + 1/2  = 7/8 


## Problem 2

    How many permutations of the integers 1
    through N consist of a strictly ascending
    sequence followed by a strictly descending
    sequence. For example, for N = 9, one such
    permutation is (1-4-5-7-9-8-6-3-2). There
    must be at least two integers in a sequence,
    and N is considered to be a member of both
    sequences. Reversals are not considered to be
    different permutations, i.e.
    (2-3-6-8-9-7-5-4-1) is the same as the above
    example.

    -- Puzzle Corner in Technology Review

TODO

    meh

## Problem 3

    Find a non-trivial (x, y, z != 0) integral
    solution of the following cubic equation:

    987,654,321x + 123,456,789y + z = (x + y + z)^3

    We want the solution with the smallest
    positive value for (x + y + z); however, this
    does not imply that x, y, and z are all
    positive.

    -- The Contest Center

TODO

    I peeked at the Bent -- still don't really understand
    the solution.  They note that the key is to realize the
    big coefficients are nearly a ratio of 8.  I saw that but
    I didn't go any further.
    
    See the Bent for solution details.


## Problem 4

    In Green Town, half the farmland must be
    maintained in a natural state by law. To
    achieve this, in each field an inside strip
    around the four sides must be left
    uncultivated to provide habitat for wildlife.
    All fields are rectangles of the same size,
    100m wide and an integral number of meters
    long. These border strips, which equal half
    the field's area, are a uniform and integral
    number of meters wide. It turns out that the
    digits in the length, when read from left to
    right, are in monotonically increasing order.
    What is the length of a field and the width
    of the border strip?

    -- Enigma by Keith Austin in New Scientist

Solution:

    Let L = length of the field (including the border)
        W = width of the border
        
    Area of field = (100-2*W)*(L-2*W)
    Area of border = L*W*2+(100-2*W)*W*2
    
    We want these two areas to be equal:
    
    L = 2*W*(W-50)/(W-25)
    
    For W,L = integers we want 2*W*(W-50)/(W-25) to be an integer
    
    A quick search for integer W yields the following possibilities:
    
    W   L
    =======
    15  105
    20  240
    23  621
    24  1248 --> digits of length in ascending order
    
    The areas of the field and border are:
    
    A(field)  = (1248-2*24)*(100-2*24) = 62400 m^2
    A(border) = 1248*100-A(field) = 62400 m^2


## Problem 5

    To increase his probability of
    throwing a 7, a gambler loaded a pair
    of dice as show in the table.

    Pip Value   Die 1   Die 2
    --------------------------
    1/2         1/4     1/12
    3/4         1/6     1/6
    5/6         1/12    1/4

    One day, he accidentally grabbed one
    standard die and one loaded die. What
    was his probability of throwing a 7 with
    one loaded die and one standard die?

    -- Mathematics Teacher

Solution:

    We should be able to do this without the computer.
    
    For the fair die the probability of each value = 1/6
    
    Assume we have one fair die and die 1.  The combinations
    that will produce 7 are:
    
    Fair   Die 1   Pr
    -----------------------
    6      1       1/6*1/4
    5      2       1/6*1/4
    4      3       1/6*1/6
    3      4       1/6*1/6
    2      5       1/6*1/12
    1      6       1/6*1/12
    -----------------------
           Total   1/6*(1) = 1/6
           
    Using Die 2 we get the same result.
    
    In general, we get the same result as with two fair dice.

    This is actually very surprising.  The loading of the second
    die has absolutely no effect on the ability to roll a 7.  It
    does, however, affect other point totals:
    
```{r}

s=outer(1:6,1:6,"+")
pfair=outer(rep(1/6,6),rep(1/6,6),"*")                  # Pr with fair die
p1=outer(rep(1/6,6),c(1/4,1/4,1/6,1/6,1/12,1/12),"*")   # Pr with die 1
p2=outer(rep(1/6,6),c(1/12,1/12,1/6,1/6,1/4,1/4),"*")   # Pr with die 2
p3=outer(c(1/4,1/4,1/6,1/6,1/12,1/12),c(1/12,1/12,1/6,1/6,1/4,1/4))   # Pr both loaded

total=unique(sort(s))
prfair=c()
pr1=c()
pr2=c()
pr3=c()
for (i in total) {
    prfair=append(prfair,sum(pfair[s==i]))
    pr1=append(pr1,sum(p1[s==i]))
    pr2=append(pr2,sum(p2[s==i]))
    pr3=append(pr3,sum(p3[s==i]))
}
a=data.frame(total=total,Pr_fair=prfair,Pr_1=pr1,Pr_2=pr2,Pr_3=pr3)
a
```

## Bonus

    Arrange the integers 1 through
    32 in a circle such that the sum of each
    pair of adjacent integers is always
    the square of an integer. Present
    your answer as a string of 33 integers
    starting and ending with 1.

    -- Puzzles 101: A Puzzle
    Master's Challenge by
    Nobuyuki Yoshigahara

TODO

    Just trial and error.
    

## Computer Bonus

    Take any positive
    integer and apply the following
    algorithm: (a) if the number is odd,
    multiply by 3 and add 1; and (b) if the
    number is even, divide by 2. If you
    continue this process, you will always
    end up with 1, no matter what number
    you start with. For example, 17
    requires 12 steps (17-52-26-13-40-20-10-
    5-16-8-4-2-1). In the range of 1 to 10,000,
    which integer requires the most steps
    and how many steps are required?

    -- Mathematical Amazements
    and Surprises by Alfred S.
    Posamentier and Ingmar Lehmann

Solution:

    Looks like it is time for exhaustive search.
    
```{r}
count_steps = function(n) {
    m=0
    while(T) {
        m=m+1
        if (n%%2 == 1) {
            n=3*n+1
        } else {
            n=n/2
        }
        if (n==1) { break }
    }
    return(m)
}

nmax=0
nmoves=0

for (i in 1:10000) {
    m=count_steps(i)
    if (m>nmoves) {
        nmoves=m
        nmax=i
    }
}
print(paste0("Maximum number of moves = ",nmoves))
print(paste0("Value for max moves = ",nmax))
```

Run time is very fast for 10K:

    For numbers up to 10K:
        Maximum number of moves = 261
        Value for max moves = 6171

    For numbers up to 1e6:
        Maximum number of moves = 524
        Value for max moves = 837799

    I would agree that this is somewhat surprising.
