---
title: "Bent Spring 2015"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Spring 2015


## Problem 1

    We start with a special Tickler.
    Tau Beta Pi is launching a new ini-
    tiative, the Chapter Endowment
    Initiative, and in recognition of that
    effort, here is a new cryptic created
    specifically to call attention to the
    initiative. Solve the following addi-
    tion/subtraction cryptic:

    DONATE + MONEY + TO + THE +
    NEW = ENDOW + MENT.

    Of course, we want the answer with
    the largest ENDOWMENT. The
    usual rules apply. Each different let-
    ter stands for a different digit, and
    each different digit is always rep-
    resented by the same letter; there
    are no leading zeros. This cryptic is
    a little trickier than usual, but it can
    be solved without a computer. Give
    it a try.

    -- Howard G. McIlvried III, Ph.D.,
    PA Gamma '53

Solution: from the cryptic arithmetic notebook

    Cryptic addition with constraint

      DONATE - MONEY - TO - THE + NEW = ENDOW + MENT

      Maximum value of ENDOWMENT

    a=make_crypt_df('donatemoneytothenewendowment')
    x1=make_num(a,'donate')-make_num(a,'money')-make_num(a,'to')-make_num(a,'the$
    x2=make_num(a,'endow')+make_num(a,'ment')
    k=(x1==x2)
    k=k&(a$d!=0)&(a$m!=0)&(a$t!=0)&(a$n!=0)&(a$e!=0)
    sum(k)
    [1] 2
    make_num(a,'endowment')[k,]
    [1] 481267483 491267493
    a[k,]
           d o n a t e m y h w
    433454 1 2 8 0 3 4 7 5 9 6
    438494 1 2 9 0 3 4 7 5 8 6

    The second solution has maximum ENDOWMENT


## Problem 2

    If Bill's age in years is appended
    to Alice's age, the resultant four
    digit integer is a perfect square.
    The same thing will be true thirteen
    years from now. What are Bill's and
    Alice's ages?

    -- Mathematics Teacher

Solution:

    Although not stated anywhere we will assume
    the Bill and Alice both have ages of 2 digits.
    We should be able to find this via exhaustive
    search:
    
```{r}

is_square = function(n) {
    m=round(sqrt(n))
    return(n == m^2)
}

for (bill in 1:99) {
    for (alice in 10:999) {
        total=alice*100+bill
        if (total>9999) { next }
        if (is_square(total)) {
            # print(paste0("Possible = ",alice,"  ",bill,"   ",total,"   ",total+13))
            total = (alice+13)*100+(bill+13)
            if (is_square(total)) {
                print(c(alice,bill))
            }
        }
    }
}

```

Alice is 19, Bill is 36.

In 13 years: Alice is 32, Bill is 49.


## Problem 3

    Three married couples, Al and
    Xenia, Bill and Yvette, and Carl and
    Zoe, are out for a hike when they
    come to a river. The good news is
    they find a boat; the bad news is
    the boat holds, at most, two people
    at a time. The problem is that the
    men are very amorous and will kiss
    any wife who is separated from her
    husband. The husbands are also
    very jealous, so how can the three
    couples cross the river without any
    illicit kissing occurring? Present
    your answer as a sequential list of
    the occupants of the boat as it makes
    multiple crossings of the river. To
    make grading easier, when there
    is an option for whom to put in the
    boat, choose the one or two eligible
    persons whose names come first
    alphabetically.

    -- The Chicken from Minsk by Yuri
    Chernyak and Robert Rose

TODO

    meh


## Problem 4

    You have a bag containing five
    marbles - three are red and two are
    blue. You agree to share them with
    your friends. One friend reaches into
    the bag, grabs a marble and leaves
    without your seeing the color. You
    then reach in and remove a marble
    and see that it is blue. What is the
    probability that the marble your
    friend took is also blue?

    -- Marilyn Savant
    in Parade Magazine

Solution:

    Exhaustive list?
    
    Pretty quick: 2 chances in 8.
    
    Analytically:
    
    If you choose Blue then of the remaining 4 marbles there
    are three red and one blue.  The chance that your friend
    took blue is 1 in 4.

    Seems way too easy.
    
    The Bent description of the solution is mind-boggling complex.


## Problem 5

    The new One World Trade Center
    (OWTC) building has an unusual
    shape -- it has a square base and a
    same-size square top, rotated 45 o rel-
    ative to the base, with sides consist-
    ing of eight isosceles triangles which
    extend the entire height of the build-
    ing. (This shape is called a square
    antiprism.) Thus, a horizontal cross
    section of the building at any height
    is an octagon, although only the
    cross section halfway up is a regular
    octagon. Let us idealize the situa-
    tion and suppose that the OWTC has
    exactly 100 evenly spaced stories
    with the base square being the floor
    of the first story and the top square
    being the ceiling of the 100th story.
    Assuming that the entire floor space
    of each story is usable, what is the
    ratio R of the total floor space of the
    100 stories to the area of the base?
    (The ceiling of the 100th story is
    not considered to be floor space.)
    Now, having considered a specific
    case, consider the general case of a
    building with N stories, and find a
    simple expression for R in terms of
    N, that is express R as a function of
    N. (Remember, the ceiling of the top
    story is not floor space.) An expres-
    sion involving a summation is not
    acceptable.

    -- Robert N. Spong, UT A '58

Solution:

    I think the bottom of the ground floor
    is a square.  Each side of the square is
    the base of an isosceles triangle and each
    corner is the tip of an inverted isosceles
    triangle.  At the ceiling of the top floor
    the roles have reversed and the ceiling is
    rotated 45 degrees wrt the base of the ground
    floor.
    
    The side of a full size square is S.
    
    Except for the bottom of the ground floor, the
    floor of each story has 8 sides.  The right-side
    up triangles are getting smaller and the upside
    down triangles are getting larger.

    The width of the rightside up triangles is
    given by:
    
        W1 = (101-N)/100*S, N=1, 2, 3, ..., 100
        
    The width of the upside down triangles is
    given by:
    
        W2 = (N-1)/100*S, N=1, 2, 3, ..., 100
        
    The W2 segments are the 45 degree corners of
    each octagon.  The total floor space is given
    by:
    
        A = (W1+2*W2/sqrt(2))^2 - 2*(W2/sqrt(2))^2

    If we substitute for W1 and W2 and then sum for
    N=1, 2, ..., 100 get the ratio against S^2 as:
    
        ratio = (3333*sqrt(2)+6667)/100 ~ 113.8157
        
    Generalizing for N story building:
    
        ratio = (sqrt(2)*N^2 + 2*N^2 - sqrt(2) + 1) / (3*N)
        
    The Bent solution is:
    
        ratio = N + k*(N^2-1)/(6*N) where k=2*(sqrt(2)-1)
        
    Which is the same.
    
    Thank maxima for all the intermediate work.


## Bonus

    Find a 4x4 magic square
    consisting of 16 different posi-
    tive semiprimes less than 100. A
    semiprime is a positive integer
    with exactly two, not necessar-
    ily different, prime factors. (For
    example, both 21 = 3x7 and 25 = 5^2
    are semiprimes.) In a magic square,
    the entries in each column, row, and
    major diagonal have the same sum
    (known as the magic sum). We want
    a magic square for which, in addi-
    tion, the four center entries, the four
    corners, and each of the four quad-
    rants also exhibit the same magic
    sum, making a total of 16 ways to get
    the magic sum. We want the square
    with the smallest magic sum for such
    a square. Present your answer as
    four rows of four integers each.

    -- Howard G. McIlvried III, Ph.D.,
    PA Gamma '53

TODO

    I keep saying I should look into magic squares and
    I never do.


## Computer Bonus

    Let S(N) equal the sum of all the divisors of
    N that are less than N. For example, S(8) = 1
    + 2 + 4 = 7, and S(30) = 1 + 2 + 3 + 5 + 6 +
    10 + 15 = 42. Define S =^2 (N) = S(S(N)) and 
    S^3 (N) = S(S^2 (N)) = S(S(S(N))), or in
    general, S^n (N) = S(S^(n-1) (N)). Call N a
    sociable number of degree n if S^n (N) = N
    and there is no m < n such that S^m (N) = N.
    An example of a sociable number of degree 5
    is 12,496, i.e., S^5 (12,496) = 12,496
    (12,496 -> 14,288 -> 15,472 ->  14,536 -> 14,264
    -> 12,496). Find the smallest sociable number
    of degree 28.

    -- Madachy's Mathematical
    Recreations by Joseph S. Madachy

Solution:

```{r}
library(gmp)

divsum=function(n) {
    if (n<2) { return(0) }
    r=rle(as.numeric(factorize(n)))
    p=1
    for (i in seq_along(r$values)) {
        p=p*(r$values[i]^(r$lengths[i]+1)-1)/(r$values[i]-1)
    }
    p=p-n
    return(p)
}

is_sociable=function(n,m) {
    s=n
    for (i in 1:(m-1)) {
        s=divsum(s)
        if (s==n) { return(FALSE) }
    }
    return(n==divsum(s))
}
```

Now try to find the smallest sociable number of degree 28:

```{r}

n=2
while (TRUE) {
    if (is_sociable(n,28)) {
        print(n)
        break
    }
    n=n+1
    if ( (n%%10000) == 0) { print(sprintf("Working, n = %d",n)) }
}

```

The smallest sociable number of degree 28 is 14316.

This was much smaller than I anticipated (feared?).
