---
title: "Bent Fall 2003"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Fall 2003


## Problem 1

    It was Mycroft's wont to send
    Sherlock little problems by telegram
    from time to time. This morning's
    read: "Anstey, Buchan, Collins, and
    Doyle played a 3-round singles tournament
    (2 matches per round). They
    finished in alphabetical order by
    name. Scoring was win 2, draw 1, loss
    0. Lion tamer won, and Musician drew
    in the first round. Neurologist won in
    last round. Total number of draws
    was. What is name and final score of
    Ornithologist?"
    "But Holmes," I protested, "he has
    omitted the total number of draws."
    "Or else the post office has," remarked
    my friend with a flash of his
    piercing intellect.
    "So the problem must be unsolvable."
    "By no means, my good Watson.
    He plainly intended to put it in. Or, if
    he did not, he plainly intended us to
    think he intended to put it in. And, as
    miserliness restrains him from
    telegraphing needless data, whereas
    pride prevents him sending insufficient
    data, it is as if he had sent it."
    What is the answer to Mycroft's
    question?
    
    -- Martin Hollis

Note: After peeking at their solution -- apparently I have no
understanding of the match play described here.  Apparently each
player only plays three times -- I don't understand the terms
match/round above.

If I rework the problem with only three "games" each:

    A>B>C>D
    A+B+C+D=12
    Each score <= 6
    
    Possibilities are:
    
```{r}
zzz=data.frame(a=NA,b=NA,c=NA,d=NA)
for (a in 0:6) {
    for (b in 0:6) {
        for (c in 0:6) {
            for (d in 0:6) {
                if ((a+b+c+d) == 12) {
                    if ((a>b) & (b>c) & (c>d)) {
                        zzz=rbind(zzz,c(a,b,c,d))
                    }
                }
            }
        }
    }
}
zzz=zzz[!is.na(zzz$a),]
zzz
```

So the only possibilties are:

    a   b   c   d
    -------------
    5   4   2   1
    5   4   3   0
    6   3   2   1
    6   4   2   0
    6   5   1   0

That is the easy part -- still can't reduce any further.

As noted in the Bent:

    A either wins all three or wins two and ties one
    D either loses all three or loses two and ties one
    
    Investigation of the
    possible outcomes with these restraints
    gives multiple possible
    outcomes with one or two ties, no
    possible outcomes for four or more
    ties, and only one possible outcome for
    three ties. Namely, A and B tie,
    C and D tie, A beats C, A beats D,
    B and C tie, B beats D, and C and D
    tie. So, the first round was A vs. D
    and B vs. C, and A is the lion tamer
    and B or C is the musician. The last
    round was A vs. C and B vs. D, and
    B is the neurologist. So, Doyle is the
    ornithologist and managed only a tie.

FWIW: this problem is nuts.

My original Solution attempt - everything wrong by factor of 2X.
Let us consider:

    A = Anstey
    B = Buchan
    C = Collins
    D = Doyle
    
    Lion Tamer: 2 in first round
    Musician:   1 in first round
    Neurologist: 2 in final round
    Ornithologist: ???

    Each person plays 6 matches (2 matches * 3 rounds) - 2 points awarded in each match
    Total of 12 matches
    Must be an even number of draws
    Must be an equal number of wins/losses
    Each person plays each other person 2 times (?)
    Each person has distinct score
    Max score for any player is 12
    Grand total = 24 points (Total of 4 points per match) by all the players
    
        A+B+C+D = 24
        A>B>C>D
        A<=12, B<=12, C<=12, D<=12
    
Try to enumerate - still have 43 feasible solutions

```{r}
zzz=data.frame(a=NA,b=NA,c=NA,d=NA)
for (a in 0:12) {
    for (b in 0:12) {
        for (c in 0:12) {
            for (d in 0:12) {
                if ( (a+b+c+d) == 24 ) {
                    if ((a>b) & (b>c) & (c>d)) {
                        zzz=rbind(zzz,c(a,b,c,d))
                    }
                }
            }
        }
    }
}
zzz=zzz[!is.na(zzz$a),]
nrow(zzz)
```


## Problem 2

    Three friends, Ann, Betty, and
    Carol, meet to celebrate the birthday
    of one of them. "How old are you now,
    Ann?" asks Betty. "Well," says Ann,
    "using our exact (not rounded) ages,
    the sum of our three ages is 80 years.
    I am twice as old as Carol, and at the
    time when I become twice as old as
    you, our three combined ages will be
    half again as large as now." The question
    is, whose birthday is it?
    
    -- The Platinic Corner

Solution:

    A+B+C = 80
    A = 2*C
    
    (A+k) = 2*(B+k) for some k
    then:
    
    A+k+B+k+C+k = 120 --> k=40/3
    
    so:
    
    (A+40/3) = 2*(B+40/3)
    A = 2*B + 40/3 --> B = (A-40/3)/2
    
    A+(A-40/3)/2+A/2 = 80
    2A - 20/3 = 80
    2A = 80+20/3
    A = 40+10/3 --> A=130/3, C=130/6, B=(130/3-40/3)/2=15
    
    Since the only "exact" age is B we conclude that
    it must be the birthday for Betty.


## Problem 3

    Al and Ben play a game involving a
    series of coin tosses. Each gambler
    picks a different series of three possible
    outcomes, and the winner is the
    one whose chosen series of outcomes
    first occurs in the string of coin tosses.
    Al announces that his series is HHH.
    If Ben makes an optimal choice for his
    series, what is his probability of winning?
    
    -- nearly impossible Brain Bafflers
    by Tim Sole and Rod Marshall

TODO


## Problem 4

    How fast, that is, how many rotations
    per year, would the Earth have
    to rotate in order for a 75 kg person at
    the equator to lift off the ground?
    Ignore atmospheric effects.
    
    -- Walter O. Stadlin, NJ Gamma'52

Solution:

    Centripetal acceleration = attraction due to gravity
    Note that mass does not matter.
    
    v^2/r = g --> v = sqrt(r*g)
    
    r ~ 6378 km, g ~ 9.8 m/s^2 --> v = 7905.972 m/s
    
    Angular velocity = v/r = 0.001239569 radians/sec
        = 17.0453 rotations per day


## Problem 5

    A right-circular cone of height H
    and radius R is rolled on a plane so
    that the apex remains at a fixed point.
    How many times will the cone revolve
    about its axis if the cone is rolled
    through a complete circle on the
    plane?
    
    -- Daryl Cooper

Solution:

    Consider a point on the edge of the base of the cone.
    When you lay the cone on the flat surface it will roll
    in a circle of radius = sqrt(H*H+R*R).  The outside
    edge of the cone will have to roll along the perimeter
    of this circle.  The number of revolutions is then:
    
    n = ( 2*pi*sqrt(H*H+R*R) ) / ( 2*pi*R )
      = sqrt(H*H+R*R)/R
      = sqrt((H/R)**2 + 1)


## Bonus

    You have a large supply of
    one-ohm resistors and some wire, and
    you wish to create a "black box," i.e., a
    box with only several terminals showing,
    which will allow you to achieve
    any integral resistance from 1
    through 10 ohms, inclusive, by proper
    choice of two of the terminals. How
    should such a box be wired (once
    wired, the wiring is fixed) if the number
    of terminals is a minimum? Using
    some wire to short terminals is not
    allowed. Remember, resistance can
    be reduced by wiring resistors in
    parallel. The solution is relatively
    simple and does not require any complicated
    wiring scheme.
    
    -- Hubert W. Hagadorn, PA E'59

TODO


## Computer Bonus

    For all values of
    N greater than 4, the decimal
    representation of N! (N factorial)
    ends with a string of zeros. We are
    interested in the digit preceding these
    zeros. What is the value of the nonzero
    digit immediately preceding the
    string of zeros at the end of
    1,000,000!?
    
    -- Howard G. McIlvried III, PA Gamma'53

Solution:

With computers/software in 2020 this is actually pretty easy.

The last character before the string of zeros is 4.
The final nonzero portion of the factorial is ...65058412544000...

This is digit 5315711 in the factorial.  There are
nearly 250000 zeros at the end!

```{r}
library(gmp)
library(stringi)

a=as.character(factorialZ(1000000))
for (i in nchar(a):1) {
    if (stri_sub(a,i,i) != "0") {
        print(paste0("Char index = ",i,"  Last Digit = ",stri_sub(a,i,i),collapse=''))
        print(paste0("Last nonzero portion of factorial = ",stri_sub(a,i-10,i+5),collapse=''))
        break
    }
}
```

