---
title: "Bent Summer 2000"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Summer 2000

## Problem 1

    Mr. Meek is pleased with his new
    extension phone number because it
    has four digits, the middle two of which
    are identical. "Like my name," he explains.
    The repeated digit is also the
    first digit of Mr. Humble's new four-digit
    number. Moreover, Meek's first
    digit is the same as the first digit of
    Mr. Lowly's new four-digit number.
    If you interchange the first and last
    digits of Lowly's number, you get
    Humble's. If you subtract Lowly's number
    from Humble's, you get Meek's. So,
    what is Meek's new number?
    
    -- New Scientist

Manual Solution:  try to solve this by hand first

    meek   = n m m *
    humble = m a b n
    lowly  = n a b m

      m a b n
    - n a b m
      -------
      n m m *

    If we assume all phone numbers are unique, then m and n must be different.
    
    Since phone numbers are not negative we can conclude that the number
    for lowly is larger than the number for humble.  In particular, we know
    the first digit in lowly (m) must exceed the first digit in humble (n).
    They can't be the same from the previous observation.  Thus m >= 1.
    
    In the subtraction, we have the last digit to consider.  We know
    that m > n so this must result in a borrow.  The second to last digit
    computation is then:
    
        (b-1) - b = m + 10 --> m = 9
        
    The second digit is the same calculation so there is a borrow out of the
    left-most digit.  This means:
    
        (m-1) - n = n --> n = 4
        
    From the right-most digit -- we now know:
    
        meek right-most digit = n - m --> 5 with a borrow of 1
    
    Making these substitutions:
    
    meek   = 4 9 9 5
    humble = 9 a b 4
    lowly  = 4 a b 9

    I do not see any way to reduce this further.  It appears a, b may be any
    digit (even the same one).  If we did an exhaustive search I suspect we
    would find 100 solution sets.

Computer Solution: seems like exhaustive search based on cryptic arithmetic.
Represent the digits with dx where the patterns are defined by the problem definition.  

    meek   = d1 d2 d2 d3
    humble = d2 d4 d5 d1
    lowly  = d1 d4 d5 d2
    
    meek   = humble - lowly
    
There are no uniqueness constraints on the digits - so we can use nested loops.

I am getting way too many solutions -- need to check this again.  There is an
implication that phone numbers are unique so I will add this.

Now I always get a unique solution for Meek but many possibilities for Humble
and Lowly.  I guess this is reasonable:

```{r}
solution_set=data.frame(meek=NA,humble=NA,lowly=NA)
for (d1 in 0:9) {
  for (d2 in 0:9) {
    for (d3 in 0:9) {
      for (d4 in 0:9) {
        for (d5 in 0:9) {
          meek   = d1*1000+d2*100+d2*10+d3
          humble = d2*1000+d4*100+d5*10+d1
          lowly  = d1*1000+d4*100+d5*10+d2
          if ((meek != humble) & (meek != lowly) & (humble != lowly)) {
            if (meek == (humble-lowly)) {
              solution_set=rbind(solution_set,c(meek,humble,lowly))
            }
          }
        }
      }
    }
  }
}
solution_set=na.omit(solution_set)
print(paste0("Number of solutions = ",nrow(solution_set)))
print(paste0("Unique solution for meek = ",unique(solution_set$meek)))
```

## Problem 2

    I know that the lifetime batting averages
    of Mike and John are 0.200 and
    0.300, but I forget which is which.
    However, in Mike's first 10 times at bat
    this year he had three hits, and in
    John's first 10 times at bat this year
    he had two. What is the probability
    that Mike's lifetime batting average is
    0.300? (Assume that the probability
    that a batter will get a hit is equal to
    his lifetime batting average, which is
    the ratio of number of hits to number
    of times at bat.)
    
    -- Byron R. Adams, TX A'58

Solution: this smacks of Bayes Theorem.

I confess that I looked at the solution before I did this but the
calculations are given below:

```{r}
## P(200 hitter gets 2 hits in 10 at bats)
p200_2 = choose(10,2)*(0.2^2)*(0.8^8)
## P(200 hitter gets 3 hits in 10 at bats)
p200_3 = choose(10,3)*(0.2^3)*(0.8^7)
## P(300 hitter gets 2 hits in 10 at bats)
p300_2 = choose(10,2)*(0.3^2)*(0.7^8)
## P(300 hitter gets 3 hits in 10 at bats)
p300_3 = choose(10,3)*(0.3^3)*(0.7^7)

## Scenario 1: 200 hitter gets 2 hits, 300 hitter gets 3 hits - Mike is 300 hitter
p1 = p200_2*p300_3
## Scenario 2: 300 hitter gets 2 hits, 200 hitter gets 3 hits
p2 = p300_2*p200_3

print(c(p1,p2,p1/(p1+p2)))

```

Conclude: there is a 63.16% probability that Mike is a 300 hitter.

Still think Bayes theorem could be used here but not sure how.


## Problem 3

    Solve this cryptic addition:
    
      OLD
     SALT
     TOLD
     TALL
    -----
    TALES
    
    -- Challenging Mathematical Teasers
    by J.A.H. Hunter

Solved in the cryptic_arithmetic notebook.

Solution:

    OLD   =   394
    SALT  =  8091
    TOLD  =  1394
    TALL  =  1099
    TALES = 10978


## Problem 4

    Sue has opened her piggy bank,
    which contains a mixture of pennies,
    nickels, dimes, and quarters. She
    quickly notes that she has at least four
    of each kind. Upon careful counting,
    she finds that she has exactly 100 coins
    with a face value of $2.73. How many
    of each kind does she have?
    
    -- Don A. Dechman, TX A'57

Solution: exhaustive search -- maybe sampling?
Let's try manual first.

Assume the following distribution of coins:

    Quarters: 4  --> 1.00
    Dimes:    4  --> 0.40
    Nickels:  4  --> 0.20
    Pennies:  4  --> 0.04
    ---------------------
    Total    16      1.64
    
    Need to add 84 coins that add up to 2.73-1.64=1.09.
    Can add a bunch of pennies as long as it leaves a
    balance that is divisible by 0.05.  Arbitrarily use a
    margin of 0.25 to allow for a quarter.  This gives
    the correct total value and then we can adjust the
    coins to get the correct total number of coins.
    
    Trade 5 pennies for 1 nickel         --> reduction of 4 coins
    Trade 2 nickels for 1 dime           --> reduction of 1 coin
    Trade 1 dime for 2 nickels           --> increase of 1 coin
    Trade 1 quarter for 2 dimes+1 nickel --> increase of 2 coins
    Trade 1 quarter for 1 dimes+3 nickels -->increase of 3 coins

    Whether or not the trade is allowed depends on maintaining the
    minimum count of 4 coins of each denomination.
    
    First attempt:

    1.09 - 0.25 = 0.84  --> 84 pennies, 1 quarter --> 101 coins
    
    Now have 5 quarters, 88 pennies, 4 dimes, and 4 nickels.
    
    Need to reduce the number of coins by 1
    
    Can't directly trade 2 nickels for a dime since we only have
    four nickels to begin with.  Instead we will drop some
    pennies to get another nickel and then adjust as needed:
    
    Drop 5 pennies and add one nickel --> 97 coins
    Drop 1 quarter and add one dime and 3 nickels --> 100 coins
    
    Final total:
    
    Quarters: 4  --> 1.00
    Dimes:    5  --> 0.50
    Nickels:  8  --> 0.40
    Pennies:  83 --> 0.83
    ---------------------
    Total     100    2.73

For grins, let's try exhaustive search:

```{r}
t1=Sys.time()
for (q in 4:trunc(273/25)) {
  for (d in 4:trunc(273/10)) {
    for (n in 4:trunc(273/5)) {
      for (p in 4:273) {
        v=q*25+d*10+n*5+p
        nc=q+d+n+p
        if (v==273 & nc==100) {
          print(c(q,d,n,p))
        }
      }
    }
  }
}
t2=Sys.time()
t2-t1
```

It only takes about a second to exhaustively search for a solution and we find
the same unique solution we found by hand.

## Problem 5

    One good way to estimate the height
    of an object is to take a known height,
    sight along a ruler until the known
    object subtends an easy length to work
    with, and then take the proportional
    height subtended by the unknown object.
    If the ruler is parallel to both
    objects, the result will be exact.
    Recently I attempted to measure
    the height of a bridge this way. There
    is a plaque on the side of the vertical
    bridge tower marking the water level
    from the last flood -- three meters
    from the ground. I backed off and held
    my ruler so that the plaque's height
    appeared to be three centimeters. The
    height of the roadway then appeared
    to be 12 cm and that of the supporting
    tower 30 cm.
    Hence, I estimated that the roadway
    is 12 m above ground and that the
    tower is 30 m high. Later I found that
    the roadway is actually 13 m above the
    ground. Clearly, the discrepancy arose
    because I did not hold the ruler precisely
    vertical. What, then, is the correct
    height of the tower?
    
    -- Adapted from Technology Review

Solution: need to draw a diagram.

From the point of the eye there will be
a nearly vertical ruler with points along the ruler at 3, 12, 30 cm.
At some (far away) distance there is a vertical lines with points at
3, 13, and and unknown height.

There is a common point at the eye that must be on a straight line
connecting 3 cm on the ruler with 3 m on the wall as well as another
straight line connecting 12 cm on the
ruler with 13 m on the wall.

Should be able to use this to calculate the location of the eye/angle of ruler.

Once that is known should be able to calculate the unknown height for the tower.


TODO


## Bonus

    As I was jogging in the predawn
    darkness, Venus was the brightest
    object in the eastern sky, and I
    wondered if it were ever brighter. I
    found that Venus and the Earth are in
    nearly circular coplanar orbits, with an
    orbital radii ratio of 0.723. Assume
    that the sunlight reflected by Venus
    towards Earth is proportional to the
    illuminated fraction of the Venus disk
    as seen from Earth. Then, at what
    value of the Earth-Sun-Venus angle
    does Venus appear brightest? Give
    your answer accurate to the nearest
    0.1 degree.
    
    -- James L. Griggs Jr., OH A'56

Solution:

We can assume circular orbits with a ratio of 0.723 for Venus and Earth.
The hard part is figuring out what fraction of Venus is illuminated based
on the angle between the Sun, Venus, and Earth.  I guess we assume the
diameter of Venus is negligible?  The real brightness will depend on the angle
and the distance (inverse square?) but the problem says only the fraction
illuminated by the sun.

I need a diagram!!!

1. When Venus is directly between the Earth and the Sun there is no part visible.
   This is clearly the minimum brightness.
   
2. When Venus is on the opposite side of the sun from Earth the fraction would be
   maximum but the Sun would be in the way.
   
3. The maximum fraction will be when Venus is just about to go behind the Sun.

This problem makes no sense without dimensions for the Sun/Venus/Earth and a
consideration of the distance between the Earth and Venus unless I
am mis-interpreting something.

TODO

I peaked at the solution.  There is an inverse square law in the solution but everything
is normalized to Earth/Sun distance = 1.0.  Need to look at the diagram and verify
these equations.

## Computer Bonus

    Find the smallest
    and largest hexadecimal integers such
    that each of their squares contains
    each of the 16 hexadecimal digits (0
    through 9 and A through F) once and
    only once. A leading zero is not allowed.
    
    -- Howard G. McIlvried III, PA Gamma'53

Solution: seems like exhaustive search

The integers in R seem very limited -- will switch to Python.

Need the decimal package to calculate sqrt() of the 16 digit hex numbers.
Unfortunately, can't communicate the results back to R since the interface
appears to use 64 bit signed integers.  These are big enough to handle the
lower end of the range below (e.g. n1 and nmin) but not the upper end
of the range(e.g. n2 and nmax get converted to -1).

Fortunately the solutions are near the ends of the search range so it is
quite feasible to linearly search up/down from the ends of the range:

```{r}
library(reticulate)
```

```{python}
import decimal
import numpy as np

n1=int(decimal.Decimal(0x1023456789abcdef).sqrt())
n2=int(decimal.Decimal(0xfedcba9876543210).sqrt())
print('Smallest = ',n1)
print('Largest  = ',n2)
print('Range  (billions)  = ',(n2-n1)*1e-9)

print('Searching for solution')
n=n1
for i in range(1000000):
    digits = [i for i in hex(n**2)[2:]]
    if len(digits)==16 and len(np.unique(digits))==16:
        print("Smallest solution = ",hex(n)," squared = ",hex(n*n))
        print('Number of items searched = ',n-n1)
        nmin=n
        break
    n=n+1
n=n2
for i in range(1000000):
    digits = [i for i in hex(n**2)[2:]]
    if len(digits)==16 and len(np.unique(digits))==16:
        print("Largest solution  = ",hex(n)," squared = ",hex(n*n))
        print('Number of items searched = ',n2-n)
        nmax=n
        break
    n=n-1

```
