---
title: "Bent Summer 2008"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Summer 2008


## Problem 1

    Peter, Quentin, Ralph, Sam, and
    Thomas went on a photography
    safari for 10 days. A different pair
    did the picture taking each day-not
    always successfully, but the expedition
    came home with photos of five
    animals, each snapped on different
    days. Thomas had no hand in
    snapping the antelope, but Quentin
    was one of the pair who got the
    elephant. No one was on teams that
    snapped both the antelope and the
    dromedary, but one proud chap can
    now brag that he snapped both the
    crocodile and the elephant. Someone
    snapped a baboon and also the
    animal that he snapped while out
    with Sam. Peter and Quentin had a
    successful day together, but Ralph
    and Thomas drew blank. Each man
    had a hand in photographing two
    animals, and Quentin, Sam, and
    Thomas a hand in all five among the
    three of them. Which pair accounted
    for each photograph?

    -- Martin Hollis

TODO

    Not sure I care about these logic puzzles.


## Problem 2

    Find three different integers, P,
    Q, R, such that P+Q, P+R, Q+R,
    P-Q, P-R, and Q-R are all squares
    of integers. The three integers may
    be positive, negative, or zero. Find
    the set {P, Q, R} with the smallest
    P+Q+R.

    -- Stephen Ainley

Solution:

    WARNING: this was logically wrong.  Q+R=0 is a
    valid solution so the arguments below should
    really be >= instead of > for R < 0.

    If the sum and differences are squares then
    they must be positive so:
    
        P>=Q>=R

    The only value that could be negative would
    be R, otherwise Q+R would be negative and
    that could not be a square.  Furthermore:
    
        Q+R>=0 --> Q >= -R
        
    So, if R is negative then P > Q >= abs(R)

    I had a logic error in my program that required Q > abs(R) but in reality
    you could have Q=abs(R) which is a key.
    
    Once the bug was fixed the solution is found quickly.
    
```{r}
is_square = function(n) {
    if (n<0) {
        return(FALSE)
    }
    m=round(sqrt(n))
    return( m*m == n )
}

check_solution = function(nmax) {
    for (r in seq(-nmax,nmax)) {
        for (q in ((r+1):nmax)) {
            for (p in ((q+1):nmax)) {
                if (is_square(p+q)) {
                    if (is_square(p-q)) {
                        if (is_square(p+r)) {
                            if (is_square(p-r)) {
                                if (is_square(q+r)) {
                                    if (is_square(q-r)) {
                                        print(c(p,q,r))
                                    }
                                }
                            }
                        }
                    }
                }
            }
        }
    }
    return(c())
}

check_solution(100)
```
 
The smallest P+Q+R is (17,8,-8)


## Problem 3

    In a stud poker game with no
    ante, on the first face-up round, high
    card made a bet with two coins.
    Each of the other players called
    without raising, e.g., made the same
    bet. Only standard U.S. coins, up to
    and including silver dollars, were
    used in the game. Second hand put
    in three coins; 3rd hand put in two
    coins and took out one in change;
    4th hand put in three coins and took
    out one in change; 5th (last) hand put
    in one coin and took out in change all
    but three of the coins then in the pot.
    Whenever coins are removed from
    the pot, they have different values
    than coins put in the pot. How much
    did 1st hand bet?

    -- Math. Puzzles for Beginners and
    Enthusiasts, Geoffrey Mott-Smith

Solution:

    I don't understand this problem.  The Bent claims
    the bet is 0.60.  I don't understand how the fourth
    or fifth hand make change.
    
    The coins are:
    
    P =   1 = penny
    N =   5 = nickel
    D =  10 = dime
    Q =  25 = quarter
    H =  50 = half-dollar
    S = 100 = silver-dollar

    First bet   = T = two coins
    Second hand = T = three coins
    Third hand  = T = two coins in and one coin out
    Fourth hand = T = three coins in and one coin out
    Fifth  hand = T = one coin in and removed all but three
        I think this means there were four in the pot
        at the end of the game.  There are 8 in the pot
        before the fifth hand so this means 5 were taken
        out in change.  It might also mean that they took
        out 6 coins and left 3 -- not sure.

    Since all the bets are the same we know that there is a total, T,
    that can be achieved with two coins or three coins or two in/one out
    or three in/one out or one coin in and five out.
    
    The fifth hand constrains the total to something less than 100.
    When fifth hand makes bet there are 8 coins in the pot.
    They put in one and pull out five.
    The fifth hand also says that 5T < 400 -> T < 80.
    
    First Hand = Possibilities for the sum of two coins (< 100):
        2   6  10  11  15  20  26  30  35  50  51  55  60  75
    
    Second Hand = Possibilities for the sum of three coins (<100):
        3  7 11 12 15 16 20 21 25 27 30 31 35 36 40 45 51 52 55 56 60 61 65 70 75 76 80 85

    Common totals:
        11 15 20 30 35 51 55 60 75

    Third Hand = Possibilties for two in/one out (0-100):

        11 = NOT POSSIBLE
        15 = D+D-N
        20 = Q+N-D
        30 = Q+D-N
        35 = NOT POSSIBLE 
        51 = NOT POSSIBLE
        55 = H+D-N
        60 = S+D-H
        75 = H+H-Q
        
    Remaining totals:
        15 20 30 55 60 75
        
    Fourth Hand = Possibilities for three in/one out:

        15 = NOT POSSIBLE
        20 = NOT POSSIBLE
        30 = NOT POSSIBLE
        55 = Q+Q+D-N
        60 = S+N+N-H
        75 = NOT POSSIBLE

    Common totals:
        55 60

    So the first bet must be either 55 or 60.

    Assume the bet is 55:
    
    First Hand:  H+N
    Second Hand: Q+Q+N
    Third Hand:  H+D-N
    Fourth Hand: Q+Q+D-N  -> Pot = H+Q+Q+H+D+Q+Q+D (220)
    Fifth Hand:  ???
    
    Assume the bet is 60:
    
    First Hand:  H+D
    Second Hand: Q+Q+D
    Third Hand:  S+D-H
    Fourth Hand: S+N+N-H  -> Pot = D+Q+Q+D+S+D+S+N+N-H
        This is not possible -- there are no H coins
        to extract
    Fifth Hand:  ???


## Problem 4

    Find all integral solutions of (x+1)**y
    = x**(y+1) + 1. Assume 0**0 is 1.

    -- Meyl, 1876

TODO


## Problem 5

    Four people each flip a fair coin
    five times.What is the exact probability
    that they each get the same
    number of heads?

    -- Adapted from Duelling Idiots and
    other Probability Puzzlers by Paul
    J. Nahin

Solution:

    If you flip a fair coin five times you can
    get 0, 1, 2, 3, 4, or 5 heads.  There are 2^5=32
    ways for these to happen and there are the
    following probabilities:
    
    P(0 Heads) = 1/32
    P(1 Head)  = 5/32
    P(2 Heads) = 10/32
    P(3 Heads) = 10/32
    P(4 Heads) = 5/32
    P(5 Heads) = 1/32
    
    The probability of all four people getting the
    same total is P^4.  Summing these over the 6 possible
    combinations yields:
    
    P = 5313/262144 ~ 2.03%


## Bonus

    Five beads are strung on
    a horizontal insulated wire in the
    shape of a unit square, with the holes
    in the beads being large enough so
    that the beads can move to any location
    on the wire (that is, they can go
    around the corners of the square).
    If each bead is given exactly the
    same negative electrical charge
    and the system is allowed to come
    to equilibrium, what are the coordinates
    of the five beads (with the
    non-symetric bead at the origin and
    an edge of the square on the x-axis)?
    We want both solutions.

    -- Daryl Cooper

TODO


## Computer Bonus

    Consider the integer
    7. Its cube is 343. The integer
    divisors of 343 are 1, 7, 49, and 343.
    Those divisors sum to 400, which is
    the square of the integer 20. Find the
    next larger integer such that the sum
    of the integer divisors of its cube is
    the square of an integer.

    -- Fermat, 1657

Solution:

    I punted and went back to sympy.  Solution time was ~3 minutes:

    Solution =  751530    Time =  184 seconds
    
    This is the only solution below 1e6

    This is very similar to Bent 2000_1 Computer Bonus.
    That was also from Fermat and asks for the smallest
    integer greater than 1 where the sum of the divisors
    of the square is a perfect cube.
    
    Here we want the sum of the divisors of a cube is a
    perfect square.
    
    The original problem used sympy but we will try gmp here:
    
```{r}
library(gmp)

```

Solution is n=???? --- having problems with gmp
Can't use sqrt with bigz numbers and non-bigz numbers
are hitting overflow.

May need to drop back to sympy.
Maybe exhaustive search is the wrong way to go.

Should I look at the prime factorization of the original
number and the square and then figure out the cube???
