---
title: "Bent Winter 2011"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Winter 2011


## Problem 1

    Find the smallest positive integer
    that has a remainder of 1 when divided
    by 2, a remainder of 2 when divided by
    3, a remainder of 3 when divided by 4,
    a remainder of 4 when divided by 5, a
    remainder of 5 when divided by 6, and
    a remainder of 6 when divided by 7.

    -- Christopher R. Oliver, AL E'08

Solution:

    I think this has something to do with the
    Chinese remainder theorem - but brute
    force will probably work.
    
```{r}
divisors=c(2,3,4,5,6,7)
remainders=c(1,2,3,4,5,6)
n=7
while(TRUE) {
    if (all(n%%divisors == remainders)) {
        print(n)
        break
    }
    n=n+2
}
```

The solution is 419.


## Problem 2

    Five pirates find a chest containing
    100 gold pieces and decide to divide
    them according to the following
    scheme. In order of seniority, each
    pirate proposes a distribution of the
    gold pieces. All of the pirates then
    vote, and if half or more of the pirates
    accept the proposal, the gold pieces
    are divided as proposed. If not, then
    the proposer must walk the plank and
    drown, and they start the process over
    again with the next most senior pirate
    making a proposal. Assuming the pirates
    are very intelligent, greedy, and
    interested in surviving, how does the
    senior pirate propose to split the 100
    gold pieces?

    -- How to Ace the Brain Tickler
    Interview by John Kador

TODO

    See the Bent description.  It involves
    the head pirate giving a token amount
    to the two pirates who can never
    get anything if the pirate count
    gets too low (e.g. the most junior
    pirate and the third in line).


## Problem 3

    My wife and I recently attended a
    party at which there were four other
    married couples, and various people
    shook hands. No one shook hands with
    himself or herself or with his or her
    spouse, and no one shook hands with
    the same person more than once. After
    the handshakes were over, I asked
    each person, including my wife, how
    many hands he or she had shaken. To
    my surprise, each had shaken a different
    number of hands. How many hands
    did my wife shake?

    -- Knotted Doughnuts and Other
    Mathematical Entertainments
    by Martin Gardner

TODO

    meh - logic

## Problem 4

    Professor Asterisk treated his
    friend, Professor Scalene, to a ride
    on the local bus. He asked the driver
    for two transfer tickets, which had
    consecutive five-digit serial numbers.
    He mentally added the ten digits and
    informed Professor Scalene that the
    sum was 62. Professor Scalene asked
    if the sum of the digits of the serial
    number on either of the tickets were
    in the range of 29 to 39, inclusive. After
    Professor Asterisk answered, Professor
    Scalene remarked, "I now know
    what the two serial numbers are."
    What were the serial numbers?

    -- Hard-to-Solve Brainteasers
    by Jaime and Les Poniachik

TODO

    Exhaustive search would work.
    

## Problem 5

    A sparrow, flying horizontally in a
    straight line, is 20 m directly below an
    eagle and 40 m directly above a hawk.
    Both the eagle and the hawk always
    fly directly at the sparrow, and they
    reach it simultaneously. If the sparrow
    is flying at a speed of 15 m/s and the
    hawk is flying at 30 m/s, how fast is the
    eagle flying?

    -- Chases and Escapes
    by Paul J. Nabin

Solution:

    I had fun with this.

    At t=0 the sparrow is at the origin:
    
        xs=0, ys=0
        
    At t=0 the hawk is on the negative y-axis:
    
        xh=0, yh=-40

    The position of the sparrow is given by:
    
    xs = 15*t
    
    We need to calculate the time for the hawk to
    reach the sparrow.
    
    The angle of attack between the hawk and the
    sparrow is given by:
    
    theta = atan((0-yh)/(xs-xh))
    
    The x,y components of the hawk's speed is:
    
    vxh = dxh/dt = vh*cos(theta)
    vyh = dyh/dt = vh*sin(theta)
    
    if tan(theta) = -yh/(xs-xh) this gets ugly fast.
    
    Isn't there a clever way to solve this other than coupled
    differential equations?

    I can solve this numerically but I would like a more elegant
    result.
    
    The Bent has a solution but I don't know if I would call it
    more elegant.  Here is there methodology.
    
    Assume hawk starts at origin and sparrow is flying vertically
    along a line x=x0.  At time=t we know:
    
        Position of sparrow = (x0,vs*t)
        Position of hawk    = (x,y)
        
    The attack angle of the hawk:
    
        dy/dx = p = (vs*t-y)/(x0-x)
        
    Solve this for t:
    
        t = p*(x0-x)/vs + y/vs
        
    So far - no problem.  The next step is to note:
    
        vh*t = S = integrate(sqrt(1+p**2),(tau,0,x))
        
    Where vh=30 (speed of hawk) and S = path length on hawk's trajectory.
    We can also solve this for t:
    
        t = (1/vh)*integrate(sqrt(1+p**2),(tau,0,x))
        
    Equating to the previous expression for t:
    
        t = p*(x0-x)/vs + y/vs = (1/vh)*integrate(sqrt(1+p**2),(tau,0,x))
    
    Differentiating both sides:
    
        dp/dx*(x0-x)/vs + p*(-1)/vs + dy/dx*(1/vs) = (1/vh)*sqrt(1+p**2)
        
        dp/dx*(x0-x)/vs + (p*(-1)/vs + p * (1/vs))   =
        
        dp/dx*(x0-x)/vs = (1/vh)*sqrt(1+p**2)
        
    Rearranging:
    
        dp/sqrt(1+p**2) = (vs/vh)*(1/(x0-x))*dx
        
    Now for the mind blowing part:  integrate the left side from 0 to p
    and the right side from 0 to x:
    
        asinh(p) = vs/vh*(log(-vh*x0) - log(vh*x - vh*x0))
    
    Since I have no idea what asinh(p) means (from sympy) I had to go to
    Wikipedia to see how to write it in logs:
    
        log(p+sqrt(p**2+1)) = vs/vh*log(x0/(x0-x))
    
    exponentiating both sides:
    
        p + sqrt(p**2+1) = (x0/(x0-x))^(vs/vh)
        
    Mind continuing to blow:  replace p by dy/dx and then multiply by dx
    
        ( dy/dx + sqrt(p**2+1) ) * dx = (x0/(x0-x))^(vs/vh) * dx
        
    Here it comes:
    
        dy + dS = (x0/(x0-x))^(vs/vh) * dx
        
    Now - integrate this all from 0 to x0:
    
        yc + Sc = x0/(1-(vs/vh))
        
    where yc = y value where hawk captures sparrow
    
    But yc=vs*tc, Sc=vh*tc so
    
        vs*tc + vh*tc = x0/(1-(vs/vh))
        
    or:
    
        tc = x0/((1-(vs/vh))*(vs+vh))

           = vh*x0 / (vh^2 - vs^2)
           
    For the case of vh=30, vs=15, x0=40 we have:
    
        tc = 1.777778 seconds (?????)
        
    The same logic applies to the eagle except for the starting
    condition.  The tc value must be the same so we can calculate
    the required speed:
    
        x0 = 20, vs=15, tc=1.777777
        
        tc = ve*x0 / (ve^2 - vs^2)
        
        1.777778 = ve*20 / (ve^2 - 15^2)
        
        ve = 21.645 m/s
        
    Compared to: vs=15 ms, vh = 30 ms.
    
    As a comparison let's try a  numerical solution.
    
    We will use the same convention (hawk starts at origin).
    
    We can use the optimizer to find when the distance between
    the hawk-sparrow or eagle-sparrow is minimized.

```{r}
library('deSolve')

# Calculate speed in x,y directions
#
# Assume state vector has two compoents:
#
# x[1] = xh (x position of the hawk)
# x[2] = yh (y position of the hawk)
#
# Input parameters have two components:
#
# parms[1] = speed of attacker
# parms[2] = initial separation
foo = function(t,x,parms) {
    vh=parms[1]   # Speed of the predator
    xs=parms[2]   # Initial separation between predator/prey
    ys=15*t       # Speed of prey is constant
    xh=x[1]       # x location of predator
    yh=x[2]       # y location of predator
    # Angle of attack
    theta = atan((ys-yh)/(xs-xh))
    # Speed in each axis
    vh_x = vh*cos(theta)
    vh_y = vh*sin(theta)
    # Return these as the derivatives
    # The ODE solver requires a list
    return(list(c(vh_x,vh_y)))
}

# Function to find the capture time of the hawk
# The speed/separation of the hawk is passed as c(30,40)
# and the unknown we are searching for is the length of
# time it takes to reach a particular state.
# The state is the minimum distance between predator-prey
shoot = function(tfin) {
    t=seq(0,tfin,length=1000)
    x=ode(c(0,0),t,func=foo,c(30,40))
    dx=40-x[nrow(x),2]
    dy=15*tfin-x[nrow(x),3]
    d=dx^2+dy^2
    return(d)
}

# Function to find the speed of the eagle
# The speed/separation of the eagle is passed as c(ve,20)
# We are searching for the optimum value of ve to make
# the final location of the eagle match the hawk/sparrow.
shoot2 = function(ve) {
    t=seq(0,topt$minimum,length=1000)
    x=ode(c(0,0),t,func=foo,c(ve,20))
    dx=(20-x[nrow(x),2])
    dy=(15*topt$minimum-x[nrow(x),3])
    d=dx^2+dy^2
    return(d)
}

# Find the capture time
topt=optimize(shoot,c(0,5))

# Calculate the optimal trajectory
t=seq(0,topt$minimum,length=500)
x=ode(c(0,0),t,func=foo,c(30,40))

# Make some plots for the hawk pursuit
plot(x[,1],x[,2],xlab='Time',ylab='X-Position')
abline(h=40,col='red')

plot(x[,1],x[,3],xlab='Time',ylab='Y-Position')
lines(x[,1],15*x[,1],col='red')

plot(x[,2],x[,3],xlab="Hawk X-Position",
     ylab="Hawk Y-Position",
     main="Hawk-Sparrow Pursuit Trajectory")
abline(v=x[nrow(x),2],col='red')
abline(h=x[nrow(x),3],col='red')

# Now that we know the time for capture we can try to find
# the speed of the eagle.  It must be faster than the
# sparrow and slower than the hawk:

veagle=optimize(shoot2,c(15,30))

print(paste0("Speed of eagle = ",veagle$minimum))

```

 Numerically:
 
    Speed of eagle = 21.645022 m/s

Compared to analytical result = 21.645007 m/s

Awfully close to the same thing.

```{r}
# Plot hawk and eagle trajectories on the same axes
# and use the correct orientation.  We need to swap
# the x-y axis and apply some offsets.
#
# The sparrow is on the +x axis
# The hawk starts 40m below
# The eagle starts 20m above

# Calculate the optimum trajectory for eagle
x2=ode(c(0,0),t,func=foo,c(veagle$minimum,20))

# Make the plots and annotations
plot(c(0,30),c(-40,20),type='n',
     xlab='X-Position',
     ylab='Height',
     main="Predator-Prey Pursuit Trajectories")
lines(x[,3],x[,2]-40,col='blue')
points(x[,3],x[,2]-40,col='blue')
lines(x2[,3],20-x2[,2],col='green')
points(x2[,3],20-x2[,2],col='green')
lines(x2[,1]*15,rep(0,nrow(x2)),col='red',lwd=3)
text(6,14,"Eagle @ 21.6 m/s")
text(6,-28,"Hawk @ 30 m/s")
text(6,-3,"Sparrow @ 15 m/s")
grid()
abline(v=tail(x[,3],1),lty='dashed')
text(29,0,'Capture')

```
 

## Bonus

    A regular dodecahedron has
    12 faces in the shape of regular pentagons.
    If some, including none or all, of
    the faces are painted red and the rest
    are painted blue, how many dodecahedrons
    that are distinguishable from
    each other are possible?

    -- Classic Puzzles
    by Gyles Brandreth

TODO

    Another one of those reflection, rotation, color
    counting puzzles.  Still don't understand these
    things.
    

## Computer Bonus

    A, B, and C are
    different nine-digit integers that each
    use the digits 1 through 9 once and only
    once. What is the value of C that satisfies
    the equation, A x B = C^2, and where
    A and B are also perfect squares?

    -- Madachy's Mathematical
    Recreations by Joseph S. Madachy

Solution:

    Should be able to exhaustively search for this
    solution.  The number of permutations is 9! which
    is reasonable.
    
    I tried to do this with gtools and gmp but the large number
    problems were very hard to work around.  The lack of sqrt for
    large integers makes it difficult to find the squares.
    
    Dropped back to using basic python
    
```{r}
library(reticulate)
```

```{python}
import itertools
import math

# n = all the relevant 9 digit numbers
n=[]
for i in itertools.permutations(range(1,10),9):
    m=0
    for j in i:
        m=10*m+j
    n.append(m)

# m = all the 9 digit values that are squares
m=[]
for i in n:
    if round(math.sqrt(i))**2 == i:
        m.append(i)

# Find solutions to A*B == C**2
n2=[i*i for i in n]
for i in range(len(m)):
    for j in range(i+1,len(m)):
        if m[i]*m[j] in n2:
            k=n2.index(m[i]*m[j])
            print("A = ",m[i])
            print("B = ",m[j])
            print("C = ",n[k])
```

Solution found as:

    A =  549386721  (23439^2)
    B =  743816529  (27273^2)
    C =  639251847

    549386721*743816529 = 408642923892911409
    
