---
title: "Bent Summer 2011"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Summer 2011

I found this group of problems to be less interesting
than most.  Not one of my favorite problem sets.

## Problem 1

    Al, Bob, Carl, Don, and Ed took a
    mathematics test with five questions.
    For a correct answer, 10 points were
    given; if the answer were wrong, either
    7, 2, or 0 points were given, depending
    on the method used and the way the
    work was presented. (No points other
    than 10, 7, 2, or 0 were awarded, and it
    is perfectly possible for two students
    to get the same wrong answer, but be
    given different points.) At least one
    student got each question right. Some
    information about the answers of different
    students is shown in the table.

         1   2  3  4 5
    Al   5   ?  11 ? 4.5
    Bob  3.5 17 5  ? 3.4
    Carl 4   43 5  ? 2.8
    Don  2.5 17 11 6 3.8
    Ed   7   17 7  7 5.2

    The total points for Al, Bob, Carl,
    Don, and Ed were 34, 19, 31, ?, and 9,
    respectively. The total points for each
    question, #1 - #5, were 14, 29, 22, 42,
    and ?. You may also find it helpful to
    know that Ed got more points for question
    5 than Bob. Determine the number
    of points awarded each student for
    each question.

    -- Brain Puzzler's Delight
    by E. R. Emmet


TODO


## Problem 2

    2010 was the 75th anniversary of
    the creation of the WPA, which provided
    millions of jobs during the Great
    Depression. Solve the following two
    cryptic addition problems simultaneously
    in base 11 (use a lower case a to
    represent the digit 10).

    USA + FDR = WPA
    USA + WPA = PARK

    Different letters represent different
    digits, the same letter always represents
    the same digit, and there are no
    leading zeros.

    -- Howard G. McIlvried III, PA Gamma'53

Solution: done in the cryptic arithmetic notebook

  Cryptic addition in Base 11

      USA + FDR = WPA
      USA + WPA = PARK

    a=make_crypt_df('usafdrwpapark',num_base=11)
    dim(a)
    [1] 19958400        9
    x_usa=make_num(a,'usa',11)
    x_fdr=make_num(a,'fdr',11)
    x_wpa=make_num(a,'wpa',11)
    x_park=make_num(a,'park',11)
    k=(x_usa+x_fdr==x_wpa) & (x_usa+x_wpa == x_park)
    k=k&(a$u!=0)&(a$f!=0)&(a$w!=0)&(a$p!=0)
    sum(k)
    [1] 1
    a[k,]
             u  s a f d r w p k
    10793931 5 10 4 3 2 0 9 1 8

    x_usa[k,]
    [1] 719
    x_fdr[k,]
    [1] 385
    x_wpa[k,]
    [1] 1104
    x_park[k,]
    [1] 1823

    Note that 's' has a value of 'a' using conventional notation.
    All of the x_* values given above are in base 10!!!


## Problem 3

    In a game of 7-card stud poker with
    deuces wild, what is the probability of
    getting five of a kind?

    -- Howard G. McIlvried III, PA Gamma'53

Solution:

    Should be able to tackle with random sampling
    
```{r}

cards=rep(1:13,each=4)
ntrials=1000000

set.seed(42)
num_wins=0
for (trial in 1:ntrials) {
    s=sample(cards,7)
    if (any(s==2)) {
        wild=which(s==2)
        s=s[-wild]
        if (max(table(s))+length(wild)>4) {
            num_wins=num_wins+1
        }
    }
}

print(paste0("Num wins = ",num_wins))
print(paste0("Pr       = ",num_wins/ntrials))

```

For 1e6 trials

    Num wins = 4572
    Pr       = 0.004572

With set.seed(42), 1e6 trials

    Num wins = 4603
    Pr       = 0.004603


## Problem 4

    Twenty-six identical coins lie on a table --
    ten heads and 16 tails. Blindfolded,
    you are told to divide the coins into two
    groups with the same number of heads
    in each group. You may move the coins
    and turn them over, but have no way of
    telling whether a coin is heads or tails.
    How do you accomplish this?

    -- All-Star Mathlete Puzzles
    by Richard I. Hess, CA B'62

TODO


## Problem 5

    Any number, real or complex, may
    be represented in the form
    
    r(cos(theta) + i sin(theta)).
    
    Considering values of theta only in
    the range -pi to +pi radians, determine
    whether i to the ith to the ith to the
    ith power, ad infinitum, approaches
    a limit, and, if so, provide r and theta to
    4 digits. As a reminder of standard
    practice, note, for example, that 3 to
    the 3rd to the 3rd means 3^27, and not
    27^3. That is, the continued exponential
    is evaluated from the top down, not
    from the bottom up.

    -- John W. Langhaar, PA A'33

TODO


## Bonus

    Given a 2x4 rectangle of
    squares (a map) with the front marked
    with the digits

        1 2 3 4
        5 6 7 8

    and the back marked with letters:

        d c b a
        h g f e

    (that is, the Nth letter of the alphabet
    is on the back of the square with the
    digit N).
    This 2x4 map can be folded (which
    includes tucks) along the boundaries
    between squares in many ways to end
    with a 1x1 by 8 layers configuration
    with the '1' square on top of the other
    seven and the '1' visible. For example,
    abcd folded onto efgh, 78 folded onto
    65, and 3 folded onto 4, ends up with
    the configuration: 1e8d3g6b. How
    many unique final configurations can
    be achieved, and what are they? To
    make grading easier, please provide
    the configurations, one per line, sorted
    in ascending order (digits less than
    letters and using only the marks on
    the upper surfaces) using the above
    numbering system.

    -- Fred J. Tydeman, CA Delta'73


TODO


## Double Bonus

    Consider a quadrilateral.
    Construct a square on each
    of its sides. Prove that the two line
    segments joining the centers of opposite
    squares are equal in length and
    mutually perpendicular.

    -- Jan Smit, CA Delta'41

TODO


