---
title: "Bent Fall 2019"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Fall 2019


## Problem 1

    The following cryptic additions are to be
    solved simultaneously, that is, letters have
    the same value in both cryptics; different
    letters are different values; * can represent
    any value. There are no zeros anywhere. The
    two sums are mirror images (i.e., reverses)
    of each other.

         ******    MIRROR
        +MIRROR   +IMAGE*
         ------    ------
         IMAGE*    ******

    -- an Enigma by Albert Haddad in New Scientist

Solution:

    Should be able to solve these with the standard
    cryptic artithmetic tools with a digit set of
    MIRROR, IMAGE + one more symbol.
    
    Need simultaneous solutions of:
    
         ******    MIRROR
        +MIRROR   +IMAGE*
         ------    ------
         IMAGE*    *EGAMI

    Only the sums are mirror images.
    
    The Bent solution:
    
        447265 = *****      274434 = MIRROR
        274434 = MIRROR     721693 = IMAGE*
        ------              ------
        721699 = IMAGE*     996127 = *EMAGI

    After working some more I get 5 solutions:
 
    The third of my 5 solutions is the one from the Bent.
    
    I can't find anything wrong with the other four solutions.
    
    Oops - I missed a constraint.  The problem says there are
    no zeros anywhere!!!
    
    When I add that constraint the solution is unique.
  
```{r}
source("rgc_lib.R")

a=make_crypt_df("MIRRORIMAGE")

x_mirror=make_num(a,"MIRROR")
x_egami=make_num(a,"EGAMI")
x_image=make_num(a,"IMAGE")

for (i in 0:9) {
    x=x_mirror+(x_image*10+i)
    k=(x<1000000) & ((x%%100000) == x_egami) & (a['M']!=0) & (a['I']!=0) & (x_image*10 > x_mirror)
    if (any(k)) {
        for (j in which(k)) {
            if (any(a[j,]==0)) { next }
            n1=x_mirror[j,]
            n2=(x_image[j,]*10+i)
            n3=as.numeric(paste0(rev(strsplit(as.character(n1+n2),"")[[1]]),collapse=""))
            if (n3>n1) {
                n4=n3-n1
                print(sprintf("i = %d",i))
                print(as.matrix(a[j,]))
                print("Left sum:")
                print(sprintf("%6d         ",n4))
                print(sprintf("%6d = MIRROR",n1))
                print(sprintf("%6d = IMAGE*",n3))
                print("Right sum:")
                print(sprintf("%6d = MIRROR",n1))
                print(sprintf("%6d = IMAGE*",n2))
                print(sprintf("%6d = *EMAGI",n1+n2))
                print("-----")
            }
        }
    }
}

```

    Matches the solution from the Bent.


## Problem 2

    An antimagic square is an NxN grid containing
    the integers from 1 through N^2, such that the
    sum of each row, column, and main diagonal is
    a different number, and those sums form a
    consecutive sequence of integers. There are
    no 3x3 antimagic squares, but there are some
    that are close in that the sequence of sums
    is consecutive except for one missing value.
    Find a 3x3 near antimagic square in which the
    three digit number represented by the top
    left (hundreds position) to bottom right
    (units position) main diagonal is a minimum.
    Present your answer as a 3x3 grid containing
    the digits 1 through 9.

    -- H.G. McIlvried III, PA G'53

TODO

    meh


## Problem 3

    On average, how often (once every N years)
    does a blue moon (2nd full moon in the same
    month) fall on New Year's Eve in Los Angeles?

    -- F.J. Tydeman, CA Delta'73

TODO

    meh


## Problem 4

    As a promotion, a pizza shop advertises a
    medium plus a large pizza, with up to five
    toppings each, for just the price of a large
    pizza. The advertisement further states that
    a patron has a choice of 1,048,576 different
    combinations of two pizzas. Assuming this
    statement is correct, if a topping can be
    used only once on a given pizza and the two
    pizzas can each have a different combination
    of toppings, how many different toppings does
    the pizza shop have?

    -- Allan Gottlieb's Puzzle Corner in
       Technology Review

Solution:

    This seems kind of obvious.
    
    1048576 = 2^20 then each pizza has 2^10=1024 combinations
    of 0 to 5 toppings.
    
    Need: C(n,0)+C(n,1)+C(n,2)+C(n,3)+C(n,4)+C(n,5) = 1024
    
    Just search for n:
    
```{r}

for (n in 5:15) {
    m=sum(choose(n,0:5))
    print(c(n,m))
}
```

    The solution is to use 11 toppings --> 1024 combinations.


## Problem 5

    Another order of merit (no ties) for A, B, C,
    D, E -- this time for Deception. In their usual
    chatty way, they are making remarks about
    their places. The remarks of those who were
    first and second are false, the rest are
    true.

        A: D was third.
        B: E was not first.
        C: I was not last.
        D: C was lower than B.
        E: B was second.

    What were their places (first to last)?

    -- Brain Puzzler's Delight
       by E.R. Emmet

TODO

    meh


## Bonus

    Consider a unit radius circle whose center is
    at the origin of a set of Cartesian
    coordinates. Construct a 2 Theta sector of this
    circle lying along the x-axis, with half the
    sector lying above the x-axis and half lying
    below the x-axis. Inscribe in this sector a
    series of circles of radii R1, R2, R3, ...,
    whose centers are on the x-axis. The first
    circle is tangent to the arc of the sector
    and its two sides. The second circle is
    tangent to the first circle and the sides of
    the sector; the third circle is tangent to
    the second circle and the sides of the
    sector; ad infinitum. Obviously, the ratio of
    the sum of the circles' areas to the sector's
    area is less than 1. For what value of Theta is
    this ratio the largest, and what is this
    largest ratio?

    -- Allan Gottlieb's Puzzle Corner in
    Technology Review

Solution:

    See the files in c:\rgc\test\play\bent\2019\3_fall
    
    Unfortunately my final solution said the angle should
    shrink to zero which bothered me:
    
    Maximum(area_ratio) = pi/4
    Optimal theta = 0

    I wonder what the Bent says ...
    
    yep -- that is what they got too ... and it didn't seem
    to bother them one bit.


## Computer Bonus

    The prime 5 exactly divides the sum of all
    the primes smaller than itself, (2+3)/5 = 1.
    The next prime with this property is 71. What
    is the next (after 71) such prime?

    -- Prime Numbers, the Most Mysterious Figures
    in Math by David Wells

Solution:

    Let's try brute force and see what happens.
    
```{r}
library(gmp)

# All primes up to 1e6
p=(2:1000000)[isprime(2:1000000)==2]

# cumsum includes each prime so need to subtract them out
s=cumsum(as.numeric(p))-p

# Find the sums that are integer multiples of the primes
k=(s%%p == 0)

# List them
t(rbind(p[k],s[k]))

```

    The primes that are factors of the sum of smaller primes are:

          Prime        Sum
    [1,]      2          0
    [2,]      5          5
    [3,]     71        568
    [4,] 369119 5536785000  <-- solution requested
    
        prime = 369119
        sum   = 5536785000
        
    I got lucky -- the solution was below 1e6.
