---
title: "Bent Summer 2009"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Summer 2009


## Problem 1

    The five offices of the Mudbridge
    Women's Guild are held by Mesdames
    Virtue, Wellbeloved, Xerxes,
    Youngblood, and Zenith. These powerful
    ladies are also very partial to a
    game of contract bridge and, on each
    weekday morning, a different one of
    the five acts as hostess to three of the
    others. The resulting fours are the
    same each week, and each lady has just
    one morning off each week -- the one
    on which she is not part of the four. All
    those whom Mrs. Virtue invites to her
    morning game invite Mrs. Wellbeloved
    to theirs. Mrs. Wellbeloved invites
    two of the ladies whom Mrs. Xerxes
    invites and, as the fourth member, a
    lady not among those invited by Mrs.
    Youngblood. Because she is president,
    Mrs. Zenith's invitations confer a
    certain distinction on the recipients.
    Who, besides herself, does not receive
    one? This problem is a bit tricky. Be
    sure to think outside the box.

    -- Tantalizer by Martin Hollis
    in New Scientist

TODO

    Ugh - I loathe logic problems.  Guessing
    games are best left to guessers.
    

## Problem 2

    Three sharpshooters fire (not simultaneously)
    at a rapidly spinning sphere. If each of them
    hits the sphere, what is the probability
    that all three shots land in the same
    hemisphere for the following two cases: (1)
    the great circle dividing the sphere into
    two hemispheres can have any orientation and
    (2) the great circle dividing the sphere
    into two hemispheres must pass through the
    sphere's poles?  Assume each point on the
    sphere has an equal chance of being hit.

    -- Daryl Cooper

Solution:

    In the first case it is a certainty.  After
    the first two shots are taken construct a
    great circle through these two points.  Regardless
    of where the third shot hits it is in the same
    half-sphere.
    
    In the second case, the first two shots will be
    in the same hemisphere (by definition).  The
    third shot will be in the same hemisphere iff
    the angle between the two extreme points is less
    than or equal to 180 degrees.  For example, if
    the three points were equidistant at the equator
    it would not be possible to separate into
    two hemispheres with one holding all three shots.
    
    In this case, we need the third shot to be within
    180 degrees of either shot 1 or shot 2 and the
    three shots will be in the same hemisphere.  Each
    of these has a P=0.5 so the overall probability
    that one or the other is true is:
    
        P = 1 - 0.5^2 = 3/4
        

## Problem 3

    Solve the following cryptic addition:
    SEVEN + THREE + TWO =
    TWELVE. We want the solution in
    which TWO is the product of two
    primes. No leading zeros; different
    letters are different digits; same letter
    is same digit. A computer is not necessary
    to solve this.

    -- Mathematical Puzzling by A. Gardiner

Solution: solved in the cryptic arithmetic notebook

      SEVEN + THREE + TWO = TWELVE
      
      TWO is the product of two primes

```{r}
source('rgc_lib.R')

a=make_crypt_df('seventhreetwotwelve')
x_seven=make_num(a,'seven')
x_three=make_num(a,'three')
x_two=make_num(a,'two')
x_twelve=make_num(a,'twelve')
k=(x_seven+x_three+x_two == x_twelve)
k=k & (a$s != 0) & (a$t != 0)
print(paste0("Number of solutions = ",sum(k)))
print(paste0('SEVEN  = ',x_seven[k,]))
print(paste0('THREE  = ',x_three[k,]))
print(paste0('TWO    = ',x_two[k,]))
print(paste0('TWELVE = ',x_twelve[k,]))
```

Need to select correct solution, check the factorization of the
candidate solutions for `TWO`:

```{r}
library(gmp)

print("Factors for 106:")
print(factorize(106))
print("Factors for 104:")
print(factorize(104))

```

Only the first solution has two prime factors.
The final solution is then:

    SEVEN  = 82524
    THREE  = 19722
    TWO    = 106
    TWELVE = 102352


## Problem 4

    Al and Ben play a game that uses a
    supply of marbles marked 1, 2, 3, etc.,
    and five saucers. Al puts marble 1
    into a saucer. Ben puts marble 2 into
    a saucer. Al puts marble 3 into a saucer,
    and so on. The only restriction is
    that the differences between any pair
    of numbers in a saucer must be different
    from the differences between
    any other pair in that saucer. For instance,
    a saucer must not contain 1, 5,
    14, and 18. If a player cannot place his
    number in any saucer, he loses. The
    game can go on for a long time, but
    what is the shortest possible game,
    and what are the numbers in the
    saucers when the game stops? Hint:
    When the game stops, all saucers
    have fewer than 4 marbles.

    -- Enigma by Stephen Ainley
    in New Scientist

Solution:  I think I solved this problem once before
by trial and error.

The only interesting part here might be to try random
solutions to see if I can solve it that way.

From the problem description there are (at most) 15 marbles
in the optimal solution with a loss at the 16th marble.
I will generate a random number of marbles between 8 and 16
and then play a "game" involving a random location for the
next marble and then see if that is feasible.  If so,
advance to next marble or try again until all of the
saucers have been tried.

```{r}
# Set ntrials to 1e6 to have a reasonable shot
# at finding a minimal set of 12/13 marbles.
# Be warned that the run time is substantial (~1 hr).
ntrials=1000
mmin=999
nmin=0

for (trial in 1:ntrials) {
    # print(paste0("Running trial = ",trial))
    s=vector("list",5)
    done=FALSE
    for (m in 1:16) {
        # print(paste0("Placing marble = ",m))
        tried=c()
        while (TRUE) {
            # Get a random saucer number
            n=sample(5,1)
            # Can I put marble m in saucer n?
            x=append(s[[n]],m)
            x=sort(outer(x,x,"-"))
            x=x[x>0]
            if (length(x)==length(unique(x))) {
                s[[n]]=append(s[[n]],m)
                break
            }
            tried=append(tried,n)
            if (length(unique(tried))==5) {
                # print("No move possible")
                done=TRUE
                break
            }
        }
        if (done) { break }
    }
    if (done & (m<16)) {
        print("Short solution found")
        print(m)
        for (i in 1:5) {print(s[[i]])}
        if (m<mmin) {mmin=m; smin=s; nmin=1}
        if (m==mmin) {nmin=nmin+1}
    }
}
if (mmin != 999) {
    print(paste0("Shortest solution found = ",mmin))
    print(paste0("Number of repeats for min solution = ",nmin))
    print(smin)
} else {
    print("No short solutions found")
}

```

After 3000K random trials the shortest game found was 12/13 moves.
Specifically, it is not possible to place marble 13.  Here is an
example:

    Saucer 1:  3   8
    Saucer 2:  4   6  11
    Saucer 3:  7  10
    Saucer 4:  5   9
    Saucer 5:  1   2  12

The other solutions are just rearranged saucers with same marble
combinations.


## Problem 5

    Find the value of:

    sqrt(2*sqrt(2*sqrt(2*sqrt(2*sqrt(2*. . .)))))

-- Crucible

Solution:

    S = sqrt(2*sqrt(2*sqrt(2*...)))
    
    S^2 = 2*sqrt(2*sqrt(2*...)) = 2*S
    
    S = 2


## Bonus

    N (an odd number) people own a bank. What is
    the minimum number of different locks that
    must be put on a safe so that when duplicate
    keys to these locks are distributed to the
    different owners, every majority contains
    a complete set of keys but no minority
    does? For example, 3 people need 3 locks with
    keys distributed as: 1-2, 2-3, 3-1; no single
    person can open the safe, but all pairs
    can. Express your answer as a function of N.

    -- Puzzle Corner by Allan Gottlieb in
    Technology Review

TODO

    I have seen this problem before - don't care
    to repeat it here.


## Double Bonus

    In 2006, a teacher said to his class, an
    American hero was born X years ago and died
    at age Y, where Y minus its reverse equals
    the square of the cube root of (X - Y).
    If half of this hero's birth year is a
    prime number, whose birthday was the teacher
    referring to, and what anniversary was 2006?

    -- Aziz S. Inan

TODO

    This is a virtual repeat of the Mozart question
    from the Bent of Winter 2008.
