---
title: "Bent Fall 2000"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Fall 2000


## Problem 1

    Smith, Jones, and Robinson make
    four statements each as follows:
    
        Smith:
        1. Jones owes me $10.
        2. Robinson owes me $5.
        3. All Robinson's statements are true.
        4. All Jones's statements are untrue.
        
        Jones:
        1. I owe no money to Smith.
        2. Robinson owes me $7.
        3. I am British.
        4. All Smith's statements are untrue.
        
        Robinson:
        1. I owe no money to anybody.
        2. Jones is a Dutchman.
        3. I always tell the truth.
        4. Two of Jones's statements are
        true, and two are false.
    
    One person made 4 true statements.
    Who? Find, for all of them,
    which statements are true and which
    are false.
    
    -- Brain Puzzler's Delight
    by E. R. Emmet

Solution:

Jones is telling the truth.
All statements by Smith and Robinson are false.

Details of each possible scenario are given below.

Assume Smith always tells the truth.

    Veracity of others:
    
    Jones:
    1. False (by Smith #1)
    
    Robinson:
    1. False (by Smith #2)
    
    Conclude this is not feasible.  Smith #3 is False since not all of
    Robinson's answers are truthful.

Assume Jones always tells the truth.

    Veracity of others:
    
    Smith:
    1. False (by Jones #1)
    2. False (by Jones #4)
    3. False (see Robinson answers - some are false)
    4. False (by assumption)
    
    Robinson:
    1. False (by Jones #2)
    2. False (by Jones #3)
    3. False (not all answers are True)
    4. False (by assumption)
    
    Conclude that this is feasible.  There are no contradictions.

Assume that Robinson always tells the truth.
If we assume they are all correct then we can assess
the veracity of the others as:

    Smith:
    1. Unknown
    2. False (per Robinson #1)
    3. True (assume Robinson is truthful)
    4. False (per Robinson #4)
    
    Jones:
    
    1. Unknown
    2. False (Robinson says he doesn't owe money to anybody)
    3. False (Robinson says he is a Dutch)
    4. False (some of Smith's statements are true)
    
    Conclude that this is not feasible.  It is not possible for Robinson #4 to be true
    since at least three of Jones statements are False.


## Problem 2

    Scalenia is a country bounded by
    three straight frontiers called A, B,
    and C, of course, each of a different
    length but each an exact number of
    kilometers long. The curious thing is
    that the length of A+B+C, and
    the lengths of A+B-C, A+C-B, and
    B+C-A, are all precisely square
    numbers of kilometers. If Scalenia
    has the smallest perimeter consistent
    with this curious fact, what is
    the length of each of its frontiers?
    -- An Enigma by Stephen Ainley

Solution: try exhaustive search?  Arbitrarily use range of 1-100.

It appears the lengths are 10, 10, and 16.

    A=10, B=10, C=16
    
    10+10+16 = 36
    10+10-16 = 4
    10+16-10 = 16
    10+16-10 = 16

```{r}
for (A in 1:100) {
  for (B in 1:100) {
    for (C in 1:100) {
      d=A+B+C
      if (trunc(sqrt(d))^2 == d) {
        d=A+B-C
        if ((d>0) & (trunc(sqrt(d))^2 == d)) {
          d=A+C-B
          if ((d>0) & (trunc(sqrt(d))^2 == d)) {
            d=B+C-A
            if ((d>0) & (trunc(sqrt(d))^2 == d)) {
              print(c(A,B,C,A+B+C))
            }
          }
        }
        
      }
    }
  }
}
```

## Problem 3

    A 12 hour AM/PM clock, whose
    display always shows six digits in
    the format hh:mm:ss, sits in a dark
    room. The digits of the LED display
    are formed by the familiar seven-element
    figures that look like
    dominos when all seven elements are
    lit. Between which two consecutive
    times is the change in the relative
    brightness of the display the
    greatest? Assume that, when it is lit,
    each element of each digit emits
    exactly the same amount of light.
    Also, assume that 0 is 6 segments lit,
    1 is 2 seg, 2 is 5 seg, 3 is 5 seg, 4 is 4
    seg, 5 is 5 seg, 6 is 6 seg, 7 is 4 seg, 8
    is 7 seg, and 9 is 6 segments lit.
    -- The Platinic Corner

Solution: should be able to enumerate this and find the solution
using brute force.  The maximum brightness shift is an increase
of 5 units at times of:

    01:59:59 to 02:00:00
    07:59:59 to 08:00:00
    11:59:59 to 12:00:00
    
Of course this is for both AM and PM.

The problem requests the *relative change in brightness* but I don't know
if this is as a percent of the initial brightness or in terms of the absolute
number of illuminated segments.  All of the time changes above have the same
change in absolute brightness (5 LED segments).

If we are concerned with a percent change in brightness then the largest
change occurs from 11:59:59 to 12:00:00 (5/26 ~ 19.2% increase).


```{r}
# Mapping of digits 0-9 to number of segments
# This is a proxy for brightness
dig2seg = c(6,2,5,5,4,5,6,4,7,6)

calc_brightness = function(h,m,s) {
  b=0
  d=trunc(h/10); b=b+dig2seg[d+1]
  d=h%%10;       b=b+dig2seg[d+1]
  d=trunc(m/10); b=b+dig2seg[d+1]
  d=m%%10;       b=b+dig2seg[d+1]
  d=trunc(s/10); b=b+dig2seg[d+1]
  d=s%%10;       b=b+dig2seg[d+1]
  return(b)
}

# Starting at 12:00:00 and going to 11:59:59
# Enumerate the brightness.
# Note there are 12*60*60 = 43200 values
xxx=rep(0,43200)
ttt=rep("00:00:00",43200)
n=1
for (h in c(12,1,2,3,4,5,6,7,8,9,10,11)) {
  for (m in 0:59) {
    for (s in 0:59) {
      xxx[n]=calc_brightness(h,m,s)
      ttt[n]=sprintf("%02d:%02d:%02d",h,m,s)
      n=n+1
    }
  }
}
# Allow for wrap-around at noon (also end of day)
xxx=c(xxx,xxx)
ttt=c(ttt,ttt)
# Check for brightness changes
dx=diff(xxx)
a=data.frame(t=ttt,Brightness=xxx)
# Location of max brightness shift (lower value)
k=which(dx==max(dx))
# Give lower and higher values of time/brightness
n=sort(c(k,k+1))
# Dump the times/brightness levels
a[n,]
```

## Problem 4

    Solve this cryptic division, where
    the right side is a repeating decimal.
    As usual, the same letter is the same
    digit throughout, different letters
    are different digits, and base - 10 is
    used: 

        EVER/NNNNN=.ONAND ONAND. . . .

    -- Thomas McNelly
    via Technology Review

Solved in the cryptic_arithmetic notebook

    EVER = 7678
    NNNNN = 33333
    ONAND = 23034

## Problem 5

    An urn contains 10 balls labeled 0
    through 9. You reach into the urn
    and pick three balls (without
    replacement) at random. What is the
    probability that you can form a
    three-digit prime number (leading
    zero not allowed) from the three
    balls?
    -- H.G. McIlvried, PA Gamma'53

Solution:

There are 720 permutations that can be formed from 10 balls taken 3 at a time.
There are 6 ways to arrange each permutation.

There are 97 primes in the range 103-983 without allowing for repeated digits.

Not sure how to put these together since the probability of each rearrangement
is not an independent quantity.

Test with Monte-Carlo:

```{r}
make_num = function(a,b,c) { return(100*a+10*b+c) }

# Get a list of three digit primes (e.g. < 1000)
# Eliminate values with leading zero (e.g. < 100)
p=read.table('primes.txt')[,1]
p=p[(p>99)&(p<1000)]

t1=Sys.time()
n_trials=10000000
n=0
for (i in 1:n_trials) {
  # Select 3 digits at random
  s=sample(0:9,3)
  a=s[1]; b=s[2]; c=s[3]
  success=0
  # Build the 6 permutations and test them
  if (make_num(a,b,c) %in% p) { success=1 }
  if (make_num(a,c,b) %in% p) { success=1 }
  if (make_num(b,a,c) %in% p) { success=1 }
  if (make_num(b,c,a) %in% p) { success=1 }
  if (make_num(c,a,b) %in% p) { success=1 }
  if (make_num(c,b,a) %in% p) { success=1 }
  n=n+success
}
t2=Sys.time()
print(paste0('Num trials = ',n_trials))
print(paste0('Num success = ',n))
print(paste0('Probability = ',n/n_trials))
print(paste0('Execution time = ',difftime(t2,t1,units="secs"),' seconds'))
```

The probability appears to be about 44.2%.


## Bonus

    In a uniform (direction and
    magnitude) gravity field, the fastest
    way for a frictionless particle to slide
    from A to B in a vacuum is along a
    cycloid. If A and B are 500 km apart
    and at the same elevation, the
    maximum depth of the cycloid is just
    over 159 km. If, however, there is a
    depth limitation of 50 km, describe
    the fastest curve. Also, with g = 9.8
    m/s/s, how long (in seconds) does it
    take for a particle started at rest to
    go from A to B?
    -- David Gluss
    via Technology Review

TODO


## Double Bonus

    Devise a graphical
    method of locating point P in the
    plane of a given triangle ABC such
    that triangles PAB, PBC, and PCA
    have equal perimeters.
    -- John Rule
    via Technology Review

TODO
