---
title: "Bent Winter 2015"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Winter 2015


## Problem 1

    Based on the following informa-
    tion, find the values of A, B, C, and
    D, which are unique positive inte-
    gers, all less than five.
    If A is 1, then B is not 3.
    If B is not 1, then D is 4.
    If B is 1, then C is 4.
    If C is 3, then D is not 2.
    If C is not 2, then D is 2.
    If D is 3, then A is not 4.

    -- Brain Workout IQ Challenges
    by Terry Stickels and
    J. J. Mendoza Fernandez

Solution:

    A=3, B=1, C=4, D=2

    We can try to reason this out instead of throwing it
    against the computer.
    
    If C = 3 then D != 2 but      \ Contradiction so C can't be 3
    if C = 3 then C != 2 so D =2  /

    What if B is 1 then C = 4.
    Since C is not 2 --> D = 2.
    A must be 3.
    
    What if B is 2 then D=4    \ Contradiction so B can't be 2
    but if C is not 2 then D=2 /

    What if B is 3 then A is not 1--> A=2, C=1 --> D=2 = Contradiction
    D=4
    
    What if B is 4 then D=4 --> Contradiction
    
    
## Problem 2

    A point in the interior of a square
    is exactly 2, 3, and 4 cm from suc-
    cessive corners of the square. What
    is the length of a side of the square,
    expressed in exact terms?

    -- adapted from The Canterbury
    Puzzles by H.E. Dudeney

Solution:

    The situation is diagrammed below in ASCII art:
    
        1----------2
        |          |
        |          |
        |   0      |
        |          |
        3----------4

    The distance from 0-3 is 2, distance from 0-1 is 3 the
    distance to one of the other two corners is 4.
    
    If the length of a side for the square is d then we can
    call the angles:
    
        1-3-0 = theta
        3-1-0 = alpha

    Then:
    
        d = 2*cos(theta)+3*cos(alpha)
        2/sin(alpha) = 3/sin(theta)
    
    Either:
    
        Distance from O to corner nearest length of 3
        d1^2 = (3*cos(alpha))^2 + (d-3*sin(alpha))^2
    or
        Distance from O to corner nearest length of 2
        d2^2 = (2*cos(theta))^2 + (d-2*sin(theta))^2
    
    We need to solve to make d1==4 or d2==4.
    
    This is getting ugly.

    What if we just sweep the length of a side and look at the other two
    distances.  This is complicated by the fact that we still don't know
    the theta, alpha angles but we do have two equations and two unknowns.
    
    Can use the law of cosines for the SSS triangle:
    
    a,b,c = sides
    A,B,C = angles opposite a,b,c
    
    cos(C) = (a^2+b^2-c^2)/(2*a*b)
    cos(A) = (b^2+c^2-a^2)/(2*b*c)
    cos(B) = (c^2+a^2-b^2)/(2*a*c)

    For our problem we will use:
    
    a = 2
    b = 3
    c = d (side of square)
    
    cos(theta) = (d^2+2^2-3^2)/(2*2*d)
    cos(alpha) = (d^2+3^2-2^2)/(2*3*d)
    
    
```{r}

# Note that foo1() and foo2() are identical except for the return values

foo1 = function(d) {
    theta=(d^2+2^2-3^2)/(2*2*d)
    alpha=(d^2+3^2-2^2)/(2*3*d)
    if (abs(theta)>1 || abs(alpha)>1) {
        d1=NA
        d2=NA
    } else {
        theta=acos(theta)
        alpha=acos(alpha)
        d1 = sqrt((3*cos(alpha))^2 + (d-3*sin(alpha))^2)
        d2 = sqrt((2*cos(theta))^2 + (d-2*sin(theta))^2)
    }
    return(d1)
}

foo2 = function(d) {
    theta=(d^2+2^2-3^2)/(2*2*d)
    alpha=(d^2+3^2-2^2)/(2*3*d)
    if (abs(theta)>1 || abs(alpha)>1) {
        d1=NA
        d2=NA
    } else {
        theta=acos(theta)
        alpha=acos(alpha)
        d1 = sqrt((3*cos(alpha))^2 + (d-3*sin(alpha))^2)
        d2 = sqrt((2*cos(theta))^2 + (d-2*sin(theta))^2)
    }
    return(d2)
}

x=seq(1,5.0,0.1)
y1=c()
y2=c()
for (d in x) {
    d1=foo1(d)
    d2=foo2(d)
    y1=append(y1,d1)
    y2=append(y2,d2)
}

d1=optimize(function(x) {abs(foo1(x)-4)},c(1,5))
d2=optimize(function(x) {abs(foo2(x)-4)},c(1,5))

print("To make distance=4 to corner nearest length 3")
print(d1,digits=15)
print("To make distance=4 to corner nearest length=2")
print(d2,digits=15)
```

The solutions are for squares of side:

    d = 4.23525477726028
    d = 4.57935453145691

The solution in the bent is closed form:

    sqrt(10+3*sqrt(7)) ~ 4.2352395366961
    
Which matches the first solution (corners are literally ordered as 2, 3, 4)
at least through 5 significant digits.

The closed form solution is fairly easy to calculate and I probably
should have tried that.  Put the corner connected to the distance of 3
at the origin of the x-y coordinates.  The length of a side of the square
is L and the location of the point is (x,y).  The distance to the three
corners is then:

    e1: x^2 + (L-y)^2 = 2^2
    e2: x^2 + y^2     = 3^2
    e3: (L-x)^2 + y^2 = 4^2

Solving is fairly simple:

    solve(e1-e2,y)
    solve(e3-e2,x)
    
    substitute the results for x,y into e2 and solve double quadratic:
    
    L^4 - 20*L^2 + 37 = 0 --> L = sqrt(10 +/- 3*sqrt(7))
    
Note that they implicitly assumed the distance of 4 is to the corner
nearest the distance of 3.  If, instead, the distance of 4 is to the
corner nearest the distance of 2 we have:

    e1: x^2 + (L-y)^2 = 3^2
    e2: x^2 + y^2     = 2^2
    e3: (L-x)^2 + y^2 = 4^2
    
Same solution technique:

    x=(L^2-12)/(2*L); y=(L^2-5)/(2*L)
    
    2*L^4 - 50*L^2 + 169 = 0
    
    L^2 = (25+/-sqrt(287))/2
    
    L = +/- sqrt( (25 +/- sqrt(287))/2 )
      ~ +/- 2.007352, +/- 4.57936
      
    Clearly the only feasibly solution is sqrt((25+sqrt(287))/2) ~ 4.57936.


## Problem 3

    A hexomino is a polygon consist-
    ing of six unit squares joined edge
    to edge along the full length of
    their sides. There are 35 different
    hexominos. How many of these can
    be folded into a unit cube without
    making any slits?

    -- More BrainMatics Logic Puzzles
    by Ivan Moscovich

TODO

    meh
    

## Problem 4

    The game of 5-Spot consists of a
    deck of cards with five capital let-
    ters on each card. Between any two
    cards, there is one, and only one,
    matching letter. If each letter occurs
    the same number of times in the
    deck, what is the maximum number
    of cards in the deck? What is one
    such deck? Assume the letters are
    consecutive, i.e., A, B, C, ... , and
    present your answer as a sentence
    of five-letter "words" with one word
    for each card in the deck. Within
    each word, the letters should be
    in alphabetical order, and then the
    words should also be in alphabetical
    order.

    -- Fred J. Tydeman, CA Delta '73

TODO

    meh
    

## Problem 5

    Consider a ten-digit integer that
    uses each of the digits 0 through 9
    exactly once and has the proCperty
    that the sum of adjacent digits is a
    prime. The smallest such number
    is 1203476589. In total, how many
    numbers have this property? One is
    not a prime number. No leading ze-
    roes are allowed. You could use your
    brain cells instead of a computer, but
    it would be tedious.

    -- adapted from Math-E-Magic
    by R. Blum, A. Hart-Davis,
    B. Longe, and D. Niederman

Solution:

    We will generate all the permutations but we can
    short cycle as soon as we find a sum that is
    not a prime.
    
    Yes - this is very ugly but it was quick to write.

```{r}
library(gmp)

n=0
for (i1 in 1:9) {
    for (i2 in 0:9) {
        if (i2 == i1) { next }
        if (!isprime(i1+i2)) { next }
        for (i3 in 0:9) {
            if (i3 %in% c(i1,i2)) { next }
            if (!isprime(i2+i3)) { next }
            for (i4 in 0:9) {
                if (i4 %in% c(i1,i2,i3)) { next }
                if (!isprime(i3+i4)) { next }
                for (i5 in 0:9) {
                    if (i5 %in% c(i1,i2,i3,i4)) { next }
                    if (!isprime(i4+i5)) { next }
                    for (i6 in 0:9) {
                        if (i6 %in% c(i1,i2,i3,i4,i5)) { next }
                        if (!isprime(i5+i6)) { next }
                        for (i7 in 0:9) {
                            if (i7 %in% c(i1,i2,i3,i4,i5,i6)) { next }
                            if (!isprime(i6+i7)) { next }
                            for (i8 in 0:9) {
                                if (i8 %in% c(i1,i2,i3,i4,i5,i6,i7)) { next }
                                if (!isprime(i7+i8)) { next }
                                for (i9 in 0:9) {
                                    if (i9 %in% c(i1,i2,i3,i4,i5,i6,i7,i8)) { next }
                                    if (!isprime(i8+i9)) { next }
                                    for (i10 in 0:9) {
                                        if (i10 %in% c(i1,i2,i3,i4,i5,i6,i7,i8,i9)) { next }
                                        if (!isprime(i9+i10)) { next }
                                        # print(c(i1,i2,i3,i4,i5,i6,i7,i8,i9,i10))
                                        n=n+1
                                    }
                                }
                            }
                        }
                    }
                }
            }
        }
    }
}
print(n)
```

There are 845 numbers with this property (10 digit numbers with adjacent digit
sums being primes).


## Bonus

    In the game of five card stud poker with
    one-eyed jacks (the jacks of spades and
    hearts) wild, what is the probability of
    getting exactly one pair? Also, what is the
    probabil- ity of getting a bust hand, i.e., a
    hand with not even a pair? Assume that a wild
    card is always used to maximize the value of
    a hand. The two one-eyed jacks are in hearts
    and spades.

-- Howard G. McIlvried III, PA Gamma '53

Solution:

    Exhaustive enumeration will be as clear as working through the
    combinatorics.

```{r}
library(gtools)

poker_hand=combinations(52,5,0:51)
```

```{r}

check_poker_hand = function(x, wild=c(41,42)) {
    # Determine the result for a poker hand, x
    # The values in x are simply integers in range 0-51
    # trunc(x/4)+1 = pip value
    # x%%4 = suit
    #
    # Wildcard codes are given by wild.
    #
    # Results are:
    has_a_straight=1
    has_a_flush=2
    has_pair_or_better=3
    has_wild_card=4
    has_wild_cards=5
    has_bust=0
    has_one_pair=10
    
    k=x %in% wild
    if (sum(k) == 0) {                    # no wildcard
        pip=trunc(x/4)+1
        if (length(unique(pip))==5) {     # no pair(s) or better
            if (length(unique(x%%4))>1) { # no flush
                s1=all(diff(sort(pip))==1) # no straight with aces low
                s2=all(diff(sort(ifelse(pip==1,14,pip)))==1)  # aces high
                if (!s1 & !s2) {
                    return(has_bust)
                } else {
                    return(has_a_straight)
                }
            } else {
                return(has_a_flush)
            }
        } else {
            if (length(unique(pip))==4) {
                return(has_one_pair)
            } else {
                return(has_pair_or_better)
            }
        }
    } else {
        # Need to make sure:
        # Only one wild card, no flush, no straight, 4 distinct pip values
        if (sum(k)>1) {
            return(has_wild_cards)
        }
        y=x[!k]
        pip=trunc(y/4)+1
        suit=y%%4
        if (length(unique(suit))==1) {
            return(has_a_flush)
        }
        if (length(unique(pip))<4) {
            return(has_pair_or_better)
        }
        # How do we check for straight?  Maximum gap is a 2 and can only be
        # one of those.
        d=sort(diff(sort(pip)))
        s1=(d[length(d)]<3) & (d[length(d)-1]<2)
        d=sort(diff(sort(ifelse(pip==1,14,pip))))
        s2=(d[length(d)]<3) & (d[length(d)-1]<2)
        if (s1 | s2) {
            return(has_a_straight)
        }
        return(has_one_pair)
    }
}
```

Initial testing:

    Total = 2598960

    Case              Hands  Bent Solution
    --------------------------------------
    Wildcard         480200
    Pair or better  1054312
    Flush              4158
    Straight           8160
    Bust            1052130   1052130

After adding the logic to properly interpret wild-cards and isolate distinct
pairs:

    Case               Hands  Probability
    -------------------------------------
    Straight           25548    
    Flush               8998  
    Better than pair  304032   
    Two Wildcards      19600 
    Bust             1052130   0.4048273
    One pair         1188652   0.4573568


## Computer Bonus

    Some pairs of "mirror image" integer"
    have mirror image squares. For example, consider 12
    and 21 and their squares, 144 and 441. How many non-
    palindromic four-digit mirror image pairs have mirror
    image squares?

    -- Technology Review

TODO

    Exhaustive search?
    
```{r}

is_palindrome = function(n) {
    x=strsplit(as.character(n),"")[[1]]
    return(all(x == rev(x)))
}

is_mirror_image = function(n1,n2) {
    x1=strsplit(as.character(n1),"")[[1]]
    x2=strsplit(as.character(n2),"")[[1]]
    if (length(x1) != length(x2)) {
        return(FALSE)
    }
    return(all(x1==rev(x2)))
}

n=0
for (i in 1000:9999) {
    if (!is_palindrome(i)) {
        j=as.numeric(paste0(rev(strsplit(as.character(i),"")[[1]]),collapse=""))
        if (j>i) {
            if (is_mirror_image(i*i,j*j)) {
                n=n+1
                print(c(i,j))
            }
        }
    }
}
print(n)
```

There are 18 such number pairs:

    1002 2001
    1003 3001
    1011 1101
    1012 2101
    1013 3101
    1021 1201
    1022 2201
    1031 1301
    1102 2011
    1103 3011
    1112 2111
    1113 3111
    1121 1211
    1122 2211
    1202 2021
    1212 2121
    2012 2102
    2022 2202

One subtlety to avoid double counting is to check j>i.  At first I
was double counting.
