---
title: "Bent Spring 2020"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Spring 2020


## Problem 1

    I started with 10 cards with a different
    digit on each card. After discarding one
    card, I used the remaining nine cards to form
    a two digit prime, a three-digit prime, and a
    four-digit prime (with no leading zeros).
    Even if I told you what the sum of these
    three primes was, you would not be able to
    deduce the digit I discarded; but if, in
    addition, I told you that the middle two
    digits of the sum were each equal to the
    digit I discarded, then you would be able to
    deduce the sum of the three primes I found.
    What was that sum?

    -- Susan Denhan in New Scientist

Solution:

    This is a cryptic arithmetic problem in disguise.
    I will generate a solution using those tools.
    
```{r}
source('rgc_lib.R')
library(gmp)

a=make_crypt_df('abcdefghi')
x2=make_num(a,'ab')
x3=make_num(a,'cde')
x4=make_num(a,'fghi')
k=(isprime(x2[,1])==2) & (isprime(x3[,1])==2) & (isprime(x4[,1])==2)

# All solutions not including hint about center digits
# k = list of index numbers into a
k=k & (a['a']!=0) & (a['c']!=0) & (a['f']!=0)
k=which(k)
print(length(k))

# Find the 4 digit sums with a repeated digit
# that is not included in the original digits
# k2 = list of index numbers into a
# zzz = matrix of the feasible solutions
zzz=c()
k2=c()
s=x2+x3+x4
for (i in k) {
    b=as.numeric(strsplit(as.character(s[i,]),"")[[1]])
    if (length(b)==4) {
        if (b[2] == b[3]) {
            dup_digit=b[2]
            missing_digit=(0:9)[which(!(0:9 %in% a[i,]))]
            if (dup_digit == missing_digit) {
                ### print(c(x2[i,],x3[i,],x4[i,],s[i,],missing_digit))
                zzz=append(zzz,c(x2[i,],x3[i,],x4[i,],s[i,],missing_digit))
                k2=append(k2,i)
            }
        }
    }
}
zzz=matrix(zzz,ncol=5,byrow=TRUE)

# Need to find a value that is unique in s[k2] but
# is not unique in s[k]

for (i in unique(s[k2,])) {
    if (sum(s[k2,]==i) == 1) {
        if (sum(s[k,]==i) != 1) {
            print(sprintf("Feasible total = %d",i))
            print(zzz[zzz[,4]==i,])
        }
    }
}

```

    The only feasible sum is given below:

    Feasible total = 8991
    
        2 digit prime   = 23
        3 digit prime   = 467
        4 digit prime   = 8501
        sum             = 8991
        discarded digit = 9


## Problem 2

    On each day of a non-leap year, I gave my
    daughter a new penny, which I obtained
    directly from the U.S. Mint and which I,
    consequently, assumed to be of standard
    weight.  I learned, however, that there had
    been a minor production problem at the Mint
    one day during the year, and a few pennies
    had been released (I don't know on which day)
    that were heavier or lighter than standard.
    Assuming my daughter's penny collection
    contains at most one off-spec penny, what is
    the minimum number of weighings with a common
    two pan balance scale needed to determine
    whether or not she has an odd penny; and, if
    she does, which is the odd penny and whether
    it is heavier or lighter than standard?

    -- The Puzzle Corner in Technology Review

TODO

    Partition into thirds - except 365/3 doesn't work.
    
    Compare first two parts and remainder -- 2 tests reduce to 1/3 of total
    
        Scales balance
            Try remainder
            If it balances --> no odd pennies
            If heavier     --> odd penny heavier in final 1/3
            If lighter     --> odd penny lighter in final 1/3

        Scales imbalance
            Swap lighter group with remainder
            If it balances --> lighter group has odd penny
            If heavier     --> impossible -- this would be three different weights
            If lighter     --> heavier group has odd penny
            
    Repeat until identify the odd penny

        365*(1/3)^(n/2) <= 1
        
        n=10.74 weighings.
        

## Problem 3

    Solve the following cubic cryptarithm. All
    the usual rules apply: the same letter always
    represents the same digit, each digit is
    represented by a unique letter, and there are
    no leading zeros.

    ABCDE = (A + B + C + D + E)^3

    -- Source Unknown

Solution:

```{r}
a=make_crypt_df('abcde')

x_abcde=make_num(a,'abcde')
s=(a['a']+a['b']+a['c']+a['d']+a['e'])^3

k=(x_abcde == s) & (a['a']!=0) & (a['b']!=0) & (a['c']!=0) & (a['d']!=0) & (a['e']!=0)

print(as.matrix(a[k,]))

```

    Solution:
    
         a b c d e
    5961 1 9 6 8 3

    sum = 27, 27^3 = 19683


## Problem 4

    Brag is a poker-like game played with a
    standard deck of 52 cards, but a hand
    consists of only three cards instead of the
    usual five. The possible hands that can be
    dealt are defined below:

    (1) Royal flush (AKQ in same suit)
    (2) Straight flush (sequence of 3
    cards in same suit)
    (3) 3 of a kind (3 cards of same
    denomination)
    (4) Straight (sequence of 3 cards
    representing at least 2 different
    suits)
    (5) Flush (3 cards in same suit but
    not in sequence)
    (6) Pair (2 cards of same
    denomination plus an unmatched 3rd
    card)
    (7) High card (3 unmatched cards -- 2
    or 3 different suits and 3 different
    denominations, not in sequence)

    Aces can be high or low for a straight. How
    many of each type of hand are there? Present
    your answer as a table with 7 rows and 2
    columns. Enter the numbers 1 to 7 in column 1
    (no need to repeat the whole definitions); in
    column 2 list the number of different ways of
    being dealt each type of hand.  Be careful
    not to double count. For example, if you
    count all the straight flushes, you will be
    counting the Royal Flushes twice.

    -- The Ultimate Book of Card Games by The
       Diagram Group

Solution:

    Exhaustive search -- yes this is very ugly.

```{r}
deck=0:51

a=combinations(52,3,deck)
pip=trunc(a/4)+2
suit=a%%4

n=rep(0,7)
for (i in seq(nrow(a))) {
    if (length(unique(pip[i,]))==1) {           # Three of a kind
        n[3]=n[3]+1
    } else if (length(unique(suit[i,]))==1) {   # Some kind of flush (royal, straight, other)
        if (all(sort(pip[i,])==c(12,13,14))) {  # Royal Flush
            n[1]=n[1]+1
        } else {
            x=sort(pip[i,])
            if (all(diff(x)==1)) {              # Straight flush
                n[2]=n[2]+1
            } else if (c(14) %in% x) {
                x=sort(ifelse(x==14,1,x))
                if (all(diff(x)==1)) {          # Straight flush (ace low)
                    n[2]=n[2]+1
                } else {
                    n[5]=n[5]+1                 # Flush
                }
            } else {
                n[5]=n[5]+1                    # Flush
            }
        }
    } else if (length(unique(pip[i,]))<3) {    # Pair
        n[6]=n[6]+1
    } else {                                   # Straight or bust
        x=sort(pip[i,])
        if (all(diff(x)==1)) {                 # Straight
            n[4]=n[4]+1
        } else if (c(14) %in% x) {
            x=sort(ifelse(x==14,1,x))
            if (all(diff(x)==1)) {             # Straight (ace low)
                n[4]=n[4]+1
            } else {
                n[7]=n[7]+1                    # Bust
            }
        } else {
            n[7]=n[7]+1                        # Bust
        }
    }
}
print(n)
```

    Results
    
    1   4       Royal Flush
    2   44      Straight Flush
    3   52      Three of a kind
    4   720     Straight
    5   1096    Flush
    6   3744    Pair
    7   16440   High Card (Bust)


## Problem 5

    What is the remainder when the number
    2^4,700,063,497 is divided by 4,700,063,497?

    -- Prime Numbers, the Most Mysterious Figures
    in Math by David Wells

Solution:

    I am pretty sure there are modulo-arithmetic rules to
    make this relatively easy to clculate but I don't know
    them so I am reduced to brute force.

    Brute force with python: remainder = 3 after crunching for 35 seconds.
    
    Maxima exploded with "out of heap" message
    
    R (gmp) barfed with "exponent e too large for pow(z,e) = z^e"
    
    If I set the modulus as shown below then things seem to work.
    When you set the modulus on a bigz value all the calculations
    are peformed using the given modulus:
    
    b=as.bigz(4700063497)
    a=as.bigz(2,mod=b)
    r=as.bigz(a^b)
    
    r = 3 with essentially no delay
    
    So gmp works but you have to treat it kindly.
    Python just sort of works.


## Bonus

    Consider a chain (loop or circle) of length L
    = n(n+1)/2 with links of integral lengths 1,
    2, 3, 4, ..., n  occurring sequentially. You
    desire to tightly wrap the chain around three
    posts that form the vertices of an
    equilateral triangle with vertices a distance
    of S = L/3 apart, so that the posts occur
    exactly at a break between two links. The
    smallest example of such a chain has n = 9
    links; for this chain, S = L/3 = 9(10)/6 =
    15; one post is between links 3 and 4, the
    second post is between links 6 and 7, and the
    third post is between links 8 and 9, so the
    three sides are (4 + 5 + 6), (7 + 8), and (9
    + 1 + 2 + 3). What is the largest chain with
    less than 1,000 links that will form an
    equilateral triangle?  What is the shortest
    chain that can be wrapped around 3 posts in
    more than one way? Make the usual simplifying
    assumptions: the chain and the corner posts
    have zero radii, there is no overlap where
    the ends of the chain join, etc.

    -- The Puzzle Corner in Technology Review

Solution:

    For a chain of n links the total length is n*(n+1)/2.
    The length of each side is 1/3 of this total
    or:
            S = n*(n+1)/6

    In addition, there must be a set of links that
    sum to S.  If there are breakpoints after link
    m1,m2, and m3 then we know:

        length m1 to m2: m2*(m2+1)/2 - m1*(m1+1)/2 = S
        length m2 to m3: m3*(m3+1)/2 - m2*(m2+1)/2 = S
        
    The third side gets tricky assuming it wraps
    around the end of the chain.  The length of the end
    plus the start must sum to S:
    
        (length m3 to n) + (length up to m1) = S 
        
        n*(n+1)/2 - m3*(m3+1)/2 + m1*(m1+1)/2 = S
        
    
    I will assume m1, find m2 (if it exists), and then find m3
    (if it exists) and then check the remainder length.
    
    I do not know how to find these breakpoints efficiently.
    I will do it with brute force - but it seems like
    there must be a better way.

```{r}
# Give a chain of n links = 1, 2, 3, ..., n
# Find the next subchain starting after link m that is
# precisely s units long
next_link = function(n,m,s) {
    k=1:n
    t=k*(k+1)/2
    which(t== (m*(m+1)/2+s))
}

for (n in 1:1000) {
    target=n*(n+1)/6
    if (target == trunc(target)) {
        for (first_link in 0:(n-1)) {
            second_link=next_link(n,first_link,target)
            if (length(second_link)==0) { next }
            last_link=next_link(n,second_link,target)
            if (length(last_link)==0) { next }
            remainder=n*(n+1)/2-last_link*(last_link+1)/2+first_link*(first_link+1)/2
            if (remainder == target) {
                print(c(n,first_link,second_link,last_link))
            }
        }
    }
}

```

    Using some ugly brute force although the runtime isn't too long:
    
    Longest chain = 989 links
        Break after 164, 594, 824
        
    Multiple solutions = 125 links
        Break after  2, 72, 102
        Break after 57, 92, 117

    All solutions with less than 1000 links:

        Chain   1st Break  2nd Break  3rd Break
        ---------------------------------------
        9       3          6          8
        90      15         54         75
        125     2          72         102
        125     57         92         117
        153     51         102        135
        189     15         110        155
        440     24         255        360
        819     56         476        671
        989     164        594        824


## Computer Bonus

    Find three consecutive triangular numbers
    whose sum is a perfect square. The first such
    set is 15 + 21 + 28 = 64 = 8^2. What is the
    third such set?  A triangular number is a
    number of the form n(n + 1)/2, where n is a
    positive integer.

    -- Elementary Number Theory by David M.
       Burton

Solution:

    Brute force
    
```{r}

T = function(n) { n*(n+1)/2 }

s=c(T(1),T(2),T(3))
n=3

num_found=0
while (TRUE) {
    m=sum(s)
    k=round(sqrt(m))
    if (k*k == m) {
        print(n)
        print(s)
        print(m)
        print(k)
        print("---------------")
        num_found=num_found+1
        if (num_found == 5) { break }
    }
    n=n+1
    s[1]=s[2]
    s[2]=s[3]
    s[3]=T(n)
}

```

    This seems awfully easy.
    

    n            T(n)                   sum    sqrt(sum)
    ----------------------------------------------------
    5   6   7    15     21     28       64      8
    14  15  16   105    120    136      361     19
    63  64  65   2016   2080   2145     6241    79  <-- requested values
    152 153 154  11628  11781  11935    35344   188
    637 638 639  203203 203841 204480   611524  782
    ----------------------------------------------------
