---
title: "Bent Winter 2013"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Winter 2013


## Problem 1

    I am participating in a Super Bowl XLVII
    betting pool. You know, the kind consisting
    of a 10x10 grid with the cells numbered 00 to
    99, where the first digit refers to the last
    digit of the AFL team's score, and the second
    digit refers to the last digit of the NFL
    team's score. Each participant pays $1 per
    square, and $25 is paid to the person holding
    the winning number at the end of each
    quarter. I paid $2 for two squares, picked
    two slips at random from the basket, and got
    00 and 88. A friend who arrived after all
    the squares were taken, wants to buy my
    numbers. What is a fair price for each of my
    numbers? The following table provides data on
    the final digits for the quarterly scores of
    the 46 previous Super Bowls.

    Final Digit   0  1 2  3  4  5  6  7 8  9
    Number of   101 21 9 56 38 10 30 77 9 17
    Appearances

-- Don A. Dechman, TX A'57

Solution:

    We should be able to enuerate the probabilities
    
```{r}
p=c(101,21,9,56,38,10,30,77,9,17)

p0 = p[1]/sum(p); p00=p0^2
p8 = p[9]/sum(p); p88=p8^2

e00=0
e88=0
for (i in 0:4) {
    e00=e00+(25*i)*(choose(4,i)*p00^i*(1-p00)^(4-i))
    e88=e88+(25*i)*(choose(4,i)*p88^i*(1-p88)^(4-i))
}
print(paste0("Expected winnings for 00 = ",round(e00,2)))
print(paste0("Expected winnings for 88 = ",round(e88,2)))

# Alternate method
print("Alternate solution method:")
print(paste0("E(00) = ",4*25*p00))
print(paste0("E(88) = ",4*25*p88))
```

Using Combinatorics:

    [1] "Expected winnings for 00 = 7.53"
    [1] "Expected winnings for 88 = 0.06"

Just quadrupling the probability (alternate solution method):

    [1] "E(00) = 7.53263823251418"
    [1] "E(88) = 0.059812145557656"


## Problem 2

    Two spherical soap bubbles, each 10 cm in
    diameter, coalesce to form a "double bubble"
    consisting of two truncated spheres joined at
    their flat surfaces with a plane circular
    interface between them. What is the area of
    the interface? Assume that the volume of the
    double bubble is equal to the sum of the
    volumes of the two original bubbles.

    -- Howard G. McIlvried III, PA Gamma'53

Solution:

    It appears that this problem is underconstrained.
    You could imagine many truncated sphere pairs that
    meet the volume constraint.
    
    I peeked at the solution.  The key is to minimize
    the surface area of the "double bubble".

    With this in place it should be possible to calculate
    a unique solution.


## Problem 3

    The newly crowned king of Cashtopia has
    decided to replace the coins in use with a
    new coinage bearing his image. Being
    interested in efficiency, he wants to have
    only three different coins. Furthermore, he
    wants the denominations chosen so that the
    average number of coins required to make
    change for any amount from 1 cent through 99
    cents is a minimum. Assuming that the need to
    make change for any value between 1 and 99 is
    equally likely, what three denominations
    should he choose? If more than one set gives
    the minimum, the king prefers the one with
    the smallest sum for the three coins.

    -- Doctor Ecco's Cyberpuzzles by Dennis E.
    Shasha, CT A'77

Solution:

    Exhaustive search should be fast enough.
    
```{r}
make_change = function(n,a,b,c) {
    na=trunc(n/a)
    n=n %% a
    nb=trunc(n/b)
    n=n %% b
    nc=trunc(n/c)
    n=n %% c
    if (n != 0) {
        return(c(-1,-1,-1))
    }
    return(c(na,nb,nc))
}

nmax=9999
for (c in 1:97) {
    for (b in (c+1):98) {
        for (a in (b+1):99) {
            ncoins=0
            for (target in 1:99) {
                x=make_change(target,a,b,c)
                if (x[1]<0) {
                    ncoins=99999
                    break
                } else {
                    ncoins=ncoins+sum(x)
                }
            }
            nmean=ncoins/99
            if (nmean<=nmax) {
                nmax=nmean
                print(paste(nmax,a,b,c))
            }
        }
    }
}
```

The best performance is obtained with coins of 22, 5, and 1.
The mean number of coins required for change is 5.3131...

The equivalent performance occurs with 23,5,1 but this sum is higher.


## Problem 4

    How many distinct trapezoidal decompositions
    (sum of consecutive positive integers) does n
    have? For example, 15 = 7+8 = 4+5+6 =
    1+2+3+4+5 has three.

    -- Technology Review

Solution: Hmmmmm .... need to work on this.

    Let us try out some numbers and see if we get
    a pattern:
    
    n  Decompositions
    ----------------------
    1  1
    2  none
    3  1+2
    4  none
    5  2+3
    6  none
    7  3+4
    8  none
    9  4+5,2+3+4
    10 1+2+3+4
    11 5+6
    12 3+4+5,
    
    I would never have gotten it.  This involves the number
    of factors and realizing that even factors don't matter.
    
    Given N = 2^k * M then the number of trapezoidal decompositions
    (TD) is 1 less than the number of factors of M.
    
    Recall that if we get the prime factors of M and they have
    multiplicities m1, m2, ... the number of factors is:
    
        (1+m1)*(1+m2)*...
        
    So: TD = (1+m1)*(1+m2)*... - 1
    
    In the case of 15 = 3*5 we have each prime factor with mi=1 so
    
    TD(15) = (1+1)*(1+1)-1 = 3 decompositions (as given in the problem).
    
    In the case of N = 90 = 2*(3^2*5)
    
    TD(45) = (1+2)*(1+1) - 1 = 5 decompositions as given in Bent solution.


## Problem 5

    The game of Spot-It consists of a deck of
    cards, with eight pictures on each card.
    Between any two cards, there is one, and only
    one, matching picture. If each picture occurs
    the same number of times in the full deck,
    what is the maximum number of cards in the
    deck?

    -- Steve Schaefer via mathrec.org

TODO

    meh
    

## Bonus

    N&Ns come in two colors: red and green; and
    there is always the same total number in each
    package (less than 50) with the same ratio of
    red to green. Al and Beth have each eaten two
    packages of N&Ns over the last couple of
    days. Al just reaches into the package and
    randomly chooses an N&N. With his first
    package, when he had eaten the last red N&N,
    he saw that he had one green left. With his
    second package, when he had eaten the last
    red, he had two greens left. He later
    calculated that the situation with the first
    package is four times as likely as the
    situation with the second package. Beth uses
    a different approach; until she runs out of
    one color, she flips a coin and eats a red if
    heads and a green if tails. With her first
    package, when she had eaten the last red N&N,
    she noted the number of greens left. With her
    second package, when she had eaten the last
    red, she had one more green left than with
    her first package. Beth also made some
    calculations and determined that the number
    of greens left with the first package is four
    times as likely as the number left with the
    second package. How many red and how many
    green N&Ns are in a package?

    -- Susan Denham in New Scientist

TODO


## Double Bonus

    What is the remainder when 2^(4,700,063,497)
    is divided by 4,700,063,497?

    -- Adapted from Prime Numbers, the Most
    Mysterious Figures in Math by David Wells

Solution:

    Brute force remainder is ???
    Does not work with gmp -- may need Python
    The exponent exceeds the gmp implementation limit
    for the pow() function.  They probably used 32
    bit integers internally!
    
    Switched to python brute force = 3

    OK - probably could have done this with gmp using
    the shortcut in the Bent.  I think they are saying:
    
    (a*b) mod m = ((a mod m) * (b mod m)) mod m
    
    They decompose the exponent into binary and then
    calculate each term mod m, the product mod m and
    have their result.

    I should probably try that.
    
    This reminds me of the modulo calculations done
    in my investigation of prime numbers.
