---
title: "Bent Spring 2018"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Spring 2018


## Problem 1

    I have found five nine-digit integers such
    that, among them, the digit 1 appears once,
    the digit 2 appears twice, 3 appears three
    times, and so forth to 9 which appears nine
    times. The first and second integers, both of
    which are exact fifth powers, between them,
    contain all nine different digits.The third
    integer is an exact fourth power. The fourth
    integer, which consists of only two different
    digits, is a palindrome divisible by 33 and
    111. The digits of the fifth integer form an
    increasing sequence in which each digit is
    equal to, or greater than, the preceding
    digit. What are the five nine-digit integers?

    -- Adapted from an Enigma by Adrian
    Somerfield in New Scientist

Solution:

    This was WAY harder than it should have been due
    to the ambiguous (?) wording of the question.  I
    assummed that the palindrome was divisible by
    the product 33*111 but it was not!  Once that
    was understood it was just cranking.

    None of the numbers can contain a 0 since all 45
    digits are accounts for with 1-9.
    
    No number can contain more than 1-1, 2-2s, 3-3s, etc.

    Need to find 9 digit 5th powers
    Need to find 9 digit 4th powers
    Need to find palindrome, 2 diff digits, divisible by 33, 111
    
```{r, eval=FALSE}

nmin=100000000
nmax=999999999

xmin=trunc(exp(log(nmin)/5))
xmax=trunc(exp(log(nmax)/5))

# Fifth powers
p5=c()
for (i in xmin:xmax) {
    m=i^5
    if (m >= 100000000) { p5=append(p5,m) }
}
p5=p5[!grepl('0',p5)]
p5=p5[!grepl('1.*1',p5)]
p5=p5[!grepl('2.*2.*2',p5)]
p5=p5[!grepl('3.*3.*3.*3',p5)]


# Fourth powers
xmin=trunc(exp(log(nmin)/4))
xmax=trunc(exp(log(nmax)/4))

p4=c()
for (i in xmin:xmax) {
    m=i^4
    if (m >= 100000000) { p4=append(p4,m) }
}
p4=p4[!grepl('0',p4)]
p4=p4[!grepl('1.*1',p4)]
p4=p4[!grepl('2.*2.*2',p4)]
p4=p4[!grepl('3.*3.*3.*3',p4)]

# Palindromes
# Need palindrome of two different digits that
# is a multiple of 33 and 111
#
# BUT - APPARENTLY - NOT SIMULTANEOUSLY!!!
#

# Check the multiples of 33
xmin=trunc(nmin/(33))
xmax=trunc(nmax/(33))
ppal=c()
for (i in xmin:xmax) {
    m=i*33
    if ((m>99999999) & (m <= 999999999)) {
        if (m%%111 == 0) {
            a=strsplit(sprintf("%d",m),"")[[1]]
            if (all(a==rev(a))) {
                if (length(unique(a))==2) {
                    ppal=append(ppal,m)
                }
            }
        }
    }
}
# Check the multiples of 111
xmin=trunc(nmin/(111))
xmax=trunc(nmax/(111))
for (i in xmin:xmax) {
    m=i*111
    if ((m>99999999) & (m <= 999999999)) {
        if (m%%33 == 0) {
            a=strsplit(sprintf("%d",m),"")[[1]]
            if (all(a==rev(a))) {
                if (length(unique(a))==2) {
                    ppal=append(ppal,m)
                }
            }
        }
    }

}
ppal=unique(sort(ppal))
ppal=ppal[!grepl('0',ppal)]
ppal=ppal[!grepl('1',ppal)]
ppal=ppal[!grepl('2',ppal)]

```

    A partial solution is found at this point:
    
    5th powers: 656356768 714924299 992436543
    4th powers: 21 values
    palindrome: 13 values
    
    We can limit the 5th powers to only the first two since only
    the second one contains a 1 and only the first contains an 8.
    
    So: first two numbers are
    
        656356768
        714924299

    This uses up all the 1s and 2s.  After removing these from the
    list of fourth powers we are only left with:
    
        688747536
    
    At this point the digit counts are:
    
        1=1, 2=2, 3=2, 4=3, 5=3
        6=6, 7=4, 8=3, 9=3
        
    All of the 6s have been used so they can be removed from the palindrome
    list which leaves only 8 values:
    
    444888444 557575755 755575557 777333777
    888555888 889898988 988898889 999777999
    
    There is only 1 available 4 so the first value is eliminated.
    There are only 2 available 5s so this eliminates 3 more.
    There are only 3 available 7s so this eliminates one more.
    There are only 5 available 8s so this eliminates two more.
    
    The only remaining candidate is:
    
        999777999
    
    At this point the solution set is:
    
        656356768 714924299 688747536 999777999
        
    The digit counts are:
    
        Digit     1 2 3 4 5 6 7 8 9 
        Count     1 2 2 3 3 6 7 3 9 
        Available     1 1 2     5
        
    These remaining digits can only be arranged as: 345588888
    
    The full solution set is:
    
        656356768 = 58^5
        714924299 = 59^5
        688747536 = 162^4
        999777999 = palindrome = 30296303*33 = 9007009*111
        345588888 = remaining digits in increasing order


## Problem 2

    At a recent bridge game, Don thought his luck
    couldn't get any worse, until he was dealt a
    hand that not only didn't have any aces or
    face cards, but included exactly two 2's,
    three 3's, and four 4's. What is the exact
    probability of being dealt such a hand? A
    bridge hand consists of 13 cards dealt from a
    well shuffled standard deck of 52 cards.

    -- Don A. Dechman, TX A'58

Solution:

    Standard combinatorics:
    
    2-2s = C(4,2)
    3-3s = C(4,3)
    4-4s = C(4,4)
    Rest = 4 cards from 5-10 = 24 total
         = C(24,4)
         
    Total number of bridge hands = C(52,13)
    
    P = C(4,2)*C(4,3)*C(4,4)*C(24,4)/C(52,13)
      = 255024/635013559600
      = 99/246511475
      ~ 4.01604e-07


## Problem 3

    At Sam's Soup, Salad & Sandwich Deli, patrons
    can order from a menu consisting of ten
    different items.They can order as many
    different items as they wish, but they can
    order at most two of any given item. To
    minimize data entry, Sam has devised an
    approach that requires sending only one
    number per order to the kitchen. He did this
    by assigning a code number consisting of a
    positive integer to each item on the menu.
    The waiter then enters only the sum of the
    codes for the items ordered.  (If two of the
    same item are ordered, the item's code is
    added twice.) If the codes have been chosen
    so that every possible order results in a
    unique number, what is a simple set of codes
    for the ten items on the menu?

    -- Adapted from Doctor Ecco's
       Cyberpuzzles by Dennis E. Shasha

Solution:

    I would just use powers of 2.

    The solution is to use powers of 3.  Apparently
    you can order 0, 1, or 2 of each item so you need
    three states to encode things.
    
    meh


## Problem 4

    In the famous Tower of Hanoi puzzle, you
    start with three pegs (P1, P2, and P3). On P1
    are n discs of different sizes, increasing in
    diameter from the smallest at the top to the
    largest at the bottom. The objective is, in
    as few moves as possible, to transfer the
    stack from P1 to P3 by moving the discs, one
    at a time, from peg to peg to peg while
    observing the rule that no disc can ever be
    placed on top of a smaller disc. For three
    pegs and n discs, the process takes 2^n-1
    moves. For example with three pegs and n = 4,
    15 moves are required; and three pegs with n
    = 18 requires 262,143 moves and (at a rate of
    one move per second) requires 72 hr, 49 min,
    and 3 sec. For Super-Hanoi, you can have more
    than three pegs. For p pegs, the objective is
    to transfer an n disc stack from P1 to Pp in
    the fewest possible moves. For example, with
    4 pegs, 4 discs can be transferred in 9 moves
    instead of 15. What is the minimum number of
    pegs needed to be able to transfer 18 discs
    from P1 to Pp in a minute or less at the rate
    of one move per second? Also, what is the
    arrangement of discs on the pegs immediately
    before the largest disc is moved from P1 to
    Pp? Transferring a single disc from one peg
    to another constitutes a move.

    -- An Enigma by Stephen Ainley
       in New Scientist

TODO

    meh
    
    They argue that you need 6 pegs
    
    If you have n empty pegs you can "store" n disks.
    They can be assembled onto 1 peg
    This leaves n-1 empty pegs so you can "store" n-1
    etc.
    
    With 5 total pegs you can only have:
    
    5, 4, 3, 2, 1 = 15 disks = not enough disks
    
    With 6 total pegs you can have:
    
    6, 5, 4, 3, 2, 1 = 21 disks which is enough
    
    You actually need a seventh peg as the destination:
    
    So, for the 18 disk game you would move off 6 disks and
    then stack 5 for a total of 11 moves.
    Repeat for 5 disks + 4 stack for 9 moves.
    Repeat for 4 disks + 3 stack for 7 moves.
    This leaves 3 disks on the original peg.
    Need to move off 2 and then transfer final disk to target. = 3 moves
    Assemble second largest and third largest = 2 moves
    Now we have 3 empty pegs
    Unpack peg with 4 disks = 7 moves
    Unpack peg with 5 disks = 9 moves
    Unpack peg with 6 disks = 11 moves
    
    Total = 59 moves < 1 minute


## Problem 5

    In a right triangle, inscribe a semicircular
    arc, such that the arc is tangent to both
    legs of the right triangle, and the center of
    the semicircle and its two endpoints lie on
    the hypotenuse of the triangle. These three
    points divide the hypotenuse into four
    segments. Starting at one end of the
    hypotenuse and moving toward the other end,
    the lengths of these segments are x cm, 120
    cm, 120 cm, and y cm. Given that x and y are
    integers, what are the maximum and minimum
    lengths the hypotenuse can have?

    -- Adapted from
       The Wall Street Journal by
        Robert W. Schweitzer, NY Z'52

Solution:

    I need to draw a picture.
    
```{r, eval=FALSE}

r=120    # radius of the circle
m=-2/3   # slope of hypotenuse - y=m*x+b
         # r = m*r + b --> b=r*(1-m)
         # x-intercept: 0 = m*x+b --> x = -b/m
b=r*(1-m)
xmax=-b/m
xh=seq(0,xmax,length=100)
yh=m*xh+b

theta=seq(0,2*pi,length=100)

xc=r+r*cos(theta); yc=r+r*sin(theta)

plot(xc,yc,type='n',asp=1)
polygon(xc,yc,border='red')
points(r,r,ps=3,col='blue')
lines(xh,yh,col='green')
lines(rep(0,length(yh)),yh,col='green')
lines(xh,rep(0,length(yh)),col='green')
lines(c(0,r),c(r,r),col='green')
lines(c(r,r),c(0,r),col='green')
text(10,max(yh)+20,"X")
text(max(xh)-20,20+20,"Y")
```

    Note that these definitions of h,w do not quite match the ones used in
    the solution in the Bent.  They assign z, w as the lengths of the legs
    outside the 120^2 area.  In the equations below we use the full leg length.
    
    Let:
    
        h = height of triangle
        w = width of triangle

    From above we can write:
    
        (h-120)^2 + 120^2 = (120+x)^2
        
            x = sqrt((h-120)^2 + 120^2) - 120
        
        (w-120)^2 + 120^2 = (120+y)^2
        
            y = sqrt((w-120)^2 + 120^2) - 120
        
        (h-120)/120 = 120/(w-120) = h/w
        
            h = 120 + 120^2 /(w-120)
            
        Need to find:
        
            (h-120) s.t. (h-120)^2 + 120^2 = n^2 if x is integer
            (w-120) s.t. (w-120)^2 + 120^2 = m^2 if y is integer
            
        The Bent asserts, and I blindly followed along, that h, w
        are integers.  How do I know that?????  If a^2+b^2 = integer
        does that necessarily imply that a,b are integers?
        
            Know: w-120 is a factor of 120^2 = 2^6 * 3^2 * 5^2
            
        How many factors for 120^2?
        
            Number of factors = (1+6)*(1+2)*(1+2) = 63
            
```{r, eval=FALSE}

# Enumerate all the possible factors of 120^2
z=c()
for (n2 in 0:6) {
    for (n3 in 0:2) {
        for (n5 in 0:2) {
            z=append(z,prod(c(1,rep(2,n2),rep(3,n3),rep(5,n5))))
        }
    }
}

# w = z+120 but only w-120 is used above so just use z
# y = sqrt(z^2+120^2)-120
y=sqrt(z^2+120^2)-120
k=(y==trunc(y))
w=z+120
h=120+120^2/(w-120)
x = sqrt((h-120)^2 + 120^2) - 120
#
print("W = ")
print(w[k])
print("H = ")
print(h[k])
print("X = ")
print(x[k])
print("Y = ")
print(y[k])

hypotenuse=240+x[k]+y[k]
print("Hypotenuse = ")
print(hypotenuse)
print("Range = ")
print(range(hypotenuse))
```

    The only integer solutions are:       
    
    W = 345 170 210 280 408 184
    Y = 135  10  30  80 192  16    

    W = 345 170 210 280 408 184
    H = 184 408 280 210 170 345
    X =  16 192  80  30  10 135
    Y = 135  10  30  80 192  16
    
    Hypotenuse:
    391 442 350 350 442 391
    
    Range: 350 442


## Bonus

    Suppose that, in a distant galaxy, there is a
    solar system in which, instead of being
    spheres, the planets are right circular cones
    (with heights equal to the diameters of their
    bases). Suppose one of these planets has the
    same total volume and mass as our Earth, but
    a uniform density; what would be the
    gravitational acceleration on a person
    standing in the center of the circular base,
    and what would be the gravitational
    acceleration on a person standing at the
    apex? Assume the Earth is a perfect sphere
    with a radius of 6,370 km and an average
    density of 5,518 kg/m3. Use a value of
    6.673E-11 N m2/ kg2 for G. Express your
    answers correct to three significant figures.

    -- Higher Mathematics for
       Engineers and Physicists
       by I.S. ans E.S. Skolonikoff

Solution:

    Volume of earth = 4/3*pi*R^3
       = 4/3*pi*6370000^3
       = 1.082697e+21 m^3
       
    Mass of earth = 5.974322e+24 kg
    
    For the conical earth: H = 2*R
    
    V = 1/3*pi*R^2*2*R = 1.082697e+21
    
    R =  8025697 m
    H = 16051394 m
    
    I did the integrations but I get a much different answer from the Bent.  I used
    a differential ring of radius=r at a height y above the axis with the peak of
    the cone at the origin and the axis of the cone vertial (y-axis in my calc).
    Since the differential element is a ring it only requires two integrals to
    perform the volumetric calculation:

    gbase: integrate(integrate(G*rho*2*%pi*r*(H-y)/sqrt(r^2+(H-y)^2)^3,r,0,(y/H)*R),y,0,H)
    gapex: integrate(integrate(G*rho*2*%pi*r*y/sqrt(r^2+y^2)^3,r,0,(y/H)*R),y,0,H)

    In calculating the attaction at the base (y=H):
    
    The distance from the ring to the point on the base is sqrt(r^2+(H-y)^2)
    The term (H-y)/d calculates the component of the gravity force along y axis
    The attraction varies as 1/d^2
    The overall denominator is then d^3.
    
    So: G*rho*2*pi*r * (H-y)/d^3 should be integrated for r and y
    
        r varies from 0 to (y/H)*R
        y varies from 0 to H

    At the center of the base I get: g = 9.494673880675947
    At the apex   of the cone I get: g = 3.920554736208179

    Even if I flip the cone upside down and try it again I get the same result.

    *****************************************************************
    ***** THIS IS NOT EVEN CLOSE TO THE SAME ANSWER AS THE BENT *****
    *****                                                       *****
    ***** I have done additional investigation and I think the  *****
    ***** Bent is wrong.                                        *****
    *****************************************************************

    The Bent has: g = 24.75 at the middle of the base, 2.75 at the apex
    
    They do a calculation involving the center of mass???
    
    They calculate the center of mass as 3/4*H with apex at origin.  I can
    get this value -- not a problem.
    
    ycm = integrate(integrate(rho*2*%pi*r*y,r,0,(y/H)*R),y,0,H)/(1/3*rho*%pi*R^2*H)
        = 3*H/4

    They use this in calculating the g values:
    
        For conical earth they calculate:
        
            R = 8026200 m  --> about the same as mine
            m = 5.9743e24 kg
    
        gbase = G*M/(1/4*H)^2 = 6.673e-11*5.974e24/(16051394/4)^2 = 24.756
        gapex = G*M/(3/4*H)^2 = 2.75
        
    Why is there a difference? and which one is correct???
    
    With a very minor adjustment to my integral I can get the correct g for a
    sphere so it should work.

    Conversely, I think their method should also work so I am confused.

    I found a reference that says the center of mass method is not correct.

    Specifically:

        Newton proved that a solid sphere of uniform density and
        mass M exerts the same force on an external point mass as
        would a point mass M placed at the center of the sphere.
        (See Problem 6a in problem section 5C).

        This does not, however, generalize to other uniform solids
        of mass M - it is not true that the gravitational force they
        exert is the same as that of a point mass M at their center
        of mass. 

    MIT courseware link:
    
        https://ocw.mit.edu/courses/
        mathematics/18-02sc-multivariable-calculus-fall-2010/
        4.-triple-integrals-and-surface-integrals-in-3-space/
        part-a-triple-integrals/
        session-78-applications-gravitational-attraction/
        MIT18_02SC_MNotes_g.pdf

    The source indicates you need to perform the integrations.


## Double Bonus

    Solve the following
    cryptic addition in base 13.

    HOWIE
    FRED
    CHUCK
    JEFF
    JUDGE

    Although the base is an unlucky
    number, the following circumstances
    negate this. The panel of JUDGEs
    is, of course, prime, as are FRED,
    CHUCK, and JEFF, while HOWIE
    is the product of 7 (lucky number),
    not necessarily different, primes. The
    usual rules apply. Use a, b, and c to
    represent the digits 10, 11, and 12.

    -- H.G. McIlvried, III, PA Gamma'53

Solution: solved in the cryptic arithmetic notebook although the
details are in a separate file.

    Cryptic addition in base 13:

      HOWIE + FRED + CHUCK + JEFF = JUDGE

      JUDGE is prime
      FRED, CHUCK, and JEFF are prime
      HOWIE is the product of 7 prime factors

    I switched to python itertools (see jeff.py)
    
    Searching on FRED, JEFF, JUDGE (7 symbols) still results in 7679 solutions
    Let's add chuck (three more symbols)
    
    This makes the problem worse since there are 13*12*11 times as many
    permutations to consider.
    
    Make need to just bite the bullet and try all the combinations even though
    it is a lot of them.
    
    sympy has isprime() and factorint() to help make the computations but it will
    still take a long time to investigate 13! permutations.  This is 6.3e9 cases.
    The run-time is about 4 hours!

    We could add some limits to the search:
    
    J, F, C, and H must be non-zero
    Can't say anything about other digits since base-13 primes can have any final digit!

    Checking on the primality constraints yields 9951 feasible solutions.
    These have been stored in `jeff_solution.txt` using numpy savetxt.
    
    Adding the equation of JEFF+FRED+CHUCK+HOWIE = JUDGE reduces
    the solution to a unique case.
    
    Maybe I should have checked the equation first and then checked primality -- not sure
    which is the most efficient.  I added the equation condition to the python code and
    the solution is actually slower so ... a more efficient algorithm may be required.
    
    I tried just generating and counting the permuations (without any calculations) and
    that would take ~30 minutes so maybe 3-4 hours isn't too bad.

    In [217]: a[k]
    Out[217]:
           j   e  f  r  d  u  g  c  h  k  o  w   i
    9798  12  11  6  1  5  4  9  7  3  2  8  0  10

    x_jeff[k]  =  28307    ( cb66 - base 13)
    x_fred[k]  =  13499    ( 61b5 - base 13)
    x_chuck[k] = 207287    (73472 - base 13)
    x_howie[k] = 103400    (380ab - base 13)
    x_judge[k] = 352493    (c458b - base 13)


