---
title: "Bent Summer 2007"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Summer 2007


## Problem 1

    The makers of Grunt, who claim
    that their aftershave is so appealing
    to women that users will have to fight
    them off, are puzzled that some men
    are using Phew, so they hired Judy
    to investigate. She boarded a bus to
    question the male passengers. The
    first man said, "I use Grunt." Then,
    the rest of the men in turn said, "I use
    Grunt, but the man you just interviewed
    uses Phew." Finally, the first
    man added, "The last man you interviewed
    uses Phew." Confused, Judy
    asked a few of the men how many of
    the male passengers used Phew. The
    tallest man said "19," and the man she
    had originally interviewed fourth after
    him said "24." The fattest man said
    "13," and the man she had originally
    interviewed fifth after him said "28."
    The nicest man said "24," and the man
    she had originally interviewed sixth
    after him said "13." If each man uses
    either Grunt or Phew, but only Grunt
    users tell the truth, how many men
    use Phew?

    -- Martin Hollis

TODO

    Aaaaargh!
    

## Problem 2

    On his trips to Monte Carlo, Sam
    always gambles exactly $50 each night
    on roulette. Usually, he just plunks
    all his chips down on black (result
    is double or nothing) and quits after
    one spin, regardless of the outcome.
    The probability of winning is 18/37.
    One night he decides to change his
    strategy by making $1 bets on black
    until he either loses the $50 or gains
    another $50. What is his probability
    of winning with his new strategy?

    -- Technology Review

Solution: Let's try a Monte-Carlo analysis

```{r}
set.seed(42)
ntrials=10000
win=0
for (trial in 1:ntrials) {
    # Play until n=100 or n=0
    n=50
    while (T) {
        if (runif(1)<=18/37) {
            n=n+1
        } else {
            n=n-1
        }
        if (n==100) {
            win=win+1
            break
        }
        if (n==0) {
            break
        }
    }
}
print(paste0('Percent wins = ',win/ntrials))
```

Winning occurs a little more than 6% of the time:

    0.0618  @ 10K trials
    0.0655  @ 10K trials (seed=123)
    0.0644  @ 10K trials (seed=42)
    0.0579  @ 10K trials (seed=2020)
    0.06401 @ 100K trials

This should be possible to calculate analytically.  I suspect the result depends
only on the relative likelihood of winning/losing 50 bets.  So, the probability
of winning/losing is;

    (18/37)^50 vs. (19/37)^50
    
I am not sure about all the other probabilities (e.g W/L cancelling) but I
suspect they all cancel out in the end.


              (18/37)^50
    P = ----------------------- ~ 0.062775
        (18/37)^50 + (19/37)^50

I think we can prove this formula by examining a simpler problem and then
calculating the results.  Assume the original bankroll is 2 and the goal
is either to win 2 or lose 2.  For two rounds of betting we get:

    B1 B2  Bankroll
    ---------------
    L  L   0 --> loser: P = (19/37)^2
    L  W   2 \ each has P = (18/37)*(19/37)
    W  L   2 /
    W  W   4 --> winner: P = (18/37)^2
    
In this case: you win (18/37)^2, you lose (19/37)^2, and you are back to
the original problem the rest of the time.  The cases where you end up back
at the beginning could still end up with winning so the total probability of
winning should be:

    winner = (18/37)^2 + 2*(18/37)*(19/37)(18/37)^2 +
                         4*(18/37)^2*(19/37)^2*(18/37)^2 +
                         ...
                         
           = (18/37)^2 * (1 + (2*18/37*19/37) + (2*18/37*19/37)^2 + ... )
           = (18/37)^2 * 1/(1-2*18/37*19/37)
           = 0.473

This is much higher than I thought but it agrees with the formula that I
suspected above.  For a bankroll of 2 we have:

             (18/37)^2
    P = --------------------- = 0.4729927
        (18/37)^2 + (19/37)^2

Extending the argument to a bankroll of n we have:

             (18/37)^n
    P = ---------------------
        (18/37)^n + (19/37)^n


Note that the longer you play the more likely you are to lose
it all.  No wonder casinos do so well.

```{r}
n=seq(2,100,2)
p=(18/37)^n / ( (18/37)^n + (19/37)^n )
plot(n,p,type='l',xlab='Initial Bankroll',ylab='Probability of Winning')
points(n,p,col='red')
```

## Problem 3

    Different letters are different digits;
    same letter is same digit throughout.

    See picture

	-- Eric R. Emmet

Solution: not solved in the cryptic arithmetic notebook

    Needs a solution.
    
              - - a -
        -----------------
    - - | d - - - - x
          k - -
          ---------------
              - -
              t -
              -----------
              m - -
              m - -
              -----------
              x - x
              h - -
              -----------
                k -


## Problem 4

    On the island of Approxima, all
    measurements are only to the closest
    inch. Alex, who makes yardsticks, has
    realized that he doesn't have to put
    marks at each inch to still be able to
    measure all integral number of inches
    from 1 through 36, inclusive. Distances
    are measured between two marks,
    or between a mark and an end of the
    ruler, or between the two ends of the
    ruler. For example, a foot ruler only
    needs marks at 1, 2, 6, and 9 inches in
    order to measure all distances from 1
    inch through 12 inches, inclusive. Using
    eight marks, where does Alex put
    them on his yardsticks?

    -- The Numerology of Dr. Matrix
    by Martin Gardner

Solution:

    We need to measure 1-36 inches (integer)
    The distances can be:
    
    (1) from the end to a marker
    (2) from the opposite end to a marker
    (3) between two markers
    (4) between two ends.
    
    I tried some guessing - but nothing worked.

OK - time to give up and try exhaustive search.  Let us assume that markers are
required at locations 1 and 2.  The remaining markers can be at any location
from 3 to 33.  That didn't work.  Modify the search to look for 8 markers over
the full range of the yardstick.  This is **SLOW** since it looks at choose(35,8)
combinations (~23.5e6) and needs to calculate 10x10 outer with logic to find the
unique values for each combination.

The two solutions that are found are actually just reflections of one solution.

```{r}
maxlen=0

t1=Sys.time()
for (m1 in 1:35) {
    for (m2 in (m1+1):35) {
        for (m3 in (m2+1):35) {
            for (m4 in (m3+1):35) {
                for (m5 in (m4+1):35) {
                    for (m6 in (m5+1):35) {
                        for (m7 in (m6+1):35) {
                            for (m8 in (m7+1):35) {
                                x=c(0,36,m1,m2,m3,m4,m5,m6,m7,m8)
                                # print(x)
                                m=length(unique(as.vector(abs(outer(x,x,"-")))))
                                if (m>maxlen) { maxlen=m; bestx=x }
                                if (m==37) {
                                    print(sort(x))
                                }
                            }
                        }
                    }
                }
            }
        }
    }
}
t2=Sys.time()
print(t2-t1)
```

There are two solution sets that work:

    end    ....... markers ......    end
    0  --- 1  3  6 13 20 27 31 35 --- 36
    0  --- 1  5  9 16 23 30 33 35 --- 36

This may be easier to understand by looking at the difference
between marks.  A length may be measured based on a difference
or the sum of adjacent differences.  For the two sets of
markers we have:

    1 2 3 7 7 7 4 4 1
    1 4 4 7 7 7 3 2 1

By summing adjacent sections near the left/right ends you can
achieve all the smaller numbers.  The center groups are only
needed to get a coarse level and then use pieces of the edges
to achieve fine tuning to the desired value.

The two sets of diffs are just reflections of each other so these
really represent the same solution.


## Problem 5

    You have eight objects of eight different
    weights, which have been partially
    ordered. That is, the order of the first
    four is known among themselves, and
    the order of the second four is known
    among themselves. What is the minimum
    number of weighings on a two-pan
    balance to ensure that the eight weights
    can be ordered in all cases?

    -- Super-Puzzles, Jean-Claude Baillif

TODO

    I am going to guess that the answer is 3.


## Bonus

    A student pilot, initially located
    at coordinates (x0,y0), rashly
    decides to fly under the Golden Gate
    Bridge, the ends of which are located at
    (-1,0) and (1,0). These end points subtend
    a visual angle Q (vertex at the
    student's airplane), and all movement
    and measurements lie in the same
    X-Y plane. He flies along a path that
    maximizes the instantaneous rate of
    increase of Q. What is the equation of
    his flight path? Assume the X-Y plane
    is parallel to a flat ocean.

    -- Byron R. Adams, TX A '58

Solution?

    I don't really understand this problem.
    
    Does the maximization of the instantaneous rate of
    increase of Q define a full flight path?
    
    The angle from the left end of the bridge (-1,0) to
    the student plane is:
    
        alpha1 = atan(y0/(x0+1))
        
    The angle from the right end of the bridge (1,0) to
    the student plane is:
    
        alpha2 = atan(y0/(x0-1))
        
    The angle at the vertex is just the difference between
    these two:
    
        Q = alpha1-alpha2
          = atan(y0/(x0+1)) - atan(y0/(x0-1))

    How do we maximize the rate of change?
    
        maximize( (dq/dx)^2 + (dq/dy)^2 )?
        
    This is maximized (e.g. d^2/dx^2 and d^2/dy^2 = 0) when:
    
        x = sqrt(1-y^2), or
        x = 0, or
        x = -sqrt(1-y^2)
    
    Or,
    
        x=0 --> fly down the y axis.
        x^2 = 1-y^2 --> x^2+y^2 = 1  --> circle with radius = 1



## Double Bonus

    If A, B, and C are the
    measures of the angles of a plane triangle,
    what are the values of L and H in

    L <= sin(3A) + sin(3B) + sin(3C) <= H
    
    when equality holds? Prove your answers.

    -- Tenth U.S.A. Mathematical
    Olympiad (1981)

Solution:

    I don't agree with the published solution.  They take the equation
    below and then diff wrt A and set it equal to zero.  The result is
    that:
    
        cos(3*A) + cos(3*A+3*B) = 0
        
    and then they conclude that A = (pi-3*B)/6 which leads to a min of 0.
    
    I think they are missing the case of A = (3*pi-3*B)/6 which lead to a
    much deeper min.  See the contour and 3D plots below for the region
    where min < 0.
    
    This is fairly obvious if you consider a very tall, narrow triangle.
    Each base angle is ~90 degrees and the apex is nearly 0.  The value
    of sin(3*A) and sin(3*B) is each nearly -1 and the apex is 0 so the
    sum must approach:
    
        limit --> -2

    You can see this in the contour plot near a=1.57 where for low values
    of b the min is very negative (angle c ~ 90 degrees) and for b near
    1.57 (about 90 degrees) the apex angle will be zero.  The same min
    occurs for low values of a with b~c~1.57.  These are all really
    the same condition just a swap of a, b and c.

    Algebraically I don't have much.  About all I can figure out is that
    
    C = pi-(A+B) so:
    
    L <= sin(3A) + sin(3B) + sin(3A+3B) <= H
    
    I can manipulate this to:
    
    L <= sin(3A)*(1+cos(3B))+sin(3B)*(1+cos(3A)) <= H
    
    But that is about it.
    
    I can explore the solution space numerically:
    
```{r}
foo = function(a,b) {
    return(sin(3*a)+sin(3*b)+sin(3*a+3*b))
}

a=seq(0,pi,length=100)
b=seq(0,pi,length=100)

zzz=outer(a,b,foo)
xxx=outer(a,b,"+")
zzz=ifelse(xxx>pi,NA,zzz)


contour(a,b,zzz,asp=1)
persp(a,b,zzz,phi=30)
```