---
title: "Bent Summer 2010"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Summer 2010

These problems all seem familiar.  I think I worked on
some of them previously.

## Problem 1

    Find all primes of the form A^4+4B^4,
    where A and B are positive integers.

    -- Puzzle Corner by Allan Gottlieb
    in Technology Review

TODO

    I couldn't solve it last time either.

    After looking at the Bent here it is:
    
    A^4 + 4*B^4 = (A^2 + 2*A*B + 2*B^2)*(A^2 - 2*A*B + 2*B^2)
    
    But: if this is a prime then one of the factors above must
    be a 1, otherwise it would be composite.  Only the second
    term can ever equal 1 and this only occurs when A=1, B=1
    which makes the first term = 5, so:
    
        A^4 + 4*B^4 = 1+4 = 5

    Sort of clever if you noted the factorization trick.


## Problem 2

    Algernon, Bertie, and Clarence had so
    often expressed their opinion about Professor
    Popoff that when he was found
    murdered (stabbed with a dagger, but
    in a thoroughly gentlemanly way) it was
    natural that they should be suspected.
    In fact, for reasons into which we need
    not now go, it may be taken as certain
    that one of them is guilty. They made
    statements as follows:
    Algernon:
    1: I hadn't seen Popoff or had any
    contact with him for a week before his
    unfortunate demise.
    2: Everything Bertie says is true.
    3: Everything Clarence says is true.
    Bertie:
    1: I have never handled a dagger.
    2: Everything Algernon says is false.
    3: Everything Clarence says is false.
    Clarence:
    1: Algernon was talking to Popoff just
    before he was killed.
    2: Bertie has handled a dagger.
    3: I have for a long time thought more
    of Popoff than is generally realized.
    Looking back on the tragic event
    now, it is interesting to see that
    Algernon and Bertie both made the
    same number of true statements. (This
    number can be anything from 0 to 3,
    inclusive).
    Who killed Popoff?

    -- Brain Puzzler's Delight
    by E.R. Emmet

TODO

    meh - Logic
    

## Problem 3

    At its main base, the Patagonian navy
    had 16 Class I gunboats with 5 guns
    apiece, some Class II gunboats with 4
    guns apiece, and some Class III gunboats
    with 3 guns apiece, when the admiral
    was ordered to crush a revolt that
    had broken out simultaneously on seven
    scattered islands. After some thought,
    he split his fleet into seven flotillas of 6
    ships each, with each flotilla having at
    least one gunboat of each class but a
    different number of guns. How many
    Class II gunboats were there?

    -- Tantalizer by Martin Hollis
    in New Scientist

Solution:

    Might be fun to attach this with random sampling.
    
    Total of 42 ships with 16 Class I gunboats.  Must have
    at least one ship of each class in each flotilla.
    
    Class I, 16 boats, 5 guns each
    Class II, ?? boats, 4 guns each
    Class III, ?? boats, 3 guns each
    
    Could generate random flotilla distributions and see
    if we can find a solution.  Alternately, we can limit
    the possibilities by noting:
    
    1. Each flotilla has 6 ships with at least one of
       each class I, II, and III.  Assume each flotilla
       starts with exactly 1 of each class then each
       flotilla has the same firepower.  We need the
       next three ships to differentiate.
       
    2. The remaining three ships can vary from 3 ships
       of class III (9 guns) to 3 ships of class I (15 shots).
       There are only 7 combinations so each flotilla has
       precisely one different combination of ships.
       
    3. The total number of class I ships is 16.  This
       means that in the final 3 ships of each flotilla
       we must include 9 ships of class I.
       
    Possibilities for the final three ships in each flotilla:
    
    Guns = 9  = 3, 3, 3
         = 10 = 3, 3, 4
         = 11 = 3, 3, 5 or 3, 4, 4
         = 12 = 3, 4, 5 or 4, 4, 4
         = 13 = 5, 5, 3 or 5, 4, 4
         = 14 = 5, 5, 4
         = 15 = 5, 5, 5

    There is only one way to get 9, 10, 14, or 15 guns so these
    are required.  This uses up 5 of the 9 remaining class I ships.
    In order to use up the other 4 ships we must pick the
    combinastions for 11, 12, and 13 guns that use class I ships.
    
    The final flotilla configurations are:
    
    3,4,5,3,3,3 = 21 guns --> 1 class II ship
    3,4,5,3,3,4 = 22 guns --> 2 class II ships
    3,4,5,3,3,5 = 23 guns --> 1 class II ship
    3,4,5,3,4,5 = 24 guns --> 2 class II ships
    3,4,5,5,5,3 = 25 guns --> 1 class II ship
    3,4,5,5,5,4 = 26 guns --> 2 class II ships
    3,4,5,5,5,5 = 27 guns --> 1 class II ship
    
    A total of 16 class I   ships
    A total of 10 class II  ships
    A total of 16 class III ships


## Problem 4

    I dreamed that I lived in a large,
    vertical, rectangular painting, where
    everything was located by its x-z
    coordinates. My mail always landed
    right at my house, at coordinates (1,0),
    from an airplane flying at a constant
    speed of 1 m/sec along the path
    z = x^2. The falling mail was subjected to
    an acceleration due to "gravity" of 1
    m/sec/sec downward, perpendicular to
    the x-axis. What was the x coordinate
    of the release point when the plane was
    flying in (1) the positive direction and
    (2) the negative direction? Ignore "air"
    resistance. Assume the initial velocity
    of the mail when it is dropped is the
    same as the airplane.

    -- Byron R. Adams, TX A'58

Solution:

    Assume +x is to the right and +z is up.
    
    Consider flying in the positive x direction:
    
    The flight angle/slope is dz/dx = 2*x
    
    Assume release at point x0.
    
    v0 = 1 m/sec (speed of plane)
    
    x0 = x0
    z0 = x0^2
    
    Flight angle = theta = atan(2*x0)
    vx0 = v0*cos(theta)
    vz0 = v0*sin(theta)

    a_x = 0
    a_z = -1 m/s/s
    
    In the x direction:  x = x0 + vx0*t
    
    When x=1: t=(1-x0)/vx0
    
    In the z direction:  z = z0 + vz0*t + 1/2*a_z*t^2
    
    Solving yields: x0 = 0.205579750059980

    When flying in the negative direction the only thing that
    changes is the sign on the initial x,z velocity.
    
    Solving yields: x0 = 1.42421701014688

    It might be interesting to look at mail trajectories based
    on the release point of the package.
    
```{r}
# Consider flying in +x direction
# When package is released it will rise to a peak, stall, and fall
v0=1
x=c()
y=c()
for (x0 in seq(0.01,1,0.01)) {
    z0=x0^2
    theta=atan(2*x0)
    vx0=v0*cos(theta)
    vz0=v0*sin(theta)
    a_x=0
    a_z=1
    # t1 = time for projectile to reach max height
    # z1 = max height
    t1=vz0/a_z
    z1=z0+vz0*t1-1/2*a_z*t1^2
    # t2 = time to fall to ground
    t2=sqrt(2*z1/a_z)
    # t = total flight time
    # xf = final x location
    xf=x0+vx0*(t1+t2)
    x=c(x,x0)
    y=c(y,xf)
}
plot(x,y,type='b',xlab='Release Point',ylab='Impact Point',main="Positive Flight")
grid()
```

```{r}
# Consider flying in -x direction in the +x area
# When package is released it will launch with a negative velocity in x,z
v0=1
x2=c()
y2=c()
for (x0 in seq(0.01,2,0.01)) {
    z0=x0^2
    theta=atan(2*x0)
    vx0=-v0*cos(theta)
    vz0=-v0*sin(theta)
    a_x=0
    a_z=-1
    # t = time for projectile to reach ground
    #
    # 0=z0+vz0*t+1/2*a_z*t^2
    # t = (-vz0 +/- sqrt(vz0^2-4*1/2*a_z*z0))/(2*1/2*a_z)
    #
    # If vz0<0 and a_z<0 then need + sign for negative t, - sign for positive
    t = (-vz0 - sqrt(vz0^2-4*0.5*a_z*z0))/a_z
    xf=x0+vx0*t
    x2=c(x2,x0)
    y2=c(y2,xf)
}
plot(x2,y2,type='b',xlab='Release Point',ylab='Impact Point',main="Negative Flight")
grid()
```

```{r}
plot(x,y,type='b',
     xlim=range(x,x2),
     ylim=range(y,y2),
     xlab='Release Point',ylab='Impact Point')
lines(x2,y2,col='red')
points(x2,y2,col='red')
grid()
```


## Problem 5

    A vertical belt sander is being used
    to polish rolling pins. The rolling pin,
    with a diameter of 5 cm and a length
    of 100 cm, is rotated at 100 rpm about
    a horizontal fixed axis while being
    pressed against the belt with a force
    of 10 Newtons perpendicular to the
    belt, which is traveling at 2 m/s. If the
    coefficient of friction is 0.5, how many
    watts are being dissipated as friction
    when (1) the rotation of the rolling pin
    is opposite to the direction of the belt
    and (2) the rotation of the rolling pin is
    in the same direction as the belt.

    -- John W. Langhaar, PA A'33

Solution:

    If the pin has a diameter of 5 cm and a rotational
    speed of 100 rpm the surface of the pin is moving
    at:
    
        100*2*pi/60*0.025 m/s = 0.2617994 m/s
        
    The friction force = 5 N
    
    The relative speed between the belt and the pin is:
    
        2 +/- 0.2618
        
    The Power dissipation is: 5*(2 +/- 0.2618) = 11.309 W or 8.691 W
    

## Bonus

    George is building a rectangular
    patio, which will be covered
    with one-foot-square concrete slabs of
    seven different colors. He has divided
    the patio into seven rectangular zones,
    each to be covered by slabs of a single
    color, with five different colors appearing
    around the perimeter of the patio
    and four different colors at the corners.
    The seven zones are all different sizes,
    but all have the same perimeter, which
    is less than 60 feet. What are the dimensions
    of the patio, and what are the
    dimensions of the seven zones?

    -- Enigma by Colin Singleton
    in New Scientist

TODO

    I peeked at the solution.  Way too complicated
    and still results in trial and error at the
    end.
    
    Not a good problem.


## Double Bonus

    Given the lengths of the N sides of an irregular
    polygon, how should the sides be arranged to maximize
    the enclosed area? Prove your answer.

    -- Puzzle Corner by Allan Gottlieb in Technology Review

Solution:

    Make a path as close as possible to a circle.  Put all
    the vertices on a circle for best result.
    
    Not sure how to prove this.
