---
title: "Bent Spring 2007"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Spring 2007


## Problem 1

    Find a primitive Pythagorean
    right triangle whose area is an
    integer consisting of repetitions
    (more than one digit long) of the
    same digit. An example of such an
    integer is 77,777.

    -- The Penguin
    Dictionary of Curious and
    Interesting Numbers by David Wells

Solution:

    Need:  a^2+b^2 = c^2 where a,b,c are integers

    Need:  1/2*a*b = d*(111...) - unknown number of 1s, unknown d
    
    We could try exhaustive search, a=1:1000, b=1:1000?
    
```{r}
for (a in 1:10000) {
    for (b in 1:1000) {
        c=round(sqrt(a^2+b^2))
        if (c^2 == (a^2+b^2)) {
            area = a*b/2
            if ((area>9) & (length(unique(strsplit(as.character(area),'')[[1]]))==1)) {
                print(c(a,b,c,area))
            }
        }
    }
}
```
    
Needed to expand range on a to find:

    a = 1924
    b = 693
    c = 2045
    area = 666666


## Problem 2

    After grading the advanced-math
    test but before handing it back,
    the teacher asked the five students
    in the class to predict how well
    other members of the class (but not
    themselves) did on the test. The
    teacher received the following five
    pairs of statements, each pair being
    made by a different student:

    i. Bill did better than Ed;
    Chuck did better than Bill.
    ii. Debbie did better than Chuck;
    Art did better than Debbie.
    iii. Ed did better than Art;
    Chuck did better than Ed.
    iv. Bill did better than Ed;
    Debbie did better than Bill.
    v. Chuck did better than Bill;
    Art did better than Chuck.

    The student with the highest grade
    was completely correct; the boy who
    came in last was completely wrong;
    and the other students had one
    correct prediction and one incorrect
    prediction. Rank the students from
    highest to lowest grade. There were
    no ties.

    -- Martin Hollis in New Scientist

Solution:

    The only feasible highest score is Debbie with order: D>A>E>C>B


    Art
    Bill
    Chuck
    Debbie (not the lowest - unless it is a boy named Debbie)
    Ed

    (i)   C>B>E    either Art or Debbie said this -- ART SAID THIS
    (ii)  A>D>C    either Bill or Ed said this    -- ED SAID THIS
    (iii) C>E>A    either Bill or Debbie said this -- BILL SAID THIS
    (iv)  D>B>E    either Art or Chuck said this  -- CHUCK SAID THIS
    (v)   A>C>B    either Debbie or Ed said this  -- DEBBIE SAID THIS

    One of these 5 statements is complely correct
    One of these is completely wrong
    The other three are half right/half wrong
    

Observations:

    D is not the lowest
    B,E not the two lowest -- if so then (i) and (iv) both completely true
    
    Chuck said statement (iv) -- it is the only possiblity for Chuck
    Art said statement (i) -- only remaining possiblity for Art
    
    
Assume:


    E is highest and (ii) A>D>C
        E>B>A>D>C - two false
        E>A>B>D>C - two false
        E>A>D>B>C - feasible <----- ART should be lowest - contradiction
        E>A>D>C>B - two true
    
    D is highest and (v) A>C>B
        D>E>A>C>B - no false
        D>A>E>C>B - feasible <----- BILL should be lowest - feasible
        D>A>C>E>B - no false
        D>A>C>B>E - two true


    A is highest and (i) C>B>E then
        A>D>C>B>E - not allowed
        A>C>D>B>E - not allowed
        A>C>B>D>E --- can't be true becaue if it is then (v) is true
        A>C>B>E>D - not allowed

    Conclude: A is not the highest

    C is highest and (iv) D>B>E
        C>A>D>B>E -- not allowed
        C>D>A>B>E -- not allowed
        C>D>B>A>E -- can't be true since (i) would also be true
        C>D>B>E>A -- can't be true since (i) would also be true

    Conclude: C is not the highest
        
    B is highest and (iii) C>E>A
        B>D>C>E>A - feasible <----- DEBBIE should be lowest - contradiction
        B>C>D>E>A - two false statements
        B>C>E>D>A - two false statements
        B>C>E>A>D - not allowed
    
    
I overwrote/erased some of the intermediate results used to identify who
made which statements between Bill, Ed, and Debbie.

## Problem 3

    The game of Quod, invented
    by Keith Still, is played on an 11 x
    11 grid of cells (like an oversized
    chessboard) minus the four corner
    squares. Players take turns placing
    markers on the cells, with the first
    player to form a square being the
    winner. As long as a player's four
    markers form the corners of a
    square, it is valid. Its sides do not
    have to be parallel to the sides of the
    board. How many different squares
    are possible?

    -- How to Cut a Cake and
    Other Mathematical Conundrums
    by Ian Stewart

TODO

    Another one where I am not sure I care.
    
    The Bent claims there are 1173 squares but they only
    consider squares that are parallel to the sides of the
    "chess board" and the problem specifically says the
    sides do NOT need to be parallel to the sides so I
    don't know if I believe the answer.
    


## Problem 4

    Let S = 1(1!) + 2(2!) + 3(3!) + ...
    + 222(222!). What is the remainder
    when S is divided by 2007?

    -- adapted from Mathematics Teacher

Solution:

    Brute force using gmp
    
    Remainder = 2006
    
    The Bent argument:
    
    Let N = 1(1!) + 2(2!) + ... + 222(222!)
    Let S = 1! + 2! + ... + 222!
    
    N+S-S = (1+1)(1!) + (2+1)(2!) + ... + (222+1)(222!) - S
          = 2! + 3! + ... + 223! - S
          = 223! - 1
          
    But 2007 = 3*3*223 so
    
    N/2007 = (223!-1)/(3*3*223)
    
    So: 223!/(3*3*223) = integer with no remainder.
    
    for a numerator of 223!-1 the remainder must be 2006


## Problem 5

    A fruit basket contains:

    - MELONS
    - PLUMS
    - APPLES
    - LEMONS
    - and a BANANA.

    If 2 MELONS are greater than 35
    PLUMS but less than 3 APPLES,
    and 99 LEMONS are greater than 16
    APPLES but less than 210 PLUMS,
    what is the size of a BANANA? The
    usual rules for cryptics apply. That
    is, the same letter always represents
    the same digit, and different letters
    represent different digits.

    -- Albert Haddad in New Scientist

Solution: solved in the cryptic arithmetic notbook

    > a[k,]
            m e l o n s p u a
    3266023 9 0 1 2 3 8 5 4 6
    
    By inspection, b=7 since it is the only missing digit:
    
    > banana=763636


## Bonus

    A bag of marbles is prepared
    by putting six black marbles in the
    bag. Then, a die is thrown, and a
    number of white marbles is added
    equal to the number shown on the
    die. You are given the bag. You know
    the procedure used to fill the bag
    but not the number of white balls
    in the bag. You draw three marbles
    without replacement, and they are
    all white. What is the probability
    that the next marble you draw will
    be black?

    -- from an old exam at
    Jesus College, Cambridge

Solution:  enumerate the possibilities?

    Given: 6B, ?W - draw three W in a row
    Find: P(next marble is black)

    My initial (naive) estimate assumed that the die
    must have been 3, 4, 5, or 6 and that each of
    these occurred with 25% probability.  While that
    may be true -- I was ignoring the draw of 3 consecutive
    W marbles -- this greatly increases the likelihood
    that the number of marbles put in the bag was
    one of the higher amounts.  Once we figure out
    the probability for each scenario then the final
    draw is simply:

    6B, 3W - 3W -> next marble black, for sure
    6B, 4W - 3W -> P=6/7
    6B, 5W - 3W -> P=6/8
    6B, 6W - 3W -> P=6/9

    Let's calculate the likelihood of each scenario:
    
    6B, 3W - Draw 3W:  P = choose(3,3)/choose(9,3) = 1/84
    6B, 4W - Draw 3W:  P = choose(4,3)/choose(10,3) = 1/30
    6B, 5W - Draw 3W:  P = choose(5,3)/choose(11,3) = 2/33
    6B, 6W - Draw 3W:  P = choose(6,3)/choose(12,3) = 1/11
    Ptotal = 303/1540
    
    So each scenario has the following likelihood:
    
    6B, 3W - Draw 3W: 55/909
    6B, 4W - Draw 3W: 154/909
    6B, 5W - Draw 3W: 280/909
    6B, 6W - Draw 3W: 420/909
    
    The probability that the next marble is black is then:
    
    P=55/909 + 6/7*154/909 + 6/8*280/909 + 6/9*420/909
     = 677/909
     ~ 0.7448


## Double Bonus

    Given a consecutive
    run of positive integers from 1
    through N, divide them into four
    sets such that for each set, no two
    numbers of that set have a sum that
    equals a third member of the same
    set. What is the maximum value of
    N, and what are the members of the
    four sets?

    -- Doctor Ecco's Cyberpuzzles
    by Dennis E. Shasta

TODO

    The Bent says the best they can do is N=66 but there is no
    proof of the upper limit.  They claim there are *thousands*
    of solution with N=66 and present one example.
    
    Basically, add numbers to set 1 until you can't then add
    to set 2.  Go back to set 1 when able and start adding again.
    Keep repeating until there is no place to put the next number.

    Well - almost.  Don't really understand where they put 7.
    