---
title: "Bent Winter 2017"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Winter 2017


## Problem 1

    Looking for some PUZZLEs to fill up your
    WINTER? What is the solution to the following
    cryptarithm?

    PUZZLE + PUZZLE + PUZZLE + PUZZLE = WINTER.

    Standard rules apply. Each different letter
    stands for a different digit, and each
    different digit is always represented by the
    same letter; no leading zeros are allowed.

    -- Jeffrey R. Stribling, Ph.D.  CA A'92

Solution: from the cryptic arithmetic notebook

      PUZZLE + PUZZLE + PUZZLE + PUZZLE = WINTER

    a=make_crypt_df('puzzlewinter')
    k=(make_num(a,'puzzle')*4 == make_num(a,'winter'))
    k=k&(a$p!=0)&(a$w!=0)
    sum(k)
    [1] 1
    a[k,]
           p u z l e w i n t r
    832790 2 3 7 1 6 9 5 0 8 4
    make_num(a,'puzzle')[k,]
    [1] 237716
    make_num(a,'winter')[k,]
    [1] 950864


## Problem 2

    Art, Bob, Cai, Dan, and Eli took a test and
    were ranked (with no ties) according to their
    performance. They were each informed of their
    own ranking, but no one was told details of
    the others' ranks. Art said he wasn't second,
    and Bob, who had recently claimed himself a
    psychic, announced confidently he was sure he
    was two places better than Dan. Cai overheard
    these remarks and came to the conclusion -- for
    reasons that need not concern us -- that one of
    them was right and the other wrong. After a
    pause for reflection Cai, who is a Tau Bate
    and hence highly intelligent, said that he
    could announce the correct ranking for all
    five people. But when he did so, he had got
    everyone in the wrong place except the bottom
    two. What is the correct order? (It is
    somewhat satisfying that Art and Bob are
    placed in the order of their truthfulness.)

    -- 101 Brain Puzzlers, E.R. Emmet

TODO

    meh


## Problem 3

    Find a scalene triangle with integral sides
    and no right angle, such that at least one of
    its angles is an integral number of degrees.
    What are the lengths of the sides of such a
    triangle with the smallest perimeter?

    -- Adapted from Technology Review

Solution:

    Apply law of cosines to find SSS for a known
    cosine?

    I am shocked -- that is exactly what the Bent did.
    They noted that the only rational values for cos(theta)
    is for theta = 60, 90, or 120 degrees.  Since we can't
    use 90 degrees we can pick 60 or 120 degrees.
    
    cos(60 degrees)  = +1/2 = (a^2+b^2-c^2)/(2*a*b)
    cos(120 degrees) = -1/2 = (a^2+b^2-c^2)/(2*a*b)
    
    c^2 = a^2+b^2 +/- a*b
    
    Need to solve this by trial and error.  The solution
    they found was a=3, b=5, c=7.  The perimeter is 15.
    
    Not sure that they rigorously showed this was the
    minimum perimeter but it probably is.
    

## Problem 4

    The end of my sink faucet bends at a right
    angle to the horizontal so that the opening,
    which has an inside diameter of 2.0 cm,
    points vertically down. The water system has
    a pressure regulator that, with the faucet
    fully open, produces a steady downward stream
    with a circular cross section at a rate of
    8.328 L/min. I noticed that the stream tapers
    down to a diameter of 0.9 cm at the point at
    which it reaches the drain. To the nearest
    millimeter, how far is the drain from the
    faucet opening? Use a value of 9.807 m/s2 for
    the standard acceleration due to gravity.

    -- Adapted from The Chicken from Minsk by
       Yuri Chernyak and Robert Rose

Solution:

    I never really thought about it before but if the
    flow rate is known at the exit from the faucet this
    defines a linear velocity of the fluid.  Since the
    fluid is falling toward the drain it is increasing
    in velocity.  As the velocity increases the cross
    sectional area has to decrease in order to maintain
    a constant flow rate.
    
    Interesting.
    
    At the exit from the faucet the cross sectional area
    is pi*(1)^2 cm^2 with a flow rate of 8328 cm^3/min.
    This requires a linear velocity of 8328/pi ~ 2651 cm/min.
    This is 44.18141 cm/sec.
    
    At the drain the cross section has been reduced but the
    flow rate is still 8328 cm^3/min.  If the diameter is 0.9 cm
    then the linear velocity is 8328/(pi*0.45^2) ~ 13091 cm/min.
    This is 218.18 cm/sec
    
    We should be able to use an energy balance to find the
    height of the faucet:
    
        1/2*0.44181^2 + g*h = 1/2*2.1818^2
        
        h = 1/2*(2.1818^2-0.44181^2)/g ~ 2.282528/g ~ 0.2327 m
        
    The faucet height is 23.27 cm above the drain.

    The Bent gave 23.3 cm as the solution.


## Problem 5

    Allen and Bill were nominated for president
    of the Golf Club. The club has 350 members,
    not all of whom voted. The tellers removed
    the votes from the ballot box and counted
    them one at a time. Allen not only won, but
    he was ahead of Bill throughout the entire
    count. If the probability of this occurring
    was exactly one in 100, how many votes did
    Allen and Bill each receive?

    -- New Scientist: Colin Singleton

TODO

    Huh?
    
    I looked at the Bent solution and I still have no
    idea what they are talking about.


## Bonus

    Consider a triangular number T which is
    simultaneously the sum of two cubes and the
    difference of two cubes (e.g., T = A^3 + B^3 =
    C^3 - D^3 = n(n+1)/2) such that D = B + 1.  The
    first such number is 91 = 33 + 43 = 63 - 53,
    and the second is 48427561 = 3603 + 1213 =
    3693 - 1223. What is the general formula for
    the ith such number?

    -- Elementary Number Theory,
       David M. Burton

TODO

    No clue.
    
    I read the solution and, okay, it is understandable
    but a bit ludicrous.


## Computer Bonus

    The prime-counting function, usually denoted
    as pi(x), returns the number of primes less
    than or equal to a positive integer x.
    Similarly, the semi-prime-counting function,
    usually denoted as pi_2(x), returns the
    corresponding number of semi-primes (integers
    with exactly two prime factors, not
    necessarily distinct). For example, p(20)=8
    and p2(20)=6. What is the minimum value of x
    such that pi_2(x)/pi(x) > pi (the mathematical
    constant 3.1415...)?

    -- Jeffrey R. Stribling, Ph.D., CA A'92

TODO

    No idea -- I would just have to go with exhaustive search.
    
    I guess that is what they did as well but they did not
    explain the result.
    
    I would have to start counting at 2 and just decide if each
    number is prime, semi-prime, or neither and then increment
    the counter and check the ratio.

    The runs VERY, VERY, SLOWLY.
    
    The values given below took over 2 hours and we still aren't there.
    Not sure if this is algorithmic or the language.
    
    The solution should be ~ 693e6
    
    Even after more than 2 hours of running the ratio below is only ~ 3.113654
    
    This is conceptually simple but the execution time is horrid.
    
    As bad as it seems -- this is faster than Python or Lua!
    
    It takes about 26-27 seconds per million values so about 120 million per hour.
    To reach 694 million is a little over five hours of run time.
    
```{r, eval=FALSE}
library(gmp)

pi_count=22727596
pi2_count=70765869

n=as.bigz(427606621)
t1=Sys.time()
while (TRUE) {
    m=length(factorize(n))
    if (m==1) { pi_count = pi_count+1 }
    if (m==2) { 
        pi2_count = pi2_count+1
        if (pi2_count/pi_count > pi) {
            break
        }
    }
    n=n+1
    if (n%%1000000 == 0) {
        t2=Sys.time()
        print(as.numeric(n)*1e-6)
        print(t2-t1)
        print(pi2_count/pi_count)
        t1=t2
    }
}
print(n)
print(pi_count)
print(pi2_count)

```


Solution found after a long, long time:

    n = 693901529
    prime_pi =    35953285  (number of primes up to n)
    prime_pi2 =  112950577  (number of semi-primes up to n)

Amazingly enough these are in exact agreement with the values
from the Bent.
