---
title: "Bent Fall 2008"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Fall 2008


## Problem 1

    When the five marriage counselors
    at Healthy Relations got divorced,
    it was a shock. When all announced
    remarriage and the brides were
    revealed to be the five ex-wives, it
    was a sensation. Still, the priggish
    Dinah was not the first to remarry
    and there were no direct swaps, so
    the decencies were preserved. The
    weddings were held on successive
    Saturdays. Peter's took place earlier
    than Anne's and later than Quentin's.
    Barbara's was later than Tristram's
    and earlier than Celia's. Peter married
    Simon's ex-wife. Barbara got
    hitched to the man whose former
    wife married Emily's ex-husband.
    Quentin paired up with the lady
    whose former husband married
    Dinah. Ronald was spliced with the
    lady whose ex-husband married
    Celia. Who remarried whom? No one
    remarried his or her ex-spouse.

    -- Martin Hollis

Solution:

    Dinah
    Anne
    Barbara
    Celia
    Emily

    Peter
    Quentin
    Tristram
    Peter
    Simon
    Ronald

    Order of Weddings (not necessarily sequential)
    
    Quentin  -> Peter   -> Anne
    Tristram -> Barbara -> Celia

    Peter married Simon's ex-wife
    Barbara married man whose former wife married Emily's ex-husband
    Quentin married woman whose ex-husband married Dinah
    Ronald married woman whose ex-husband married Celia

    No one remarried ex-spouse, no direct spouse swaps
    
    Need a way to represent this graphically!!!
    

## Problem 2

    Place eight queens on a standard
    chessboard so that the number of
    squares not attacked is a maximum.
    What is this number? Describe the
    positions of the queens, labeling
    columns a-h and rows 1-8.

    -- Technology Review


Solution:  I think I looked into this a long time ago
using grumpy.  I think the answer is 11.  I even made
a gui to play with this and some code to search for
the solution.  Ignoring reflections there are only
choose(64,8) combinations to explore.  This is about 4e9.
If you take advantage of reflections this can be drastically
reduced.

In the end there are 7 unique patterns that generate the
optimal solution of 11 squares not attacked by the 8 queens.

I checked on grumpy under ~/temp/puzzles/queens and there
were the files.  I resurrected them and pulled them into
a subdirectory called queens on this machine.

Even on this machine the exhaustive search is going to take
a long time.  I had to port from Python 2 to 3 but it is
running very slow.  May want to port this further to LuaJIT.

Furthermore, it was not running correctly.  Over the years the
syntax on the nonzero() function has changed and it returns a
tuple instead of a list so the counting logic was fubar.

I fixed this, after several hours of wasted effort, and started
up the search again.  After a short while I killed the search
and confirmed that it was generating the correct output.  I
started it once more and I will allow it to run to conclusion.

I was able to improve the pruning code a bit so maybe it will
run a bit faster.  Oh yes -- much faster.  The new results
match the old ones.

BTW the solution they give in the Bent is the same as my solution
zero except it is flipped top-to-bottom.

From the old `readme.txt` file:

    Problem:
    --------
    Position 8 queens on a chessboard such that that number of
    squares that is not attacked is a maximum
    
    Seems simple enough but the number of combinations is large
    C(64,8)~4.4 Billion.  With reflections, translations, rotations,
    etc. I am sure the number is much smaller ... but there doesn't
    seem to be an obvious alorithm to search out just the unique
    configurations.
    
    queens3.py performs exhaustive search
    queens.tcl provides a board to try the problem manually.
    
    After exhaustive search:
    ------------------------
    The best achievable score is 11.
    There are 48 configurations that achieve this.
    Many are reflections or rotations.
    They have not been removed.
    
    Assume the squares are numbered 0-63 row-wise starting at the
    upper left corner of the board.  The 8 queens should be positioned
    as follows to achieve this score:


     0  1  2  9 10 11 40 41
     0  1  2  9 10 32 40 41
     0  2  9 11 32 33 40 41
     0  4  5  8  9 13 16 17
     0  4  5  9 12 13 16 25
     0  5  8  9 13 16 17 25
     1  2 10 11 32 33 40 41
     1  3 16 19 24 25 32 33
     1  6  8 10 14 17 48 49
     1  6  9 13 15 22 54 55
     
     ... etc ...
     
    38 47 50 51 54 58 59 63
    40 41 48 49 53 56 60 61
    46 47 50 54 55 58 59 63


## Problem 3

    The result of adding the date of
    the last Monday of last month and
    the date of the first Thursday of
    next month is 38. If both dates are
    in the same year, what is the current
    month?

    -- Hard-to-Solve Brainteasers by
    Jamie and Lea Poniachik

Solution:

    The current month must be August.

    The latest a first Thursday can occur is on the seventh
    of a month.  The only way to get a total of 38 = n+7 is
    when n=31 so we conclude:
    
    Last Monday of previous month is 31
    First Thursday of next month is 7
    
    Current month begins on Tuesday and the last day of the
    month is a Thursday.  The dates in the current month
    must be:
        Tuesday = 1
        Tuesday = 8
        Tuesday = 15
        Tuesday = 22
        Tuesday = 29
        Wednesday = 30
        Thursday = 31 --> current month has 31 days
        
    The previous month has 31 days and the current month has
    the same number (31 days).  If both dates occur in the
    same year these must be July and August so the current
    month must be August.


## Problem 4


    Starting with a standard 52-card
    deck and drawing cards randomly
    without replacement, what is the
    expected number of cards that must
    be drawn until each of the four suits
    is represented at least once?

    -- The Theory of Gambling and Statistical
    Logic by Richard A. Epstein

Solution:

    Let's try Monte Carlo and then (maybe) combinatorics.
    
```{r}
ntrials=100000

n=0
set.seed(2020)
for (trial in 1:ntrials) {
    s=sample(52)
    s=trunc((s-1)/13)+1
    m=c(0,0,0,0)
    for (i in s) {
        m[i]=m[i]+1
        if (all(m>0)) {
            n=n+sum(m)
            break
        }
    }
}
print(n/ntrials)
```

For 100e3 trials the mean number of cards = 7.66367

    Seed   ntrials   Num Cards
    ==========================
    42     100e3     7.67276
    123    100e3     7.66918
    2020   100e3     7.65759

I don't know if it is worth trying to do this analytically.

They go through the method in the Bent.  Basically you sum up the probabilities
from draw=4 through 40 or mess with expectation equations.

The result = 4829/630 = 7.665079

To three digits I got the same result with a very quick Monte-Carlo run!


## Problem 5

    A stick of length L is standing
    vertically against a wall, with a small
    beetle sitting on top of the stick. The
    stick slides down the wall at a constant
    rate in such a way that the top
    maintains contact with the wall and
    the bottom maintains contact with
    the floor.
    The beetle starts to crawl down
    the stick at a constant speed just as
    the stick starts to slide. If the beetle
    reaches the bottom of the stick just
    as the top of the stick reaches the
    floor, provide the equation of the
    beetle's path in terms of Cartesian
    coordinates only.

    -- Jeffrey R. Stribling, CA A ?92

TODO

I must admit I am intrigued but I don't know how to
solve this.  The fact that the stick slides down the
wall at a constant rate might make this more tractable.

The speed of the beetle is directed along the length of
the stick so we have a vector with speed and direction.

I looked at their solution.  One of the keys is to realize
that the distance between the top of the ladder and the
floor is the same as the distance of the beetle from the
end of the ladder (has to be for them to arrive at the
same time).  This allows you to work similar triangles and
get rid of some variables.

They give the equation of the beetle as:

    u = (1-v^0.5)*(1-v)^0.5
    
    where u = x/L, v=y/L


## Bonus

    A spherical bead of mass M
    is free to slide along a frictionless,
    horizontal wire. Hanging from this
    bead at the end of a massless wire of
    length L is a lighter sphere of mass
    m. If the lighter sphere is given a
    small displacement in a direction
    parallel to the horizontal wire, what
    is the period of the resulting oscillation?

    -- John R. Sellars

TODO

I must admit I am intrigued but I don't know how to
solve this.  I could try to write the equations of
motion and see where that leads.  We will probably
need to approximate sin(x)~x for small x and the result
will be a harmonic oscillator with the smaller mass
acting as a pendulum but the pivot point moves due
to tension in the wire.

I looked at their solution and that is what they did.
I ended up with a 4th degree ODE so I should probably
work through this to make sure I could.

The end result should be:

    The period of oscillation is T=2*pi*[(L/g)*(M/(m+M))]^0.5
    
    The equation is: x'''' + [g*(m+M)/(M*L)]*x'' = 0
    
    They make the small angle approximation sin(theta) ~ theta
    
    Clearly this has a sinusoidal response with repeated roots
    at zero that disappear when the boundary conditions are
    applied.  The oscillation frequency is w0^2 = g*(m+M)/(M*L).
    



## Double Bonus

    The infinite horn
    of plenty for x > 1 is generated by
    revolving the curve y =1/x about
    the x-axis. (For this problem, let x
    and y be in feet.) Your job is to paint
    the inner surface of the horn using
    the new nanotech paint. Nanotech
    paint comes in one gallon cans, can
    be applied as an infinitely thin layer
    and dries that way, but it still takes
    one gallon of paint for each 10,000
    square feet of surface area. How
    would you paint the inner surface of
    the horn, and how much paint would
    it require?

    -- John L. Bradshaw, PA Alpha'82

Solution:

    OOOps -- when I started solving this I was rotating around
    the y-axis.  If I use the x-axis instead the results
    are a bit more interesting.

    Around the X-axis the volume is finite:
    
    V = integrate(pi*(1/x)**2,(x,1,oo)) --> pi
    
    The surface area of the horn, however, is not finite:
    
    A = integrate(2*pi*(1/x),(x,1,oo)) --> infinity
    
    So - paradoxically you can fill the horn with pi cubic feet of
    paint you can never paint the internal surface since it is infinite.
