---
title: "Bent Summer 2015"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Summer 2015


## Problem 1

    There was a young and adventur-
    ous man who found among his great-
    grandfather's papers a map on a
    piece of parchment that revealed the
    location of a hidden treasure. The
    instructions read:

    "Sail to ___ latitude and ___ longi-
    tude where thou wilt find a deserted
    island. [When this Tickler first ap-
    peared in a 1955 Bent, the island's
    location was given as 16 o N 16 o W,
    but that is obviously incorrect, and
    the true coordinates, which were
    known to the young man at the time
    he found the map, have since been
    lost.] There lieth a large meadow,
    not pent, on the north shore of the
    island where standeth a lonely oak
    and a lonely pine. There thou wilt
    see also an old gallows on which we
    once were wont to hang traitors.
    Start thou from the gallows and
    walk to the oak counting thy steps.
    At the oak thou must turn right
    by a right angle and take the same
    number of steps. Put here a spike
    in the ground. Now must thou re-
    turn to the gallows and walk to the
    pine counting thy steps. At the pine
    thou must turn left by a right angle
    and see that thou takest the same
    number of steps, and put another
    spike into the ground. Dig halfway
    between the spikes; the buried trea-
    sure is there."
    The instructions were quite clear
    and explicit, so our young man char-
    tered a ship and set sail. He found
    the island, the field, the oak and the
    pine, but to his great sorrow the
    gallows were gone. Too long a time
    had passed since the document had
    been written; rain and sun and wind
    had disintegrated the wood and re-
    turned it to the soil, leaving no trace
    even of the place where it once had
    stood. The young man, not having
    been a Tau Bate, gave up in despair.
    Perhaps YOU can find the gallows'
    former location for him.

    -- One, Two, Three...Infinity
    by George Gamow

TODO

    I never tried this problem but there is apparently
    a typo and they weren't really looking for the
    location of the gallows.  They really wanted the
    location of the treasure.


## Problem 2

    Consider five points arranged as
    the vertices of a regular pentagon
    and labeled A through E. Identical
    resistors are used to connect each
    of the following pairs of points: A-B,
    A-C, A-D, B-C, B-D, B-E, C-D, C-E,
    and D-E. Note that the pair A-E
    is not connected. The resistance
    between points A and E is 1 ohm.
    When three of the nine resistors are
    removed, the resistance between
    points A and E remains 1 ohm.
    Which resistors are removed, and
    what is the value of one of the resis-
    tors?

    -- Hubert W. Hagadorn, PA E '59

Solution:  If I can't solve this I should be forced to
turn in my diploma.   Brave statement.  Fortunately, I
succeeded.

    ASCII Art below:
  
                C
      +----------o--------------+
      |         / \             |
      |        /   \            R
      |       R     R           |
      |     A/       \  B       | E
      R     o----R----o----R----o
      |      \       /          |
      |       R     R           |
      |        \   /            R
      |         \ /             |
      +----------o--------------+
                D

    I could ground the E node and drive the A node with a 1 A source.
    The voltage AE is the resistance.  This requires solving 4 equations
    in 4 unknowns.  Maxima can do this easily or I could try some
    manipulation of the resistors.  Especially Delta-Wye transformations.
    
    The equation approach:
    
    At node A: (Va-Vc)/R+(Va-Vb)/R+(Va-Vd)/R-1=0
    At node B: (Vb-Vc)/R+(Vb-Va)/R+(Vb-Vd)/R+Vb/R=0
    At node C: (Vc-Vd)/R+(Vc-Va)/R+(Vc-Vb)/R+Vc/R=0
    At node D: (Vd-Vc)/R+(Vd-Va)/R+(Vd-Vb)/R+Vd/R=0
    
    Solve for Va, Vb, Vc, Vd.
    
    Set Va=1 -> solve for R.
    
    Va = 2/3*R = 1 --> R = 3/2 (First Solution).


    Attempt at removing resistors (SUCCESS)
    
    Let us go back to the first circuit and look at the solutions:
    
    Va=2/3*R, Vb=R/3, Vc=R/3, Vd=R/3
    
    Since Vb=Vc=Vd we can remove resistors bridging these nodes:
    
                C
                 o--------------+      (Second Solution)
                /               |
               /                R
              R                 |
            A/          B       | E
            o----R----o----R----o
             \                  |
              R                 |
               \                R
                \               |
                 o--------------+
                D
                
    In this case we get:
    
    At node A: (Va-Vc)/R+(Va-Vd)/R+(Va-Vb)/R-1=0
    At node B: (Vb-Va)/R+Vb/R=0
    At node C: (Vc-Va)/R+Vc/R=0
    At node D: (Vd-Va)/R+Vd/R=0
    
    Solving yields the same values as the original circuit.

    Note that we can also solve this by inspection since there are
    three parallel branches of 2*R between nodes A and E.  This
    gives a total resistance of 2/3*R as desired.

    The solution in the Bent is to visualize this as a three
    dimensional surface with A and E at opposite ends.  The B,C, and D
    note are around the middle.  There are resistors connecting
    all the edges of the solid.  Clearly the loop B,C,D has no driving
    force and is symmetrical wrt AE so these resistors carry no current
    (same as my solution) and can be eliminated.
    
    This leaves three parallel branches connecting AE, each has a
    resistance of 2*R so:
    
      Req = 2/3*R = 1 --> R=3/2.


## Problem 3

    A new theater, with exactly 200
    seats, has just opened. Every night
    200 people, each with an assigned
    seat, line up. However, the first
    person in line is an acoustic engi-
    neer, who is interested in assessing
    the sound quality throughout the
    theater. The engineer picks a seat
    at random, which may or may not
    be his assigned seat. As the rest of
    the people enter in order, if their as-
    signed seat is vacant, they sit in it. If
    not, they pick any unoccupied seat.
    What is the exact probability that
    the last person in line sits in their as-
    signed seat?

    -- The Everything Brain Strain
    Book by Jake Olefsky

Solution:

    They make the argument that it is 50% regardless
    of the size of theatre.  I couldn't really follow
    their argument -- I will just try some Monte Carlo
    to confirm.
    
```{r}
ntrials=10000
nseats=200

set.seed(1)
n=0
for (trial in 1:ntrials) {
    s=sample(nseats,nseats-1)
    x=1:nseats
    # Engineer can steal any seat he/she wants
    engineer=sample(nseats,1)
    x=x[-which(x==engineer)]
    for (i in s[1:(nseats-2)]) {
        # If seat is available then sit in it
        if(i %in% x) {
            k=which(x==i)
        } else {
            # Pick a random seat from those remaining
            k=which(x==sample(x,1))
        }
        x=x[-k]
    }
    # If the last person has the correct seat
    # available then consider that a success.
    if (s[nseats-1] %in% x) { n=n+1 }
}
print(sprintf("Trials = %d,  Success = %d   Pr = %10.6f",ntrials,n,n/ntrials))
```

    Trials   Seed   Pr
    ----------------------
    10000    1      0.4987
    10000    42     0.5023
    10000    123    0.5065
    10000    2020   0.4937

    Mean = 0.5003  --> conclude that there is no reason to doubt the
    Bent claim that the probability is 50%.
    

## Problem 4

    The DryFroot Company makes
    one pound (16 oz) boxes of Froot-
    Mix. Each box is made to the same
    recipe and has a positive integral
    number of ounces of apples, bananas,
    cranberries and dates. The com-
    pany also makes one pound boxes of
    MixedFroot, and although the recipe
    is different, each box still contains
    an integral number of ounces of the
    same four fruits. Finally, DryFroot
    produces giant boxes of TootyFroot-
    ies by mixing four pounds of Froot-
    Mix with three pounds of Mixed-
    Froot. In each of the three products,
    the weights of the four fruits are dif-
    ferent. The respective boxes list the
    ingredients in descending order by
    weight as follows: FrootMix: dates,
    cranberries, bananas, apples; Mixed-
    Froot: apples, bananas, cranberries,
    dates; TootyFrooties: bananas, cran-
    berries, dates, apples. How many
    ounces of each fruit are there in a
    box of each product?

    -- An Enigma by Keith Austin in
    New Scientist

TODO

    meh
    

## Problem 5

    Solve the following cryptic equa-
    tion:
    TBP = sqrt(PUZZLE + 1)
    All the usual rules apply. There are
    no leading zeros. Each letter repre-
    sents a different digit, and the same
    letter always represents the same
    digit.

    -- Adapted from Allan Gottlieb's
    Puzzle Corner in Technology Review

Solution: from the cryptic arithmetic notebook

    Cryptic square root:

      TBP = sqrt(PUZZLE + 1)

    a=make_crypt_df('tbppuzzle')
    k=(make_num(a,'tbp')^2 == (make_num(a,'puzzle')+1))
    k=k&(a$t!=0)&(a$p!=0)
    sum(k)
    [1] 1
    a[k,]
           t b p u z l e
    550374 9 0 8 2 4 6 3


## Bonus

    A dozen POWs share a cell.
    What is important to them is that
    they escape as soon as possible.
    They have found a way to remove
    the bars on the window of their cell
    so they can climb through and, once
    on the ground, plan to head due
    north, climb over the east-west fence
    near the window, and disappear. But
    they can only hope to escape unseen
    if they do it in the dark, one man at a
    time, and with each man having from
    the time he starts, a clear two min-
    utes with all four guards at least 200
    meters away from a point directly
    below their window.
    The guards each walk an east-
    west route, two inside the fence and
    two outside. Robinson starts 300 me-
    ters west of the prisoners' window,
    marches to a point exactly outside
    their window, and turns around and
    marches back. Smith starts outside
    their window, marches 400 meters to
    the east, turns around, and marches
    back.
    Outside the fence, Phillips starts
    500 meters west of their window,
    marches 550 meters to his turn
    around point, 50 meters east of their
    window. Quinn starts at Phillips'
    turn around point, marches 500 me-
    ters east and turns around.
    The guards march back and forth
    on their beats at a uniform speed of
    100 meters per minute. (Assume no
    time is spent in turning.)
    Because it is winter, the POWs
    reckon that it is dark enough for
    their purposes at 6 p.m. until 6 a.m.
    The guards of one shift begin their
    routes at the above indicated loca-
    tions at precisely 6 p.m.
    How many prisoners can success-
    fully escape? At what time(s) should
    they start their attempts? Express
    your answer in minutes.

    --Brain Puzzler's Delight
    by E.R. Emmet

TODO

    Seems like ugly problem in modulo arithmetic
    and looking for collisions in the times when
    all the guards are in the proper position.


## Computer Bonus

    A k-digit automorphic, or circular, number
    is a number whose square ends in the same k
    digits as the number itself.  That is, N is a
    k-digit automorphic number if N^2 = N (mod 10^k).
    For example, 76 is a 2-digit automorphic
    number because 76^2 = 5776. K-digit
    automorphic numbers are known to come in
    pairs. What is the leftmost (that is, most
    significant) digit of each of the two
    1,001-digit, base-10, automorphic numbers?

    -- J.C. Rasbold, OH A'83

Solution:

    Brute force?  Seems like it will take forever.  And it practically will
    if I don't find a better search strategy.
    
    OK - I gave up and looked at the solution.  It would have taken me forever.
    There are two keys:
    
    1. For N digit circular number the last N-1 digits must be a circular number.
       You can start with the 1 digit values: 1, 5, 6 and then try the 10 leading
       digits with each (9 leading digits?) to find the only two digit values:
       25 and 76.  Note the last digit is in the set {1,5,6}.  You can continue
       to try leading digits until you build up a value of 1001 digits.  Not sure
       if this is a tree or if you can just stick with one value and crank.  Note
       that 1 is sort of special since it does not generate any larger circular
       numbers.  The other values do: 5 -> 25, 6 -> 76.
       
    2. Apparently Wikipdedia has a page on this and has a generating equation
       of (3*N^2 - 2*N^3) mod (10^(2*k)).  I don't think this generates all the
       values but it should be sufficient.  Once you reach a number with more than
       1001 digits the final 1001 digits must be circular (see #1 above).  As
       noted above, 1 is a special value and is a fixed point for this function.
    
The original brute force code is included for posterity -- but don't run it.    

```{r, eval=FALSE}
library(gmp)

is_circular = function(n) {
    n1=as.bigz(n)
    n2=as.character(n1*n1)
    n1=as.character(n1)
    num_digits=nchar(n1)
    n2=substr(n2,num_digits+1,nchar(n2))
    if (n1==n2) {
        return(TRUE)
    }
    return(FALSE)
}

m=0
n1=as.bigz(paste0("1",strrep("0",1000),collapse=""))
while(TRUE) {
    if (is_circular(n1)) {
        print(n1)
        break
    }
    n1=n1+1
    m=m+1
    if ((m%%1000000) == 0) { print(sprintf("Working, m/1e6 = %d",m/1000000)) }
}
```

The revised code based on the generator from Wikipedia runs blindingly fast:

```{r}
library(gmp)

gen_next = function(n) {
  n1=as.bigz(n)
  m=nchar(as.character(n1))
  m2=as.bigz(10)^(2*m)
  n2=(3*n1^2-2*n1^3)%%m2
  return(n2)
}

n=5
while (1) {
  m=nchar(as.character(n))
  if (m>1000) { break }
  n=gen_next(n)
}
n25=substr(as.character(n),m-1000,m)

print(sprintf("Using 5 as seed, result = %s ...",substr(n25,1,10)))

n=6
while (1) {
  m=nchar(as.character(n))
  if (m>1000) { break }
  n=gen_next(n)
}
n76=substr(as.character(n),m-1000,m)

print(sprintf("Using 6 as seed, result = %s ...",substr(n76,1,10)))

```


The resulting values start with:

    Using 5 as seed, result = 3127812540 ...
    Using 6 as seed, result = 6872187459 ...
