---
title: "Bent Winter 2006"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Winter 2006


## Problem 1

    One hundred parishioners attended
    our church picnic, and 90 had
    a hot dog, 80 had potato salad, 70 had
    cake, and 60 had ice cream. However,
    no one ate all four items. How many
    had both cake and ice cream?

    -- Classic Mathemagic by
    R.A.Blum, A. Hart-Davis, B. Longe,
    and D. Niederman

Solution:  30 people ate Cake and Ice Cream

    If no one had all four items
    then everyone skipped (at least) one item.


    10 people skipped the hot dog
    20 people skipped the potato salad
    30 people skipped the cake
    40 people skipped the ice cream
    
    Total = 100 people skipped something so
    each person skipped exactly one item.
    
    The four meals (along with counts) were:
    
    n1 = Hot dog and Potato Salad and Cake
    n2 = Hot dog and Potato Salad          and Ice Cream
    n3 = Hot dog and                  Cake and Ice Cream
    n4 =             Potato Salad and Cake and Ice Cream

    We know:
    
    n1+n2+n3 = 90
    n1+n2+n4 = 80
    n1+n3+n4 = 70
    n2+n3+n4 = 60
    
    Solving yields: n1=40, n2=30, n3=20, and n4=10
    
    Cake and Ice Cream = n3+n4 = 30


The Bent solution was actually quite interesting.  They make the
same observation about no overlap in the skipped foods and then
make a series of bars that represent the foods and the skipped
foods as follows:

    - = food consumed, X = food skipped

    Hot Dog       |X---------|
    Potato Salad  |-XX-------|
    Cake          |---XXX----|
    Ice Cream     |------XXXX|

    Since each group is independent we can just look at the Cake and
    Ice Cream lines and see that only 30 people ate both.  This is
    very elegant.
    
    This also makes it possible to solve the problem by inspection.
    Since no person skipped more than one food we know:
    
    10 people skipped the hot-dog      -> they ate Cake and Ice Cream
    20 people skipped the potate salad -> they ate Cake and Ice Cream
    The other people skipped either Cake or Ice Cream so they did not
    eat both.
    
    Trivially, then, the total eating Cake and Ice Cream is 30.


## Problem 2

    Al, Sam, and Ralph play a challenge
    tennis tournament in which
    two of them play a set, with the
    winner staying on the court to play
    the person who sat out. In the course
    of the tournament, Al played 15
    sets, Sam played 14 sets, and Ralph
    played 9 sets. Who played in set #13?

    -- Richard I. Hess, CA B '62

Solution:  Al-Sam play in set #13

    The total number of sets must be
    half of the total of the three players:
    
    Number of sets = (15+14+9)/2 = 19 sets

    Who could Al have played and how many times.
    
    Al vs Sam = 14 times -> only leaves 1 set for Ralph
    Al vs Sam = 13 times -> leaves 3 sets for Ralph
    Al vs Sam = 12 times -> leaves 5 sets for Ralph
    Al vs Sam = 11 times -> leaves 7 sets for Ralph
    Al vs Sam = 10 times -> leaves 9 sets for Ralph
    
    Conclude:
    
    Al played Sam 10 times
    Al played Ralph 5 times
    Sam played Ralph 4 times
    
    What sequence could generate this?
    The following sequence works -- is this unique?
    
    (The Bent argues that Ralph plays 9 times and nobody sits more than
    one set in a row so the only way for this to occur is if Ralph
    plays in all the even sets.)
    
    Al-Sam play all odd sets
    Ralph plays even sets with alternating opponents starting with Al.
    
    Al plays 15 sets:    WWL-WWL-WWL-WWL-WWX
    Sam plays 14 sets:   LWW-LWW-LWW-LWW-LX
    Ralph plays 9 sets:  all losses
    
    1.	Al-Sam
    2.	Al-Ralph
    3.	Al-Sam
    4.	Sam-Ralph
    5.	Al-Sam
    6.	Al-Ralph
    7.	Al-Sam
    8.	Sam-Ralph
    9.	Al-Sam
    10.	Al-Ralph
    11.	Al-Sam
    12.	Sam-Ralph
    13.	Al-Sam
    14.	Al-Ralph
    15.	Al-Sam
    16.	Sam-Ralph
    17.	Al-Sam
    18.	Al-Ralph
    19.	Al-Sam


## Problem 3

    In the following equations, each
    letter always has the same value, but
    different letters may have the same
    value, and values may be positive,
    negative, zero, integer, or fractional.
    Given that 
    
        U+N=1
        D+E+U+X=2
        T+R+O+I+S=3
        Q+U+A+T+R+E=4
        C+I+N+Q=5
        S+I+X=6
        S+E+P+T=7
        H+U+I+T=8
        N+E+U+F=9
        D+I+X=10
        O+N+Z+E=11
        D+O+U+Z+E=12
        T+R+E+I+Z+E=13
        Q+U+A+T+O+R+Z+E=14
        Q+U+I+N+Z+E=15,
    
    what is the value of S+A+I+N+T+T+R+O+P+E+Z?

    -- C.W. Haigh in New Scientist

Solution: solved in the cryptic arithmetic notebook

    Solution set:
    
    u   4.5
    n  -3.5
    d  -7.0
    e   4.5
    x   0.0
    t -13.5
    r   0.5
    o  10.0
    i  17.0
    s -11.0
    q  -7.5
    a  15.5
    c  -1.0
    p  27.0
    h   0.0
    f   3.5
    z   0.0
    
    The desired sum: SAINTTROPEZ = 33


## Problem 4

    On my last birthday, I received
    a present packed in a rectangular
    box, 10 cm by 20 cm by 30 cm. The
    box was wrapped by an inelastic
    continuous loop of very narrow ribbon
    that crossed all six faces at least
    once, held in place with transparent
    tape, with the total length of ribbon
    on each face being at least 1 cm.
    What is the shortest ribbon that
    meets these requirements? Express
    your answer accurate to three decimal
    places.

    -- James M. Garnett III, MS '65

Solution:

    Seems like the shortest piece would be
    a loop that just looped over the top of
    opposite corners, ran at an angle down each
    side, and make long angular runs along the
    bottom.
    
    Need a diagram to figure this out and then
    some calculus to minimize.

    The Bent hand-waves the minimization step and
    just asserts the shortest distance possible.
    

## Problem 5

    Five friends, Andrew, Bernie,
    Cliff, Don, and Ed, each has a son
    and a daughter. Their families are
    so close that each friend's daughter
    is married to the son of one of his
    friends. As a result, the daughterin-
    law of the father of Andrew's sonin-
    law is the sister-in-law of Bernie's
    daughter. And, the son-in-law of
    the father of Cliff's daughter-in-law
    is the brother-in-law of Don's son.
    Furthermore, the daughter-in-law
    of the father of Bernie's daughter-inlaw
    has the same mother-in-law as
    the son-in-law of the father of Don's
    son-in-law, and the son-in-law of the
    father of Ed's son-in-law has the
    same father-in-law as the daughterin-
    law of the father of Cliff's daughter-
    in-law. However, the situation is
    simplified by the fact that no daughter-
    in-law is the sister-in-law of the
    daughter of her father-in-law. Who is
    married to whom?

    -- 100 Games of Logic by Pierre Berloquin


Solution: exhaustive search for all the combinations
and then check for contradictions.  Just reading this
problem hurts my head.  **I give up.**

    According to the Bent solution, there are five
    statements.
    
    S1 implies something about A's daughter:
    either A's daughter is married to B's son or A's son
    is married to B's daughter.
    
    S2 means C's son married to D's daughter or
    C's daughter married to D's son.
    
    This leaves 8 possibilities
    
    S3 means B's daughter-in-law's brother is D's
    son-in-law -> eliminates 5 of the possibilities.
    
    S4 means C's daughter-in-law's brother is E's
    son-in-law -> eliminates 2 more.
    
    The only remaining possibility is:
    
    A's son is married to B's daughter
    B's son is married to C's daughter
    C's son is married to D's daughter
    D's son is married to E's daughter
    E's son is married to A's daughter


## Bonus

    A mathematics professor
    smokes a pipe. He carries two identical
    matchboxes, originally containing
    20 matches each. When he lights
    his pipe, he selects a matchbox at
    random and lights his pipe with one
    match and discards the used match.
    There will come an occasion when
    he first selects a matchbox with only
    one match in it. At this point, what is
    the expected number of matches in
    the other box? Express your answer
    accurate to three decimal places.

    -- Adapted from An Introduction
    to Probability Theory and
    Its Applications by William Feller

Solution: I think this is the same as the coffee ticket
problem from the Riddler.


    The problem is similar to flipping a coin.  We can
    calculate the probabilities of flipping a coin and
    having it come up H 20 times in a row -- this would
    be like picking every match out of the same box and
    ending up with a box with only 1 match.  This
    probability is:
    
    0.5^20 = 0.5^20
    
    Now, consider flipping 20 H and one T.  The T can
    not be in the final position.  The total number of
    items is then 20+1-1, and there is one T so the
    total probability is:
    
    Pr(1) = choose(20,1)*0.5^20*(1-0.5)^1 = 20*0.5^21
    
    In general, picking n tails in a row:
    
    Pr(n) = choose(20+n-1,n)*0.5^20*(1-0.5)^n
          = choose(19+n,n)*0.5^(20+n)
    
    The mean value is then: sum(n*Pr(n),n=0 to 19)
    
    This could occur for either box so we need to double the probability.
    
    Finally, this is the mean number picked from the other box so we need
    to subtract this from 20 to get the number left over:
    
```{r}
n=0:19
20-2*sum(n*choose(19+n,n)*0.5^(20+n))
```

Conclude: we expect 5.014828 matches left in the other box.

Can we get this by Monte-Carlo?

```{r}
n_trials=100000

zzz=rep(0,n_trials)
for (trial in 1:n_trials) {
    n_box1=20
    n_box2=20
    while(T) {
        s=sample(2,1)
        if (s==1) {
            if (n_box1==1) {
                n=n_box2
                break
            }
            n_box1=n_box1-1
        } else {
            if (n_box2==1) {
                n=n_box1
                break
            }
            n_box2=n_box2-1
        }
    }
    zzz[trial]=n
}

print(paste0("Mean remaining = ",mean(zzz),collapse=""))
```

Ideal result = 5.014828

Monte-Carlo results, 1e5 trials per test:

    5.00191
    5.01455
    5.0147
    4.99383
    5.01945

Which are all in the same range as the analytical solution.

The distribution is far from gaussian:

```{r}
hist(zzz)
abline(v=5.014828,col='red',lwd=3)
grid()
```

## Computer Bonus

    Solve the following
    cryptic multiplication problem.

    GEORGE = HW * BUSH

    -- H.G. McIlvried III, PA Gamma'53

Solution: Somehow missed this in cryptic arithmetic notebook.
Not really sure why this is a computer bonus but I will use
the computer in any event:

```{r}
source('rgc_lib.R')

a=make_crypt_df('georgehwbush')

x_george=make_num(a,'george')
x_hw=make_num(a,'hw')
x_bush=make_num(a,'bush')
k=(x_george == x_hw*x_bush) & (a$g != 0) & (a$h !=0) & (a$b !=0)

print(paste0("Number of solutions = ",sum(k),collapse=""))
print(t(a[k,]))
print_num(a,'george',k)
print_num(a,'hw',k)
print_num(a,'bush',k)
```

Solution:

    george = 764076
    hw = 82
    bush = 9318
