---
title: "Bent Summer 2019"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Summer 2019


## Problem 1

    After a church service, a priest and a cantor
    had a conversation.  The priest asked, "Do
    you know the ages of the three visitors we
    had today?" The cantor said, "I don't know."
    The priest said, "What if I told you that the
    product of their ages is 2652?" The cantor
    did some calculations and then replied, "I
    still don't know." The priest then leans in
    and whispers another hint into the cantor's
    ear. The cantor responded, "I still don't
    have enough information." The priest said,
    "By the way, thank you for that book of Brain
    Ticklers you gave me for my birthday. I see
    by the card you sent me that you know how old
    I am. I wonder if those three visitors will
    still enjoy challenging their minds when they
    reach my age." The cantor then said, "Oh! Now
    I know their ages and anyone listening to us
    should still be able to determine your age."
    Assuming that all ages are integral, how old
    is the priest?

    -- Grant L. Bilker, son of Scott L. Bilker,
       NJ G'91

Solution:  I still don't get this.

    I have read (and reread) the Bent solution and it makes
    no sense.  The key to their argument is what hint did
    the Priest tell the Cantor and the fact that it still
    lead to ambiguity.
    
    Since we don't know what was said it could just as
    readily been something like: "The youngest visitor was 1".
    There are 12 age combinations that meet this "hint".
    
    They make some magical pronouncement about a subset
    with cardinality > 1 ... blah, blah, blah but it is
    all nonsense.


```{r}

# Generate all feasible age combinations
ages=c()
for (i1 in 1:2652) {
    for (i2 in i1:2652) {
        if (2652 %% (i1*i2) == 0) {
            i3=2652/(i1*i2)
            if (i3 >= i2) {
                ages=rbind(ages,c(i1,i2,i3))
            }
        }
    }
}
ages=data.frame(unique(ages))
colnames(ages)=c("Visitor1","Visitor2","Visitor3")
ages$priest=apply(ages,1,function(x) {max(x)+1})

print(sprintf("Number of possible age combinations = %d",nrow(ages)))

print(as.matrix(ages))
```

    I submit that, as long as the priest can be very old,
    that any of these combinations solves the problem.


## Problem 2

    You are playing a game where you toss a
    circular disk randomly on a floor made up of
    square tiles with sides of unit length. You
    win if your disk lands so as to cover parts
    of at least two tiles, as long as it doesn't
    cover a corner of a tile. You get to pick the
    diameter of the disk.  What is the maximum
    probability of winning and what diameter disk
    achieves this?

    -- Brian Barwell and Richard Hess, CA B'62,
       in Mental Gymnastics

Solution:

    I remember this problem -- need to dig up my solution.

    See directory c:\rgc\test\play\bent\2019\2_summer
    for files involving circular_disk and/or coin_flip.

    The solution is:

    d   = 4/(4+pi) ~ 0.56 units
    Likelihood of a win is also 4/(pi+4) ~ 0.56

    Given:

    Disk of diameter=d, radius=r

    Winning area is perimeter of unit square with area along each edge
    excluding the corners.  The area along the edge is r*(1-2*r).  The area
    in the corner is what is left of the square of r by r when a circular
    arc of the quarter circle is subtracted (loss if you hit a corner).
    This occurs along all four edges.  The result is:

    winner  = 4*( r*(1-2*r) + (r*r - pi*r*r/4) )
            = 4*r*(1-2*r)+4*r*r-pi*r*r
            = r*(4*(1-2*r)+4*r-pi*r)
            = r*(4 - 4*r - pi*r)

    The value of winner is maximized when diff(winner,r)=0 is solved.

    diff(winner,r) = 4-8*r-2*pi*r

    so

    4-8*r-2*pi*r = 0

    or

    r   = 4/(8+2*pi)
        = 2/(4+pi)

    thus, the best result is obtained with:
        d   = 2*r = 4/(4+pi) ~ 0.56 units
    Likelihood of a win is also 4/(pi+4) ~ 0.56

    In other words, slightly more than half the tosses will result in a win.
    
    This diameter maximizes the likelihood of winning.  We know this is a
    maximum since diff(diff(winner,r),r) = -8-2 pi < 0.

    Note that this formula breaks down for r>0.5.


## Problem 3

    How many ways are there of walking up a
    flight of 13 stairs if you take either 1 or 3
    stairs with each step?

    -- Eustice Solves a Problem

Solution:

    Let's try to work up to this:
    
    1 stair:  1 way
        1
    
    2 stairs: 1 way
        1+1
    
    3 stairs: 2 ways
        1+1+1
        3
        
    4 stairs: 3 ways
        1+1+1+1
        1+3
        3+1

    5 stairs: 4 ways
        1+1+1+1+1
        1+1+3
        1+3+1
        3+1+1
        
    6 stairs: 6 ways
        1+1+1+1+1+1
        1+1+1+3
        1+1+3+1
        1+3+1+1
        3+1+1+1
        3+3

    For n stairs:
        1+(solutions for n-1 stairs)
        3+(solutions for n-3 stairs)

    Let s(n) = number of solutions for n stairs
    
        s(n) = s(n-1)+s(n-3)
        
    s(1)=1, s(2)=1, s(3)=2, then
    s(4)=s(3)+s(1)=3
    s(5)=s(4)+s(2)
    etc.
    
    Can also solve the recurrence relationship:
    
    1-z^-1-z^-3=0 --> z^3-z^2-1=0 but the roots are ugly
    
    Easier to do this numerically:
    
```{r}

s=c(1,1,2)
for (n in 4:13) {
    m=s[3]+s[1]
    print(c(n,m))
    s[1]=s[2]
    s[2]=s[3]
    s[3]=m
}
```
    
The solution for 13 stairs is 88 ways.    


## Problem 4

    Jan has 27 wooden 1-inch cubes whose faces
    have been painted with red, white, and blue
    paint. Using these, she can assemble a 3-inch
    cube that is red on all sides. She can
    rearrange them to form a 3-inch cube that is
    white on all sides. She can also arrange them
    to form a 3-inch cube that is blue on all
    sides.  How many of the 1-inch cubes have 2
    red sides, 2 white sides, and 2 blue sides?

    -- Tom Stluka, brother of Frederick Stluka,
       PA Delta'82

TODO

    meh


## Problem 5

    In the poker game of showdown, each player is
    dealt one card face up from a well shuffled
    standard deck of 52 cards and the highest
    card wins. If there are five players, exactly
    which card would give a player just over a
    50% chance of winning the pot? Assume the
    hierarchy of suits in descending order is
    spades, hearts, diamonds and clubs.

    -- Don Dechman, TX A'57

Solution:

    This also seems familiar.  I recall doing this
    with Monte Carlo.

    The Bent does it with Combinatorics.
    
    Pick the Nth ranked card such that the probability
    there are 4 cards lower is greater than 50%.
    
    If you get card number N there are N-1 cards below
    it and there are 51 total cards to draw from.  The
    four other cards drawn should satisfy:
    
        C(N-1,4)/C(51,4) > 0.5
        
    40   0.329136
    41   0.365706
    42   0.405242
    43   0.447899
    44   0.493838
    45   0.543221 <- Smallest N for p>0.5
    46   0.596218
    47   0.653001
    48   0.713745
    49   0.778631
    50   0.847843
    
    The card rankings are:
        1-4 = 2
        5-8 = 3
        etc.
        37-40 = J
        41-44 = Q
        45-48 = K
        49-52 = A

    So the card ranked 45 is the lowest K = K of clubs.


## Bonus

    Three logicians, A, B, and C, each wears a
    hat with a positive integer on it. Each
    logician sees the numbers on the other hats,
    but not their own. Each knows the numbers are
    positive integers and that one is the sum of
    the other two. They take turns in order,
    starting with A, to identify their number, or
    pass if they can not. A, B, and C all pass on
    their turn in each of the first two rounds.
    In the third round, A and B pass, but C
    correctly asserts her number is 89. What are
    the other two numbers? Provide a quick
    outline of C's logic used to determine her
    number was 89.

    -- Donald Aucamp, Puzzle Corner in Technology
       Review

TODO

    I remember this but I don't recall the trick.
    After reading the solution I still don't understand
    the trick.


## Computer Bonus

    Consider the sequence of rational fractions,
    3/1, 13/4, 16/5, 19/6, 22/7, 179/57, 201/64,
    223/71, 245/78, 267/85, 289/92, 311/99,
    333/106, 355/113, 52163/16604,...  Pi/Qi...,
    of successively closer approximations to p.
    In this sequence, each entry has a smaller
    error than any other entry with a smaller
    value of Qi. A ratio Pi/Qi is a member of the
    sequence if, and only if, it meets this
    criterion.  Thus, this sequence includes all
    the convergents resulting from expressing p
    as a continued fraction, but many other
    values as well. Find the smallest Pi such
    that its units digit is 0, the units digit of
    Pi+1 is 1, the units digit of Pi+2 is 2, etc.
    up to the units digit of Pi+7 is 7. Present
    your answer as the first value of this
    series.

    -- Timothy J. Slegel, PA A'80

TODO

    Still don't understand this.
