---
title: "Bent Winter 2012"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Winter 2012


## Problem 1

    How many primes appear in the following
    sequence: 9; 98; 987; 9,876; ...;
    9,876,543,210; 98,765,432,109;
    987,654,321,098; and so on?

    -- Daryl Cooper

Solution:

    None.
    
    The only ones that can be prime are the ones that
    end in 9, 7, 3, or 1.
    
    9           SOD = 9  --> divisible by 3
    987         SOD = 24 --> divisible by 3
    9876543     SOD = 42 --> divisible by 3
    987654321   SOD = 45 --> divisible by 3
    
    All the other numbers are prepended by 9876543210 which
    has a SOD = 45 so these are also divisible by 3.


## Problem 2

    Find two three-digit integers that sum to a
    four-digit integer such that each of the
    digits 0 through 9 is used, with no leading
    zeros. What are the maximum and minimum
    values for the four-digit integer?

    -- The Playful Brain by
    Richard Restak and Scott Kim

TODO

    Doesn't seem like much fun.


## Problem 3

    Bill has a problem. He has locked his Tau
    Beta Pi pin in a strong box and forgotten the
    combination. The lock uses all the digits 1
    through 6 in some order. He has tried the
    following three combinations: 4-5-6-1-3-2
    which has one digit correct; 6-2-3-4-5-1
    which has two digits correct; and 2-3-1-6-4-5
    which has three digits correct. What is the
    combination?

    -- The Everything Brain Strain Book by Jake Olefsky

TODO

    meh - logic.

## Problem 4

    Mac is idly tossing two dice when he decides
    to see how many tosses it would take to go
    from 1 to 6 in order. The rules are to toss
    the two dice until one or the other or both
    show a 1. Then, toss until a 2 shows.
    However, if both a 1 and a 2 show at the same
    time, both can be used. He then tosses for a
    3 and similarly for other numbers. What is
    the expected number of tosses to go from 1 to
    6? Once you have the answer, get a couple of
    dice and play the game a few times to see how
    closely the average matches your answer.

    -- Howard G. McIlvried III, PA Gamma'53

Solution:

    This seems to be begging for Monte Carlo
    
```{r}
ntrials=100000
n=0

set.seed(42)
for (trial in 1:ntrials) {
    target=1
    while(T) {
        s=sample(6,2,replace=T)
        n=n+1
        # if (trial == 1) { print(c(n,target,s)) }
        if (target %in% s) {
            target = target+1
            if (target %in% s) {
                target = target+1
            }
        }
        if (target>6) { break }
    }
}

print(paste0("Number of rolls = ",n))
print(paste0("E(n) = ",n/ntrials))
```

My solution did not agree with the Bent.  OOOps - my sampling
was not correct -- needed replace=T.  After making the correction
things seem to match:

    For set.seed(42)   with ntrials = 100K, E(n) = 17.04754
    For set.seed(123)  with ntrials = 100K, E(n) = 16.99581
    For set.seed(2020) with ntrials = 100K, E(n) = 17.01562
    
    The Bent has 17.0414 using combinatorics
    
I will try one sim of 1e6 trials to see the result:

    For set.seed(42)   with ntrials = 1000K, E(n) = 17.036943

Well - this is close enough to their answer that they seem
equivalent.


## Problem 5

    A distinct positive integer has been assigned
    to each letter of the alphabet such that the
    letters of PLUTO add to 40; URANUS 36;
    NEPTUNE 29; SATURN 33; JUPITER 50; MARS 32;
    Earth 31; MOON 36; VENUS 39; MERCURY 33; and
    SUN 18. What is the value of PLANETS? Get a
    pencil and paper and solve this one without
    computer help!

    -- C. W. Haigh in New Scientist

TODO

    Tried this before and couldn't solve it.
    

## Bonus

    Start with a checker in the center of a 5x5
    checkerboard. The objective is to move the
    checker from square to square so that each
    square is visited exactly once, except the
    center square to which you must return on
    your 25th move. The only legal moves are to a
    square that is two squares away diagonally or
    to a square that is three squares away
    horizontally or vertically. How many unique
    paths can the checker trace? Rotations,
    reflections, path reversals, and combinations
    thereof, are not considered different paths.
    This problem can be solved without computer
    help.

    -- William W. Verkuilen, WI B'92

TODO

    Don't care
    

## Computer Bonus

    Beginning with any multi-digit number, N,
    start a sequence by writing its digits in
    order, starting with the left-most digit. The
    next number in the sequence is the sum of the
    digits of N. Continue with a Fibonacci-like
    process, where the next term in the series is
    the sum of the previous n terms, where n is
    the number of digits in N. If N appears in
    this sequence, then N is a Keith number,
    named after Michael Keith, who discovered
    them. For example, 197 is a Keith number
    because 1, 9, 7, 17, 33, 57, 107, 197, ...
    contains N. The four smallest Keith numbers
    that are also prime numbers are 19, 47, 61,
    and 197. What is the next smallest prime
    Keith number?

    -- Fred J. Tydeman, CA Delta'73

Solution:

    Exhaustive search ... this takes a little while to run.
    Suppressed evaluation for html version.
    
```{r eval=FALSE}
library(gtools)
library(gmp)

is_keith = function(N) {
    x=baseOf(N)
    n=length(x)
    while(TRUE) {
        m=sum(x)
        if (m==N) { return(T) }
        if (m>N)  { return(F) }
        x[1:(n-1)]=x[2:n]
        x[n]=m
    }
}

N=197
while(T) {
    N=nextprime(N)
    if (is_keith(N)) {
        print(paste0("Prime Keith = ",N))
        break
    }
}


```

The next smallest Prime Keith = 1084051

