---
title: "Bent Summer 2003"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Summer 2003


## Problem 1

    Two park rangers at a Civil War
    battle site are assigned the job of
    counting a pile of cannonballs
    arranged in the shape of a regular
    tetrahedron. The first ranger says,
    "We will have to disassemble the
    pile to count the cannonballs in the
    middle." The second ranger, a Tau
    Bate, replies, "That won't be
    necessary. Just tell me the number n
    of balls along one side of the bottom
    layer, and I can calculate the total
    number in the stack." What closedform
    formula did he use?
    
    -- John R. Sellars

Solution: (seems too easy)

    The number of balls on each side should
    increase by one on each layer.  The top
    row is 1, the next is 2, etc.  Need to
    calculate:
    
    summation(m**2,(m,1,n)) using sympy
    
    In [306]: summation(m**2,(m,1,n))
    Out[306]: n**3/3 + n**2/2 + n/6

    In [307]: factor(summation(m**2,(m,1,n)))
    Out[307]: n*(n + 1)*(2*n + 1)/6

Well I obviously don't know what a tetrahedron is.
I used a pyramid with 4 sides and a bottom.  They
wanted a pyramid with 3 sides and a bottom.  In that
case:

    Top row = 1
    Next    = 3
    Next    = 6
    ... triangular numbers ...
    
    So - if you give the number along one side then that is the layer number
    and the number on the layer is n*(n+1)/2
    
    So - sum these up:
    
    In [406]: summation(m*(m+1)/2,(m,1,n))
    Out[406]: n**3/6 + n**2/2 + n/3

    In [407]: factor(summation(m*(m+1)/2,(m,1,n)))
    Out[407]: n*(n + 1)*(n + 2)/6

Which was the requested result.


## Problem 2

    An urn contains ten balls, numbered
    1 through 10. Three balls are
    drawn at random without replacement.
    What is the probability that
    the values on the three balls can be
    sides of the same triangle?
    
    -- Unknown

Solution: Monte-Carlo, Permutations, analytical

Given sides = a,b,c

Can form a triangle iff:

    a+b > c, and
    a+c > b, and
    b+c > a

Look at Monte-Carlo first:

```{r}
num_trials=1000000

n=0
for (i in 1:num_trials) {
    s=sample(1:10,3)
    a=s[1]; b=s[2]; c=s[3]
    if ( (a+b)>c & (a+c)>b & (b+c)>a ) { n=n+1 }
}
n/num_trials
```

Solution by Monte-Carlo suggests probability = 0.41571.

Try this be exhaustive enumeration:

```{r}
library(gtools)

a=permutations(10,3,1:10)

k1=(a[,1]+a[,2]>a[,3])
k2=(a[,1]+a[,3]>a[,2])
k3=(a[,2]+a[,3]>a[,1])
k=k1&k2&k3

paste0("Number of possible drawings  = ",nrow(a),collapse='')
paste0("Number of feasible triangles = ",sum(k),collapse='')
paste0("Probability = ",sum(k)/nrow(a),collapse='')
```

So the exhaustive probability = 300/720 = 5/12 = 0.41666...

Finally, we try to enumerate them:

    Total possibilities = choose(10,3)*3! = 720 permutations
    Total triangle permutations = 50*3! = 300 possible triangles
    
    Details below:
    
    Note that we fail to form a triangle if:
    
        a+b>c, or
        a+c>b, or
        b+c>a

    Consider only the choose(10,3)=120 combinations.  We can figure
    out which are permissible and then multiply by 3! to account for
    permutations.  Since each value is unique, assume a < b < c
    
    Consider each case for a = 1, 2, ..., 8.  Note that a can not
    exceed 8 since b=9, c=10 will be the only combination.
    
    In general:
    
        a in range [1,8]
        b in range [a+1,9]
        c in range [b+1,min(a+b-1,10)]
    
    Enumeration of the solutions:
    
    a=1 --> no solution since c > b and c <= b.
    
    a=2, b=3, c=4  \
    a=2, b=4, c=5   | 7 combinations
    a=2, b=5, c=6   |
    ...             |
    a=8, b=9, c=10 /
    
    a=3, b=4, c=5,6  \
    a=3, b=5, c=6,7   | 11 combinations
    ...               |
    a=3, b=8, c=9,10  |
    a=3, b=9, c=10   /
    
    a=4, b=5, c=6,7,8   \
    a=4, b=6, c=7,8,9    | 12 combinations
    a=4, b=7, c=8,9,10   |
    a=4, b=8, c=9,10     |
    a=4, b=9, c=10      /

    a=5, b=6, c=7,8,9,10 \
    a=5, b=7, c=8,9,10    | 10 combinations
    a=5, b=8, c=9,10      |
    a=5, b=9, c=10       /

    a=6, b=7, c=8,9,10   \
    a=6, b=8, c=9,10      | 6 combinations
    a=6, b=9, c=10       /
    
    a=7, b=8, c=9,10     \  3 combinations
    a=7, b=9, c=10       /
    
    a=8, b=9, c=10       - 1 combination

    Total number of combinations = 7+11+12+10+6+3+1 = 50 combinations
    Total triangle permutations = 50*3! = 300 possible triangles
    

## Problem 3

    Solve the following cryptics. Each
    letter has the same value in both
    cryptics, and each letter also
    represents a different digit.
    FIVE - FOUR = ONE
    FIVE + FIVE = EVEN
    
    -- M. Yagi, CA I'96

Solution: solved in the cryptic arithmetic notebook

    FIVE = 2374
    FOUR = 2190
    ONE  = 184
    EVEN = 4748


## Problem 4

    Given a one-meter cube, drill a
    hole of diameter 0.25m with its axis
    along a diagonal of the cube. What is
    the volume of the material removed?
    
    -- SW Missouri State U. Math. Dept.

Solution: this seems hard.  A lot like the water pipe through the
    manhole but in three dimensions.  If we ignore the slight distortion
    on the ends this is essentially a cylinder of the specified diameter
    with a length equal to the diagonal of the cube.
    
    Cube diagonal length = sqrt(3)
    cylinder radius = 0.25/2
    
    Approximate volume = L * pi*r**2
        = sqrt(3)*pi*0.125**2
        = 0.085 m^3

    To get a precise solution I think you need to view this in a plane
    parallel to the diagonal of the cube -- one that bisects the cube and
    the drill hole.  That would make this problem a lot like
    the water pipe/manhole problem and it would be "simple" to make
    the calculation.
    
    I am not very good with 3D geometry.

    Their solution is very complex (IMHO) and yields V = 0.0754 m^3 after
    subtracting out the corners of the cube.


## Problem 5

    If the Earth's movement
    matched the Gregorian calendar
    perfectly, how many degrees per
    day would the Earth rotate relative
    to the "fixed" stars? Give the
    answer exactly.
    
    -- Byron R. Adams, TX A'58

TODO

I don't even understand this problem.  Even after reading their
solution I don't understand what is going on.

Here is their analysis:

    Length of a year = 365+97/400 days (97 leap years in 400?)
    
    Number of rotations in a year = 1+365+97/400
    
    Rotations per day = (1+365+97/400)/(365+97/400)
    
    deg per day = 360*Rotations
                = 360*(1+365+97/400)/(365+97/400)
                = 5859880/16233
                ~ 360.9856465225158


## Bonus

    This is a variation of a
    problem that appeared in the Winter
    2003 column. A package-handling
    company will ship a package with a
    length plus girth of up to 108 inches.
    A customer wishes to mail a rightcircular
    cone of maximum volume
    consistent with the given limitation.
    If the girth is defined as the shortest
    loop of string through which the
    cone will pass and the length is
    defined as the maximum dimension
    of the cone, what are the altitude
    and base diameters of the cone?
    
    -- R. Wilson Rowland, MD B'51

Solution:

    I almost had this -- they say the girth around the cone is the
    same a flattening the cone into a pizza slice (not a simple triangle).
    They solve the resulting equations by trial and error to get:
    
        See their solution.  Need to work through this in detail
        to understand the girth values they are using.
    
    I am not sure that you can actually flatten a cone without distorting
    the surface of the cone.
    
    My attempt:

    The girth could be the circumference of the circular base
    or it could be the size of the eqilateral triangle composed
    of the diameter of the base and the two edges of the cone.
    
    For a very short cone the circumference of the base will
    not define the girth.  It will be shorter to put the cone
    through the loop "sideways".  For a taller cone, however, the girth
    will be defined by the circumference of the base.
    
    The critical height is: h=r*sqrt(pi/(pi-2)) where r=radius of base
    
    The length of the cone is the longest dimension.  For the very
    short cones the length will actually be defined by the diameter
    of the base.  For a long "skinny" cone the longest dimension will
    be the length of the side (sqrt(h**2+r**2)).  The critical
    height is: h=r*sqrt(3) where r=radius of base.
    
    Interestingly, these two critical values are very close to each other.
    
    So:
    
    h < r*sqrt(pi/(pi-2)) --> girth = 2*r+sqrt(h^2 + r^2) else 2*pi*r
    h < r*sqrt(3)         --> length = 2*r else sqrt(h^2 + r^2)
    
    Thus:
    
    h < r*sqrt(pi/(pi-2))  --> girth = 2*r+sqrt(h^2+r^2), length = 2*r
    h > r*sqrt(3)          --> girth = 2*pi*r, length=sqrt(h^2 + r^2)
    otherwise              --> girth = 2*pi*r, length=2*r

    This is very ugly -- there must be a better way.
    
    What about Monte-Carlo?
    
    
## Double Bonus

    What are the
    smallest and largest perfect squares
    that use each of the digits 0 to 9 once
    and only once. A leading 0 is not
    allowed.
    
    -- Technology Review


Solution: exhaustive search (awfully easy)

    Range is 32043 to 99066
    
    32043^2 = 1026753849
    99066^2 = 9814072356
    
    FYI there are 87 squares using all 10 digits exactly once.

```{r}
n1=trunc(sqrt(1e9))
n2=trunc(sqrt(1e10))

zzz=c()
for (i in n1:n2) {
    n=length(unique(strsplit(as.character(i^2),'')[[1]]))
    if (n==10) {
        # print(i)
        zzz=c(zzz,i)
    }
}
range(zzz)
```

