---
title: "Bent Summer 2018"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Summer 2018

    These problems look very familiar.  I would suspect I have solved
    them before.


## Problem 1

    A man and his daughter make a 64 km trip with
    their saddle horse which travels 16 km per
    hour but can carry only one person at a time.
    The man walks at the rate of 6 km per hour
    and the daughter at the rate of 8 km per
    hour. The trip is made in the following
    manner. The two start out at the same place
    and time with one walking and the other
    riding. After the rider has gone a certain
    distance, the rider dismounts, ties the
    horse, and immediately starts walking ahead.
    When the walker comes up to the horse, the
    walker mounts and rides forward another
    certain distance, whereupon they dismount,
    and the above process is repeated. The two
    reach the halfway point at the same time, at
    which point they take a halfhour break to
    feed the horse. They then proceed in the same
    manner as in the first half of the journey.
    At what time does the pair arrive at their
    destination if they start at 6 a.m.?

    -- Ingenious MathematicalProblems
       and Methods by L. A. Graham

Solution:

    tgw=time girl walks, rate = 8 km/hr
    tgr=time girl rides, rate = 16 km/hr
    tfw=time father walks, rate = 6 km/hr
    tfr=time father rides, rate = 16 km/hr
    
    [1] tgw*8 + tgr*16 = 32 --> girl completes first half of trip
    [2] tfw*6 + tfr*16 = 32 --> father completes first half of the trip
    [3] tgw+tgr = tfw+tfr   --> they arrive at same time

    Need one more equation???
    
    Can I make the simplifcation that there is only one exchange?
    I don't think it makes any difference.
    
    d = distance for the exchange
    
    Father: t = d/16 + (32-d)/6
    Girl:   t = d/8  + (32-d)/16
    
    These times must be the same so: d = 20
    
    tfr = 20/16, tfw=12/6 --> t = 20/16+12/6 = 13/4
    tgw = 20/8,  tgr=12/16 --> t = 13/4
    
    First half journey takes 3.25 hours
    Break for 0.5 hours
    Second half takes 3.25 hours
    
    Total time = 7 hours
    
    Start at 6 am then should finish at 1pm


## Problem 2

    If a decimal integer ends in a particular
    digit, repeated enough times, the hexadecimal
    representation of the number will end with
    either 1 or 3 repeated hex digits. For
    example, a decimal number ending in a string
    of 1's will produce a hexadecimal number
    ending in the repeated digits "1C7," and a
    decimal number ending in a string of 6's will
    produce a hexadecimal number ending with the
    repeated digit "A" (A stands for the value
    10). How long is the smallest decimal integer
    consisting entirely of 6's which has a
    hexadecimal representation that ends with a
    string of 25 consecutive A's?

    -- Franklin J. Gulian, DE A'83

TODO

    Last time I brute forced it with Sympy.


## Problem 3

    Twice the larger of two numbers is three more
    than five times the smaller and the sum of
    four times the larger and three times the
    smaller is 71. What are the numbers?

    -- Mean Girls, 2004

Solution:

    n1 = larger number
    n2 = smaller number
    
    2*n1        = 5*n2+3 
    4*n1 + 3*n2 = 71

    Solving: n1 = 14, n2 = 5
    
    Seems awfully easy.


## Problem 4

    Mary stood beside a large pile of turnips
    (fewer than 20,150), which she was to
    distribute evenly among the people in the
    group in front of her. Since it was unlikely
    that the number of turnips would divide
    evenly among the people, she was given
    permission to add or subtract turnips to
    produce an even division to the nearest
    integral value.  A quick division of the
    number of turnips by the number of people
    gave an answer between 99 and 100, but closer
    to 99. However, realizing that everyone would
    prefer 100 turnips to 99, she decided to do
    the calculations a different way. First, she
    assumed everyone would get 99 turnips, so she
    divided the number of turnips by 99. Then,
    she assumed everyone would get 100 turnips,
    so she divided the number of turnips by 100
    and found that this gave an answer closer to
    the actual number of people than the division
    by 99, so she added some turnips to the pile
    and gave everyone 100 turnips.  How many
    people were there and how many turnips were
    in the original pile?

    -- Keith Austin in New Scientist

Solution:

    t = number of turnips
    p = number of people

    t < 20150
    99 < t/p < 99.5
    abs(t/99)-p > abs(t/100)-p

    Find, p and t

    Clearly, the maximum number of people would be for maximum
    turnips at the lowest qotient:
    
    p<t/99 = 20150/99 = 203.5
    
    At the other extreme we really don't know anything.
    Let us sweep p from 1 to 203 and see what we get for t:
    
```{r}
p=1:203
tmin=p*99
tmax=p*99.5
# if tmax*99.5 is an integer then it is half-way between
# 99 and 100.  The problem says it is closer to 99 so need
# to reduce it by 1.  In the Bent they just dismiss all
# even values for p since it wouldn't meet the strictly
# closer to 99 than 100 constraint -- I don't agree.
tmax=ifelse(tmax==trunc(tmax),tmax-1,trunc(tmax))
# Limit the maximum number of turnips
tmax=ifelse(tmax>20150,20150,tmax)

# These are all the feasible solutions
a=data.frame(p=p,tmin=tmin,tmax=tmax)

# Find the solution that meets the div by 99 and div by 100 test
k=apply(a,1,function(x) { t=seq(x[2],x[3]); any(abs((t/99)-x[1]) > abs((t/100)-x[1]))})

print(a[k,])

```

    The only solution is:
    
        p = 201 people
        t = 19999 turnips
        
        Need to add 101 turnips


## Problem 5

    My uncle's ritual for dressing each morning
    except Sunday includes a trip to the sock
    drawer, where he (1) picks out three socks at
    random, then (2) wears any matching pair and
    returns the odd sock to the drawer or (3) if
    he has no matching pair, returns the three
    socks to the drawer and repeats steps (1) and
    (3) until he completes step (2). The drawer
    starts with 16 socks each Monday morning
    (eight blue, six black, two brown) and ends
    up with four socks each Saturday evening.  On
    which day of the week is he least likely to
    get a pair from the first three socks chosen?
    On average, which day of the week requires
    the greatest number of times that my uncle
    grabs three socks from the drawer?

    -- Richard Hess, CA B'62, in The
       Mathemagician and Pied Puzzler

Solution:  Let us try Monte-Carlo

```{r}
ntrials=100000

total=rep(0,6)
firstmiss=rep(0,6)

set.seed(42)

for (trial in 1:ntrials) {
    
    # Load up the sock drawer
    # Encode the sock colors as 0=blue, 1=black, 2=brown
    drawer=c(rep(0,8),rep(1,6),rep(2,2))

    for (dow in 1:6) {
        n=1
        while (T) {
            s=sample(length(drawer),3)
            if (drawer[s[1]]==drawer[s[2]]) {
                drawer=drawer[-c(s[1],s[2])]
                total[dow]=total[dow]+n
                break
            } else if (drawer[s[1]]==drawer[s[3]]) {
                drawer=drawer[-c(s[1],s[3])]
                total[dow]=total[dow]+n
                break
            } else if (drawer[s[2]]==drawer[s[3]]) {
                drawer=drawer[-c(s[2],s[3])]
                total[dow]=total[dow]+n
                break
            }
            if (n==1) { firstmiss[dow]=firstmiss[dow]+1 }
            n=n+1
        }
    }
}

print("P(miss/match on first try):")
print(firstmiss/ntrials)
print(1-firstmiss/ntrials)
print("Mean number of tries:")
print(total/ntrials)

```

    Summary of results with 100K random trials
    
    Pr(miss on first try)
    
    Seed  Mon     Tues    Wed     Thurs   Fri     Sat
    -----------------------------------------------------
    1     0.17218 0.18419 0.19917 0.21069 0.21925 0.21386
    42    0.17218 0.18357 0.19750 0.21237 0.21860 0.21427
    123   0.16982 0.18208 0.19763 0.21169 0.22025 0.21791
    2020  0.17009 0.18171 0.19807 0.21034 0.21872 0.21630
    
    It is subtle but Friday appears to be the most likely day to
    miss on the first try at getting a pair.

    Mean(number of tries)    
    
    Seed  Mon     Tues    Wed     Thurs   Fri     Sat
    -----------------------------------------------------
    1     1.20687 1.22919 1.25365 1.27908 1.30813 1.35661
    42    1.20714 1.22686 1.24983 1.28107 1.30655 1.35774
    123   1.20488 1.22378 1.24980 1.28038 1.30880 1.36255
    2020  1.20548 1.22471 1.25185 1.28023 1.30399 1.36050
    
    Saturday requires (on average) the most number of tries.

    A single run of 1e6 trials performed with seed = 42

    Pr(miss on fist try):
    
    Seed  Mon      Tues     Wed      Thurs    Fri      Sat
    -----------------------------------------------------------
    42    0.171385 0.184345 0.197512 0.210901 0.219584 0.214613 (miss)
          0.828615 0.815655 0.802488 0.789099 0.780416 0.785387 (match)

    Mean(number of tries):
    
    Seed  Mon      Tues     Wed      Thurs    Fri      Sat
    -----------------------------------------------------------
    42    1.206727 1.227657 1.251155 1.279814 1.307513 1.357621

    The Bent used combinatorics and provided the following to 4 digits.
    
    Day   Match    Expected
    -----------------------
    Mon   0.8286   1.2069
    Tue   0.8162   1.2270
    Wed   0.8027   1.2511
    Thu   0.7892   1.2801
    Fri   0.7805   1.3074
    Sat   0.7849   1.3586

    The match is spectacularly close considering how subtle the difference
    is between the match probabilities.


## Bonus

    Suppose you are given four non-standard but
    otherwise fair, dice: one blue, one green,
    one red, and one white. For simplicity,
    assume that the four dice have the numbers 1
    through 24 on their faces. After a long
    sequence of rolling pairs of these dice, you
    conclude the following: When rolled
    simultaneously, two thirds of the time the
    blue die shows a higher value than the green
    die. When rolled simultaneously, two thirds
    of the time the green die shows a higher
    value than the red die. When rolled
    simultaneously, two thirds of the time the
    red die shows a higher value than the white
    die. Now, if the blue and white dies are
    rolled simultaneously, what is the least
    probability P that the blue die will show a
    higher value than the white die on a given
    roll? For such P, give an example of the
    distribution of the numbers 1 through 24 on
    the four die.

    -- Puzzle Corner by Allan Gottlieb
       in Technology Review

TODO

    I remember working this out once upon a time.
    Don't want to do that again.
    
    The interesting thing is that it is all about
    matchups!  The solution they provide is:
    
    Blue:   6,  7,  8,  9, 23, 24
    Green:  3,  4,  5, 20, 21, 22
    Red:    1,  2, 16, 17, 18, 19
    White: 10, 11, 12, 13, 14, 15

    Blue beats Green: 24 out of 36
    Green beats Red:  24 out of 36
    Red beats White:  24 out of 36
    White beats Blue: 24 out of 36


## Double Bonus

    The game of Sprouts, invented by J.H. Conway
    and M.S. Patterson, is a pencil and paper
    game played as follows. Any number of points
    are placed on the paper, and the players take
    turns connecting two of the points. Each move
    consists of two parts: (a) drawing a line
    between two points or looping back to the
    starting point and (b) placing a new point on
    the line anywhere except at an endpoint.
    There are three rules: (1) a line cannot
    cross itself or any other line; (2) a line
    cannot pass through an existing point but
    must start and end on existing points (can be
    the same point); and (3) no point can have
    more than three lines that start or end on
    it. The player who cannot make a move loses.
    A little thought will show that the game must
    always end. In how many different
    configurations can a two-point game end?
    Isomorphic transformations, such as order of
    play, rotations, and reflections are not
    different games.

    -- The World Book of Math Power

TODO

    meh
