---
title: "Bent Summer 2001"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Summer 2001

## Problem 1

    They were one of those Tau Beta
    Pi bi-spousal couples. One day, she
    went on an errand. That evening, he
    asked her how long the errand had
    taken. Knowing he also liked puzzles,
    she responded, "When I left, the
    hands on my watch were aligned,
    with the hour hand between 8 and 9
    and the minute hand between 2 and
    3. When I returned, the hands were
    again aligned, but this time the hour
    hand was between 2 and 3, and the
    minute hand was between 8 and 9.
    You tell me how long I was gone." To
    the nearest second, how long did the
    errand take?
    
    -- Justo T. Del Puerto

Solution:

    Measure all angles in degrees, clockwise from 12.

    Let m = number of minutes past the hour
    
    The angle of the minute hand is:
    
        Minute angle = (m/60)*360
    
    When hour hand is between 8 and 9 the angle of the hour hand is:
    
        Hour angle = 240+(m/60)*30
    
    When leaving:
        240+(m/60)*30 = (m/60)*360 + 180
        2 = (m/60)*11
        m = 120/11 (between 2 and 3)
        
        Time = 8 hours + 120/11 minutes = 8.1818...
        
    When coming home (by analogy to above):
    
        Hour angle = 2/12*360 + (m/60)*30
        
        60+(m/60)*30 + 180 = (m/60)*360
        2            +  6  = (m/60)*11
        m = 60*8/11 = 480/11 (between 8 and 9)
        
        Time = 2 hours + 480/11 minutes = 2.7272...
    
    We assume she left in the morning (about 8) and returned in afternoon (about 3).
    If we use 24 hour time we get an elapsed time of:
    
    (14+480/11/60) - (8+120/11/60) = 6.5454... hours
        = 392.7272... minutes
        = 23563.64    seconds

Note: their solution is a lot shorter but it gives the same result.  They make the
observation that the hands are in opposition at 6AM and 6PM and 10 (?) times in
between.  This means that the hands are in opposition once every 12/11 hours.
From ~8AM to ~3PM there are 6 such intervals so the time is 6*12/11 = 72/11 hours.
This is the same result.
        
## Problem 2

    A palindrome is a number that
    reads the same forward and backward.
    How many integers in the range from
    1 through 10^10 are palindromic?
    
    -- Edward C. Creutz, PA Gamma'36, and
    Michael J. Creutz, CA Beta'66

Solution:

A value of 10^n has n+1 digits so 10^10 has 11 digits.  Numbers smaller
than this have 1-10 digits. So:

The integers can have 1-10 digits.  Consider each pattern where
a must be non-zero (9 possibilities), b, c, d, and x can be any digit
(e.g. 10 possibilities):

    Num digits: pattern       Number
    --------------------------------
    10 digits: abcdxxdcba     90000
    9 digits:  abcdxdcba      90000
    8 digits:  abcddcba       9000
    7 digits:  abcxcba        9000
    6 digits:  abccba         900
    5 digits:  abxba          900
    4 digits:  abba           90
    3 digits:  axa            90
    2 digits:  aa             9
    1 digit:   a              9
    -------------------------------
    TOTAL                    199998

## Problem 3

    Solve the following cryptic
    multiplication, where FLY is exactly
    divisible by GO. GO x FLY = KITES
    
    -- Don A. Dechman, TX Alpha'57

Solution: solved in the cryptic_arithmetic notebook

    GO    = 27
    FLY   = 594
    KITES = 16038


## Problem 4

    On the first day of spring on the
    level plains of Ecuador, a certain
    tribe of tall Indians celebrates the
    fact that they can see the sun for a
    longer time than a neighboring tribe
    of short Indians. In fact, how much
    longer, in seconds, is the time between
    sunrise and sunset for the tall tribe
    (whose eyes are 200 cm above the
    ground) as compared to the short
    tribe (whose eyes are 140 cm above
    the ground)?
    
    -- Byron R. Adams, TX Alpha'58

TODO

I think we need to make some assumption about the radius of the
earth and the fact that the earth is (approximately) a sphere???

I peeked at the answer: the assume a perfect sphere of radius = 6378 km.
They also consider the earth's rotation rate of 360 degrees per day.
They make a small angle approximation for cos(theta).

The answer they are looking for is: 3.556 seconds

## Problem 5

    In order to help deal with the
    changing postage rates, an efficiency
    expert tried to design a 2 x 3 pane of
    stamps in such a way that by using a
    single stamp or a connected set of
    them, all possible postage rates 1, 2, 3,
    ..., N could be met exactly. The first
    attempt yielded the arrangement
    below, which can satisfy the values
    from 1 to 32 except for 18; therefore,
    N = 17 for this arrangement. The
    maximum would be 63, of course
    (1+2+4+8+16+32), but there is no way
    to reach that total while meeting the
    requirement for all sets of stamps to
    be contiguous. Find the largest N and
    show its 2 x 3 array.

        +----+----+----+
        | 10 |  9 |  7 |
        +----+----+----+
        |  2 |  3 |  1 |
        +----+----+----+
    
    -- Adapted from New Scientist

TODO

I gave up and looked at the solution.  They indicate that
the theoretical maximum is 40 (there are 6 singles, 7 pairs,
... etc. up through 1 group of 6) but there are no feasible
solutions above 36.  Both of the following achieve this
feasible maximum:

    1 - 2 - 15
    4 - 6 - 8
    
    1 - 3 - 17
    8 - 2 - 5


## Bonus

    Riley has ten spherical beads,
    1 cm in diameter. Two are red, two are
    orange, two are yellow, two are green,
    and two are blue. The beads are to be
    strung to make a circular bracelet.
    How many different bracelets can be
    formed?
    
    -- Howard G. McIlvried III, PA Gamma'53

Note: after solving the problem to my satisfaction I looked at their solution
and they have a smaller number of unique bracelets.  They consider a palindrome
to be the same bracelet so I need to add that to the calculations.  They also
talk about two types of palindromes ???

Their solution is very complicated.  Not sure I would have ever found an analytical
solution.  The solution in the next issue of the Bent is longer than one page
(of a narrow column).

My computer solution (including palindromes) = 11352 unique combinations
Their solution = 5736 combinations

After adding the palindrome check, my Python computer solution yields:

    Number of solutions =  5736
    Solution time (secs) =  1719.3

The number of unique solutions agrees with their solution.

Fixing the Lua version took longer than I anticipated -- but it finally runs
with similar (corrected) results:

    Number of solutions = 5736
    Solution time (secs) = 297.3

Note the Lua runtime is about 5 minutes versus the nearly 3 day run time of my
attempt to solve this in R.


Solution:

I could simply enumerate all the 10-digit 5-tuples and then pick the possible
combinations (e.g. 2 of each digit).  After that, I could throw away any rows
that are shifted versions of previous rows.

This seems rather ham-fisted but it should work.

I tried it and it ran for freaking ever!  Just generating the data takes a minute.
Attempting to eliminate the shifted rows took over an hour and wasn't close to being
done.  After checking this out some more I realized it would run for more than 3 days!!!
I was able to speed it up a bit - but still more than 2 day run time.

Need to be smarter about this -- or use a different language.



```{r}
library(gtools)

t1=Sys.time()

# Generate all the digit permutations
a=gtools::permutations(5,10,repeats.allowed=TRUE)

# Select only the ones with 2 of each digit
k=apply(a,1,function(x) {sum(x==1)==2 & sum(x==2)==2 & sum(x==3)==2 & sum(x==4)==2})
b=a[k,]

rm(a)

t2=Sys.time()

print(paste0('Setup = ',difftime(t2,t1,units='secs'),' seconds'),collapse='')

print(dim(b))

## Commented out - just a feasibility test
if (FALSE) {
  t3=Sys.time()
  
  for (i in 1:100) {
    # print(i)
    for (j in (i+1):nrow(b)) {
      x=b[j,]
      for (k in 1:9) {
        x=c(x[-1],x[1])
        if (all(x==b[i,])) { b[j,]=0; break }
      }
    }
    k=apply(b,1,sum)
    b=b[k!=0,]
  }
  t4=Sys.time()
  print(paste0('Runtime = ',difftime(t4,t3,units='secs'),' seconds'),collapse='')
}
```

A couple of observations:

    If there are m different colors for each of the n beads then the number of possible
    combinations is m^n.  We can generate all possible combinations as a number of
    n digits in base m (m-ary n-tuple).
    
    Not sure how to calculate the number of combinations with 2 of each color.
    
    Note that a unique pattern will eliminate up to (n-1) rows since they will be
    circular shift(s) of the unique pattern.
    

Let's try a smaller problem to see if we can gain any insight.
Start with two colors:

    Assume we have 4 beads of two different colors = aabb
    Enumerate all the binary combinations and then select the ones that have two of each.
    Finally, select the ones that are unique (even with circular shift)

    
    bead1 bead2 bead3 bead4   Result
    -------------------------------
    a     a     b     b       Starting pattern
    a     b     a     b       Second pattern
    a     b     b     a       Eliminate - circular shift of starting pattern
    b     a     a     b       Eliminate - circular shift of starting pattern
    b     a     b     a       Eliminate - circular shift of second pattern
    b     b     a     a       Eliminate - double circular shift of starting pattern

Doing the same thing up to more colors/beads (2 each color) yields:

     4 beads, 2 colors -->      6 possible --> 2   unique combinations
     6 beads, 3 colors -->     90 possible --> 16  unique combinations
     8 beads, 4 colors -->   2520 possible --> 318 unique combinations
    10 beads, 5 colors --> 113400 possible --> ??? unique combinations

Trying brute force on 10 beads, 5 colors seems to take a really long time.
May need to switch languages or get a smarter algorithm.  The difficulty
is in checking each row against all shifts of all other rows.  The
algorithm is O(n^2) and n is very big.

** WARNING -- the first version of Lua code is wrong **

I rewrote this in Lua -- a lot of work but it runs faster.  At least it
finishes a couple of orders of magnitude faster but it still takes a
long time to run (about 10 minutes).  With LuaJit:

    10 beads, 5 colors --> 30144 unique combinations

The source is locate in `bracelet.lua`.  The unique solutions are given
in the text file `bracelet_solution.txt`.  The 5 colors are denoted by
digits 1-5.  The columns in the output represent the pattern.  It is assumed
that we needed to eliminate any rows that are circularly equivalent.

Need more work on this.

It should be possible to enumerate these manually and use cmbinatorics
but I'm not sure how.

Even so - the LuaJit execution time of 10 minutes versus R of ~3 days?????
Need to figure this out for future reference.

** Python was easy to write **

Should also look into Python itertools ... nothing really makes this easier except
in the generation of the n-tuple.  The itertools.product call makes it trivial to
parameterize this instead of hard-coded nested loops.  In addition, np.roll makes
it easy to check for cyclic rotation of vectors.  Unfortunately, a quick check shows
that it is still very slow.

Well -- I found out about collections.deque for quick rotation and ran the *same* tests
as the Lua program and now I have a different answer.  For the Python program
I get:

    10 beads, 5 colors --> 11352 unique combinations

Comparing the results I think the Python answer is correct.  There are rows in the
Lua solution that are shifted versions of prior rows.  For example, check out rows
0 and 3281 (zero-based) in the Lua solution.

The Python code is in file `bracelet.py`.

** Fixing the Lua Code **

Examining the Lua code for errors I think it was only testing 8 shifts instead of 9
when looking for duplicates.  I changed the limit to 9 and we will try to rerun this
when the python code has finished (I added some instrumentation and timing code).

    Python:
    Number of solutions =  11352
    Solution time (secs) =  1757.561486005783

    LuaJIT after fixing shift counter:
    Number of solutions = 22680
    Solution time (secs) = 492.128
    
Something still seems wrong.  Need to compare results and figure out why.
Problem still exists at the first line where the full number of shifts is required.
I think I have the shift and comparison code reversed.  Either I need to increase
the shift counter once more or swap the shift/check logic.  The current logic
actually makes one useless comparison so I will do the shift and then compare.
That should work correctly.

    LuaJIT after fixing shift/check logic:
    Number of solutions = 11352
    Solution time (secs) = 338.531

The solution for LuaJIT is now identical to the Python code.

Note that after writing this I modified the Lua and Python code to check for
palindromes to match the solution from the Bent.  In that case, the number
of feasible bracelets is reduced to (about) half the value given above.

## Double Bonus

    A chemist had twelve
    10-gram weights, one of which he
    knew was inaccurate. By conducting
    only three weighings on his equal-arm
    balance, he determined which weight
    was inaccurate and whether it was too
    light or too heavy. How did he do it?
    
    -- Maurice W. Widener, TX Alpha'48

TODO

I gave up and looked at the solution.  They indicate that there are
two possiblities for each of the 12 weights for 24 possibilities.

With three weighings you can get - left heavy, right heavy, or balance.

This is 3^3 possibilities = 27 results.

It should be possible to map the 27 weight results against the 24
possible weights to get a unique result.  They suggest the following
as one solution with weights 1, 2, ..., 9, A, B, C:

    1 2 3 4   vs.   5 6 7 8
    1 5 6 9   vs.   2 7 A B
    3 7 8 A   vs.   2 6 B C

Need to confirm that this works.  They give one example for

    Right-heavy
    Balanced
    Right-heavy
    
    This means weight 3 is light.

