---
title: "Bent Spring 2005"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Spring 2005


## Problem 1

    A large flower bed in front of our
    courthouse is laid out in the form of a
    regular polygon of N sides. A walk,
    composed of N identical stone slabs
    in the shape of isosceles trapezoids,
    surrounds the flower bed. The radius
    of the inscribed circle tangent to the
    inner edges of the walk is 19 feet,
    and the radius of the circumscribed
    circle through the outer vertices of
    the walk is 22 feet. The ratio of the
    area of the flower bed (not including
    the walk) to the area of the walk is
    almost exactly three (off by less than
    0.001). What is the value of N?

    -- Adapted from Technology Review

Solution:

A garden with 42 sides provides the requested ratio.

This seems fairly straightforward if I understand the definition
of an isosceles trapezoid.  I assume they mean this.


      <-a->
      -----   -
     /     \  h
     -------  -
     <--b-->
     
     Area = a*h+(b-a)/2*h
          = a*h/2 + b*h/2
          = h/2 * (a+b)
          
    The small side (a) is toward the center of the flower bed.
    The long side (b) is outward.  Each trapezoid lines up with
    a pizza slice of the polygon.
    
    For a regular polygon with N sides the central angle of each
    pizza slice is 2*pi/N.
    
    For an inscribed circle:
    
    The distance from the center to the center of the edge is R.
    The length of a side is L=R*tan(pi/N).
    
    For a circumscribed circle:
    
    The distance from the center to the vertex is R.
    The length of a side is 
    
    For the inside circle:  R = 19
    For the outside circle: R = 22
    
    Using the given data we can tabulate the areas and ratios as
    a function of N.
    
```{r}
n=5:50               # range of polygon sides
theta=pi/n           # Angle for half a pizza slice
r1=19                # Radius of the inscribed circle
s1=2*r1*tan(theta)   # Length of inside walkway segment
a1=n*0.5*s1*r1       # Area of the garden
r2=22                # Radius of the circumscribed circle
s2=2*r2*sin(theta)   # Length of outside walkway segment
h =r2*cos(theta)-r1  # width of walkway (height of trapezoid)
a2=n*0.5*h*(s1+s2)   # Area of the walkway

a=data.frame(N=n,garden=a1,walkway=a2,ratio=a1/a2)
a
```

A garden with 42 sides provides the requested ratio.

    N = 42
    Garden area = 1136.235
    Walkway area = 378.631
    Ratio = 3.000905
    
    Each walkway trapezoid has dimensions of:
       Inside edge = 2.847706
       Outside edge = 3.288124
       Width of walkway = 2.938484
       area = 9.015018


## Problem 2

    At a certain gambling establishment,
    the game of Lucky 10 is played
    as follows. The house provides a pot
    of $10. The player then tosses two
    coins simultaneously. If two heads
    appear, the player takes $1 from the
    pot; if two tails, he adds $1 to the pot.
    If he gets a head and a tail, it is a
    draw, and he neither adds to nor
    withdraws from the pot. After 10
    two-coin tosses (including draws), if
    there is exactly $10 in the pot, the
    player takes it. Otherwise, the house
    gets whatever is in the pot. How
    much should the player pay the
    house (to the nearest penny) to make
    this a fair game?

    -- Adapted from
    Joseph-Louis Lagrange (1770)

Solution: start with Monte-Carlo and then try analytics.

```{r}
ntrials=1000000
winnings=0

for (i in 1:ntrials) {
    p=10
    for (j in 1:10) {
        d=sample(c(0,1),2,replace=T)
        if (sum(d)==2) {
            p=p-1
        }
        if (sum(d)==0) {
            p=p+1
        }
    }
    if (p==10) {
        winnings=winnings+10
    } else {
        winnings=winnings+(10-p)
    }
}

print(paste0("Average winnings = ",winnings/ntrials,collapse=""))
```

For 1e6 trials the average winnings are 1.76

Note tha the analytical result below yields 1.761971 as a result of
184756 pots won out of total of 1048546 pots.


We can try this analytically but it gets difficult due to the draws.  There must be
some clever way to look at this problem that limits the number of combinations
that you need to consider.

Let's just try brute force:

    Round 1:  P(Win)=1/4 P(Loss)=1/4 P(Draw)=1/2, Pot = 9, 10*2, 11
    Round 2:  Same win probabilities so pot should change by -1,0*2,+1

        Pot = 8,9,9,10, 9,10,10,11, 9,10,10,11, 10,11,11,12
            = 8,9*4,10*6,11*4,12
            
    Round 3:
    
        Pot = 7,8*4,9*6,10*4,11, 8*2,9*8,10*12,11*8,12*2, 9,10*4,11*6,12*4,13
            = 7,8*6,9*15,10*20,11*15,12*6,13
    
    Showing this as a matrix:
    
    
                                          1  1  1  1  1  1  1  1  1  1  2
    Round   0  1  2  3  4  5  6  7  8  9  0  1  2  3  4  5  6  7  8  9  0
      0     0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0
      1     0  0  0  0  0  0  0  0  0  1  2  1  0  0  0  0  0  0  0  0  0
      2     0  0  0  0  0  0  0  0  1  4  6  4  1  0  0  0  0  0  0  0  0
      3     0  0  0  0  0  0  0  1  6  15 20 15 6  1  0  0  0  0  0  0  0
      ...
      
      In general, the entry at round m, pot value n is a recurrence of the
      form:
                x(m,n)=x(m-1,n-1)+2*x(m-1,n)+x(m-1,n+1), n=[10-m,10+m]
                
      We might be able to solve this closed form but we can likely code
      this more easily:
      
```{r}
# Unfortunately, R is 1-based so need to shift everything by 1
# This means the pot where the player wins (10) is actually at index 11.
a=rep(0,21); a[10]=1; a[11]=2; a[12]=1

zzz=a
for (r in 2:10) {
    b=a*0
    for (j in (11-r):(11+r)) {
        if (j==1) {
            b[j]=0+2*a[j]+a[j+1]
        } else if (j==21) {
            b[j]=a[j-1]+2*a[j]+0
        } else {
            b[j]=a[j-1]+2*a[j]+a[j+1]
        }
    }
    zzz=rbind(zzz,b)
    a=b
}
rownames(zzz)=1:nrow(zzz)

# By symmetry the only pot that pays off is when pot=10.
# All the other pots cancel out (e.g. wins=losses).  The
# payout for the pot is 10 and the mean gain is then:
print(10*zzz[10,11]/sum(zzz[10,]))
print(paste0("Pots that win = ",zzz[10,11]))
print(paste0("Total pots    = ",sum(zzz[10,])))

```

This matches the Monte-Carlo simulation.  Not sure if there is
a simple function to generate the pot counts as a function of
round number:

     2
     6
     20
     70
     252
     924
     3432
     12870
     48620
     184756 

It appears that this can be directly calculated as choose(2*m,m) where
m is the round number.  The total number of pots is just 4^m = 2^(2*m).

For the case of m=10 we get:

    Payout = choose(20,10)/2^20 = 1.761971
    

## Problem 3

    One morning, Ann left home and
    cycled north at a steady speed. At
    the same time, Bill left the same
    home and cycled south at a steady
    speed 1 mph faster than Ann's. An
    hour later, their mother realized that
    they had both left without their
    lunches, so she set out at a steady
    speed of 10 mph and cycled after
    Ann, gave her her lunch, then turned
    and cycled after Bill, give him his
    lunch, and finally reached home
    exactly 5 hours after she started.
    How long would it have taken her if
    she had cycled to Bill first and then to
    Ann?

    -- Susan Denham in New Scientist

Solution:

End result -- doesn't matter -- takes 5 hours either way.

Since the mother cycled for 5 hours @ 10 mph she travelled 50 miles.

    Let a = ann's speed
        b = a+1 = bill's speed
        m = 10  = mom's speed
        
    For the first leg: 0<t<t1
    
    For the first hour ann runs away from mommy: d = a
    After this the mother is catching up.  Eventually they
    meet at t=t1:
    
        Ann's distance = d1 = t1*a
        Mom's distance = d1 = (t1-1)*m
        
        So: t1*a = (t1-1)*10 --> t1=10/(10-a)

    Mom turns around and heads back home.  It takes her (t1-1)
    to get there.  So at t=t1+t1-1= 2*t1-1:

    Bill has been running away south.
    At t=2*t1-1 bill has a headstart of:
    
        d2 = (a+1)*(2*t1-1)
        
    So -- bill is far from mom.  It takes a while for mom to overcome
    the headstart -- assume it takes a time of t2:
    
        mom's distance = bill's distance + bill's headstart
    
        10*t2 = (a+1)*t2 + (a+1)*(2*t1-1)
        (10-a-1)*t2 = (a+1)*(2*t1-1)
                 t2 = (2*t1-1)*(a+1)/(9-a)
                 
    Once Mom catches Bill she can go back home.  This takes the same
    amount of time and distance so:
    
        t3=t2
        
    Total time = 6 = 2*t1-1 + 2*t2 = 2*t1-1 + 2*(2*t1-1)*(a+1)/(9-a)
    
    and, from the first leg: t1=10/(10-a)
    
    so: 6 = 2*10/(10-a)-1+2*(2*10/(10-a)-1)*(a+1)/(9-a)
    
    solving for a = -sqrt(385)/2 + 27/2
                  ~ 3.68929156482571

    Sanity check:
    
        t1 = 1.584 --> Mom catches up in 0.584 hours
        Mom back home at time = 2*t1-1 = 2.169 hours
        
        Meanwhile Bill has headstart of: 4.689*2.169 = 10.170
        
        To catch him:
        
        10*t = 4.689*t+10.170 --> t = 10.170/(10-4.689) ~ 1.915 hours
        
        Back home = 1.915 hours
        
        Total travel time = (1.584-1)*2+2*1.915 = 4.998 hours
        
    So -- we know the speeds.  Unfortunately that was not the question.  The
    question is how long does it take if she goes after Bill first?  I am going
    to guess that it is faster since you want to catch the fastest one first.
    But I could be completely wrong (I suspect it doesn't matter):
    
    bill has 1 hour head start.  Once Mom gets going:
    
        10*dt = b*(dt+1) --> dt = b/(10-b) = 0.882988
        
    Mom needs to get back home in same amount of time.  Total head start for
    ann is:
    
        t = 1+2*0.882988 = 2.766 --> headstart = 10.20449
        
    Now catch ann:
    
        10*dt = a*dt + 10.20449 --> dt = 10.20449/(10-a) = 1.617 hours
        
    Time to get back home is the same = 1.617 hours.
    
    Mom's travel time = 2*0.883 + 2*1.617 = 5 hours.
        
    Takes the same amount of time!


## Problem 4

    Consider a four-digit integer M
    with all different nonzero digits. Generate
    two other four-digit integers, A
    and D, by rearranging the digits of M
    so that A has the digits arranged in
    ascending order and D has the digits
    arranged in descending order.
    If D - A = M, what is the value of M?

    -- nearly impossible Brain Bafflers by
    Tim Sole and Rod Marshall

Solution:  We could generate all the candidate numbers
    and then check for compliance with these constraints.
    
    all permutations of 9 digits taken 4 at a time = 3024 possibilities
    
    take each row as M and then construct A, D.
    
```{r}
library(gtools)

check_m = function(m) {
    n=sum(m*c(1000,100,10,1))
    a=sum(m[order(m)]*c(1000,100,10,1))
    d=sum(rev(m[order(m)])*c(1000,100,10,1))
    return( (d-a)==n )
}

a=permutations(9,4,1:9)

s=apply(a,1,check_m)
a[s,]
```

The only 4 digit combination that works is:

    m = 6174
    a = 1467
    d = 7641
    

## Problem 5

    In 2004, the largest known prime
    was 2^(24,036,583) - 1. If this number were
    written in base 10, how many digits
    would it contain, and what would be its
    leftmost and rightmost three digits?

    -- Adapted from Technology Review

Solution:

    Number of digits ~ 7235733 (need to round up)
    Left-most  = 299...
    Right-most = ...407


    The number of digits can be approximated with log (base-10).
    
    num_digits ~ log10(2^(24036583))
               ~ 24036583 * log10(2)
               ~ 7,235,732.476
               
    How do we calculate the left/right-most digits?
    
    We can guess at the left-most digits by throwing away the exponent and just
    inverting the mantissa.
    
        Number starts with 2994...
    
    For the least significant digits:
    
    If we add 1 then this is a power of 2.
    The powers of 2 follow a pattern.  The right-most digit is:
    
    2, 4, 6, or 8 depending on the modulus of the exponent.
    
    For 2^n  if n%%4==0,1,2,3 --> last digit = 6,2,4,8
    
    For the given problem: n=24036583 so n%%4 = 3 --> last digit is 8.
    
    For the given prime: 2^n-1 --> last digit is 7
 
    Looking only at terms that are +3, modulo-4 we find that the final
    three digits of the powers of 2 repeat every 25 values (e.g. every 2^100).
    I found this looking at bigz values of 2^n for a wide range of n.  The
    solution in the Bent uses a theorem from Euler.
    
    Starting with 2^7, 2^11, 2^15, ... we have:
 
    128,  48, 768, 288, 608,
    728, 648, 368, 888, 208,
    328, 248, 968, 488, 808,
    928, 848, 568,  88, 408,
    528, 448, 168, 688,   8
    
    Starting at 2^107 this pattern repeats
    
    For given n this is many, many of the patterns.
    
    We have 4*i+3 = n --> i=6009145
    
    Modulo-25 this is 20 -- so we need the 20th pattern (1 based) = 408
    
    So -- the final 3 digits of 2^n are: 408
    
    After subtracting the 1 --> final digits of the prime are 407
    

## Bonus

    Consider an infinite plane
    ruled with parallel lines one meter
    apart. An ellipse with a major axis of
    0.75 m and a minor axis of 0.50 m is
    thrown down at random on the plane.
    What is the probability that the ellipse
    will fall on one of the lines?

    -- Joseph Whitworth (1803-87)

Solution:

    By Monte-Carlo: Pr ~ 0.643
    
    Needs more work -- Bent has closed form solution that is 0.631.

    I think the axis specifications imply a=0.75/2 and b=0.50/2.
    Assume the parallel lines are horizontal.
    
    We need to consider the random location of the center of ellipse
    and the angle of rotation of the ellipse.  The entire problem
    repeats every +/- 0.5 units along the y-axis so we only need to
    analyze this range.
    
    When the center is in the range [-0.25,+0.25] the ellipse will
    land on one of the lines

    When the center of ellipse is more than 0.75/2 = 0.375 units from
    the origin then the ellipse can not fall on a line.
    
    In the intermediate area [0.25 to 0.375] the likelihood of falling
    on a line will depend on the rotation angle of the ellipse.  Again,
    due to symmetry we only need consider one quadrant of rotation.

        (x/a)^2 + (y/b)^2 = 1
        
    The critical angle of rotation will occur when the distance from
    the center of the ellipse to the point on the ellipse where it
    crosses one of the lines, d:
    
        x^2 = a^2 * (1 - (y/b)^2)

        d^2 = x^2 + y^2 = a^2 * (1-(y/b)^2) + y^2
            = a^2 - y^2*(a/b)^2 + y^2
            = a^2 + (1 - (a/b)^2) * y^2
            
        y = sqrt((d^2 - a^2)/(1-(a/b)^2))

    The rotation angle that makes the ellipse hit the line will be:
    
        theta = asin(y/d)
        
    The actual rotation must be at least this large so the likelihood that
    the ellipse will cross a line is:
    
        Probability = (pi/2-theta)/(pi/2) = 1 - 2*theta/pi
    
    Not sure I can do this symbolically but I can do Monte-Carlo.
    
```{r}
a=0.75/2
b=0.50/2

ntrials=1000000
nhits=0
for (i in 1:ntrials) {
    d=runif(1)/2
    if (d<=b) {
        nhits=nhits+1
    } else if (d<a) {
        y0=sqrt((d^2-a^2)/(1-(a/b)^2))
        theta=asin(y0/d)
        p=1-2*theta/pi
        nhits=nhits+p
    }
}

print(paste0("Pr(hit a line) = ",nhits/ntrials))
```

The initial simulation suggests Pr = 0.538 but this seems way too low.
The central region alone provides Pr = 0.5 so I would expect more
contribution from the range 0.25 to 0.375.

I think I was miscalculating the Pr hitting a line by a factor of 2.
Mental error used pi/4 instead of pi/2 for 90 degrees.

After modifying the code: Pr ~ 0.643 which seems reasonable.

Need to look at this some more.

The Bent solution derives this "closed form" in terms of Elliptic
integrals and gets Pr ~ 0.63127.  They say the general problem is
just the radio of the perimeter of the ellipse and the perimeter of
a circle that just touches the horizontal lines.


## Computer Bonus

    Some ways to
    express a power of 10 as the product of
    two factors without using any zeros
    are 10 = 2 x 5, 10^2 = 4 x 25, and 10^3 = 8
    x 125. What is the largest power of 10
    which can be so expressed?

    -- The Mathematics Teacher

Solution:  isn't the answer just 1e7?

    Searching up through 10^10000 I agree with the Bent
    solution that the only powers are:
    
    ... 6, 7, 9, 18, and 33.

    My first attempt at this completely missed the question.
    See below.
    
    Need to do this again.  They weren't asking for
    the smallest power of 10 that requires a zero.
    They were asking for the highest power that
    CAN be expressed without a zero.

    For example, in the table 10^8 requires a zero but
    10^9 does not.

    In general, the prime factors of 10 = 2 * 5
    Any power of 10^n = 2^n * 5^n
    
    Thus, any factors of 10^n must be some combination
    of 2^m and 5^k.  If we mix the 2 and 5 terms we
    immediately get 2*5 = 10 --> a zero.
    
    Conclude, must keep 2^n and 5^n separate such that
    10^n = (2^n) * (5^n)
    
    The powers are given as:
    
    N   2^N   5^N
    1   2     5
    2   4     25
    3   8     125
    4   16    625
    5   32    3125
    6   64    15625
    7   128   78125
    8   256   390625   <- Zero is unavoidable in 5^8
    9   512   1953125
    10  1024  9765625

    Final conclusion: need to limit to 10^7 = 2^7 * 5^7 = 128 * 78125
