---
title: "Bent Summer 2014"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Summer 2014


## Problem 1

    Our local baseball park is rather
    oddly shaped. It is exactly 260 feet
    from first and third bases to the foul
    poles and from second base to the
    fence in dead center field. The out-
    field fence itself is in the form of an
    arc of a circle. In order to improve
    the production of home runs, it has
    been decided to reconstruct this
    fence so that the fair ball portion
    of the field will be in the shape of a
    quadrant of a circle with a radius of
    350 feet (distance from home plate to
    the outfield fence). Thus, the dis-
    tance from home plate to the fence
    along the foul lines will be the same
    as before, but the distance from
    home to the center field fence will
    be shorter. How much area will be
    removed from the fair ball portion
    of the park? Express your answer to
    the nearest square foot.

    -- Larry M. Lesser, son of member

Solution:

    Need to determine the location of the center and
    the radius of the original circle.  If we know
    the coordinates of the three points on the circle
    we should be able to project perpindicular
    bisectors and find the intersection.
    
    I am going to assume the foul lines are 90 degrees
    apart and the infield is a square -- is that known?
    
    If so then the three points are located at:
    
    Point 1: x=90+260=350, y=0
    Point 2: x=0,          y=90+260=350
    Point 3: x=y=(90*sqrt(2)+260)/sqrt(2)
    
    Midpoint 1-3: x13=(x1+x3)/2, y13=(y1+y3)/2
    Slope 1-3: m13=(y3-y1)/(x3-x1)
    Midpoint 2-3: 
    Slope 2-3: m23=(90*sqrt(2)+260-350)/(90*sqrt(2)+260)
    
    For the perpindicular bisector the slopes are -1/m13, and -1/m23.
    
    We have point and slope -- need equation of line:
    
    y0=m*x0+b --> b=(y0-m*x0) ==> y=m*x+(y0-m*x0)
    
    Line 1: y=-1/m13*x+(y13+1/m13*x13)
    Line 2: y=-1/m23*x+(y23+1/m23*x23)
    
    Point of intersection = center of circle:
    
    -1/m13*xc+(y13+1/m13*x13) = -1/m23*xc+(y23+1/m23*x23)
    xc*(1/m23-1/m13) = (y23+1/m23*x23)-(y13+1/m13*x13)
    xc = ((y23+1/m23*x23)-(y13+1/m13*x13))/(1/m23-1/m13)
    yc = -1/m13*xc+(y13+1/m13*x13)
    
    radius of circle:
    
    r = sqrt((x1-xc)^2+(y1-yc)^2)

    This is all rather ugly but should be sufficient:
    
    First circle centered at xc,yc and radius=r
    Second circle centered at 0,0, and radius=350
    
    Find the area between first and second circles.
    This is the same as finding the fair area of the original
    field and the fair area of the second field and the difference.

    Could do this symbolically or numerically.
    
```{r}
# Point 1
x1=350; y1=0
x2=0;   y2=350
x3=(90*sqrt(2)+260)*cos(pi/4)
y3=(90*sqrt(2)+260)*sin(pi/4)

x13=(x1+x3)/2; y13=(y1+y3)/2; m13=(y3-y1)/(x3-x1)
x23=(x2+x3)/2; y23=(y2+y3)/2; m23=(y3-y2)/(x3-x2)

xc=((y23+1/m23*x23)-(y13+1/m13*x13))/(1/m23-1/m13)
yc= -1/m13*xc+(y13+1/m13*x13)

r1=sqrt( (x1-xc)^2 + (y1-yc)^2 )
r2=sqrt( (x2-xc)^2 + (y2-yc)^2 )
r3=sqrt( (x3-xc)^2 + (y3-yc)^2 )

print(sprintf("Right Field Foul Pole: x1 = %10.4f   y1 = %10.4f",x1,y1))
print(sprintf("Left  Field Foul Pole: x2 = %10.4f   y2 = %10.4f",x2,y2))
print(sprintf("Dead Center Field:     x3 = %10.4f   y3 = %10.4f",x3,y3))

print("Original layout - center of circle")
print(sprintf("xc = %10.4f   yc = %10.4f",xc,yc))

print("Original layout - radius of circle")
print("Distance to Foul poles and dead center should all be the same")
print(sprintf("Right Field   r1 = %10.4f",r1))
print(sprintf("Left  Field   r2 = %10.4f",r2))
print(sprintf("Center Field  r3 = %10.4f",r3))

# We have center and radius of original circle
#
# Calculate area of original field
# Calculate area of new field
# Calculate difference

# Area of the original field
# This is shaped like an ice cream cone
# Need area of circular arc and the two triangles
# on either side.  The circular arc extends beyond
# 90 degrees so we need to calculate the total angle:
theta1=asin(xc/r1)
theta2=pi/2+2*theta1
area1=(r1^2/2)*theta2
# Each triangle goes down fowl line out to xc,yc
area2=1/2*350*xc
area_original=area1+2*area2

# Area of the revised field is just a quarter circle
# with the new radius
area_new=pi*350^2/4

print(paste0("Original area = ",area_original))
print(paste0("Revised  area = ",area_new))
print(paste0("Reduction = ",area_original-area_new))
```

    Summary of Solution:
    
    Original fence defined by circle centered at: 69.514,69.514
    with a radius of 288.9716.
    
    Original fair area = 110200.775
    Revised  fair area =  96211.275
    Reduction          =  13989.5003


## Problem 2

    A jeweler dies and bequeathes
    his hoard of 45 jewels (9 diamonds,
    9 rubies, 9 sapphires, 9 opals, and
    9 zircons) to his five children. The
    diamonds are worth $1,000 each,
    the rubies $750, the sapphires $500,
    the opals $250, and the zircons are
    worthless. His will specifies that the
    jewels be divided so that each child
    gets the same total value and the
    same total number of jewels. Fur-
    thermore, each child gets at least one
    of each kind of gem, but that no two
    children get identical combinations of
    gems. Finally, each kind of gem is to
    be distributed as evenly as possible.
    That is, the total number of three-
    of-a-kinds among the five children
    is to be a minimum. How should the
    jewels be distributed?

    -- Mathematical Recreations,
    by Maurice Kraitchik

TODO

    meh

## Problem 3

    On the game show What's in a
    Name?, N contestants write their
    names on slips of paper and drop
    them into a bowl. The slips are
    mixed, and each person draws a slip.
    If a contestant gets his own name,
    he drops out. The remaining contes-
    tants put their slips back in the bowl,
    and the process is repeated. The
    prize money is split evenly among
    those going out on the last round
    (which leaves no players). What is
    the expected number of rounds for
    everyone to have dropped out? For
    large values of N, what is the prob-
    ability that the total prize is split
    between exactly two contestants?

    -- Mathematics Magazine

Solution:

    We can try Monte-Carlo with specific values of N and
    try to generalize to a formula or just use a plot.
    
```{r}
ntrials=10000

x=c()
y=c()
z=c()
for (N in seq(10,100,10)) {
  n=0
  m_final=0
  for (trial in 1:ntrials) {
    m=N
    while(T) {
      s=sample(m)
      prev_m=m
      m=sum(s != 1:m)
      n=n+1
      if (m==0) {
        m_final=m_final+(prev_m==2)
        break
      }
    }
  }
  x=append(x,N)
  y=append(y,n/ntrials)
  z=append(z,m_final/ntrials)
}

plot(x,y,type='b')
plot(x,z,type='b')
```

    It appears the expected number of rounds for N contestants is N rounds.
    About 80% of the time the final round consists of two contestants.
    
    The Bent said they wrote a program to calculate the precise probability
    as 0.80189.  Close enough.
    

## Problem 4

    Four logicians went on a camping
    trip. Among their supplies were 11
    donuts. At their evening campfire,
    each logician ate one donut. During
    the night, one or more of the logi-
    cians, unbeknownst to the others,
    raided the donuts, so that in the
    morning the box was empty. The
    following exchange then took place.
    Ann asked, "Beth, did you eat more
    donuts than I did?" Beth replied,
    "I don't know. Carol, did you eat
    more donuts than I did?" Carol an-
    swered, "I don't know." Diane then
    remarked, "Now I know how many
    donuts each of us ate." If no logician
    asked a question to which she knew
    the answer, how many donuts did
    each logician eat? Assume the logi-
    cians are honest, at least when it
    comes to answering questions truth-
    fully.

    -- Hard-to-Solve Brainteasers,
    by Jaime and Lea Poniachik

TODO

    meh


## Problem 5

    Consider six one-ohm resistors
    that are connected to form a tetrahe-
    dron. What is the resistance between
    any two vertices?

    -- Allan Gottlieb's Puzzle Corner
    in Technology Review

Solution:

    The network is shown below:
    
         +------+------+
         |      |      |
         |      R      |
         R      |      R
         |  +-R-+-R-+  |
         |  |       |  |
         +--+---R---+--+

    If I could remember how to do delta-wye
    transformation this would be trivial.  Maybe
    I should look into that.
    
    If I knew how to do that I would transform this
    to:
    
         +------+------+-------
         |      |      |
         |      R      |
         R      |      R     <== R seen from here
         |      +      |
         |      Ry     |
         +--Ry--+--Ry--+-------
    
    R seen from right side is:
    
    Req = ((R+Ry)//(R+Ry)+Ry)//R = R/3*(R+3*Ry)/(R+Ry)


    So -- how do we do the delta-wye transformation?

        +---------------- A     +------------- A
        |                       |
        +---R---+               Ry
        |       |               |
        R       +-------- B     +-----Ry------ B
        |       |               |
        +---R---+               Ry
        |                       |
        +---------------- C     +------------- C
        
    We want to choose Ry such that the impedances are the same
    on the two networks.
    
    From Wikipedia: Ry = R/3 when all the R are equal.
    
    So: substituting into the above we have:
    
    Req = ((R+Ry)//(R+Ry)+Ry)//R = R/2 = 1/2 Ohm

    They had a more elegant answer in the Bent.  Go
    back to the original diagram:
 
               
         +------+---------------+
         |      |               |
         |      +------A        |
         |      R       <=      |
         R      +------B        R
         |      |               |
         |  +-R-+-R-+           |
         |  |       |           |
      C--+--+---R---+-----------+--D
                ^   
                +--- By symmetry - no current in this R
                     No current means it is open circut

    Then this reduces to:
    
         +---+---+-----
         |   |   |
         2R  2R  R   <=
         |   |   |
         +---+---+-----

    Req = 1/(1/R+1/(2*R)+1/(2*R)) = R/(1/1+1/2+1/2) = R/2 = 1/2 Ohm


## Bonus

    Two Tau Bates play a game.
    They start with a sheet of paper
    on which are written the integers
    from 1 to 100 (inclusive). They take
    turns crossing off numbers. Once a
    number is crossed off, it is dead and
    cannot be used again. Other than the
    opening move, the chosen number
    must be either an exact divisor of the
    previous choice or an exact multiple
    of it. For example if one player chose
    45, the other player could choose 1,
    3, 5, 9, 15, or 90 (provided they were
    still available). The opening move
    must be an even number. The first
    player who cannot follow the rules
    loses. Can the player crossing off the
    first number force a win? If so, what
    opening move should be made?

    -- Professor Stewart's Hoard of
    Mathematical Treasures,
    by Ian Stewart

TODO

    meh


## Computer Bonus

    Arrange the integers from one to N in
    alphabetical order. To avoid ambiguity, use
    no hyphens or and's.  Consider spaces to
    alphabetically precede all letters. For
    example, 837,301 would be spelled eight
    hundred thirty seven thousand three hundred
    one; also, 1,807 is one thousand eight
    hundred seven, not eighteen hundred seven.
    Define F(n,m) as the mth integer in the
    alphabetical list of the integers from one to
    n, and define G(n,m) as the position of the
    integer m in this list.  (a) What are the
    values of F(n,m) and G(n,m) when n =
    1,000,000 and m = 1,000? When n = m =
    1,000,000?  (b) What are the ten smallest
    values of n for which F(n,n) = G(n,n) = n?

    -- Allan Gottlieb's Puzzle Corner
    in Technology Review

TODO


    I thought about doing this but I really don't
    see the point.
