---
title: "Bent Fall 2013"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Fall 2013


## Problem 1

    At exactly noon, Bob left the Red Lion at
    Upper Darby and set off by foot on the trail
    to the Purple Cow in Lower Merion. At exactly
    noon, Carl left the Purple Cow and set off by
    bicycle on the same trail for the Red Lion.
    When they met, Bob had covered 4 miles. After
    a 10 minute chat, Carl gave Bob the bike and
    walked on to the Red Lion, where he drank a
    bottle of beer in 3 minutes and set off again
    for the Purple Cow. Bob, as soon as Carl left
    him, cycled on to the Purple Cow, drank a
    bottle of beer in 3 minutes, and cycled off
    again toward the Red Lion. This time they met
    7 miles from the Purple Cow and discussed
    football for 10 minutes. Then Bob
    relinquished the bike and walked on to the
    Red Lion, and Carl cycled on to the Purple
    Cow. Each drank another bottle of beer, in
    the same time as before, and off our heroes
    went again with Carl still riding and Bob
    afoot. They collided with each other 2 miles
    from the Red Lion. If they each maintain
    their own individual walking and riding rates
    for the entire episode, how far is the Red
    Lion from the Purple Cow?

    -- A Tantalizer by Martin Hollis
    in New Scientist

TODO

    may do this -- seems like fun.
    

## Problem 2

    Ann collects stamps. She has half as many
    from Canada as from Japan; one tenth as many
    from France as from Denmark; one fifth as
    many from Libya as from Egypt; five times as
    many from Haiti as from France; two fewer
    from Libya than from Canada; three times as
    many from India as from Canada; half as many
    from Australia as from Korea; one fewer from
    Korea than from India; four times as many
    from Brazil as from France; and twice as many
    from Guatemala as from Libya. Ann has a total
    of 303 stamps. How many Canadian stamps does
    Ann have, and how many French stamps?

    -- Logic Puzzles to Bend
    Your Brain by Kurt Smith

Solution:

    The only integer solution is: C=13, F=4
 
    All of the variables can be expressed in terms of C or F:
 
    C = J/2  --> J = 2*C
    F = D/10 --> D = 10*F
    L = E/5  --> E = 5*L = 5*(C-2)
    H = 5*F
    L = C-2
    I = 3*C
    A = K/2  = (I-1)/2 = (3*C-1)/2 -> C is odd
    K = I-1  = 3*C-1
    B = 4*F
    G = 2*L  = 2*(C-2)
    C+J+F+D+L+E+H+I+A+K+B+G=303
    
    C=?   F=?

    C+J   +L     + E       + I   + A         + K     + G
    C+2*C +(C-2) + 5*(C-2) + 3*C + (3*C-1)/2 + 3*C-1 + 2*(C-2)
    17*C-17+(3*C-1)/2
    (17+3/2)*C - (17+1/2)
    
    F + D    + H   + B
    F + 10*F + 5*F + 4*F
    20*F
    
    37/2*C - 35/2 + 20*F = 303
    
    The only integer solution is: C=13, F=4
    ---------------------------------------
    A = (3*C-1)/2 = 19
    B = 4*F = 16
    C = 13
    D = 10*F = 40
    E = 5*(C-2) = 55
    F = 4
    G = 2*(C-2) = 22
    H = 5*F = 20
    I = 3*C = 39
    J = 2*C = 26
    K = 3*C-1 = 38
    L = C-2 = 11


## Problem 3

    Consider all possible cryptic additions with
    only two addends, where the addends may be
    strings of any length. An example of such a
    cryptic is: ABCA + DCDCD = ABCDE. The
    solution to this cryptic is: 5275 + 47474 =
    52746. We are concerned with cryptics that
    have a unique, that is only one, solution in
    base 10. For such a cryptic, define its
    N-value as the number resulting from deleting
    the + and = signs from its solution. For
    example, the N-value of the above cryptic is
    52,754,747,452,746. Cryptics, such as A + B =
    C, with more than one solution, have no
    N-value. Of all possible N-values, what are
    the four smallest?

    -- An Enigma by Keith Austin in New Scientist

TODO

    meh


## Problem 4

    What is the probability that a positive
    integer, N, chosen at random, will have no
    repeated prime divisors? That is, if p is a
    prime divisor of N, then p^2 is not a divisor
    of N?

    -- A Biography of the World's
    Most Mysterious Number
    by Alfred S. Posamentier
    and Ingmar Lehmann

Solution:

    OK - I gave up and looked at the Bent solution.
    I would have never gotten it.
    
    Result: Pr = 6/pi^2 ~ 0.60793
    
    The argument goes like this:
    
    Given a prime number, p, the probability that N has
    p as a prime factor is 1/p (every pth integer has p
    as a prime factor).  The probability that p^2 is
    a factor is 1/p^2.  So:
    
    Pr(N is not divisible by p^2) = 1-1/p^2
    
    If N has a number of prime factors, the probability
    that none of them are repeated is:
    
    P = (1-1/p[1]^2)*(1-1/p[2]^2)*(1-1/p[3]^2)*...
    
    Invert both sides:
    
    1/P = 1/(1-1/p[1]^2) * 1/(1-1/p[2]^2) * 1/(1-1/p[3]^2) * ...
    
    But:
            1/(1-x^2) = 1 + x^2 + x^4 + x^6 + ...
            
    So:
    
    1/P = 1/(1-1/p[1]^2) * 1/(1-1/p[2]^2) * 1/(1-1/p[3]^2) * ...

        = (1+1/p[1]^2+1/p[1]^4+...) * 
          (1+1/p[2]^2+1/p[2]^4+...) * 
          (1+1/p[3]^2+1/p[3]^4+...) * ...

    Letting p[1]=2, p[2]=3, p[3]=5, etc. we have:
    
    1/P = (1+1/2^2+1/2^4+...) * 
          (1+1/3^2+1/3^4+...) * 
          (1+1/p5^2+1/5^4+...) * ...
    
    Multiplying this out:
    
    1/P = (1 + 1/2^2 + 1/3^2 + 1/2^4 + 1/5^2 + 1/(2^2*3^2) + ...)
        = (1 + 1/2^2 + 1/3^2 + 1/4^2 + 1/5^2 + 1/6^2 + ...)
        = sum(1/n^2,n,1,inf)
        = pi^2/6
        
    So:  P = 6/pi^2 ~ 0.6079271018540267

    We can check it out with a small numerical test.  We will try
    1000000 random samples from a range of 1-1e9.
    
```{r, eval=FALSE}
library(gmp)

x=sample(1000000000,1000000)
n=sapply(x,function(x) {max(rle(as.numeric(factorize(x)))$lengths)==1})
print(mean(n),digits=10)
```

One random trial -- amazingly close = 0.607675

    
## Problem 5

    A young Tau Bate is contemplating her
    financial future. Confident in her
    engineering skills, she assumes that her
    salary will increase 5% annually for an
    indefinite time. On the advice of a financial
    planner, she intends to invest 10% of her
    total income (salary plus dividends) at the
    end of each year in a respected mutual fund.
    The fund pays out a flat annual dividend of
    10% at the end of each year. The principle is
    to remain invested indefinitely. For year 1,
    her income is simply her starting salary S,
    and she invests 10% of that. In year 2, her
    total income is 1.05 S plus 10% of her
    accrued investment, which is 0.1 S. At the
    end of the year, she adds 10% of her year 2
    income to her investment. Find closed
    formulas for her total income for year N, and
    for the total amount invested at the
    beginning of that year.

    -- Adapted from George Boole, 1860

TODO

    I think this is a rather standard investment
    formula.  May need the sum of a couple of
    terms.
    

## Bonus

    Find positive integers a, b, c, and d
    which simultaneously solve the following
    two equations:

    a^2 + b^3 = c^4
    a^4 + b^6 = d^7

    Find the solution with the smallest value of
    d. Note that zero is not a positive integer.

    -- Allan Gottlieb's Puzzle Corner
    in Technology Review

Solution:

    I cheated and looked at their solution.
    
    Substitute multiples of another variable as:
    
        a=A*F^12, b=B*F^8, c=C*F^6, d=D*F^7
        
    into first equation:
    
        A^2*F^24 + B^3*F^24 = C^4*F^24
        or:
        A^2 + B^3 = C^4
    
    Substitute into the second equation:
    
        A^4*F^48 + B^6*F^48 = D^7*F^49
        or:
        A^4 + B^6 = D^7 * F
        
    Finally, wave a magic want and find a solution to the first
    of these equations:
    
        A^2 + B^3 = C^4
        
    They magically come up with a solution as:
    
        A = 41472 = 2^9 * 3^4
        B =  1728 = 2^6 * 3^3
        C =   288 = 2^5 * 3^2

    I have no idea how they came up with this solution.  There is a hint
    in the exponents - but I don't understand it.
    
    Once this is known you can subsitute into the second equation:
    
        A^4 + B^6 = D^7 * F
        2^36 * 3^16 + 2^36 * 3^18 = D^7 * F
           = (2^35*3^14)*(2^1*3^2) + (2^35*3^14)*(2^1*3^4)
           = (2^35*3^14)*(2^1*3^2 + 2^1*3^4)
           = (2^5*3^2)^7 * (2*3^2 + 2*3^4)
           = 288^7 * 180
        D = 288
        F = 180
        
    Finally, back-substituting:
    
        d = D*F^7 =  1763193692160000000
        c = C*F^6 =  9795520512000000
        b = B*F^8 =  1904249187532800000000
        a = A*F^12 = 47976111050506371405143433281536


## Computer Bonus

    In a children's game called Beetle, the
    object is to draw a beetle based on the
    results of throwing a die. Each face of the
    die allows adding a different part to the
    beetle: 1-body, 2-head, 3-eye (two needed),
    4-feeler (two needed), 5-leg (six needed),
    6-tail. You must have a body before you can
    add any other parts, and you must have a head
    before you can add eyes or feelers. Calculate
    the exact expected number of die tosses
    required to complete the beetle.

    -- World's Greatest Dice Games

Solution:

    May need to try Monte-Carlo?
    
```{r}
ntrials=10000

t1=Sys.time()
set.seed(42)
n_total=0
for (trial in 1:ntrials) {
    n=0
    n_body=0
    n_head=0
    n_eye=0
    n_feeler=0
    n_leg=0
    n_tail=0
    # Get a body
    while (n_body==0) {
        s=sample(6,1)
        n=n+1
        if (s==1) {
            n_body=1
        }
    }
    # Add parts until done
    while (TRUE) {
        s=sample(6,1)
        n=n+1
        if (s==2) {
            # Add the head
            n_head=1
        }
        if (s==3) {
            # Add an eye
            if (n_head==1) {
                n_eye=n_eye+1
            }
        }
        if (s==4) {
            # Add a feeler
            if (n_head ==1) {
                n_feeler=n_feeler+1
            }
        }
        if (s==5) {
            # Add a leg
            n_leg=n_leg+1
        }
        if (s==6) {
            # Add a tail
            n_tail=1
        }
        # Check if done
        if ((n_head==1) & (n_eye>=2) & (n_feeler>=2) & (n_leg>=6) & (n_tail>0)) {
            break
        }
    }
    n_total=n_total+n
}
t2=Sys.time()
print(paste0("Total tosses = ",n_total))
print(paste0("Mean  tosses = ",n_total/ntrials))
print("Simulation time = ")
t2-t1
```


Results of Monte-Carlo

    Seed  Trials  Result
    ---------------------
    42    10K     44.3886
    42    100K    44.30266
    123   100K    44.19435
    2020  100K    44.29614
    1     100K    44.19622
    42    1000K   44.254233
    123   1000K   44.215643
    2020  1000K   44.240336
    

The have a closed form solution in the Bent as:

    44 + 1006625/3981312 tosses
    
    ~ 44.25284
    
They argue these are the state transitions in a FSM.
As a caution, however, there are 141 states and the
transition paths make it complicated to analyze.  They
wrote a computer program to calculate the expected
value (EV) and said it "converged" to the value given
above.

As an alternative, they suggested Monte-Carlo.

