---
title: "Bent Winter 2002"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Winter 2002


## Problem 1

    Two dice, one red and one white,
    are thrown. The number shown on
    the red die is divided by the number
    shown on the white die. What is the
    expected value of this quotient?
    
    -- William A. Whitworth, 1901

Solution: Should just be able to enumerate these
without much difficulty.  Consider each case
of the red die (1:6) and the white die (1:6)
for a matrix of 36 values.  The mean value
should be the desired result.

           ...... Red ............
    White  1   2   3   4   5   6
    1      1   2   3   4   5   6
    2      1/2 2/2 3/2 4/2 5/2 6/2
    3      1/3 2/3 3/3 4/3 5/3 6/3
    4      1/4 2/4 3/4 4/4 5/4 6/4
    5      1/5 2/5 3/5 4/5 5/5 6/5
    6      1/6 2/6 3/6 4/6 5/6 6/6

Now - is there a clever way of summing these or
do we just use brute force?  We can sum the
numerators on each row as 1+2+3+4+5+6 = 21 and
then find a common denominator to add the row sums.

    Total = 21*(1/1+1/2+1/3+1/4+1/5+1/6)

The least common denominator is 60, so:

    Total = 21/60*(60+30+20+15+12+10)
          = 21/60*147

The mean result is:

    Mean = Total/36 = 21/60*147/36
         = (21*147)/(60*36)
         = (3*7*3*7*7)/(2*2*3*5*2*2*3*3)
         = 7*7*7/(2*2*3*5*2*2)
         = 343/240
         ~ 1.429167
         
Just FYI - this is one line of R code

```{r}
mean(outer(1:6,1:6,"/"))
```

## Problem 2

    Two painters accept a job painting
    a room. Al can do the flat work in
    two hours and the trim work in one
    hour. Ben can do the flat work in
    three hours and the trim work in two
    hours. What is the minimum time, in
    minutes, that the two painters can
    paint the room if they divide the
    work in the optimum manner?
    
    -- William S. Alderson, MI E'43

Solution: this is like `distance=rate*time`

    Intuitively it seems like the work should
    be partitioned based on the relative
    ability of the workers.  For example:
    
    If they are equal speed then each does
    half the work.
    
    If one is 2X faster then they should
    get 2X of the job.

    When there are two jobs -- how do we
    optimally partition this?

    There must be an easier way to figure this out but
    I am failing to see it.

    Al flat rate  = 1/2 job per hour
    Al trim rate  = 1/1 job per hour
    Ben flat rate = 1/3 job per hour
    Ben trim rate = 1/2 job per hour

    If Ben did the trim and Al the flat then the
    whole job would take 2 hours.  We should be
    able to improve on that.
    
    Let Al do all the trim in 1 hour and Ben
    would complete 1/3 of the flat.  This leaves
    2/3 of the flat with both working on it:
    
    t*(1/2+1/3) = 2/3
    t*(3/4+1/2) = 1
    t*(5/4) = 1
    t = 4/5 hours
    
    Total = 1.8 hours

    Is this the best we can do?  Let's try analytically:
    
    pf = percent of flat work job performed by Al
    pt = percent of trim work job performed by Al

    Al_time = pf/(1/2)+pt/(1/1)
    Ben_time = (1-pf)/(1/3)+(1-pt)/(1/2)
    
    Al_time = Ben_time
    
    2*pf + pt = 3*(1-pf)+2*(1-pt)
    5*pf       = 5 - 3*pt
    pf         = 1 - 3/5 * pt
    
    So: Al_time = 2*(1-3/5*pt) + pt
                = 2 - 6/5*pt + pt
                = 2 - 1/5*pt
                
    I want to minimize this so need maximum pt = 1, pf = 2/5

    Al_time = 9/5 = 1.8
    Ben_time = 3/5*1/3 + 0 = 9/5 = 1.8
    
    I guess this confirms my original guesstime for full time on trim.

    Should I try Monte-Carlo just for grins???

```{r}
# Uniform random percent of each job performed by Al
# Note that I added the underscore to prevent name
# collision with R functions
p_f=runif(100000)
p_t=runif(100000)
al_time = 2*p_f + p_t
ben_time = 3*(1-p_f) + 2*(1-p_t)

# We aren't done until the larger of these two have finished
job_time = ifelse(al_time>ben_time,al_time,ben_time)

# The best solution is the min of this value
print(paste0('Best time = ',min(job_time)))

# Allocation for this optimum
print(paste0('Portion of flat work done by Al = ',p_f[which.min(job_time)]))
print(paste0('Portion of trim work done by Al = ',p_t[which.min(job_time)]))

```

Which confirms the analysis above.  Al does all the trim and 40% of the flat work.
Ben does the remaining 60% of the flat work.

## Problem 3

    Walking near a pond, a 45 kg boy
    finds a hemispherical concrete shell
    2.5 cm thick and tries to use it as a
    boat. With the boy aboard and
    crouching low for stability, the shell
    floats with 10 cm of freeboard. If the
    specific gravity of the concrete is 2.5,
    what is the outside diameter of the
    hemisphere?
    
    -- Craig K. Galer, MI A'77

Solution: this goes back to a previous problem involving the volume of the
lower portion of a hemisphere along with some interesting calculations.

For a full hemisphere:

    R = outside radius of the hemisphere
    Vo = volume of the outside of the hemisphere
    Vi = volume of the (hollow) inside of the hemisphere
    Mh = mass of the hemisphere
    
    Vo = 2/3*pi*R^3
    Vi = 2/3*pi*(R-2.5)^3
    Mh = (Vo-Vi)*2.5
    
    Total Mass of "boat" = Mh + 45 kg
    
    Volume of Water displaced = lower part of hemisphere (all but top 10 cm).
    
    For a sphere of radius = R, height of h from bottom
    
    V = pi*R*h^2 - pi*h^3/3

    Volume of water = pi*R*(R-10cm)^2 - pi*(R-10cm)^3/3
    
    If we assume water is 1000 kg per cubic meter and put everything in kg, meters:
    
    (pi*R*(R-0.1)^2 - pi*(R-0.1)^3/3)*1000 = (Vo-Vi)*2.5*1000 + 45
            = (2/3*pi*R^3 - 2/3*pi*(R-0.025)^3)*2.5*1000+45
            
    2000*pi*R^3/3 - 100*pi*R^2 + pi/3 = 125*pi*R^2 - 25*pi*R/8 + 5*pi/192 + 45
    
    2000*pi*R^3/3 - 225*pi*R^2 + 25*pi*R/8 - 45 + 59*pi/192 = 0
    
    R = 0.436911963870549 meters
    
    Diameter = 2*R ~ 0.873823927741098 meters
 
    (OK - I confess I used sympy in a separate ipython window for the heavy lifting)
    
    Check for sanity:
    
```{r}
foo = function(r) {
    return(2000*pi*r^3/3 - 225*pi*r^2 + 25*pi*r/8 - 45 + 59*pi/192)
}

r=uniroot(foo,c(0,1))$root
print(paste0("Numerical solution for r     = ",r))

r=0.436911963870549   # outside radius of hemisphere (from sympy)
r2=r-0.025            # inside radius of hemisphere
freeboard=0.1         # height from water to top of hemisphere
boy_mass=45           # mass of the passenger

water_volume = pi*r*(r-freeboard)^2 - pi*(r-freeboard)^3/3

displaced_mass = water_volume*1000
concrete_mass = (2/3*pi*r^3-2/3*pi*r2^3)*2.5*1000

total_mass = concrete_mass + boy_mass

print(paste0('Outside radius of hemisphere = ',r))
print(paste0('Outside diameter of hemisphere = ',2*r))
print(paste0('water volume     = ',water_volume))
print(paste0('displaced mass   = ',displaced_mass))
print(paste0('concrete mass    = ',concrete_mass))
print(paste0('total craft mass = ',total_mass))

```

We should also be able to do this graphically:

```{r}
spec_gravity=2.5   # Given in problem definition
freeboard=0.10     # Given in problem definition
mass_boy=45        # Given in problem definition

water_density=1000 # kg per m^3 (assumed water density)

r=seq(0.1,1,length.out=200)   # Range of hemisphere radii to be explored

outside_volume=2/3*pi*r^3
inside_volume=2/3*pi*(r-0.025)^3
concrete_mass=(outside_volume-inside_volume)*spec_gravity*water_density
total_mass = concrete_mass + mass_boy

## For bottom part of hemisphere (height = h)
##
##    V = pi*R*h^2 - pi*h^3/3

displaced_volume = pi*r*(r-0.1)^2-pi*(r-0.1)^3/3
buoyancy=displaced_volume*water_density

plot(r,total_mass,type='l',xlab='Radius',ylab='kg - force',xlim=c(0.1,1),ylim=c(0,400))
lines(r,buoyancy,col='red')
legend(0.25,350,c('Total mass','Buoyancy'),lty="solid",col=c("black","red"))
grid()
```

Clearly, for small radii the buoyancy is too small to support the mass of the shell
and passenger.  When the shell gets *large enough* then the boat will float.

To get a meaningful result we would need to zoom in the plot but it appears that
a radius of (approximately) 0.44m is sufficient.


## Problem 4

    In how many different ways can
    eight queens be placed on a chess
    board such that no queen threatens
    another queen and no queen occupies
    a square that is on either major
    diagonal? A solution is considered
    different if it is not a rotation or
    reflection of another solution.
    
    -- David H. Westwood, MN A'43

TODO


## Problem 5

    During a recent game of fivecard-
    draw poker, played with a
    standard deck of cards, I was dealt a
    hand with the following characteristics.
    It contained no aces or face
    cards and had no two cards of the
    same value. All four suits were
    present. The total value of the odd
    cards equaled the total value of the
    even cards. There were no three-card
    straights. The total value of the black
    cards was 10, and the total value of
    the hearts was 14. And the card with
    the lowest value was a spade.
    Precisely what were the five cards?
    
    -- nearly impossible Brain Bafflers
    by Tim Sole and Rod Marshall

Solution:  Note that when I checked the solution I realized my final
solution was correct but in my process I made a mistake.  I missed the
fact that the two Hearts could be 10+4 so I have added that to the
analysis.  There is no feasible hand with these cards so it did not
affect the final solution.

    Hand = 9H, 5H, 2S, 8C, 4D
    
    Details:

    Since there are only 5 cards covering four suits
    there can be no more than 2 cards of any one suit.
    
    All cards in range 2-10.  No pairs.
    
    Since the total value of the hearts is 14 we can
    conclude that there are two hearts, either:
    
        10+4 of Hearts, 9+5 of Hearts, or 8+6 of Hearts
        1 club
        1 spade
        1 diamond
        
    Total value of odd == total value of even
    No three card straights
    Total club+spade = 10, spade is lowest
    
        2 of Spades + 8 of Clubs, or
        3 of Spades + 7 of Clubs, or
        4 of Spades + 6 of Clubs
        
    Assume we have 8+6 of Hearts:  then we have a contradiction.
    Can't have any of the Spade/Club combinations.
    
        2 of Spades + 8 of Clubs --> pair of 8s is disallowed
        3 of Spades + 7 of Clubs --> 6,7,8 straight is disallowed
        4 of Spades + 6 of Clubs --> pair of 6s is disallowed
        
    Assume we have 9+5 of Hearts
        2 of Spades + 8 of Clubs --> feasible
        3 of Spades + 7 of Clubs --> feasible
        4 of Spades + 6 of Clubs --> 4,5,6 straight is disallowed

    Assume we have 10+4 of Hearts
        2 of Spades + 8 of Clubs --> feasible
        3 of Spades + 7 of Clubs --> feasible
        4 of Spades + 6 of Clubs --> pair of 4s is disallowed

    Assume we have 9+5 of Hearts, 2 of Spades, 8 of Clubs then
        Sum of odd == sum of even --> Need 4 of Diamonds --> feasible
        
    Assume we have 9+5 of Hearts, 3 of Spades, 7 of Clubs then
        Sum of odd = 9+5+3+7 = 24 --> not feasible

    Assume we have 10+4 of Hearts, 2 of Spades, 8 of Clubs then
        Sum of even = 26 --> not feasible

    Assume we have 10+4 of Hearts, 3 of Spades, 7 of Clubs then
        Sum of odd == sum of even --> 10 vs. 14 --> not feasible
        This would require 4 more counts for an odd card

    Final Conclusion:
    
        9+5 of Hearts
        2   of Spades
        8   of Clubs
        4   of Diamonds
        
        Heart total = 14
        Black total = 10
        Odd total = 14
        Even total = 14
        No three card straights
        Lowest card is a Spade


## Bonus

    Ann and Bev play a game of
    random tic-tac-toe. They start with a
    tic-tac-toe grid with the boxes
    labeled 1 through 9. Next, they
    shuffle a deck of cards consisting of
    the ace, 2, 3, 4, 5, 6, 7, 8, and 9 of
    spades. They then take turns
    drawing cards, without replacement,
    and placing their initials in the
    squares corresponding to the card
    drawn until a winner is determined.
    If Ann draws first, what is her
    probability of winning, ignoring
    games that end in a draw?
    
    -- The Surprise Attack in
    Mathematical Problems
    by L. A. Graham

Solution:  Consider the tic-tac-toe grid as shown:

    +---+---+---+
    | 1 | 2 | 3 |
    +---+---+---+
    | 4 | 5 | 6 |
    +---+---+---+
    | 7 | 8 | 9 |
    +---+---+---+

    A win occurs when one player has any of the following
    eight combinations:
    
    1-2-3
    4-5-6
    7-8-9
    1-4-7
    2-5-8
    3-6-9
    1-5-9
    3-5-7

    No one can win until they have three numbers.
    
    Ann gets to three numbers first.  There are 84 possible 3 number
    combinations from 9 numbers.  Of these, 8 will win so the
    probability that she wins is 8/84.
    
    Now it gets complicated since the remaining probabilities are
    conditional on Ann failing.
    
    I tried to do this analytically and it was nonsense.

    I suspect Bayes Theorem could be useful - but I will punt.

    What about Monte-Carlo?
    
```{r}
num_trial=10000

win_code = function(s) {
    r=0
    if (all(c(1,2,3) %in% s)) {r=1}
    if (all(c(4,5,6) %in% s)) {r=2}
    if (all(c(7,8,9) %in% s)) {r=3}
    if (all(c(1,4,7) %in% s)) {r=4}
    if (all(c(2,5,8) %in% s)) {r=5}
    if (all(c(3,6,9) %in% s)) {r=6}
    if (all(c(1,5,9) %in% s)) {r=7}
    if (all(c(3,5,7) %in% s)) {r=8}
    return(r)    
}

result=c()
for (i in 1:num_trial) {
    ann=c()
    bev=c()
    s=sample(1:9)
    for (k in 1:9) {
        if (k%%2 == 1) {
            ann=c(ann,s[k])
            r=win_code(ann)
            if (r > 0) {
                result=c(result,1)
                break
            }
        } else {
            bev=c(bev,s[k])
            r=win_code(bev)
            if (r>0) {
                result=c(result,2)
                break
            }
            
        }
    }
}
print(paste0("Ann wins = ",sum(result==1)/length(result)))
print(paste0("Bev wins = ",sum(result==2)/length(result)))
print(paste0("Draws    = ",num_trial-length(result)))
```

So Ann wins about 2/3 of the time and Bev wins 1/3 of the time.
This ignores the draws.  It is sort of interesting that draws
only happen about 13% of the time with random square selection.

They actually solve this analytically by looking at every possible
tic-tac-toe grid and counting the solutions.  The likelihood that
Ann wins (discounting ties) is precisely 0.67.


## Double Bonus

    Using only a pair of
    compasses, locate the midpoint
    between two given points, using
    fewer than 10 compass operations.
    
    -- Elvin Greene, NJ Gamma'52

TODO

I looked at the solution and I could not follow it mentally.
Need to make a drawing and try it.

