---
title: "Bent Summer 2012"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Summer 2012


## Problem 1

    Bingchester is an important junction where
    train lines cross thus: In addition, roads
    parallel the train tracks. A, C, D, and E
    represent stations on the lines whose
    distances from Bingchester by road are: A, 4
    miles; C, 7 miles; D, 10 miles; and E, 7
    miles. The areas between A, C, D, and E are
    covered by impenetrable woods so it is
    impossible for anyone to get from one of
    these places to another without passing
    through Bingchester either by road or by
    train.  An extract from the timetable reads
    thus:

    A dep: 9:15 C dep: 9:10
    B arr: 9:23 B arr: 9:27
    B dep: 9:25 B dep: 9:30
    C arr: 9:44 A arr: 9:38
    D dep: 8:55 E dep: 8:58
    B arr: 9:26 B arr: 9:22
    B dep: 9:29 B dep: 9:25
    E arr: 9:53 D arr: 9:56

    Trains run every 15 minutes, precisely on
    time, so that the time of a previous or
    following train can be determined by
    subtracting or adding 15 minutes.  The most
    famous attraction in Bingchester is the
    Moaning Lisa, hung in the train station.
    Unfortunately, someone has defaced the
    picture by adding a beard and mustache. The
    picture was seen unadorned by several people
    at 9:25:30, and the damage was discovered at
    9:28:30.  Five men are suspected of the
    desecration, and one is certainly guilty. The
    five suspects make the following true
    statements:

    Paul: I saw Ron in A at 9:14 and I was in E at 9:52.
    Quentin: I was in E at 9:01, and in C at 9:58.
    Ron: I was in D at 10:09. I left my bicycle at A.
    Sam: I was in C at 8:56, and in D at 10:03.
    Ted: I was in A at 9:40, and in E at 8:59.

    They all have bicycles which they can peddle
    at a steady speed of 15 m.p.h. They can take
    their bicycles on the train, but no one rides
    anybody else's bicycle. Two of their bicycles
    were found at Bingchester after the defacing.
    Whose bicycles were they? What were the exact
    movements of all five suspects? Who defaced
    the Moaning Lisa?

    -- Brain Puzzler's Delight by E. R. Emmet

TODO

    TLDR
    

## Problem 2

    Solve the following cryptic multiplication,
    where each different letter represents a
    different digit: 

    ABCDEF = BCDEFA x M.

    -- The Crucible

Solution: solved in the cryptic arithmetic notebook

  Cryptic multiplication:

      ABCDEF = BCDEFA * M

    a=make_crypt_df('abcdefm')
    x_abcdef=make_num(a,'abcdef')
    x_bcdefa=make_num(a,'bcdefa')
    x_m=make_num(a,'m')
    k=(x_abcdef == x_bcdefa * x_m)
    k=k & (a$a != 0) & (a$b != 0) & (a$m != 0)
    sum(k)
    [1] 1
    > a[k,]
           a b c d e f m
    559534 9 2 3 0 7 6 4

      ABCDEF = BCDEFA * M
      923076 = 230769 * 4


## Problem 3

    On an analog watch, in less than half a
    second after the precise time that the second
    hand passes one of the twelve hourly marks,
    the minute hand passes over the hourly hand.
    If this occurs before noon, at what time do
    the minute hand and hour hand coincide?

    -- Source Unknown

Solution:

    The hour hand moves 30 degrees per hour.
    The minute hand moves 360 degrees per hour.
    
    Let t = hours after midnight
        H = angle of the hour hand
        M = angle of the minute hand
        n = number of times minute hand has lapped hour hand

    H = t*30
    M = t*360 - 360*n
    
    These angles are the same for the following times:
    
    t = 360/330 * (1:10)
    
    We can convert these to seconds and then human readable times.
    We want the time that is just after an integer number of 5 seconds
    
```{r}
library(hms)

t=(360/330*(1:10))*3600
print(hms(t))
```

The times where the hour and minute hand coincide are:

    01:05:27.272727
    02:10:54.545455
    03:16:21.818182
    04:21:49.090909
    05:27:16.363636
    06:32:43.636364
    07:38:10.909091
    08:43:38.181818
    09:49:05.454545  <-- within 0.5 seconds
    10:54:32.727273

The only time within 0.5 seconds of the second hand reaching
one of the hour marks is 09:49:05.454545.


## Problem 4

    What is the minimum number of people, born on
    random days in 1981, that need to be in a
    room to have at least 50% probability that
    either there are (at least) two pairs of
    persons who have the same birthday or there
    are three (or more) people who share a common
    birthday or both?

    -- D.A. Dechman, TX A'57

Solution:

    Sounds like Monte-Carlo is easier than analytics.
    We use a 365 day year.
    
    After a couple of quick runs we will make 100K trials
    at each group size in the region of the solution:
    
```{r}
ntrials=100000
set.seed(42)

x=c()
y=c()
t1=Sys.time()
for (group_size in 30:40) {
    n=0
    for (trial in 1:ntrials) {
        s=sample(365,group_size,replace=T)
        t=rev(sort(table(s)))
        if ((t[1]>2) | (t[2]>1)) {
            n=n+1
        }
    }
    x=append(x,group_size)
    y=append(y,n/ntrials)
}
t2=Sys.time()
plot(x,y,type='b',
     xlab="Group Size",
     ylab="Probability",
     main="Birthday Party: Triple or Multiple Pairs")
abline(h=0.5,col='red')
grid()
print(t2-t1)
a=data.frame(n=x,pr=y)
m=lm(pr~n,a)
n_min=(0.5-m$coefficients[1])/m$coefficients[2]
print(paste0("Min group size for Pr=0.5 is ",n_min))
a
```

Using the results of the simulation and a linear regression we can
predict the required number as:

    n = (0.5 + 1)/0.01277 ~ 117.4 people

Well -- the Bent says it is 36 so this isn't even close.  Did
I misunderstand the problem?  Yup.  I was solving for groups larger
than 3 and larger than pairs.  The value above is for a single
group of 4 (or more) or two (or more) groups of 3 or both.

Repeating with the correct logic

    n = (0.5 + 0.5939)/0.03067 ~ 35.7 people



## Problem 5

    What is the minimum number of knights that
    can be placed on a standard 8x8 chess board,
    so that every square (including those
    occupied by knights) is threatened by a
    knight and what is such a configuration? A
    square is threatened if a knight can move to
    that square on its next move. A knight moves
    two squares in one direction and one square
    in another direction (perpendicular to first
    direction) to end up on a square of opposite
    color. The move can occur even if intervening
    squares are occupied. Present your answer as
    an 8x8 grid with "o" representing an
    unoccupied square and "N" representing a
    square occupied by a knight. Hint: the
    minimum is less than 16.

    -- Amusements in Mathematics by H. E. Dudney

TODO

    Didn't do it but would probably just try
    exhaustive search with increasing numbers
    of Knights.
    
    Bent claims you need 14 in one of three
    fundamental configurations.


## Bonus

    A rigid, uniform rod one meter long is held
    leaning against a frictionless, vertical wall
    at a 30 degree angle to the horizontal by a
    peg at point A (where the stick touches the
    ground) on a frictionless, horizontal
    surface. If the peg is suddenly removed, what
    is the horizontal speed of the center point
    of the rod as it passes point A?

    -- Allan Gottlieb's Puzzle Corner in Technology Review

Solution:

    It should be possible to determine the energy
    balance -- just not sure if the wall does work
    on the rod or if it is just a conversion of potential
    to kinetic energy.
    
    The Bent solution is rather long -- I will try
    to see if it can be simplified.  The final answer
    should be 36.9 cm/sec.
    
    use:  m = mass/unit length
          g = 980 cm/s^2

    As the rod slides down the wall it loses potential
    energy and gains kinetic energy.  If we ignore the
    work done by the wall we have:
    
    PE = L*m*g*(1/2*L*1/2) = L^2*m*g/4
    
    KE = 1/2*(m*L)*v^2
    
    Equating we have: ... a mess that doesn't work ...
    
    I am breaking down and looking at the Bent solution.
    I want to work through this because this seems like the
    kind of problem I should be able to solve but it has been
    a very long time since I took Dynamics...
    
    ... I should try to solve this one of these days ...
    

## Computer Bonus

    In how many ways (order matters) can 14
    married couples be seated in chairs numbered
    consecutively 1 to 28 about a round table in
    such a manner that there is always one man
    between two women and none of them is ever
    next to his own wife?

    -- 100 Great Problems of
    Elementary Mathematics
    by Heinrich Dorrie

TODO

    I looked at the solution.  They make the observation
    that there are 14! ways to seat the women in the even
    numbered chairs and 14! ways in the odd.  Pick one of
    the arrangement.
    
    For any particular arrangement - they used a computer
    program to calculate 10,927,434,464 ways to arrange the
    men for each configuration of women.
    
    The total = 2*14!*10927434464 = 1905270127543015833600
        ~ 1.90527e21


