---
title: "Bent Spring 2006"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Spring 2006

The prequel has a plea to stop using the computer and
start exercising the gray cells.

## Problem 1

    A 10-digit number consists of two
    4s, two 3s, two 2s, two 1s, and two 0s.
    The 4s are separated by four digits,
    the 3s by three digits, the 2s by two
    digits, the 1s by one digit, and the 0s
    by no digits. What are the smallest
    and largest such numbers? Leading
    zeros are not allowed.

    -- Adapted from The Mathematics Teacher

Solution:

        1 2 1 4 2 3 0 0 4 3  --> smallest number
        4 1 3 1 2 4 3 2 0 0  --> largest number


    These are like rigid puzzle pieces that must
    fit together.  The pieces are:
    
        4 - - - - 4
        3 - - - 3
        2 - - 2
        1 - 1
        0 0
        
    In attempting to make the largest number we will
    start with 4.  In this case, the next digit can
    not be 3 since the trailing 3 will fall on top of the
    trailing 4.  Let's try a 2 instead:
    
        4 2 - - 2 4

        3 - - - 3
        1 - 1
        0 0

    Unfortunately, there is no way forward.  The 00 could
    fill the gap but then the 3---3 makes for an 11 digit
    number.
    
    Try 41:
    
        4 1 - 1 - 4

        3 - - - 3
        2 - - 2
        0 0

    The 3 can fill the first empty slot:
    
        4 1 3 1 - 4 3
        2 - - 2
        0 0

    The 2 can fill the next slot:
    
        4 1 3 1 2 4 3 2
        0 0

    The 00 can be placed at the end:
    
        4 1 3 1 2 4 3 2 0 0  --> largest number
        
    In attempting to make the smallest number we will
    start with 1.  In this case the next digit can not
    be 0 since it will overlap with the trailing 1. so try 2:
    
        1 2 1 - 2

        4 - - - - 4
        3 - - - 3
        0 0
    
    The empty slot can not be 0 or a 3.  If it is 3 then the
    string starting with 4 will result in 11 digits. So:
    
        1 2 1 4 2 - - - 4

        3 - - - 3
        0 0
    
    Filling in the final digits:
    
        1 2 1 4 2 3 0 0 4 3 --> smallest number


## Problem 2

    Lay out a line of length 2 along
    the x-axis of a coordinate system,
    with the center of the line at the
    origin. Above this line, construct an
    isosceles triangle with the line as a
    base and base angles of 75 deg. Below
    the line construct another isosceles
    triangle (with its vertex pointing
    down) with the line as a base and
    base angles of 45 deg. What is the radius
    of the inscribed circle of the lanceolate
    figure formed by the sides of the
    two triangles, and where is its center
    (distance from the origin along the yaxis)?
    Express your answer exactly
    in terms of surds.

    -- Technology Review

Solution:  need a picture to visualize this:

    Symbolically:
    
        y0 = (2*sin(75)-sqrt(2))/(2*cos(75)+sqrt(2))
           = 2-sqrt(3)
           
        d  = sqrt(2)*(sin(75)+cos(75))/(sqrt(2)+2*cos(75))
           = (3-sqrt(3))/sqrt(2)

    Numerically:
        y0 = 0.2679492
        d = radius = 0.8965755


    Height/1 = tan(theta)
    Height = tan(theta)
    
    Upper triangle has height of tan(75)
    Bottom triangle has height of -tan(45)


```{r}

# Make an empty region to plot on
plot(c(-3,3),c(-2,4),type='n',asp=1)

# Base of the triangle
lines(c(-1,1),c(0,0),lwd=3)

# Upward pointing triangle
y_top=tan(75*pi/180)
lines(c(-1,0),c(0,y_top))
lines(c(1,0),c(0,y_top))

# Downward pointing triangle
y_bot=tan(45*pi/180)
lines(c(-1,0),c(0,-y_bot))
lines(c(1,0),c(0,-y_bot))

# Draw in the circle after solving the problem
if (T) {
    x0=0
    y0 = 2-sqrt(3)
    r = (3-sqrt(3))/sqrt(2)

    theta=seq(0,2*pi,length=100)
    x=x0+r*cos(theta)
    y=y0+r*sin(theta)
    polygon(x,y,border='red',lwd=2)
}

```

Need to find a point that is equi-distance from the long sides of
the two triangles.  By symmetry this is somewhere along the y-axis.
The distance to the lines will then be the radius of the inscribed
circle.

Given a point: x0,y0 and a line y=m*x+b what is the distance from the
point to the line?

[Wikipedia has the answer](https://en.wikipedia.org/wiki/Distance_from_a_point_to_a_line)

    Turns out this is a bit ugly.
    
    if a*x+b*y+c = 0 then the distance to (x0,y0) is:
    
        d = |a*x0+b*y0+c|/sqrt(a^2+b^2)

    We know the center of the circle is at (0,y0):

    For the upper triangle, left side:
    
        y = tan(75)*x + tan(75) --> tan(75)*x + (-1)*y + tan(75) = 0
        
        d = |tan(75)*0 + (-1)*y0 + tan(75)|/sqrt(tan(75)^2+1)
          
          = (3.732051 - y0) / 3.863703

    For the lower triangle, left side:
    
        y = -tan(45)*x - tan(45) --> tan(45)*x + (1)*y + tan(45) = 0
        
        d = |tan(45)*0 + (1)*y0 + tan(45)|/sqrt(tan(45)+1)

          = (1+y0) / sqrt(2)
          
    These two must be the same:
    
        (3.732051 - y0) / 3.863703 = (1+y0) / sqrt(2)
        
        y0 = (3.732051/3.863703-1/sqrt(2))/(1/sqrt(2)+1/3.863703)
           = 0.2679493
           
    The radius of the circle is just the distance to the line:
    
        d = radius = 0.8965756
        
    Doing this symbolically instead:
    
        y0 = (2*sin(75)-sqrt(2))/(2*cos(75)+sqrt(2))
           = 2-sqrt(3)
           
        d  = (1+2-sqrt(3))/sqrt(2)
           = (3-sqrt(3))/sqrt(2)


## Problem 3

    A and B each put 10 coins in the
    pot. A then takes a coin from the pot
    and tosses it, while B calls heads or
    tails. If B's call is correct, he takes
    the coin and keeps it; otherwise, A
    keeps the coin. B then takes a coin
    from the remaining pot and tosses it
    while A calls heads or tails. The play
    continues in this way until one player
    has accumulated 10 coins, whereupon
    he wins the game and takes all the
    coins remaining in the pot. At one
    point in the play, A has six coins,
    while B has only four. What is A's
    probability of winning? When the
    situation occurs in a game such that
    one player has won six coins while
    his opponent has won four coins,
    what is the expected value of the
    winnings (net number of coins) of the
    player with six coins?

    -- Attributed to the Chevalier de Mere,
    a 17th century gambler

Solution:

    The pot initially has 20 coins.
    
    When A=6, B=4 there are 10 coins remaining.
    
    What is the probability that A reaches 10 first?
    
    We will try Monte-Carlo and then think about analytics:
    
```{r}
ntrials=100000

win=0
win_sum=0
for (trial in 1:ntrials) {
    A=6
    B=4
    while(T) {
        s=sample(c(0,1),1)
        if (s==0) {
            A=A+1
            if (A==10) {
                win=win+1
                win_sum=win_sum+(20-A-B)
                break
            }
        } else {
            B=B+1
            if (B==10) {
                break
            }
        }
    }
}
print(paste0("Pr(win) = ",win/ntrials))
print(paste0("Mean(winnings) = ",win_sum/win))
```

Monte-Carlo results:

    Pr(win) = 0.74501
    Mean(winnings) = 3.320103

Note that these winnings only occur if A actually wins.  The times when
A does not win the winning are 0.  Weighting this gives:

    0.74501*3.320103 = 2.4735 ???
    
According to the Bent this should be 2.000.
The Bent does everything using combinatorics.


## Problem 4

    You have two two-liter bottles,
    one (marked W) containing one liter
    of pure water and one (marked A)
    containing one liter of pure alcohol.
    The two things you can do are to
    pour liquid from either bottle down
    the drain or pour liquid from one
    bottle into the other, but you must
    always keep at least one liter of liquid
    in bottle A. What is the minimum
    alcohol concentration that can be
    achieved in bottle A? Assume that
    there is no volume change on mixing
    water and alcohol.

    -- Daryl Cooper

Solution:

    If we add 1/m L Water to Alcohol and then dump the
    excess A+W back to 1L each time:
    
    In the limit as m->infinity this approaches exp(-1).


    The obvious answer is you can get 50% by pouring
    all the water into the alcohol.  Of course this
    is quite boring.
    
    What if we try some things:
    
    1. Add half the water to alcohol
       Throw away 1/3 of the A+W mixture
       Remaining alcohol = 2/3
       Add the remainder of the water:
            Alcohol concentration =  (2/3)/(1+1/2) = (2/3) / (3/2) = 4/9 < 1/2
            
    2. Add 1/4L the water to the alcohol
       Throw away 1/5 of the A+W mixture
       Remaining alcohol = 4/5
       
       Add 1/4L water
       throw away 1/5 of the mixture
       Remaining alcohol = (4/5)^2
       
       then (4/5)^3 finally (4/5)^4 = 0.4096
       
     ...
     
     m=integer
     
     In general, add 1/m L water to alcohol
     Remove 1/m L of the mixture --> remaining is m/(m+1)
     ...
     repeat this m times
     ...
     Final result = (m/(m+1))**m

    In the limit as m->infinity this approaches exp(-1)


## Problem 5

    A modern aluminum sculpture
    consists of a hollow cylinder that is
    capped on one end by a solid hemisphere.
    The cylinder has an outer
    diameter of 100 cm and thickness of
    1 cm, and the hemisphere has the
    same diameter as the outside of the
    cylinder. If, on a level surface, the
    sculpture balances in stable equilibrium
    at any point on its hemispherical
    surface, how long is the cylinder, and
    what is the minimum ceiling height
    in the museum to permit the sculpture
    to assume any stable position?

    -- R. Wilson Rowland, MD B '51

Solution:

    Cylinder height: L = 177.667 cm
    The maximum distance will be when they line up = 234.569 cm.
    This is the minimum ceiling height.


    As the sculpture is tipped to one side ...
    The balance point of the hemisphere and the floor
    can be used as the reference.  The amount of mass
    the left/right of this reference must be equal.
    Clearly, the easiest condition to understand is
    when the cylinder is perfectly vertical.  This will
    be stable and will define the minimum ceiling
    height.
    
    Need a picture to show this.

    When the scupture is tilted to one side will this
    really be a stable position?  Iff the mass balance
    works out.  The unstable part comes if the center
    of mass goes outside the possible balance point.
    
    Where is the center of mass of a hemisphere?
    
    If we measure from the flat side of the hemisphere then
    the center of mass is at: 3/8*R
    
    The mass of the hemisphere is: rho*2/3*pi*R**3
    
    We can think of all the mass of the hemisphere to be
    located at precisely this location.
    
    When the cylinder is tilted at an angle of theta from
    the vertical the center of mass is moved off center
    by:
    
        3/8*R*sin(theta)
        
    This creates a moment of mass * distance:
    
        M_hemisphere = rho*2/3*pi*R**3 * (3/8)*R*sin(theta)

    Meanwhile, the cylinder has defined radii with a total mass
    of:
    
        Mass = pi*(R**2-(Rinner**2))*rho*L
        Center of Mass = L/2
        
        When inclined by theta: Moment = L/2*sin(theta)*rho*L*pi*(R**2-(Rinner**2))
        
    This will balance when:
    
        2/3*R**3*(3/8)*R = L/2*L*(R**2-Rinner**2)

        1/4*R**4 = 1/2*L**2*(R**2-Rinner**2)
        
        R=100 cm / 2 = 50 cm
        Rinner = 49 cm
        
        L = 177.667 cm
        
    The longest possible dimension is from the outside of the hemisphere to an
    outside upper edge of the cylinder.
    
    From the point where the center of the cylinder joins the hemisphere the distance
    to the outside upper edge of the cyclinder is:
    
        sqrt(50^2 + 177.667^2) = 184.569 cm
        
    The distance from this point to the outside of the hemisphere = 50 cm.
    
    The maximum distance will be when they line up = 234.569 cm.
    
    If you want to allow for any angle you need to support this height.
    
## Bonus

    Select three points at random
    from within a right circular cone
    of altitude h. What is the expected
    value of the distance of the closest
    point from the base?

    -- John W. Langhaar, PA A '33

Solution:

    I think they multiplied the derived probability by three to
    account for the fact that there are three possible points that
    might be the lowest.  It is the only thing that makes sense.

    Given a height, y, above the base of the cone, the volume above
    that height is:

        Vtop = 1/3*pi*(R*y/h)**2*(h-y)

    The total volume is:
    
        V = 1/3*pi*R**2*h
        
    The ratio is the probability of picking a point in the top volume:
    
        Pr = Vtop/V = (y/h)**2*(h-y) / h

    Given an arbitrary set of altitudes: y1, y2, y3
    
    What is the probabiity that y2>y1 and y3>y1?
    
        YUCK!
        
Here is what they did in the Bent:

    Consider a (differential) circular disk at a height h above the base.
    
    The probability of a point being in this disk is:
    
        Pr = pi*(R*(h-y)/h)**2*dy / (1/3*pi*R**2*h)
           = 3*(h-y)**2/h**3
        
    So -- the probability of being between y and the top of the cone is:

        Pr = integrate(3*(h-tau)**2/h**3,(tau,y,h))
        Pr =  (h-y)**3 / h**3
           =  (1-y/h)**3
           
    The probability of two points being higher:
    
        Pr = (1-y/h)**6
        
    Total probability is then:
    
        Pr(point is in the disk)*Pr(both other points are higher)
        
        = 3*(h-y)**2/h**3 * (1-y/h)**6
        
    The expected value is:
    
        integrate(y * 3*(h-y)**2/h**3 * (1-y/h)**6, (y,0,h))
        
        = h/30
        
    This does NOT match the result from the Bent: h/10
    
    I am going to hand-wave that there are three points so this Pr applies
    to all three.  The total probability is then:
    
        3*Pr = h/10
    
    The Bent makes the following derivation:
    
    differential slice of the cone with diameter = x at distance of y from base
    
    Pr(in slice) = pi*(x/2)**2*dy / (1/3*pi*(D/2)**2*h)
                 = 3*x**2/ (D**2*h) * dy
                 
    Also: x/D = (h-y)/h --> x=D*(h-y)/h = D*(1-y/h)
    
    Pr (point being above y) = x**2 * (h-y)/(D**2 * h) (ratio of two cones)
    
        = (1-y/h)**3 (substitute for x = f(y))
        
    For two points: Pr = (1-y/h)**6
    
    Expected value is then (presented without any explanation):
    
        E = (9/D**2) * integrate(x**2*y*(1-y/h)**6,(y,0,h))
        
    I understand Pr(in slice) = 3*x**2/(D**2*h)
    
    I would write this based on results above as:
    
        integrate(3*x**2/(D**2*h)*y*(1-y/h)**6,(y,0,h))
        
        = integrate(3/h * (x/D) **2 * y * (1-y/h)**6, (y,0,h))
        = integrate(3/h * (1-y/h)**2 * y * (1-y/h)**6, (y,0,h))
        = integrate(3/h * y * (1-y/h)**8,(y,0,h))
        = h/30
        
        which is the same result I got.
        
    I think they snuck in the factor of 3 without any explanation.

I can try to verify this by Monte-Carlo.  We will arbitrarily use a cone with
base radius = 1 and height = 1.

```{r}
ntrials=10000

# Sample random points in x,y until we get a point inside the cone
# Do this until we get 3 points.

zzz=c()
for (trials in 1:ntrials) {
    # Get three points
    x=c()
    y=c()
    n=0
    while(T) {
        x0=runif(1); y0=runif(1)
        if ((1-y0)>=x0) { x=c(x,x0); y=c(y,y0); n=n+1}
        if (n==3) { break }
    }
    zzz=c(zzz,min(y))
}
hist(zzz)
print(summary(zzz))
```

Interestingly, the median is about h/10 (0.1097) but the mean is 
signficiantly higher (0.14312) due to long upper tail.

So -- what the heck does expected value really mean?


## Double Bonus

    For many years,
    the Roman Catholic church used the
    Julian calendar, which has a leap year
    every year that is divisible by four,
    making the average calendar year
    longer than the sidereal year and
    causing the date of the first day of
    spring to change gradually. To correct
    this, Pope Gregory decreed that
    Thursday, October 4, 1582, would be
    followed by Friday, October 15. He
    also declared that years divisible
    by 100 would be leap years only if
    divisible by 400. For any year since
    1582, if one printed two 12-month
    calendars, one Julian and the other
    Gregorian, with dates for the days
    of the month, at least some of the
    dates would not fall on the same day
    of the week. What is the first year
    for which each day of each month will
    fall on the same day of the week for
    both calendars?

    -- Colin Singleton in New Scientist


TODO

I don't really care but the Bent explanation is
that the two calendars will be in sync iff the
number of days between them is divisible by 7.

They were forced 10 days apart in 1582.

The two calendars are the same except in years
that are divisible by 100 (by not 400).  This
occurs in 1700,1800,1900, and 2100 to make the
total number of days different = 14.  So, in
the first non-leap year (2101)  the two will
be in sync and stay there until 2200.

