---
title: "Bent Spring 2016"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Spring 2016


## Problem 1

    Solve the following cryptic multiplication.
    Each different letter stands for a different
    digit, and each different digit is always
    represented by the same letter; there are no
    leading zeros.  TRY*THIS = TICKLER

    -- H. G. McIlvried III, PA Gamma'53

Solution: solved in cryptic arithmetic notebook

      TRY * THIS = TICKLER

    a=make_crypt_df('trythistickler')
    k=(make_num(a,'try')*make_num(a,'this') == make_num(a,'tickler'))
    k=k&(a$t != 0)
    sum(k)
    [1] 1
    a[k,]
            t r y h i s c k l e
    3620205 9 8 6 2 0 3 7 4 1 5

    TRY = 986
    THIS = 9203
    TICKLER = 9074158

## Problem 2

    What is the largest integer n such that
    7^2048 - 1 is exactly divisible by 4^n?

    -- 2015 Quant Quiz, National Museum of
       Mathematics

Solution:

    Should be able to brute force this with gmp.
    
```{r}
library(gmp)

n1=as.bigz(7)^2048-1

for (n in 1:2874) {
    n2=as.bigz(4)^n
    if (n1%%n2 == 0) { maxn=n }
}
print(sprintf("Maximum value of n = %d",maxn))
```

The maximum value of n = 7


## Problem 3

    On their birthday, twins Jackie and Jill
    received from their uncle a set of 12 wooden
    cubes and a two-pan balance. Each cube was a
    different color and integral number of cm (1
    through 12 cm) on a side, but they all were
    the same density. The colors of the cubes
    were azure, brown, crimson, drab, ebony,
    fuchsia, gold, heliotrope, ivory, jade,
    khaki, and lilac, not in order of size. Their
    uncle challenged the twins to find
    combinations that balanced some of the cubes
    against others. Four of the balances the
    twins discovered were: (1) the drab and gold
    cubes balanced the brown and heliotrope
    cubes; (2) the khaki cube balanced the azure,
    ebony, and lilac cubes; (3) the ebony, gold,
    jade, and khaki cubes balanced the azure,
    brown, drab, fuchsia, heliotrope, and ivory
    cubes; and (4) the brown cube balanced the
    drab, ivory, and khaki cubes. When they were
    called in to supper, they ran in quickly,
    forgetting to take their cubes with them. It
    rained that night and washed the colors off
    all the cubes. The twins were very unhappy
    and asked to have the cubes repainted in the
    same way as they were originally. Match the
    size of each cube to its original color.

    -- Madachy's Mathematical Recreations by
       Joseph S. Madachy

TODO

    meh


## Problem 4

    On their next birthday, the twin's uncle gave
    them a big pile of unpainted 1-cm cubes and
    suggested they glue some of the cubes
    together to make larger cubes. Jill quickly
    produced a larger cube (without any voids).
    Not to be outdone, Jackie formed an even
    larger cube (also without any voids). Then,
    each twin painted some of the faces of her
    large cube. As happened the previous year,
    when they were called in to supper, they left
    their cubes outside, and the next morning
    found that an overnight rain had dissolved
    the glue (but had not washed off any paint).
    Each twin examined the pile of small cubes
    that had made up her large cube, and each
    found that exactly 100 of her small cubes had
    no paint on any of their faces. How large was
    each twin's cube, and how many of its faces
    had each twin painted?

    -- Adapted from Mathematical Puzzling by A.
       Gardiner

TODO

    meh


## Problem 5

    You have a copy machine that lets you scale
    your original from 100% (no enlargement) to a
    maximum of 155% in increments of 1%. How can
    you use this machine to produce a copy of a
    square picture such that the length of the
    side of the final copy is exactly double the
    length of the side of the original? (A
    setting of xxx% means that the ratio of the
    length of the side of the copy to the length
    of the side of the original is x.xx.)

    -- Mathematics Teacher

Solution:  can try exhaustive search with 2, 3, 4, ...
enlargements.

    Apparently you can not shrink the size - just
    expand - so we need to look for products.

```{r}

n=c()
for (a in 100:154) {
    for (b in (a+1):155) {
        d=a*b
        if ((d %% 10000) == 0) {
            n=append(n,c(a,b))
        }
    }
}
print(n)
```

    Well -- two attempts won't work.
    Let's try more.  May need to go recursive.
    Not really sure how to approach this problem.
    Let us try three
    
```{r}
n=c()
for (a1 in 100:155) {
    for (a2 in 101:155) {
        for (a3 in 101:155) {
            d=a1*a2*a3
            if (d==2000000) {
                n=append(n,c(a1,a2,a3))
            }
        }
    }
}
n=matrix(n,ncol=3,byrow=T)
print(n)

```

Apparently it can be done in three tries: 125%, 125% and 128%.


## Bonus

    The number
    
    365 = 13^2 + 14^2 = 10^2 +11^2 +12^2
    
    is the smallest integer which is both
    the sum of the squares of two consecutive
    positive integers and the sum of the squares
    of three consecutive positive integers. The
    second smallest such integer is 
    
    35,645 = 133^2 + 134^2
           = 108^2 + 109^2 + 110^2. 
           
    What is the tenth smallest such integer
    and what is the exact ratio of the
    (n+1)st such integer to the nth such
    integer as n gets very large? We want the
    ratio in terms of simple mathematical
    functions, such as surds. A computer is
    not necessary, except to do the math.

    -- Time Travel and Other Mathematical
    Bewilderments by Martin Gardner

Solution:

    Exhaustive search.
    
    We can generate candidates as: n^2+(n+1)^2+(n+2)^2 = i
    Check to see if this is the sum of consecutive squares.
    
    This is taking a really, really long time.  Still only have the
    first seven after ~ 60 minutes of running:
    
    365              = 5 73
    35645            = 5 7129
    3492725          = 5 5 139709
    342251285        = 5 1321 51817
    33537133085      = 5 29 29 1061 7517
    3286296790925    = 5 5 337 2909 134089
    322023548377445  = 5 149 432246373661


    About the only pattern I see is that they are all multiples of 5.  They also
    have only a few prime factors but not sure if that is relevant.
    
    Eventually I gave up and looked up the solution.  The key is writing the
    expressions properly, reorganizing them, and then recognizing that this
    is a Pell equation:
    
    N = n^2 + (n+1)^2 = (m-1)^2 + m^2 + (m+1)^2
    
        (2*n^2 + 2*n + 1 = 3*m^2 + 2) * 2
      
        (4*n^2 + 4*n + 1) + 1 = 6*m^2 + 4
      
        (2*n+1)^2 + 1 = 6*m^2 + 4
      
        (2*n+1)^2 - 6*m^2 = 3
      
    Let p = 2*n+1, then:
    
        p^2 - 6*m^2 = 3  which is a Pell equation.
        
[Pell Equation on Wikipedia](https://en.wikipedia.org/wiki/Pell%27s_equation)

    Unfortunately, all of these results apply to p^2-6*m^2=1.
    The three on the RHS makes this a problem.
    
    Considering p^2-6*m^2=1:
    
    There are tables of primitive solutions for this equation.
    Once the primitive solution is known there are recurrence
    relations to generate infinite additional solutions:
    
    The smallest solution is: p=5, m=2
    
    Additional terms may be solved from the relationship:
    
    p(k+1) = p(1)*p(k)+n*m(1)*m(k)
    m(k+1) = p(1)*m(k)+m(1)*p(k)

    These generate the terms:
    
    p(k)      m(k)      n(k)
    ---------------------------
    5         2         2
    49        20        24
    485       198       242
    4801      1960      2400
    47525     19402     23762
    470449    192060    235224
    
    Now we nee to figure out how to apply this to the real problem:
    
[A more general Pell Equation](https://math.stackexchange.com/questions/1719280/does-the-pell-like-equation-x2-dy2-k-have-a-simple-recursion-like-x2-dy2)

    If we have:
    
        X^2 - d^Y^2 = k, k != 1 then we need to find a solution.
        
    After that we can solve the related Pell equation:
    
        X^2 - d*Y^2 =1 for (u,v) then
        
    We can recurse the first solution as:
    
        x(n+1) = u*x(n) + d*v*y(n)
        y(n+1) = v*x(n) + u*y(n)

    In our case we have a solution of the first equation:
    
    p=3, m=1 (any solution - found by trial and error)
    
    For the second equation we have u=5, v=2 (from Wikipedia)
    
    So:

        p(2) = 5*3 + 6*2*1 = 27
        m(2) = 2*3 + 5*1   = 11
        N = (m-1)^2 + (m^2) + (m+1)^2 = 365
        
    and:
    
        p(3) = 5*27 + 6*2*11 = 267
        m(3) = 2*27 + 5*11 = 109
        N = 35645
        
    and
    
        p(4) = 5*267 + 6*2*109 = 2643
        m(4) = 2*267 + 5*109 = 1079
        N = 3492725
        
    etc.
    
    Tabulating yields:
    
    N
    -------------------------
    1  365
    2  35645
    3  3492725
    4  342251285
    5  33537133085
    6  3286296790925
    7  322023548377445
    8  31555021444198565
    9  3092070077983081805
    10 302991312620897818205

    The algebraic ratio is harder to explain so I will just adopt the Bent
    notation:
    
        p0^2 - 6*m0^2 = 1   (Pell's equation)
        
    and
    
        r^2 - 6*s^2 = 3   (the generalized equation)
        
    To generate additional solutions:
    
        (p[0]+m[0]*sqrt(6))^n * (r+s*sqrt*6)) = (p[n]+m[n]*sqrt(6))
        
        Where p[0]=5, m[0]=2, initial state r=3, s=1

    For the next (r,s) in the solution sequence we have:
    
        Ratio = (p[0]+m[0]*sqrt(6))^(n+1) * (r+s*sqrt(6)) / (p[0]+m[0]*sqrt(6))^(n)
        
              = (p[0]+m[0]*sqrt(6)) * (r+s*sqrt(6))
              
              = (5 + 2*sqrt(6)) * (r+s*sqrt(6))
              
    This is the ratio for m[n]/m[n+1].  The number we are finding is N ~ m^2 so
    
        Ratio^2 = 49 + 20*sqrt(6) ~ 97.98979485566355

    This problem was hard!

Preserved for prosperity but it would never have finished before the expiration of
the life of the universe!

Do not actually try to run this.

```{r, eval=FALSE}
library(gmp)

isqrt=function(n) {
    xold=as.bigz(n)
    while(T) {
        xnew = as.bigz((xold+as.bigz(n/xold))/2)
        if (xnew == xold) { break }
        if (xnew == (xold+1)) {
            break
        }
        xold=xnew
    }
    return(xnew)
}


nlist=c()
n=as.bigz(1)
while(T) {
    n1=n^2+(n+1)^2+(n+2)^2
    n2=as.bigz((isqrt(2*n1-1)-1)/2)
    if (n2^2 + (n2+1)^2 == n1) {
        nlist=append(nlist,n1)
        print(nlist)
        if (length(nlist)>=10) { break }
        n=n*9.88
    } else {
        n=n+1
    }
}
```


## Computer Bonus

    What are the smallest and next smallest bases
    for which the integer ba exactly divides
    abcde? For bases 11 and higher, letters are
    used to represent digits greater than 9, that
    is, a is used to represent 10, b to represent
    11, c for 12, etc. Present your answer as
    base 10 integers.

    -- Don A. Dechman, TX A'58

TODO

    Never would have found this.
    
    The bases are 1182 and 14194.
