---
title: "Bent Winter 2010"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Winter 2010

As a note -- these problems seem remarkably familiar.
I think I may have worked these before.

## Problem 1

    Al's job is testing bowling balls. He
    has two identical bowling balls and
    is to test their impact resistance by
    dropping them out of windows on various
    floors of a 100-story building. He
    is to determine from which exact floor
    a dropped bowling ball will shatter
    on impact with the pavement below.
    Al knows nothing about the strength
    of the balls. They may shatter when
    dropped from the first floor or not
    until dropped from the 100th floor.
    What is the minimum number of ball
    drops needed to guarantee that Al
    can uniquely determine the floor from
    which the balls will shatter? Balls
    that do not shatter may be dropped
    again. Both balls may be destroyed
    during the test. Include a brief outline
    of how the testing is done.

    How to Ace the Brain Teaser
    Interview by John Kador

Solution:

    I think this is the problem where you
    want to minimize the maximum penalty.
    Turns out you go up 14 then 13 ...
    such that the total number of potential
    drops is minimized (and constant).

    If you start at 14 and the ball shatters
    then you need to go back to 1, 2, ... 13
    for a total of 14 drops.
    
    If you start at 14 and the ball survives,
    then you have used up one drop so you go
    up to 27 and repeat the process.
    
    Each time the maximum exposure is 14.
    
    The starting floors for a really tall
    building are:
    
        n, 2*n-1, 3*n-2, 4*n-3, 5*n-4,
        6*n-5, 7*n-6, 8*n-7, 9*n-8, 10*n-9,
        11*n-10,
        
    If n = 12 we get stalled at 78 floor
    If n = 13 we get stalled at 91 floor
    If n = 14 we make it to 105 floor
    
    So the smallest n that reaches the top
    of the 100 floor building is 14.


## Problem 2

    Our local Soggy Center Donut
    shop makes six different kinds of
    doughnuts -- namely barbeque, garlic,
    pepperoni, jalapeno, broccoli, and onion.
    Each day I stop and buy a different
    selection of a dozen doughnuts.
    If I always buy at least one of each
    different kind, how many days will it
    take to exhaust all the possibilities?
    Assume that the shop always has at
    least seven of each different kind on
    hand.

    -- Adapted from Introductory Combinatorics
    by Richard A. Brualdi

Solution:

    Lots of room for error here -- but I'll try.
    
    The most you can get of any one type is 7 since
    you need to get at least one of each flavor.
    
    I could try exhaustive enumeration.  This is only
    6 nested loops of 1:7 = 7^6 = 117649 possible cases.
    The total for the 6 loops has to be 12.
    
```{r}
n=0
for (n1 in 1:7) {
    for (n2 in 1:7) {
        for (n3 in 1:7) {
            for (n4 in 1:7) {
                for (n5 in 1:7) {
                    for (n6 in 1:7) {
                        if ((n1+n2+n3+n4+n5+n6) == 12) {n=n+1}
                    }
                }
            }
        }
    }
}
n
```

There are 462 combinations

The run-time is almost instantaneous

    Can I do this analytically?
    
    Assume n1=1 then there are 11 donuts to be split between
    5 different flavors.
    
    choose(11,5) = 462 --> same as above
    
    Is this just luck?


## Problem 3

    If we write an integer in the decimal
    system, its representation either
    contains at least one digit 5 or it does
    not. Find the smallest and the largest
    values of N such that for the integers
    between 1 and N inclusive, exactly
    half contain at least one digit 5.

    -- Adapted from Keys to Infinity
    by Clifford A. Pickover

Solution:  we can just count them?

Turns out that brute force is not a good solution method.
We can find the lower limit that way but not the upper.

Ooops - I think I misread the problem.  It says EXACTLY half.
This search is lengthy but not ridiculous.


```{r}
n=0
ilow=0
ihigh=0
for (i in 1:10000000) {
  n=n+grepl('5',i)
  if (n==(i/2)) {
    print(paste0("i = ",i))
    if (ilow == 0) { ilow = i }
    ihigh=i
  }
}
print(paste0("Low = ",ilow))
print(paste0("High = ",ihigh))
```

Over the range of 1-10e6 the limits are:

    Low  =  590488
    High = 5314408


## Problem 4

    Given that TEN is one more than
    a perfect square that is divisible by 9,
    NINETY is divisible by 9, and there
    are SIX perfect squares between
    TEN and NINETY, what is the
    value of SENT? The usual rules for
    cryptics apply.

    -- Susan Denham in New Scientist

Solution: solved in the cryptic arithmetic notebook

      TEN is one more than a perfect square that is divisible by 9
      NINETY is divisible by 9
      SIX perfect squares between TEN and NINETY

      What is the value of SENT?


    1. There are a limited number of 3 digit numbers that can be TEN.
       If we use the perfect square, divisible by 9, and distinct digits
       we have the following possibilities:

        TEN = 145, 325, 730, 901
   
       Since N begins NINETY it can't be zero we can remove 730

       Thus: TEN = 145, 325, 901

    2. There are a lot of 5 digit numbers divisible by 9 but if we add
       constraints for the values of TEN and limits on non-zero digits
       it becomes manageable.

       A brute force search of number of perfect squares between the
       limits of TEN and NINETY requires less than 200 sequences.

       We create a list of perfect square (0:775)^2 and then count the
       number between TEN and NINETY for each feasible sequence.

    Solution:

        TEN = 901
        NINETY = 151092
        SIX = 358

        SENT = 3019

    Note that there are only 7 symbols used (TENIYSX)


## Problem 5

    Fill in the following cross-number puzzle with 13
    different three-digit perfect squares. No leading zeros.
    Read top-to-bottom and left-to-right
    as in a crossword puzzle.

    -- Richard I. Hess, CA B'62

    The puzzle grid looks like this:

    +-+-+-+-+-+-+-+-+-+-+-+
    | |X| | | |X| | | |X| |
    +-+-+-+-+-+-+-+-+-+-+-+
    | | | |X| | | |X| | | |
    +-+-+-+-+-+-+-+-+-+-+-+
    | |X| | | |X| | | |X| |
    +-+-+-+-+-+-+-+-+-+-+-+

TODO

    I don't think I ever solved this.

    There are 22 candidate values:
    
    100 121 144 169 196 225
    256 289 324 361 400 441
    484 529 576 625 676 729
    784 841 900 961

    The digit counts by position are:
    
               ...... Digit ......
               0 1 2 3 4 5 6 7 8 9
    ------------------------------               
    Hundreds:  0 5 3 2 3 2 2 2 1 2
    Tens:      3 0 6 0 3 1 3 2 3 1
    Units:     3 5 0 0 4 2 4 0 0 4    
    ------------------------------               

    This information could be used to reduce the number of possible
    numbers to be tried.
    
    The vertical numbers that also start another number, for example,
    can not have 0 as the Tens digit.  In addition, since no number has
    a 3 as the tens digit the numbers that start at a Tens digit can
    not start with 3.
    
    The vertical numbers that end another number can not have 2,3,7, or 8
    as the Tens digit.

    These are all relatively weak constraints and don't get to a unique
    solution.
    
    I don't see an algorithmic solution.  This smacks of trial and error
    or using a Monte Carlo approach.
    
    Not sure I care.


## Bonus

    If 12 one-ohm resistors are soldered together to
    form a cube, with each resistor being the edge of the cube,
    then the equivalent resistance of the network between
    two corners at the ends of a body diagonal is 5/6 ohm.
    It is also possible to solder 32 one-ohm resistors
    together in the network equivalent of a four-dimensional
    hypercube. The vertices of a four-dimensional hypercube
    can be designated using the coordinates (w,x,y,z) where
    each variable has a value of 0 or 1. Then adjacent vertices
    are those in which only one variable has a different value;
    for instance, (0,1,0,1) and (0,1,1,1) are adjacent vertices.
    One can show that the equivalent resistance of the fourdimensional
    hypercube between a body diagonal, say
    (0,0,0,0) and (1,1,1,1), is 2/3 ohm.
    Now, consider 192 resistors soldered together to form
    the network equivalent of a six-dimensional hypercube.
    What is the equivalent resistance between corners at the
    ends of a six-dimensional body diagonal?

    -- John L. Bradshaw, PA A'82

Solution:

    I would probably just create a Spice deck and use it to solve
    this problem.  There may be some clever way to do this with
    combinatorics.


## Computer Bonus

    Thirty-six is the smallest number,
    greater than one, that is both a triangular number and
    a perfect square. What are the next four numbers that
    are both triangular and a perfect square? Triangular
    numbers are numbers of the form n(n + 1)/2, the first few
    being 1, 3, 6, 10, and 15.

    -- The Colossal Book of Mathematics by Martin Gardner

Solution:

    Should be able to search for this by generating the triangular
    numbers and then checking to see if they are perfect squares.
    
```{r}
n=0
# Check the first 10000 triangular numbers
for (i in 37:100000) {
    t=i*(i+1)/2
    m=trunc(sqrt(t))
    if (t == (m*m)) {
        print(t)
        n=n+1
        if (n==10) { break }
    }
}
```

The next four values after 36 are:

        1225
       41616
     1413721
    48024900

BTW:  the next value is 1631432881.
