---
title: "Bent Spring 2013"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Spring 2013

Generally not a good problem set.

## Problem 1

    Joe has wired 100 bulbs, labeled 1 to 100,
    into an electrical circuit along with a
    button switch. He starts with all the bulbs
    unlit. When he pushes the button, every light
    lights. When he pushes it a second time,
    every second light (i.e., lights 2, 4, 6,
    etc.) goes off. On the third push, every
    third light changes status, that is, if it is
    off, it turns on, and if it is on, it turns
    off. On the fourth push the same thing occurs
    for every fourth light, and so on for the
    fifth through hundredth pushes. At this
    point, how many lights are lit?

    -- The Electric Toilet Virgin Death Lottery
    and Other Outrageous Logic Problems, by
    Thomas Byrne and Tom Cassidy

Solution:

    Try brute force - It says there are 10 lights on.
    The lights that remain on are the perfect squares:
    
    > which(bulbs)
     [1]   1   4   9  16  25  36  49  64  81 100
    
```{r}
bulbs=rep(T,100)
for (n in 2:100) {
    for (i in seq(n,100,n)) {
        bulbs[i]=!bulbs[i]
    }
}
print(sum(bulbs))
```


## Problem 2

    As part of an experiment to study the
    alertness of students, the letters A to F
    were permuted and briefly shown (in the form
    of a six letter "word") to seven students.
    After an hour, the students were asked to
    write down the order of the letters. What the
    students wrote was: Greg, BCDAEF; Hal,
    DAEFBC; Ivan, ABEFDC; Jill, BCFDEA; Kate,
    AEBDFC; Lila, CFEABD; and Mel, DCAEFB. The
    students were then asked a series of
    questions that each answered based on what he
    or she had written as his or her recollection
    of the order of the letters. For each
    possible pair of letters, they were asked if,
    reading from left to right, one letter came
    before another. A typical question was, "Does
    A come before D?" (The letters in the
    questions were always in alphabetical order.)
    Each student got a different even number of
    questions correct. (No one had them all
    wrong.) What was the correct order of the
    letters?

    -- Susan Denham in New Scientist

TODO

    meh


## Problem 3

    A typical roll of NECCO(R) wafers contains 40
    wafers with a random distribution of eight
    different flavors. Suppose NECCO(R) decides to
    make sample rolls containing only 12 wafers.
    What is the probability that a twelve-wafer
    roll will have at least one wafer of each
    flavor? Assume that the rolls are made up
    from batches of wafers that contain equal
    numbers of each flavor and that the wafers in
    each roll are selected at random.

    -- Howard G. McIlvried III, PA Gamma'53

Solution:

    Try Monte-Carlo
    
```{r}
ntrials=100000

set.seed(42)
n=0
for (trial in 1:ntrials) {
    s=sample(8,12,replace=T)
    if (length(unique(s))==8) {
        n=n+1
    }
}
print(paste0("n = ",n))
print(paste0("Pr = ",n/ntrials))
```

Results of random sampling:

    Seed   Trials Pr
    ======================
    42     100K   0.09333
    42     1000K  0.092724
    123    1000K  0.093373
    2020   1000K  0.093218

    Bent analytical solution = 0.09330642
    
    I do not understand the analytical solution.
    See below for details.
    
My attempt - analytically:

    Since there are 8 flavors and 12 wafers there are 8^12 possible
    packages.  This is ~65e9 which is a bit much for brute force
    counting.

    If there were only 8 wafers this would be simple:
    
    (8/8)*(7/8)*...*(2/8)*(1/8) = factorial(8)/8^8
                                = 40320 / 16777216
                                ~ 0.002403259

    With a couple of extra tries the odds should improve
    substantially.  Effectively we have 4 extra wafers
    that can be distibuted throughout the 12.  This
    should improve things as C(12,4)= 495
    
    Multiplying gives a number > 1 so this isn't right.
    
    Let us try a different way:
    
    We will try a toy problem and then try to generalize.
    Assume 2 flavors and 4 wafers
    
    a b c d y = 14/16 = 7/8
    0 0 0 0 0
    0 0 0 1 1
    0 0 1 0 1
    0 0 1 1 1
    0 1 0 0 1
    0 1 0 1 1
    0 1 1 0 1
    0 1 1 1 1
    1 0 0 0 1
    1 0 0 1 1
    1 0 1 0 1
    1 0 1 1 1
    1 1 0 0 1
    1 1 0 1 1
    1 1 1 0 1
    1 1 1 1 0

    The odds of getting the first "flavor" is 1.
    The odds of getting the second "flavor" is 1-(1-1/2)^3 = 7/8
    
    Multiplying = 1*7/8 = 7/8 --> confirms result above.

    We can repeat this for 3 flavors and 5 wafers.  This requires 3^5 = 243
    rows so we will check it and just report the results.
    
    Num flavors  Num cases
    ----------------------
    1            3
    2            90
    3            150

    We can repeat this for 4 flavors and 7 wafers.  This is 4^7 = 16384 rows.
    
    Num flavors  Num cases
    ----------------------
    1            4
    2            756
    3            7224
    4            8400

    For 5 flavors and 7 wafers
    
    Num flavors  Num cases
    ----------------------
    1            5
    2            1260
    3            18060
    4            42000
    5            16800
    
    For 5 flavors and 8 wafers
    
    Num flavors  Num cases
    ----------------------
    1            5
    2            2540
    3            57960
    4            204120
    5            126000

    Still not seeing the pattern -- needs more work.

I eventually gave up -- checked out the Bent
    
    Prepare to have your mind blown:
    
    The Bent gives the solution as the number of flavors
    missing as:
    
        C(8,8-n)*(8-n)^12
    
    Apparently this only works for n=1,2, ..., 7 but they
    don't specifically state that so they enumerate the
    number of possibilities as:
    
        8^12 = 68719476736 (68.71948e9)
    
    For each of the values 1:7 we can tabulate the numbers.
    They give the following table:
    
    n   Possibilities
    ------------------
    1   110730297608 ~ 110.7e9
    2    60949905408
    3    13671875000
    4     1174405120
    5       29760696
    6         114688
    7              8    
    
    This makes no sense to me.  How can the number of possibilities
    exceed the total number of permutations???  In any event, they
    sum up the values with an alternating series (no explanation)
    as:
    
    N(at least one missing) = sum((-1)^(n+1)*choose(8,8-n)*(8-n)^12,n,1,7)
    
    This sums to: 62307508096
    
    The Pr (at least one missing) = 0.9066936

    So: Pr (none missing) = 0.09330642

    I looked in the following volume of the Bent but they didn't add
    any more explanation other than to say this was one of the harder
    problems.
    

## Problem 4

    Consider a double row of regular hexagons,
    ten in the top row and nine in the bottom
    row, arranged like the cells in a honeycomb.
    Starting at the upper left hexagon and
    zigzagging down and up, label the hexagons A
    through S. How many different paths are
    there, starting at A and ending at S, where a
    path consists of a series of ten to 19
    letters indicating the order in which the
    hexagons are visited? You can only move from
    one hexagon to another hexagon that shares a
    common edge. No hexagon can be visited more
    than once for a given path.

    -- Why Do Buses Come in
    Threes? by Rob Eastaway and
    Jeremy Wyndham


TODO

    My head hurts.
    

## Problem 5

    Find three different nonzero digits such that
    each of the six permutations of the digits,
    read as six three-digit integers, is a
    semiprime. A semiprime is an integer that is
    the product of two, not necessarily
    different, primes. For example, 121 = 11*11
    and 143 = 11*13 are semiprimes, while 153 =
    3*3*17 and 105 = 3*5*7 are not.

    -- Richard England in New Scientist

Solution:

    Brute force
    
```{r}
library(gtools)
library(gmp)

a=combinations(9,3)

for (i in seq(nrow(a))) {
    b=permutations(3,3,a[i,])
    k=apply(b,1,function(x) {length(factorize(100*x[1]+10*x[2]+x[3]))==2})
    if (all(k)) {
        print(a[i,])
    }
}

```

The solution set is:

    [1] 1 7 8

    N    Factors
    -------------
    178   2  89
    187  11  17
    718   2 359
    781  11  71
    817  19  43
    871  13  67


## Bonus

    Al, Beth, Carl, and Dawn are sitting around a
    table at a bar, as Al tries to guess Beth's
    age. They all know she is at least 21, or she
    wouldn't have been allowed into the bar. Al
    asks Beth five questions, pausing for
    contemplation after each question:

    1. Is your age a multiple of 17?
    2. Is your age a multiple of 3?
    3. Is your age a prime number?
    4. Are you older than I am? (Beth knows Al's age.)
    5. Have you celebrated your 51st birthday?

    At this point, Al announces that he has
    deduced Beth's age, but Beth tells him he is
    wrong. Carl, whose age is a prime number, has
    been listening to this conversation and is
    able to correctly deduce Al's age. From his
    knowledge of Beth, he surmises that she has
    not answered all the questions truthfully and
    guesses that she has alternated correct and
    incorrect answers. He knows that Beth is
    older than he is, and although he has guessed
    correctly how many of Beth's answers are
    incorrect, he has assumed the wrong ones. So,
    when he announces what he has deduced as
    Beth's age, Beth tells him he is also wrong.
    Finally, Dawn, who has also been listening in
    and is sharper than Carl, guesses correctly
    which of Beth's answers are incorrect. Now,
    knowing that Beth is younger than she is,
    Dawn is able to correctly announce Beth's
    age. What are the ages of Al, Beth, Carl, and
    Dawn, and what are Al's and Carl's incorrect
    guesses? It may help to know that Dawn's age
    is divisible by 13 and they all know that
    their ages are all different.

    -- Brain Puzzler?s Delight
    by E.R. Emmet


TODO

    Yuck


## Computer Bonus

    You are floating in a sea of 7s on a raft
    with the number 101. You discover that you
    can take a 7 and insert it into your raft to
    enlarge it (getting 7101; 1701; 1071; or
    1017). Unfortunately, every time you do this,
    the raft divides itself by its smallest prime
    factor (leaving in the above case 2367; 567;
    357; or 339).If the raft goes below 100, it
    will sink. What is the maximum number of
    insertions you can make before you sink?
    Remember, 1 is not a prime.

    -- Technology Review


TODO

    Seems like a prime candidate for recursive programming.

    Glad I didn't work on this.  Turns out the answer is
    that there are lots of ways you can get an infinite
    loop so the raft can stay floating forever.

    101 --> 1701 (divide by 3) --> 567
    567 --> 7567 (divide by 7) --> 1081
    1081 --> 10817 (divide by 29) --> 373
    373 --> 3737 (divide by 37) --> 101
    
    and the loop goes forever.
