---
title: "Bent Spring 2001"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Spring 2001

## Problem 1

    A local eatery has an electronic
    "slot machine" that offers various
    games. In one of them, called
    PrimeTime, after a coin is inserted,
    the digits 1 through 9 appear in
    random order around a circle. Then,
    an arrow, like the minute hand on a
    clock, spins and stops, pointing to the
    space between two of the digits. If
    the two-digit number formed in a
    clockwise direction by the two digits
    on either side of the space the arrow
    points to has a two-digit prime
    factor, you win the jackpot. For
    example, if the arrow stopped
    between 9 and 2, you would win
    because 92 has the prime factor 23.
    On the other hand, if your number
    were 81, you would lose. I recently
    played the game; and, as the arrow
    was spinning, I realized to my
    dismay that no matter where the
    arrow stopped, I would lose. Starting
    with 1, what was the clockwise order
    of the digits?
    
    -- Susan Denham in New Scientist

Solution:

Could generate all the permutations and then check them.  Seems excessive.

Could enumerate all the two digit primes and their multiples to look for patterns.
For example, I might find that a 9 can never be preceded by the number 8 (since 89 is prime).
If I can find any digits that only permit a single successor/predecessor then I will have
part of the sequence.  The sequence we are looking for should be revealed from this
pattern analysis:

```{r}
is_prime = function(n) {
  if (n<4) {
    if (n==2 | n==3) { return(TRUE) }
    return(FALSE)
  }
  if (n%%2 == 0) { return(FALSE) }
  for (i in 3:sqrt(n)) {
    if (n%%i == 0) { return(FALSE) }
  }
  return(TRUE)
}

p=(1:100)[sapply(1:100,is_prime)]
p=p[p>9]

print("Two digit primes:")
print(p)

# Generate all the integer multiples of these primes up to 9*11=99
# Remove all the results above two digits
#
# These results are Poison (we want to avoid them)
p=as.vector(outer(1:9,p,'*'))
p=p[p<100]

print("Two digit primes and multiples:")
p

print("Patterns based on final digit:")
for (i in 1:9) {
  print(c(i,sort(p[p%%10 == i])))
}

print("Patterns based on initial digit:")
for (i in 1:9) {
  print(c(i,sort(p[trunc(p/10) == i])))
}

```

From the results above we conclude:

    --> indicates a conclusion based strictly on the patterns above.
    Other comments and eliminated digits occur based on these primal
    patterns and the observation that a digit can only occur before or
    after another digit (not in both places - e.g. 2 relative to 7):

    1 must be preceded by 8                 must be -8-1-
    2 must be preceded by 1, 3, 7
    3 must be preceded by 6  -->            must be -6-3-
    4 must be preceded by 1, 5
    5 must be preceded by 1, 3, 7
    6 must be preceded by 1, 3, 5, 9
    7 must be preceded by 2  -->            must be -2-7-
    8 must be preceded by 1, 9
    9 must be preceded by 4  -->            must be -4-9-

    1 must be followed by 2, 4, 5, 6, 8
    2 must be followed by 7
    3 must be followed by 2, 5, 6
    4 must be followed by 9
    5 must be followed by 4, 6
    6 must be followed by 3
    7 must be followed by 2, 5 -->          must be -2-7-5-
    8 must be followed by 1
    9 must be followed by 8                 must be -9-8-

Merging some of the partial patterns:

    -4-9-8-1-  as a result of -4-9- and -9-8- and -8-1-
    -2-7-5-    as a result of 2 must precede 7 so it can't also follow
    -6-3-      primal pattern

But 4 must be preceded by 1, 5 so:

    -2-7-5-4-9-8-1-  as a result of -2-7-5- and -4-9-8-1-
    -6-3-

And, 6 must be preceded by 1,3,5,9:

    -2-7-5-4-9-8-1-6-3- ...
    
If we circularly shift to start with 1:

    ... -1-6-3-2-7-5-4-9-8- ...

Then the game can only generate numbers with single digit prime factors:

    16 = 2*2*2*2
    63 = 3*3*7
    32 = 2*2*2*2*2
    27 = 3*3*3
    75 = 3*5*5
    54 = 2*3*3*3
    49 = 3*3*7
    98 = 2*3*3*7
    81 = 3*3*3*3

Just for grins, let's try the permutation method:

```{r}
library(gtools)

t1=Sys.time()
# a is the permutation matrix with 1 in the left-most column
# The remaining columns are all permutations of the digits 2-9
a=cbind(1,gtools::permutations(8,8,2:9))
# b is the two digit number made with digits from adjacent columns of a.
# This also wraps around from the end of the a matrix back to column 1.
b=10*a+cbind(a[,2:9],a[,1])

# Check each of the permutations to see if they contain any of the
# target values (in p).  If so - then that is not what we want.
k=apply(b,1,function(x) {any(x %in% p)})
t2=Sys.time()

print(t2-t1)

# Find all the rows that are "losers" -- e.g. have no two digit prime factors
a[which(k==FALSE),]

```

Which yields the same answer and runs very quickly.  It is actually amazing that there is
only one combination that yields this result.

As a corollary: what is the expected payout from this game?  What percent of the time can
you actually get a two digit prime factor?

We can use the results from above to calculate this:

```{r}
m=apply(b,1,function(x) {sum(x %in% p)})
sum(m)/length(b)
```

Amazingly, 5/8 of the time you will actually have a two digit prime factor!

This also means that 3/8 of the time you will not have a two digit prime factor.
If we assumed the 9 spaces on the wheel were statistically independent (not a
good assumption), then we could estimate the likelihood of an entire wheel
that has no two digit prime factor as:

```{r}
(3/8)^9*nrow(a)
```

So we expect that about 6 rows of the permutation matrix has no two digit
prime factor.  Actually the result is 1 which is remarkably close to the random
assumption calculation -- at least it is within an order of magnitude.


## Problem 2

    Don't let the wording of this
    cryptic fool you into thinking it is
    solvable in base 10. It isn't.
    
    TEN x TEN - TEN = NINETY
    
    Your job is to solve it in the
    smallest base for which a solution
    exists.
    
    -- Journal of Recreational Mathematics

Solution:

There are only 5 symbols used in the equation above.  This means that
the base must be at least base 5.  We could exhaustively search for
solutions to the cryptic problem but the number of permutations increases
rapidly with the base number but it should be feasible up to base 20 or so:

```{r}
d=data.frame(base=5:20)
d$perm=choose(d$base,5)*factorial(5)
plot(d$base,d$perm,type='b',log="xy")
grid()
```

Let's try this:

```{r}
# Get the code to play with cryptic arithmetic problems
source('rgc_lib.R')
```

We will verify this doesn't work in base-10:

```{r}
a=make_crypt_df('tenninety',10)
x10=make_num(a,'ten',10)
x90=make_num(a,'ninety',10)
k=(x10*x10-x10 == x90) & (a$t != 0)
sum(k)
```
I tried several base values up through 32 without any solutions.
Need to think this through - let's try symbolically:

    TEN*TEN-TEN = TEN*(TEN-1) = NINETY
    
    TEN ~ sqrt(NINETY)  --> TEN is just slightly more than sqrt(NINETY)

    number of digit in NINETY = 6, TEN = 3.  TEN must exceed sqrt(base) to add digit.

    TEN = T*b^2 + E*b + N
    NINETY = N*b^5 + I*b^4 + N*b^3 + E*b^2 + T*b + Y
    
So:

    TEN*TEN-TEN = (T*b^2 + E*b + N)*(T*b^2 + E*b + N) - (T*b^2 + E*b + N)

        T^2       b^4
        2*E*T     b^3
        2*N*T+E^2 b^2
        2*N*E     b
        N^2

    NINETY = N*b^5 + I*b^4 + N*b^3 + E*b^2 + T*b + Y
    
TODO -- need to work on this some more


## Problem 3

    I recently visited a small aviary
    and asked the keeper how many
    birds he had. He replied, "I have a
    total of 401 birds of five different
    species. The number of auks is a
    fourth power, although they will be
    breeding soon. I have more than ten
    boobies, and their number is both a
    palindrome and triangular. The
    number of cuckoos is divisible by 13.
    The number of doves, which is more
    than the number of auks, is prime
    and a Fibonacci number. More than
    five-eighths of the four species I
    have mentioned so far are auks and
    cuckoos. How many eagles do I
    have?" (A triangular number has the
    form n(n + 1)/2, where n is a positive
    integer. Such numbers can be
    arranged in triangles, like bowling
    pins.)
    
    -- The Platinic Corner

Solution:

    Total number of birds = 401
    Five species, each has count less than 401
    
    auks = fourth power > 1 (if breeding) --> [16, 81, 256]
    boobies = triangular, palindrome > 10 --> [55, 66, 171]
    cuckoos = divisible by 13
    doves   = prime, Fibonacci > auks
      prime, Fibonacci = [89, 233] --> auks can not be 256.
    auks+cuckoos > 5/8*(auks+cuckoos+doves+boobies)
    
        boobies >= 55
        doves   >= 89
        
        auks+cuckoos > 5/8*(auks+cuckoos+doves+boobies)
        3*(auks+cuckoos) > 5*(doves+boobies) >= 5*(55+89)
        auks+cuckoos >= 240
    
    This is starting to get confusing -- let the computer do the search.
    
    eagles  = unknown = 401-(auks+boobies+cuckoos+doves)
    
```{r}
a=data.frame(auks=NA,boobies=NA,cuckoos=NA,doves=NA,eagles=NA)
for (auks in c(16,81)) {
  for (boobies in c(55,66,171)) {
    for (cuckoos in (1:30)*13) {
      for (doves in c(89,233)) {
        if ((auks+cuckoos) > 5/8*(auks+cuckoos+doves+boobies)) {
          if ((auks+cuckoos+doves+boobies)<401) {
            a=rbind(a,c(auks,boobies,cuckoos,doves,401-(auks+boobies+cuckoos+doves)))
          }
        }
      }
    }
  }
}
a=a[!is.na(a$auks),]
row.names(a)=1:nrow(a)
a
```

The number of eagles must be 7.

The number of boobies must be 55, number of doves must be 89.
The number of auks and cuckoos can vary as noted in the solution.


## Problem 4

    An urn contains nine balls labeled 1
    through 9. Draw out the balls one at
    a time and line them in the order
    drawn to form a nine-digit number.
    What is the probability that the
    number so formed is divisible by 11?
    
    -- Technology Review

Solution: try Monte-Carlo

```{r}
n_trials=1000000
n=0
for (i in 1:n_trials) {
  s=0
  for (j in sample(1:9,9)) {s=10*s+j}
  if (s%%11 == 0) { n=n+1 }
}
n/n_trials
```

I could also look at the divisibility rules on Wikipedia
and see if there is a tricky way to find divisibility by 11.

I could also check out all the 9 digit permutations.  This should be feasible since
there are only 9! permutations to consider:

```{r}
library(e1071)
a=permutations(9)

s=rep(0,nrow(a))
for (i in 1:ncol(a)) { s=10*s+a[,i] }

n=sum(s%%11 == 0)

n/nrow(a)

```

This is the result of exhaustive search.
It compares well with the result from Monte-Carlo:

    Exhaustive = 0.08730159
    Sampling   = 0.087654    (using 1e6 Monte-Carlo trials)


## Problem 5

    A pentomino is a plane figure
    constructed from five unit squares
    joined along their edges. A complete
    set of pentominoes consists of one of
    each of the 12 different such figures.
    Rotations and reflections are allowed
    but do not count as different figures.
    Starting with a complete set whose
    squares are the size of chessboard
    squares, what is the smallest number
    of pentominoes that can be placed on
    a chessboard so that none of the
    remaining pentominoes can be placed
    on the board? Pentominoes must be
    placed so that they are congruent
    with the squares of the chessboard,
    and they cannot overlap. To see what
    a complete set of pentominoes looks
    like, see the Brain Ticklers column in
    the Fall 1999 issue of THE BENT. We
    like to assign the letter names F, I,
    L, N, P, T, U, V, W, X, Y, and Z,
    which the pentominoes at least
    somewhat resemble.
    
    -- Mathematics Teacher

TODO

## Bonus

    An anti-aircraft gun with
    muzzle velocity v can be fired at any
    azimuth or elevation by vertical and
    horizontal rotation about its fixed
    location. What is the total volume of
    sky that can be reached by a shell
    from the gun, i.e., what is the
    envelope of coverage? Assume the
    land in the vicinity of the gun is level,
    and ignore wind and air resistance.
    Express your answer as a function of
    v and the gravitational constant g.
    
    -- Richard W. de Nobel, OH Alpha'52

Solution:

    We should be able to calculate the volume using
    slices in the horizontal direction and then calculating
    the distance (radius) achievable at the optimal
    firing angle.  These will be differential discs
    that can be summed from 0 to ymax to yield the
    total volume.
    
    Assume the target height is y0.  We need to find
    the largest x that can be achieved at this height.
    
    Under uniform acceleration:
    
        v = a*t
        x = 1/2*a*t^2
    
    Assume firing angle is theta, then:
    
    vx0=v*cos(theta)
    vy0=v*sin(theta)
    
    Time to reach max altitude: t1=vy0/g
    Max altitude: ymax=1/2*g*(vy0/g)^2 = vy0^2/(2*g)
    Time to fall: t2=sqrt(2*(ymax-y0)/g)
    Total flight time = t1+t2
    Distance (x-axis) = vx0*(t1+t2)
    
    Substituting:
    
      xmax = v*cos(theta)*(v*sin(theta)+sqrt(2*(v^2*sin(theta)^2/(2*g)-y0)/2))
      
    Want to maximize wrt theta (need sympy -- this is ugly):
    
    Unfortunately, this gets ugly and stays ugly!
    
    Sympy ran for several minutes trying to integrate this so I killed it.
    
    The sympy expressions seemed correct for y0=0 using some numerical
    values for v, g.
    
    Conceptually this works -- but maybe need a different way to find
    the volume.
    
    For a given x (radius) there should be a maximum altitude.  Is it possible
    to find this and then sum up differential cylinders?
    
    I gave up -- but this would have been the way to go.
 
    This is the approach taken in the Bent and they
    express y=f(x,theta) and then maximize y wrt theta with dy/dtheta=0.
    Solving this gives the relationship between x and theta which can be
    used to eliminate theta and get an equation for x and ye (envelope).
    
    Once this relationship is known we can just integrate the cylinders from 0
    to the maximum possible firing distance.
    
    From the Bent:
    
    x = v*t*cos(theta)
    y = v*t*sin(theta) - 1/2*g*t^2
    
    elimination of t leads to:
    
    t=x/(v*cos(theta))
    y=v*x/(v*cos(theta))*sin(theta) - 1/2*g*x^2/(v*cos(theta))^2
     = x*tan(theta) - x^2*g/(2*v^2*cos(theta)^2)
     
    Let: h = v^2/(2*g) = maximum height when firing straight up
    
    y = x*tan(theta) - x^2 * 1/(4*h*cos(theta)^2)
    
    The maximal y for a given x defines a point on the envelope.
    To find the maximal y we need dy/dtheta = 0
    
    this is maximized when x=2*h/tan(theta) or
    
    theta = atan(2*h/x)
    
    so:
    
    ye = h-x^2/(4*h)  (yes - I used sympy to simplify this)
    
    so, now:
    
    we can constuct differential cylinders (as I suspected above) and
    calculate the volume:
    
    integrate(2*pi*x*ye,(x,0,xmax))
    
    where xmax = maximum possible x value which occurs for theta=45 degrees.
    
    xmax:  vy0=v*sin(45*pi/180)=v/sqrt(2)
           vx0=v*cos(45*pi/180)=v/sqrt(2)
           t = 2*vy0/g
           xmax = 2*v/sqrt(2)*v/sqrt(2)/g
                = v^2/g
                = 2*h (from above)
                
    so: V = integrate(2*pi*x*(h-x^2/(4*h)),(x,0,2*h))
          = 2*pi*h**3
          
    or:
    
        v = 2*pi*((v^2)/(2*g))^3
          = 2*pi*v^6/(8*g^3)
          = pi*v^6/(4*g^3)

We should be able to do this with reticulate and sympy:

```{r}
library(reticulate)
```


```{python}
import sympy
from sympy import *

x,y,theta,t,g,v = symbols('x y theta t g v',positive=True)

xpos=v*t*cos(theta)
ypos=v*t*sin(theta)-g*t**2/2

# Solving the first equation for t:
tval=solve(xpos-x,t)[0]

# and substituting into the second eliminates t.
# Need to find the value of theta that maximizes this height for a given x
thetaval=solve(diff(ypos.subs(t,tval),theta),theta)[0]
pprint(Eq(S('theta_for_ye'),thetaval))

# If we substitute back into equation for y then we have the desired function
# This is the height of the envelope (aka differential cylinder of radius x)
ye=simplify(ypos.subs(t,tval).subs(theta,thetaval))
pprint(Eq(S('ye'),ye))

# Now we need to integrate differential cylinders over x from 0 to xmax.
# xmax occurs when the firing angle is 45 degrees
# The time the projectile is in the air is 2*time for projectile to reach peak.
# The speed in the x direction is constant
tmax=2*(v*sin(pi/4)/g)
xmax=v*cos(pi/4)*tmax
pprint(Eq(S('xmax'),xmax))

# Now we can integrate
pprint(Eq(S('Volume'),integrate(2*pi*x*ye,(x,0,xmax))))

print()
print("Alternate expression:")
print()

# Alternately
pprint(Eq(S('Volume'),pi/4*S('xmax')**3))
```

I am actually starting to wonder what this envelope looks like.  Using the results
from above we should be able to just plot it:

```{r}
v=100
g=32.2
xmax=v^2/g

x=seq(-xmax,xmax,length=200)
y=-g*x^2/(2*v^2)+v^2/(2*g)

plot(x/xmax,y/xmax,type='b')
abline(h=0,col='red',lwd=3)
abline(v=0,col='red',lwd=3)
```

Looking again at the equations shows the envelope is quadratically related to
the maximum firing range:

    ye = (xmax - g*x^2)/2
    
We could probably use this calculate the volume as well -- but we will leave that
for another day.


## Double Bonus

    A 9-ft. by 12-ft. rug
    has a centered 1-ft. by 8-ft. hole, with
    the 8-ft. dimension parallel to the 12-ft.
    side. Show how to cut the rug into
    two pieces that can be fit together to
    form a 10-ft. by 10-ft. rug with no
    hole.

    -- Author not specified

TODO

    I have seen this problem before.  Make a staircase
    and shift it one step.
    