---
title: "Bent Summer 2006"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Summer 2006


## Problem 1

    Although there are an infinite number
    of primes, we can find sequences
    of integers of any desired length,
    none of which are prime. Give an expression,
    in terms of N, for the first
    number of a string of at least N consecutive
    composite integers. This expression
    does not have to identify the
    smallest such number.

    -- Daryl Cooper

TODO

    I had to look at the solution.
    
    They used (N+1)! + i which is divisible
    by i = 2 to N+1 which is N terms.


## Problem 2

    An archivist, baker, clergyman,
    and doctor played three rubbers of
    bridge (bridge is a four-person game
    with two partners on one side versus
    two partners on the other side). By
    coincidence, the score of each rubber
    was an exact multiple of 100. Each
    partnered each other once, and, at a
    penny a point, the doctor emerged $5
    up. The clergyman and his brother
    lost the largest rubber by 800. They
    are a bit vague about other details,
    but the scoresheet shows that Peter
    finished $19 down, having fared better
    with Ronnie for partner than with
    Quentin. It shows too that Quentin did
    better overall than his father's only
    brother and that Sam did worse overall
    than the baker. The doctor, by the
    way, sent a greetings telegram when
    the archivist was born. And, also by
    the way, when Peter was Sam's current
    age, he could open beer bottles
    with his teeth. Who does what for a
    living?

    -- Martin Hollis

TODO

    meh


## Problem 3

    Solve the following cryptic multiplication:
    WIT * WILL = THIRST
    with different letters representing
    different digits.

    -- New Scientist

Solution: from the cryptic arithmetic notebook

Standard cryptic multiplication:

    WIT * WILL = THIRST
    
```{r}
source('rgc_lib.R')

a=make_crypt_df('witwillthirst')
x_wit=make_num(a,'wit')
x_will=make_num(a,'will')
x_thirst=make_num(a,'thirst')
k=(x_wit*x_will == x_thirst)
k=k & (a$w != 0) & (a$t != 0)
print(paste0('Number of solutions = ',sum(k)))
print(t(a[k,]))
print_num(a,'wit',k)
print_num(a,'will',k)
print_num(a,'thirst',k)
```

    [1] "Number of solutions = 1"
      137063
    w      2
    i      3
    t      5
    l      1
    h      4
    r      0
    s      8
    [1] "wit = 235"
    [1] "will = 2311"
    [1] "thirst = 543085"


## Problem 4

    A point P lies in the same plane as
    a unit square. Let the vertices of the
    square, taken counterclockwise, be A,
    B, C, D. Also, let the distances from P
    to A, B, and C, be u, v, w, respectively.
    What is the greatest distance that P
    can be from D if u^2 + v^2 = w^2?

    -- 34th Annual High-School Mathematics Examination

Solution:

    I need a picture.  Maybe I can draw it with ASCII Art
    I have (incorrectly) solved this twice so I will adopt
    the same convention as the Bent so we can all follow along:

    C (0,1                          B (1,1)
    +-------------------------------+
    |                               |
    |                               |
    |                               |
    |                   o           |
    |                   P (x,y)     |
    |                               |
    |                               |
    |                               |
    |                               |
    |                               |
    |                               |
    |                               |
    |                               |
    |                               |
    +-------------------------------+
    D (0,0)                         A (1,0)

    Using equations:
    
    Assume P=(x,y) then
    
    PA = u: u^2 = (x-1)^2 + y^2
    PB = v: v^2 = (x-1)^2 + (y-1)^2
    PC = w: w^2 = x^2 + (y-1)^2
    
    PD = z: z^2 = x^2 + y^2
    
    Given: u^2 + v^2 = w^2, so
 
        (x-1)^2 + y^2 + (x-1)^2 + (y-1)^2 = x^2 + (y-1)^2
        
        2*(x-1)^2 - x^2 + y^2 = 0
        
        y^2 = x^2 - 2*(x-1)^2 = -x^2 + 4*x - 2

    The distance we are looking for is:
    
        z^2 = x^2 + y^2
        
            = 4*x - 2

    The distance is maximized as x increases.
    
    The other constraint is that y^2 = -x^2 + 4*x-2 >= 0
    
    Solving for the critical value of x: y = 0 when x=2+/-sqrt(2)
    
    Since we want to maximize x we pick x=2+sqrt(2)
    
    Final solution:  point is on the +x axis at y=0
    
        x = 2+sqrt(2)
        y = 0
        
        Distance to origin = x = 2+sqrt(2)

        Distance PA = 1+sqrt(2)
        Distance PB = sqrt(4+2*sqrt(2))
        Distance PC = sqrt(7+4*sqrt(2))


## Problem 5

    Every pair of communities in a
    county is linked directly by exactly
    one mode of transportation: bus,
    train, or airplane. All three modes of
    transportation are used in the county
    with no community being serviced by
    all three modes and no three communities
    being linked pairwise (e.g., in a
    triangle) by the same mode. What is
    the maximum number of communities
    in the county? What is the maximum
    if only two modes of transportation
    are used and a community can be
    served by both modes? Besides the
    numbers, provide a labeled sketch of
    the networks.

    -- Tenth U.S.A. Mathematical Olympiad

TODO

    meh


## Bonus

    A hotel has rooms with cipher
    locks that open (indicated by a green
    LED) when the last four digits of
    any sequence keyed provide the correct
    combination. Thus, keying 01234
    (five key strokes) tests both 0123 and
    1234. If the combination is not known,
    what is the minimum number of keystrokes
    to guarantee that the lock can
    be opened?

    -- Technology Review

Solution:

    I think this is just a grey code so it
    should only take 3 presses + 1 for each
    code thereafter.  If there are 10000 codes
    then the length should be 10003.

    They called it a de Bruijn sequence in the Bent.


## Double Bonus

    Let a(0) + b(0) + c(0) = 1 with 
    0 <= a(0), b(0), c(0) < 1 and
    
    a(n+1) = a(n)^2 + 2b(n)c(n) with corresponding
    equations for b(n+1) and
    c(n+1). What is the limit of a(n) as n
    
    approaches infinity? Prove.

    -- G. R. Wachtell

Solution:

    The Bent says the solution is 1/3 but the proof
    was omitted due to space constraints.

    A simple numerical investigation supports the 1/3 fixed point.
   
```{r}

a=1/29; b=1/4; c=1-(a+b)

for (n in 1:10) {
    anew = a^2 + 2*b*c
    bnew = b^2 + 2*a*c
    cnew = c^2 + 2*a*b
    a=anew; b=bnew; c=cnew
}

c(a,b,c)

```

    Even a few iterations appears to be enough.
    
    We can assume the equation has a fixed point and then
    solve for it.  If there is a fixed point then a(n+1) = a(n).
    
    So:
    
    a(n+1) = a(n) = a and
    
    a = a^2 + 2*b*c
    b = b^2 + 2*a*c
    c = c^2 + 2*a*b
    
    maxima finds 8 solutions but the only one with all solutions positive
    and results less than 1 are:
    
    [a,b,c] = [1/3, 1/3, 1/3]

