---
title: "Bent Spring 2002"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Spring 2002


## Problem 1

    Pam lives on Rhombus Road,
    which has houses numbered from 3
    to 99, inclusive. Ron is anxious to
    learn her house number so he can call
    on her, but Pam won't tell him. His
    buddies, Art, Bill, Carl, and Don,
    have a piece of information each
    about the house number, but none
    knows the actual number: Art
    knows whether the house number is
    a perfect square; Bill knows whether
    it is a multiple of 3; Carl knows
    whether it has one or two digits; and
    Don knows whether it is a multiple of
    5. Ron knows the type of information
    each has. After questioning and getting
    answers from Art, Bill, and Carl,
    Ron thinks that Don's information
    might confirm what Pam's number is.
    After getting an answer from Don,
    Ron says, "Now, if I knew whether
    her number is odd or even, I'd be
    sure." After wheedling the parity of
    her number from Pam, Ron is sure
    he knows her number. He buys flowers
    and shows up at the address only
    to find an abandoned house. Later,
    he learns that Art and Bill had both
    lied. What is Pam's house number?
    
    -- adapted from Brain Puzzlers
    Delight by E.R. Emmet

Solution: Must be house number 9.

    Pam hour number is in the range [3,99]

    Art knows if it is a perfect square
    Bill knows if it is multiple of three
    Carl knows if it has 1 or 2 digits
    Don knows if it is a multiple of 5.

    Ron thinks he needs Don's answer after
    getting answers from Art, Bill, and Carl.

    After Don's answer he needs to know if
    the number is odd or even - Pam tells him.

    He gets the answer wrong.

    Art and Bill both lied.

    What is the house number?

    Possible perfect squares: 4,9,16,25,36,49,64,81

    Answer possibilities

    Art     Bill    Carl    Possibilities   Conclude
    ------------------------------------------------
    No      No      1       5, 7, 8
    No      No      2       Many (56)
    No      Yes     1       3, 6            Don not needed
    No      Yes     2       Many (28)
    Yes     No      1       4               Don not needed
    Yes     No      2       16,25,49,64
    Yes     Yes     1       9               Don not needed
    Yes     Yes     2       36,81           Don not needed

    Some of the cases do not require Don's input so we
    can ignore these.  Of the remaining cases, 2 have so
    many possibilities that Don wouldn't be good enough to
    determine a final result.  This leaves two cases:

    Art     Bill    Carl    Possibilities   Conclude
    ------------------------------------------------
    No      No      1       5, 7, 8
    Yes     No      2       16,25,49,64

    In the first case, if Don says it is not a multiple of
    5 then he would need Pam's parity input to decide between
    house numbers 7 and 8.  This is a possibility.

    In the second case, no answer from Don can isolate the
    answer to only odd/even.  If Don says it is a multiple of
    5 then we don't need Pam.  If it is not a multiple of 5 then
    it could be 16, 49, or 64.  Pam's input may not be definitive.

    Conclude: we have the first case with a number that is not
    a multiple of 5  (not sure which one -- but it is either house 7 or 8).

    If Art and Bill both lied then the truth is:

    Art     Bill    Carl    Possibilities   Conclude
    ------------------------------------------------
    Yes     Yes     1       9

    Must be house number 9.


## Problem 2

    Once when I was visiting Upper
    Slobovia where the currency is denominated
    in oseps (each osep is
    worth 100 ovatnecs) and the coins in
    circulation have values of 5, 10, 20,
    and 50 ovatnecs.  I was counting my
    change and discovered that I had exactly
    10 oseps worth of coins. Also,
    to my surprise, I found that I had a
    prime number of each kind of coin.
    Then, I realized that I had the minimum
    total number of coins I could
    have under these conditions. How
    many coins did I have? Remember,
    1 is not a prime number.
    
    -- The Platinic Corner

Solution:

    50 ovatnec coins = 7
    20 ovatnec coins = 31
    10 ovatnec coins = 2
    5 ovatnec coins  = 2
    Total = 1000 ovatnecs

Details: let us try this by hand

    Coins = 5, 10, 20, and 50 ovatnecs
    Prime number of each kind of coin --> Minimum of 2 of each
    2*50+2*20+2*10+2*5 = 170 ovatnecs
    
    50 ovatnecs --> (1000-2*(5+10+20))/50 --> maximum of 17
    20 ovatnecs --> (1000-2*(5+10+50))/20 --> maximum of 41
    10 ovatnecs --> (1000-2*(5+20+50))/10 --> maximum of 83
    5 ovatnecs  --> (1000-2*(10+20+50))/5 --> maximum of 167

    Want to minimize total number of coins so probably want to maximize
    use of the highest value coins.

    Note: for the 5 ovatnecs coins the total value always ends in 5 except
    for the case of 2 coins.  All the other totals end in zero so we must
    have exactly 2 of these coins.
    
    Remaining coins must total 990 ovatnecs.  The following subtotals are
    possible:

    > 50*primes[primes<990/50]
    [1] 100 150 250 350 550 650 850 950
    > 20*primes[primes<990/20]
    [1]  40  60 100 140 220 260 340 380 460 580 620 740 820 860 940
    > 10*primes[primes<990/10]
    [1]  20  30  50  70  110 130 170 190 230 290 310 370 410 430 470
         530 590 610 670 710 730 790 830 890 970

    Can't use max number of 50 ovatnec coins since this only leaves
    40 ovatnecs and that is needed for the 20 ovatnec coins.  There is
    nothing left for the 10 ovatnec coins.
    
    Try successively lower 50 ovatnec totals and all possibilities

    50 ovatnecs   20 ovatnecs   10 ovatnecs   5 ovatnecs
    -----------------------------------------------------
    850           40            100 needed    10
    850           60            80 needed     10
    850           100           40 needed     10
    850           140           0 needed      10
    650           40            300 needed    10
    650           60            280 needed    10
    650           100           240 needed    10
    650           140           200 needed    10
    650           220           120 needed    10
    650           340           0 needed      10
    550           40            400 needed    10
    550           60            380 needed    10
    550           100           340 needed    10
    550           140           300 needed    10
    550           220           220 needed    10
    550           260           180 needed    10
    550           340           100 needed    10
    550           380           60 needed     10
    350           40            600 needed    10
    350           60            580 needed    10
    350           100           540 needed    10
    350           140           500 needed    10
    350           220           420 needed    10
    350           260           380 needed    10
    350           340           300 needed    10
    350           380           260 needed    10
    350           460           180 needed    10
    350           580           60 needed     10
    350           620           20 needed     10  --> feasible
    
    Need:
    
    50 ovatnec coins = 7
    20 ovatnec coins = 31
    10 ovatnec coins = 2
    5 ovatnec coins  = 2
    
That was too hard.  Must be an easier way.

I looked at their solution.  Here is what they did:

    Let N=number of 5 ovatnec coins, D=10, Q=20, H = 50

    Since the total is 1000 we must have an even number of
    the 5 ovatnec coins.  Since 2 is the only even prime then
    we must have 2 of these.  Then:

        50*H + 20*Q + 10*D = 990
        5*H  + 2*Q  + D    = 99

    or

        2*Q + D = 99 - 5*H

    We want the minimum number of coins so we want the most
    possible large denomination coins.

    Since H<20 we can try successively smaller primes.  If we
    assume H is odd then 2*Q+D must be even - so - D must be 2.

    If this is the case then:

        2*Q + 2 = 99 - 5*19 --> Q = 1 --> not prime
        2*Q + 2 = 99 - 5*17 --> Q = 6 --> not prime
        2*Q + 2 = 99 - 5*13 --> Q = 16 --> not prime
        2*Q + 2 = 99 - 5*11 --> Q = 21 --> not prime
        2*Q + 2 = 99 - 5* 7 --> Q = 31 --> PRIME!!! (max H, min total)
        2*Q + 2 = 99 - 5* 3 --> Q = 41 --> PRIME!!!

    Thus, for minimum coin count:

        N = 2, Q = 31, D = 2, H = 7 --> total = 42 coins

        N = 2, Q = 41, D = 2, H = 3 --> total = 48 coins X not minimum

    There may be a solution for H = 2 then 2*Q+D and D must be odd.
    This is more difficult to calculate:

        2*Q + D = 99 - 5*2 = 89

    Prime solutions:

        D   Q   Total (not included 2D and 2H)
        --------------------------------------
        83   3  86
        79   5  84
        67  11  78
        43  23  66
        31  29  60
        7   41  48
        3   43  46

    Since all of these totals exceed 42 the original solution
    is the minimum.


Verify with exhaustive search

```{r}
primes=read.table('primes.txt')[,1]
zzz=c()
for (c50 in primes[primes<1000/50]) {
    for (c20 in primes[primes<1000/20]) {
        for (c10 in primes[primes<1000/10]) {
            for (c05 in primes[primes<1000/5]) {
                t=50*c50+20*c20+10*c10+5*c05
                if (t==1000) {
                    print(paste(c50,c20,c10,c05,c50+c20+c10+c05))
                }
            }
        }
    }
}
```

The minimum number of coins is obtained with:

    50 ovatnecs = 7  -->  350
    20 ovatnecs = 31 -->  620
    10 ovatnecs = 2  -->   20
    5 ovatnecs  = 2  -->   10
    -------------------------
    Total         42 --> 1000

## Problem 3

    Solve the following cryptic addition:
    
    YELLOW + YELLOW + RED = ORANGE
    
    The usual rules apply: each different
    letter represents a different digit;
    the same letter always represents
    the same digit; and there are no
    leading zeros.
    
    -- Technology Review

Solution: in cryptic_arithmetic notebook

    YELLOW = 143329
    RED    = 846
    ORANGE = 287504


## Problem 4

    Consider the letters from A
    through M. If all possible 13-letter
    permutations (no repeated letters)
    of these letters are arranged in alphabetical
    order, what is the
    1,234,567,890th permutation in the
    list?
    
    -- adapted from
    Mathematics Teacher

Solution: my original solution was the zero based counting

    Zero based = [C, H, M, D, B, E, L, I, J, K, A, F, G]

    Note the Bent solution is not the same:

          [C, H, M, D, B, E, L, I, J, G, K, F, A]

    I think they may be correct and my solution is actually off
    by one.

    If I use my code and let n=1234567890-1 I get their solution.

    I think this is the difference between starting the count at
    zero and starting the count at 1.  If I ask my code for
    permutation 1 it actually swaps the last two letters.
    If I ask for permutation 0 I get the original sequence.

    The solution is the same up to the last four characters.

    The solution method they outline is the same one I developed.

    The next Bent acknowledges this as the hardest problem but they
    do not acknowledge that they have an error.

    For each choice of the first letter there are 12! permutations to
    follow.

    Axxxx  removes 12! possibilities
    Bxxxx  removes 12! possibilities

    This leaves 1234567890 - 2*12! = 276564690 permutations
    This is not enough to cover all the Cxxx permutations so
    the first character must be C.

    We can eliminate C from the list and start over with 27656490
    permutations of the set ['A','B','D',...]

    We remove all the Axxx for 11! possibilities, etc. until we
    determine the next character and remove it from the set.

    Repeat this for all the remaining positions removing 10!, 9!, etc.
    possiblities per letter as the set length is reduced.


    Example code:

    n=1234567890  --> should be one less as noted above

    zzz=['A', 'B', 'C', 'D', 'E', 'F', 'G', 'H', 'I', 'J', 'K', 'L', 'M']

    for i in range(12,-1,-1):
        m=(n//math.factorial(i)); print(m,zzz[m]); del zzz[m]
        n=n%math.factorial(i)
    
    2  C
    6  H
    10 M
    2  D
    1  B
    1  E
    6  L
    3  I
    3  J
    3  K
    0  A
    0  F
    0  G

## Problem 5

    At a recent party at my house, I
    burned two cylindrical candles. I lit
    the red candle, which was 2 cm
    longer than the white candle, at
    4 p.m., and I lit the white candle 15
    minutes later. At 8 p.m. I noticed
    that they were the same length. The
    red candle burned out at midnight,
    and the white candle burned out an
    hour later. If each candle burned at
    its own steady rate, what were their
    original lengths?
    
    -- nearly impossible Brain Bafflers
    by Tim Sole and Rod Marshall

Solution:

    The red and white candles are the same length at 8 pm.
    The red candle burns to extinction in 4 hours.
    The white candle burns to extinction in 5 hours.

    The burn rate of the red candle = 5/4 * white burn rate.

    The white candle height at time 0 = H
    The white candle total burn time = 9-1/4 hours = 8+3/4 hours
        = 35/4 hours
    White candle burn rate = H / (35/4) = 4*H/35

    The red candle height at time 0 = H+2
    The red candle total burn time = 8 hours
    Red candle burn rate = (H+2)/8

    So: (H+2)/8 = 5/4 * 4*H/35 = H/7
    or: H+2 = 8/7 * H
    or: 2 = 1/7 * H
    or: H = 14

    White candle length = 14
    Red candle length = 16

    Sanity check:

    Red candle burn rate = 16/8 = 2 cm / hour
    White candle burn rate = 4/5 * 2 = 8/5 cm / hour
    White candle burn time = 14 / (8/5) = 35/4 hours [QED]


## Bonus

    A satellite is in an equatorial
    circular orbit 400 km above the
    surface of the earth. To place the
    satellite in geosynchronous orbit, its
    engine must be fired twice, once to
    place it into an elliptical transfer orbit
    with apogee at the geosynchronous
    altitude, and again at this
    apogee to place it in circular geosynchronous
    orbit. By how much, in
    m/s, must the magnitude of the
    satellite's velocity be increased at
    each burn?
    
    -- C.K. Galer , MI A'77

Solution:  I gave up and looked at the answer.  Very complex.

    First burn delta V = 2399 m/s
    Second burn delta V = 1457 m/s

Many questions:

1. I am not sure I understand why you can substitue ellipse axis
   in place of radius -- but I will accept it.

2. Also -- why is geosynchronous period 86164 instead of 86400?
   I guess this is 24 hours minus the fraction of a day the
   earth moves during this time.  If we assume a year is 365.24
   days then `86400*(1-1/365.24) ~ 86163.44`.  The value is given
   as precisely 86164 so maybe this is a better estimate of the
   length of a year.

3. Why is the potential energy negative?

The total energy of the system changes when a burn occurs but it
remains constant between burns.  This means that as the radius
increases the speed must decrease.  If the potential energy was
a positive value then as radius increases the value decreases.
If the total energy was constant this would require an increase
in velocity -- which is clearly wrong.  I guess this means the
sign convention adopted below is correct.

Many additional pieces of information are required:

    Mass of Earth = 5.983e24 kg
    Radius of Earth = 6378 km
    G = 6.67e-11 m^3/kg s^2
    For geosynchronous orbit: T = 86164 seconds (???)

    Basic strategy:

        E = K + U, K=Kinetic energy, U=Potential energy

        K = 1/2 * m * v**2
        U = - G*M*m/r**2 * r = -G*M*m/r

        Centripetal force = Gravity
        
        m*v**2/r = G*M*m/r**2

        Solving the last one for v**2 = G*M/r, so

        K = 1/2 * m * G*M/r = G*M*m/(2*r)
        U = -G*M*m/r

        E = K+U = G*M*m/(2*r) - G*M*m/r = -G*M*m/(2*r)

    For an elliptical orbit:

        Same equations apply for E except replace r with a where
        a is the major semi-axis. (Why???)

        For any orbit: K = E - U

            1/2*m*v**2 = -G*M*m/(2*a) + G*M*m/r
                       = G*M*m*(1/r-1/(2*a))
                  v**2 = G*M*  (2/r-1/a)
                  v    = sqrt(G*M*(2/r-1/a))

        For a cirular orbit: a=r so

                  v    = sqrt(G*M/r)

    Then they go nuts with subscripts.  i = initial orbit, g =
    geosynchronous orbit, t is the transfer orbit, a = apogee,
    and p is the perigee.

    In the initial orbit:

        vi = sqrt(G*M/ri)

            G, M given above
            ri = Radius of earth+height of orbit

        vi = sqrt(6.67e-11 * 5.983e24 / (6378e3 + 400e3)
           = 7673 m/s

    In the geosynchronous orbit:

        v = 2*pi*r/T --> vg = 2*pi*rg/Tg = sqrt(G*M/rg)
        (2*pi/Tg)**2 * rg**2 = G*M/rg
        rg **3 = G*M/(2*pi/Tg)**2 = G*M*Tg**2/(4*pi**2)

        But: Tg = 86164 so

        rg = ((G*M*86164**2)/(4*pi**2)) ** 1/3
           = 42181 km

        vg = sqrt(G*M/rg)
           = 3076 m/s

    In between (the transfer orbit): we slow down as we increase
    the radius.  We are trading kinetic for potential energy.

    The elliptic axis is 1/2 average of the two radii.

        a = 1/2 * ((6378+400)+42181) = 24480 km

    During the transition we have two different radii to
    consider: the min radius (perigee) and max radius (apogee).
    These must be considered separately:

    Initial orbit: ri= 6778 km, vi=7673 m/s
    Final   orbit: rg=42181 km, vg=3076 m/s

        v = sqrt(G*M*(2/r - 1 /a)

    At low point:  v = sqrt(G*M*(2/6778e3 - 1/24480e3))
                     = 10072 m/s

    At high point: v = sqrt(G*M*(2/42181e3 - 1/24480e3))
                     = 1619 m/s

    So - to get started in the transfer process we need to speed
    up from 7673 m/s to 10072 m/s = 2399 m/s.

    When we get out to geosynchronous orbit we will have slowed
    way down (to 1619 m/s) and we need to speed up the geosynchronous
    speed: go from 1619 m/s up to 3076 m/s = 1457 m/s

    So:

    First burn delta V = 2399 m/s
    Second burn delta V = 1457 m/s

    Sanity check for the transfer burn doing the calculations at
    full floating point precision:

        E = 1/2*m*v**2 - G*M*m/r
    so:
        E/m = 1/2*v**2 - G*M/r

    Transfer perigee: E/m = -8151100.807458311
    Transfer apogee:  E/m = -8151100.807458315

    Which agrees up to the final digit!


## Double Bonus

    Ann and Brandon
    play the following game. First, Ann
    puts $1 in the pot. Then, they take
    turns tossing a coin, with Brandon
    tossing first. If Brandon gets heads,
    he takes $1 from the pot; if he gets
    tails, he adds $1 to the pot. Then,
    Ann tosses the coin. If Ann gets
    heads, she takes $1 from the pot; but
    if she gets tails, she neither takes
    anything from the pot nor adds anything
    to the pot. They continue taking
    turns tossing the coin until the
    pot is empty, at which time the game
    ends. Is this a fair game? If so, why?
    If not, why not?
    
    -- Nikolaus Bernoulli (1687-1759)

Solution:  I cheated and saw the solution.  Very clever.

    The game is fair.

    This is a zero sum game.  That means that if it
    is fair for one player then it is fair for the
    other.  The result for Ann is difficult to
    determine but the result for Brandon is clearly
    a fair game.

    Observe: if it is fair for Brandon then it is fair for Ann.
    The rules for Ann could be arbitrarily complex and it would
    not change the results for a two person game.

