---
title: "Bent Winter 2016"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Winter 2016


## Problem 1

    Given one red cube and a supply of green
    cubes of the same size, what is the maximum
    number of green cubes that can simultaneously
    touch the red cube on at least part of a face
    under the condition that the surface of the
    red cube is completely covered by the
    touching green cubes? Touching means contact
    over a finite area; contact at only a corner
    or an edge is not considered touching.  

    -- Wheels, Life and Other
    Mathematical Amusements
    by Martin Gardner

TODO

    meh


## Problem 2

    Joan has drawn a right triangle on a 76.2 mm
    x 127 mm (standard 3"x5") index card and cut
    it out. She then cut the triangle into two
    right triangles and further cut each of those
    triangles into two right triangles. She
    discovered that all 12 sides of the four
    triangles were integral numbers of
    millimeters long. What was the size of the
    original triangle?

    -- An Enigma by Bob Walker
    in New Scientist

Solution:

    Generate all the right triangles that can fit
    on the card and then see how many can be split
    into triangles that are also on the list.
    
    Repeat this for each of the two triangles.
    
    Question: how do you partition a right
    triangle into two right triangles?  One
    way is to start at the right angle and
    create a line perpindicular to the hypotenuse.
    Are there any other ways?
    
```{r}
triangle=c()
for (a in 1:76) {
    for (b in a:127) {
        c=round(sqrt(a^2+b^2))
        if ((a^2+b^2)==c^2) {
            # add it to the list
            triangle=append(triangle,c(a,b,c))
        }
    }
}
triangle=matrix(triangle,ncol=3,byrow=T)
print(dim(triangle))

```

The legs of the original triangle (e.g. the first two columns)
must become the hypotenuse of the new triangle.

```{r}

k=apply(triangle,1,function(x) {(x[1] %in% triangle[,3]) & (x[2] %in% triangle[,3])})

print(sum(k))
```

Of these 13 triangles, only 5 can be parsed such that they produce two
right triangles with integer sides.  They are all multiples of 3,4,5 as
shown below:

    15   20   25  -->  9,12,15 and 12,16,20
    30   40   50  --> 18,24,30 and 24,32,40
    45   60   75  --> 27,36,45 and 36,48,60
    60   80  100  --> 36,48,60 and 48,64,80
    75  100  125  --> 45,60,75 and 60,80,100
    
In order to divide the triangles a second time, the values in the decomposition
list must also be one of the five possible dissectable triangles.  The only
triangle that meets this constraint is the last one.  The solution is then:

    75,100,125
        45,60,75
            27,36,45
            36,48,60
        60,80,100
            36,48,60
            48,64,80
            
Interestingly, two of the final four triangle are identical.


## Problem 3

    For dates from January through September, a
    digital 24-hour watch displays the date
    (month and day) and time as M:DD:hh:mm:ss.
    Occasionally, such as at 23:46:57 on August
    19, each of the digits 1 through 9 is used
    exactly once. How many times does this occur
    each year?

    -- Puzzles 101: A PuzzleMaster's
       Challenge by Nob Yoshigahara

Solution:

    Exhaustive enumeration is reasonable.  There are
    less than 24 million seconds in 3/4 of a year.

    The critical part is getting the number of days
    right in each month:

    I tried direct brute force and it was not good.
    Very slow.
    
    Try something else.  Find the times during the
    day when there are 6 distinct digits in the
    range 1:6.
    
    Can marry these with month/day separately.

```{r}

month_length = c(31,28,31,30,31,30,31,31,30)

n=1
times=c()
for (hour in 0:23) {
    for (minute in 0:59) {
        for (second in 0:59) {
            t1=sprintf("%02d%02d%02d",hour,minute,second)
            t2=strsplit(t1,"")[[1]]
            if (length(unique(t2))==6 && !any(t2=='0')) {
                times=append(times,t1)
                n=n+1
            }
        }
    }
}
print(sprintf("Number of cases with 6 distinct time digits = %d",n))

nfinal=0
tfinal=c()
for (month in seq_along(month_length)) {
    print(sprintf("Checking month = %d",month))
    for (day in 1:month_length[month]) {
        d=sprintf("%d%02d",month,day)
        d2=strsplit(d,"")[[1]]
        if (length(unique(d2))==3 && !any(d2=='0')) {
            for (t1 in times) {
                x=c(d2,strsplit(t1,"")[[1]])
                if (length(unique(x))==9) {
                    nfinal=nfinal+1
                    tfinal=append(tfinal,x)
                }
            }
        }
    }
}
tfinal=matrix(tfinal,nrow=9)
print(sprintf("Number of cases with 9 distinct date/time digits = %d",nfinal))
```

There are 768 times where this occurs.  The first such occurrence is:

    3/26 17:48:59

The last such occurrence is:

    9/28 17:56:43

It might be interesting to try to visualize this.


## Problem 4

    Find three three-digit integers that meet the
    following conditions.  The nine digits making
    up the three integers are all different. At
    least one of these integers is divisible by
    two, at least one is divisible by three, at
    least one by four, at least one by five, at
    least one by six, at least one by seven, at
    least one by eight, at least one by nine; and
    all three integers are divisible by eleven.
    What are the three integers?

    -- An Enigma by Susan Denham
    in New Scientist

Solution:

    Seems like a candidate for exhaustive search.
    Yes - the code is ugly but it was quick to write.
    Only multiples of 11 are tried so there is no
    explicity check for divisibility by 11.
    
    This leads to a total of 85320 combinations to consider.
    Of this total only 312 have 9 distinct digits.
    The remaining nested loops only get executed for these cases.

```{r}

for (n1 in seq(110,968,11)) {
    for (n2 in seq(n1+11,979,11)) {
        for (n3 in seq(n2+11,990,11)) {
            n=c(n1,n2,n3)
            if (length(unique(strsplit(paste0(n,collapse=""),"")[[1]])) != 9) {
                next
            }
            if (sum((n%%2)==0)>0) {
                if (sum((n%%3)==0)>0) {
                    if (sum((n%%4)==0)>0) {
                        if (sum((n%%5)==0)>0) {
                            if (sum((n%%6)==0)>0) {
                                if (sum((n%%7)==0)>0) {
                                    if (sum((n%%8)==0)>0) {
                                        if (sum((n%%9)==0)>0) {
                                            print(n)
                                        } # divisible by 9
                                    } # divisible by 8
                                } # divisible by 7
                            } # divisible by 6
                        } # divisibile by 5
                    } # divisibile by 4
                } # divisible by 3
            } # divisible by 2
        } # for n3
    } # for n2
} # for n1

```

The solution is: 165, 308, 792


## Problem 5

    Place the integers 1 through 12 on a regular
    hexagon such that six of these integers are
    at the vertices and the other six are at the
    midpoints of the sides, and each of these
    midpoint integers is the sum of the two
    adjacent vertex integers. How many different
    ways can this be done? Now, repeat for the
    integers 1 through 16 on a regular octagon.
    How many different ways can this be done for
    the octagon? Rotations and reflections are
    not considered to be different arrangements.

    -- An Enigma by Bob Walker in New Scientist

TODO

    meh


## Bonus

    A Bingo card consists of a 5 by 5 grid with
    numbers in each cell, except for the center
    square, which is marked "Free." The five
    columns are labeled B, I, N, G, and O and
    contain, respectively, numbers in the ranges
    1 through 15, 16 through 30, 31 through 45,
    46 through 60, and 61 through 75. The caller
    has 75 balls, numbered 1 through 75, and
    randomly draws balls and calls numbers until
    someone yells "Bingo!" Bingo is achieved when
    a player fills in a column, row, or major
    diagonal. Consider the case of a single
    player with a single card. With seven or
    fewer numbers having been called, what is the
    exact ratio of the probability of getting a
    Bingo using the "Free" cell to the
    probability of getting a Bingo not using the
    "Free" cell?

    -- D.A. Dechman, TX A'57 and
    F.J. Tydeman, CA Delta'73

TODO

    meh


## Computer Bonus

    Using the ten digits, 0 through 9, once and
    only once, create four integers A, B, C, and
    D such that A * B = C * D = N, where N is a
    minimum. Then, find four other integers where
    N is a maximum. Leading zeroes are not
    allowed. An apparent solution is 86 * 259 =
    74 * 301 = 22274, but this is not correct
    because 22274 is neither the minimum nor the
    maximum value of N.

    -- The Canterbury Puzzles by H. E. Dudeney

Solution:

    I could do this with exhaustive enumeration.  All
    of the permutations of the 10 digits give 10! = 3628800
    rows.  For each row need to consider all the possible
    numbers of digits = 210 total combinations.
    
    The grand total is 762e6 conditions to consider.
    
    The vast majority will fail the multiplication test.
    
```{r}
library(gtools)

digits=permutations(10,10,0:9)

```

    Now we need to partition these into the integers and check
    the conditions noted above.  This takes a LONG time to run
    so it has been commented out:
    
```{r, eval=FALSE}

to_int = function(x) {
    n=0
    if (x[1]==0) { return(999999) }
    for (i in x) { n = 10*n+i }
    return(n)
}

check_digits = function(x) {
    n=c()
    for (alen in 1:7) {
        for (blen in 1:(8-alen)) {
            for (clen in 1:(9-(alen+blen))) {
                a=to_int(x[1:alen])
                b=to_int(x[(alen+1):(alen+blen)])
                c=to_int(x[(alen+blen+1):(alen+blen+clen)])
                d=to_int(x[(alen+blen+clen+1):10])
                if (a*b == c*d) {
                    if (a*b < 6970) {
                        print(c(a,b,c,d,a*b))
                    }
                    n=append(n,c(a,b,c,d,a*b))
                }
            }
        }
    }
    return(n)
}

n=c()
m=0
for (i in 1:nrow(digits)) {
    n=append(n,check_digits(digits[i,]))
    m=m+1
    if ((m%%100000) == 0) { print(sprintf("Checking m = %d  n = %d",m,length(n))) }
}
print(sprintf("Number of solutions = %d",length(n)))
```


The range of solutions is: 2716 to 58560

According to the Bent the range is: 6970 to 58560.

Why do I find lower values?  It appears that the problem does not
allow for leading zeros.  Of course, this is clearly stated but I
missed it.

If we eliminate the cases of leading zeros then the solution
range is: 6970 to 58560.

