---
title: "Bent Spring 2011"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Spring 2011


## Problem 1

    Solve the following cryptic multiplication,
    where each different letter
    stands for a different digit, the same
    letter always stands for the same digit,
    and there are no leading zeros. An
    * can stand for any digit.

            RNAGEL
           *******
           -------
            RNAGEL
           LRNAGE
          ELRNAG
         GELRNA
        AGELRN
       NAGELR
      RNAGEL
      ------------
      R**N*AG**E*L

    What is the value of RNAGEL?

    -- 150 Puzzles in Crypt-
    Arithmetic by Maxey Brooke

Solution: done in cryptic arithmetic notebook

    Cryptic multiplication with unknowns

                  R N A G E L
                * * * * * * *
                ---------------
                  R N A G E L
                L R N A G E
              E L R N A G
            G E L R N A
          A G E L R N
        N A G E L R
      R N A G E L
      -----------------------
      R * * N * A G * * E * L 

      The asterisks are wildcards.  There are no leading zeros.
      What is the value of RNAGEL?

    1. Create all the values of RNAGEL and then check for multiplication
       by a single digit to generate the cyclic shifts and non-zero leading
       digits.  There are only two feasible solutions:

    > a[k,]
          r n a g e l
    20534 1 4 2 8 5 7
    22423 1 5 3 8 4 6

    2. Based on the final sum we conclude that N must be less than 5
       since there is no carry out to affect the most-significant digit.

    Conclude that the first solution is valid:

    > a[k,]
          r n a g e l
    20534 1 4 2 8 5 7

    Working the problem backwards we can fill in all the wild-cards:

                  R N A G E L
                1 3 2 6 4 5 1
                ---------------
                  R N A G E L
                L R N A G E
              E L R N A G
            G E L R N A
          A G E L R N
        N A G E L R
      R N A G E L
      -----------------------
      R 8 9 N 9 A G 1 0 E 0 L 
      1 8 9 4 9 2 8 1 0 5 0 7


## Problem 2

    A spider is chasing an ant. The spider
    is crawling counterclockwise at a
    speed of 701 cm/min on the circumference
    of a circle with a diameter of 100
    cm. The ant is crawling at a speed of
    700 cm/min, also counterclockwise, on
    a semicircle consisting of the upper
    half of the spider's circle plus a diameter.
    At the start of the chase, the ant
    is at the left end of the diameter and
    about to crawl along it, and the spider
    is at the other end of the diameter and
    ready to start crawling along the circumference
    of the circle (to which it is
    restricted). They commence crawling
    at the same instant. How many complete
    circuits of the circle must the
    spider make before it catches the ant?
    Idealize the problem by treating the
    spider and the ant as points.

    -- John H. Cook

Solution:

    The spider can only catch the ant under a very
    narrow range of conditions on the upper half of
    the circle.  The ant must start the upper half
    of the circle when the spider is nearly at the
    upper half of the circle.  In the distance of
    the upper half circle the excess speed of the
    spider must be sufficient to catch the ant.
    
    L = length of upper half circle = pi*D/2 = 50*pi
    Speed of ant = Sa = 700 cm/min
    Speed of spider = Ss = 701 cm/min
    
    Time for ant to crawl upper half circle = pi*D/(2*Sa)
    Distance travelled by spider during this time = pi*D/2*Ss/Sa
    
    So - at the time the ant starts the upper half circle if the
    spider is beyond 2*pi*D/2 - pi*D/2*Ss/Sa but less than pi*D/2
    from the start -- it should catch the ant.
    
    The ant starts the upper half circle at times of:
    
    t = D/Sa+(pi*D/2+D)/Sa*N, N = 0, 1, ...
    
    The location of the spider at these times is:
    
    t*Ss
    
    Need this value (modulo 2*pi*D/2) to be in the capture
    interval.
    
```{r}
D=100
Sa=700
Ss=701

xlow=2*pi*D/2+pi*D/2*(1-Ss/Sa)

print(paste0("Lower target = ",xlow))

t=D/Sa
for (n in 1:2000) {
  xs=(t*Ss) %% (2*pi*D/2)
  if ( (xs >= xlow) ) {
    print(paste0("Captured for n(ant) = ",n))
    print(paste0("t = ",t))
    print(paste0("xs = ",xs))
    print(paste0("n(spider) = ",(t*Ss)/(2*pi*D/2)))
    break
  } else {
    # print(paste0("n = ",n))
    # print(paste0("t = ",t))
    # print(paste0("xs = ",xs))
  }
  t = D/Sa+(pi*D/2+D)/Sa*n
}
```

    The spider has nearly completed 1005 circuits when the
    ant starts the upper half circle.  The spider will catch
    the ant on cycle 1006.
    
    Alternately, we can look at the spider position as the reference
    and calculate where the ant will be located at the start of
    each circuit:

    
```{r}
D=100
Sa=700
Ss=701

# Required headstart for ant to avoid capture.
# Time for spider to cover upper half circuit
tupper = pi*D/2/Ss
# Distance ant travels in this time
dant = tupper*Sa
# headstart required for ant to avoid capture
headstart = pi*D/2 - dant

print(paste0("Required ant headstart = ",headstart))

for (n in 0:2000) {
  d = 2*pi*D/2*n   # distance travelled by spider
  t = d/Ss                # time for start of the circuit
  xa= (t*Sa) %% (D + pi*D/2)  # position of the ant
  # print(paste0("n = ",n,"  t = ",t,"  d = ",d,"  xa = ",xa))
  if (xa<D) {
    # ant is safe -- it is on the diameter
    # print(paste0("Ant is on the diameter"))
  } else {
    if (xa>(D+headstart)) {
      # ant is safe -- it has a big enough head start
      # print(paste0("Ant position = ",xa))
    } else {
      # ant is captured
      print(paste0("Captured for n = ",n))
      print(paste0("t = ",t))
      print(paste0("d = ",d))
      print(paste0("xa = ",xa))
      break
    }
  }
}

```

Which is the same result as above.  The spider completes 1005 circuits
and captures the ant on the next loop.


## Problem 3

    Doris, the librarian, wishing to fill
    an empty bookshelf, asked her assistants
    how many books it would
    take to fill the shelf exactly. Al said
    it would take 2 catalogs, 3 dictionaries,
    and 3 encyclopedias; Bob said it
    would take 4 catalogs, 3 dictionaries,
    and 2 encyclopedias; and Connie said
    it would take 4 catalogs, 4 dictionaries,
    and 3 encyclopedias. As it turned
    out, only two of these estimates were
    correct. Desiring to fill the shelf with
    only one kind of book, Doris found
    that only one of the three types would
    exactly fill the shelf and that 15 of
    that type were required. Which type
    of book did Doris use? All the catalogs
    are the same width, all the dictionaries
    are the same width, and all the encyclopedias
    are the same width.

    -- Logical Deduction Puzzles
    by George J. Summers

Solution:

    It appears that 15 catalogs fills one shelf.
    

    Make assumption about who is telling the truth
    and then see if the equations work.
    
    C = width of catalogs
    D = width of dictionaries
    E = width of encyclopedias
    
    Al says:     2*C+3*D+3*E = 1
    Bob says:    4*C+3*D+2*E = 1
    Connie says: 4*C+4*D+3*E = 1

    Assume Al and Bob tell truth

      2*C+3*D+3*E = 1
      4*C+3*D+2*E = 1
      ---------------
      2*C - E = 0 -> E = 2*C
      

    Assume Al and Connie tell truth
    
      2*C+3*D+3*E = 1
      4*C+4*D+3*E = 1
      ---------------
      2*C+D = 0 -> Impossible
    
    Assume Bob and Connie tell truth

      4*C+3*D+2*E = 1
      4*C+4*D+3*E = 1
      ---------------
            D+E   = 0 -> Impossible
      
    Conclude that Connie is not correct.  So we know
    
    2*C+3*D+3*E = 1 \ 2*C - E = 0 --> E = 2*C
    4*C+3*D+2*E = 1 /
    
    15*C = 1, or 15*D = 1, or 15*E = 1
    
    Since 15*E = 30*C and only one kind of book would
    fill the shelf it could not be encylopedias.
    
    If it is catalogs then 15*C = 1 so: 3*D = 1-8/15 = 7/45
    
    so - an integer number of dictionaries will not fill shelf.
    
    If it is dictionaries then 15*D = 1 so: 8*C +3/15 = 1
    thus:
    
        C = 1/8*(1-3/15) = 1/8*12/15 = 1/2*3/15 = 1/10
        
    In this case 10 catalogs would fill the shelf which contradicts
    the given information that only one type of book works.


## Problem 4

    Take the aces, kings, queens, and
    jacks from a deck of cards and arrange
    the 16 cards in a four-by-four
    square array such that no two cards
    of the same rank or same suit are in
    the same row, the same column, or
    the same major diagonal. How many
    such arrangements are possible, if rotations
    and reflections are considered
    to be the same arrangement?

    -- A Gardner's Workout
    by Martin Gardner

TODO

    Still don't understand reflections, rotations, et. al.
    

## Problem 5

    In the town of Isobar, every rainy day is
    followed by a sunny day. Every sunny day is
    followed by either a rainy day or a sunny day
    with equal probability. To the nearest day,
    what is the expected number of sunny days in a
    365-day year? On a rainy day, it rains all
    day; on a sunny day, it is sunny all day.

    -- The Surprise Attack in Mathematical Problems by
    L.A. Graham

Solution:

    Number of Sunny days = S
    Number of Rainy days = R
    
    S = R + 1/2*S --> S = 2*R
    
    S+R = 365 --> R = 121.7, S = 243.3
    
    This seems too easy.


## Bonus

    Suppose that, instead of being a sphere, the
    Earth were a right circular cylinder (with a
    diameter equal to its height) of the same
    total volume and mass. What would be the
    gravitational acceleration on a person
    standing in the center of one of the circular
    faces? Assume the earth is a perfect sphere
    with a radius of 6,370 km and a uniform
    specific gravity of 5.518. Use a value of
    6.674E-11 m3/s2 kg for G, the universal
    gravitational constant.

    -- Howard G. McIlvried III, PA Gamma'53

Solution:

    Used this as a test vehicle to dust off Maxima.
    I had a lot of trouble doing this problem when
    I started -- I was trying to derive g for a
    spherical earth using polar coordinates with
    a differential volume elemnt and I
    never did get that to work.  My integral
    kept coming out negative.  Somehow I don't
    think the signs on the trig functions were
    quite right.
    
    Switching to a circular ring and cartesian
    coordinates worked much better.

    In the end, the result is:
    
    integrate(integrate(2*%pi*r*rho*G*h/(sqrt(h^2+r^2)^3),r,0,R),h,0,H);
    
    To make the cylindrical earth match the parameters of the
    problem we need R=H/2 and H=(16/3)^(1/3)*Radius of spherical earth
    
    Making these substitutions yields:
    
    g(cylinder) = 9.83659584535665
    
    Using a similar approach for the spherical earth yields:
    
    a: R-y;
    d: sqrt(r^2+a^2);
    assume(y<R);
    assume(R>0);
    integrate(integrate(G*rho*2*%pi*r*a/d^3,r,0,sqrt(R^2-y^2)),y,-R,R);
    
    g(sphere) = 9.826433967748834
    
    Amazingly close:
    
    Ratio = 1.001034136864011


## Computer Bonus

    Within the set of prime numbers is a subset of
    primes, the sums of whose digits are also
    primes; we will call such numbers double
    primes. For example, 23 is a double prime
    because it is a prime number whose digits sum
    to 5, which is also prime; while 13 is not a
    double prime since it is a prime whose digits
    sum to 4, which is not prime. Let PN be the
    Nth smallest prime, and let DN equal the
    number of double primes less than or equal to
    PN. For N = 1, 2, 3, 4, and 5, DN/N = 1.
    However, as more primes are considered, DN/N
    generally (although not continuously)
    decreases.  For example, when N = 9, DN/N =
    2/3. Determine the value of N such that DN/N <
    1/e for the first time, where e is the base
    for the natural logarithms.

    -- Samuel L. SanGregory, OH M'88

Solution:

    We shall try exhaustive search.  Not sure what else
    to try:

Sample R code -- runs for several hours to get solution:

  xxx ```{r}
    library(gtools)
    library(gmp)
    
    n=0
    dn=0
    m=1
    while (TRUE) {
      m=nextprime(m)
      n=n+1
      if (isprime(sum(baseOf(m)))) { dn = dn+1 }
      if (dn/n < 1/exp(1)) {
        print("Done")
        print(paste0("m = ",m))
        print(paste0("n = ",n))
        print(paste0("dn = ",dn))
        print(paste0("dn/n = ",dn/n))
        break
      }
      if (n%%10000 == 0) {
            print(paste0("Trying n = ",n,"   Ratio = ",dn/n))
      }
    }
  xxx ```

Solution takes a **LONG** time (several hours):

    m = 967718839
    n = 49288471
    dn = 18132215
    dn/n = 0.367879437769534
