---
title: "Bent Summer 2016"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Summer 2016


## Problem 1

    The planet Topsis is unusual in that each
    "hemisphere" is a right-circular cone of
    height 3000 km and base radius 4000 km. The
    two cones are joined base to base, so that
    the equator is the base circle and the
    vertical distance between the poles is 6000
    km. Topsy and Turvy are two cities located on
    the planet's equator 180 degrees apart. To
    the nearest km, what is the shortest distance
    between the two cities, assuming that one
    must travel along the surface of the planet?

    -- Superior Mathematical Puzzles by Howard P.
       Dinesman

Solution:

    Need the equivalent of a great circle route over
    the surface of the right circular cone.
    
    Need to write expression for arc length and
    then minimize it.  I could also try random walk.
    
    Maybe a quick web search will help.
    
    Seems like you can "flatten" the cone into a
    circular segment and then just draw a straight
    line on the flattened cone.
    
    If the height is 3000 km and the base radius is 4000 km
    then the length of a side is 5000 km.  When the cone is
    cut along one side and flattened the circle has a
    radius of 5000 km.  The length of the outside of the circle
    is just the circumference of the original cone base
    which is 2*pi*4000 km.  For a circle of radius 5000 km
    this will subtend and angle of 2*pi*4000/5000 = 1.6*pi radians.
    
    Points that were originally 180 degrees apart will be on the
    outer edge of the circle separated by an arc length of
    4000*pi.  This is a subtended angle of 0.8*pi when the
    radius is 5000 km.  The half angle is 0.4*pi.
    
    The right triangle has hypotenuse of 5000 km and half
    angle of 0.4*pi so half the distance is 5000*sin(0.4*pi).
    The full distance is twice that so:
    
    d = 10000*sin(0.4*pi) ~ 9510.565 km


## Problem 2

    A fair coin is flipped ten thousand times.
    What is the expected value of the absolute
    difference between the number of heads and
    the number of tails? Express your answer to
    four significant figures.

    -- Kevin M.T. Stewart, NJ Delta'77

Solution:

    We shall try Monte-Carlo
    
```{r}
ntrials=100000

n=0
set.seed(1)
for (trial in 1:ntrials) {
    flips=sample(2,10000,replace=T)
    n1=sum(flips==1)
    n2=sum(flips==2)
    n=n+abs(n1-n2)
}
print(n/ntrials) 

```

    Results
    
    Seed     trials   result
    --------------------------
    unknown  100000   79.97836
    42       100000   80.27992
    123      100000   80.27554
    2020     100000   80.07424
    1        100000   79.55706


    Analytically:

    P(T=49999) = C(10000,4999)/2^10000
    P(T=N)=P(H=N) so Ptotal = 2*P
    absolute difference is (5000-N)*2
    
    Together: p = sum(2*(5000-N)*2*C(10000,N)/(2^10000),N,4500,4999)
    
    For all n from 4500 to 5500: mean = 79.78646139382153

    For various number of terms:
    
    n     Value
    -----------------------
    4000  79.78646139382153
    4500  79.78646139382153
    4600  79.78646139382063
    4700  79.78646026183222
    4800  79.76080088969908
    4900  69.20379096537465


## Problem 3

    The digits 1, 2, and 3 can be combined in 27
    different ways to form a three digit number,
    e.g., 111, 112, 123, 232, 321, ..., 333.
    Arrange nine 1's, nine 2's and nine 3's
    around a circle so that each of the 27
    combinations of these three digits appear
    exactly once when reading the arrangement in
    a clockwise direction. For example, the
    sequence 1112321 would give the first five
    values above. Express your answer as a 27
    digit number, starting with 1112 and
    proceeding around the circle.

    -- Problem Solving Through Recreational
       Mathematics by Bonnie Averbach and Orin Chein

TODO

    meh


## Problem 4

    Five soccer playing countries were entered in
    2016 European Consolation Cup and played the
    usual one game against each of the others.
    Each nation played two games at home, two
    games away, and there were no draws. Exactly
    three of the games were won by the away side.
    No two countries won the same number of
    games. Azerbaijan won two. Belarus was away
    at Azerbaijan and at Cyprus. Cyprus played
    away at Azerbaijan. Denmark was home against
    Estonia. How many games did each nation win?

    -- A Tantalizer by Martin Hollis in New
       Scientist

TODO

    meh


## Problem 5

    Starting with 1, write the positive integers
    in continuous order without separating them
    as if you were writing a single number:
    1234567891011121314... The first and tenth
    digits in this number each are clearly 1. In
    addition to those first two, what are the
    hundredth, thousandth, ..., through trillionth
    digits in this number? That is, what digits
    occupy places 10n, for n=0 through 12?
    Express your answer as a sequence of 13
    digits.

    -- Mathematical Recreations by Maurice
       Kraitchik

TODO

    meh


## Bonus

    Alf, Bert, Charlie, and Doug were the first
    to volunteer as the crew for the British
    spacecraft. At the time with which our story
    is concerned, they were moving through space
    at a speed, and in a direction, such that the
    hands of their special chronometer, which
    allowed for local variations in time, had,
    since liftoff, been moving exactly twice as
    fast as those of the Greenwich Mean Time
    (GMT) clock. In the following exchange, all
    references to the passing of time as recorded
    by the GMT clock are in italics (time,
    minutes, hours, etc.), and all references to
    time recorded by the ship's chronometer are
    in a regular font (time, minutes, hours,
    etc.). The four space travelers speak, with
    their statements starting at 5 minute
    intervals:

    Doug: Oh, Alf! Oh, Bert! That all should come
    so clear! In less than half an hour,
    tomorrow's here, and our today is one with
    yesteryear.

    Alf: Yes, your today perhaps, but not so
    mine! You watch the clock; myself, I watch
    the time. 10 minutes since I heard the hour
    sound; and when another 23 minutes have
    passed, then 2 1/2 hours will have elapsed,
    since that moment we boarded, 34 minutes
    before we left the ground.

    Bert: Tomorrow and tomorrow and tomorrow!
    When will they both arrive? In more than 2
    hours. Alas! Before the time is what the time
    is now, only a half an hour must pass. Oh,
    Alf and Charlie, ere then I do pray, repent
    of the lies you've told this day. (In case I
    fail to make my meaning plain, you're liars
    both. Try not to lie again!)

    Charlie: I'm not a very clever chap. I cannot
    tell the time. I cannot add, nor yet
    subtract, much less compute a prime. I never
    learnt to cut a rug, some say that I'm
    uncouth. But unlike those liars Bert and
    Doug, Alf and I do tell the truth.

    Everything a speaker says is either all true
    or else everything he says is false. Prior to
    launch, the ship's chronometer is
    synchronized with GMT, and each speaker knows
    the veracity of every other speaker. What was
    the time (GMT) when they lifted off?

    -- Brain Puzzlers Delight
       by E.R. Emmet

TODO

    meh


## Computer Bonus

    In 2015, after a long career teaching
    mathematics, Dr. E. Trig retired. It was his
    habit each year to ask his students to find
    the smallest number which, when multiplied by
    the year, would result in an integer all of
    whose digits were either 0's or 1's. For
    example, 5,568,978 times 1995 equals the
    11-digit number 11,110,111,110. In 1945, the
    first year he asked, most of his students
    were able to find a suitable four digit
    number. Of the years for which Dr. Trig posed
    his question, which one had the longest
    required number and what was that number?

    -- An Enigma by Susan Denham
       in New Scientist

Solution:

    Exhaustive search?

    By far the worst case is for 1998.  The required multiplier is:
    
    556111667222778333889445
    
    Which yields a product of:
    
    1111111111111111111111111110
    
    I switched to Python.  This one particular solution consumes more than
    99% of the total solution time.
    
    Originally I tried brute force with R:
    
    1945 was fast
    1946 took a long time
    1947 was fast
    1948 is taking a long, long time
    by the time it got to 1951 I had manually switched to Python
    and then I terminated the R program.

    Python:
    1951 564326551
    
    R is still thinking about it
    
    I think I need a better generator function.  Increment by 1
    is very slow.
    
    Could I just generate all the k-bit binary numbers and interpret as decimal?
    
    I can do this in python with bin() and int() but I don't know how to do
    this in R.
    
    I rewrote in python and it runs blindingly fast until 1998 and then it stalls.
    Not sure what is going on -- eventually it finished and the rest were fast.
    The required multipliers are given below:
    
    Year Multiplier                  Solution Time
    -------------------------------------------------------------------
    1945 5198                        ( 0.0  seconds)
    1946 5190185                     ( 0.0  seconds)
    1947 51413                       ( 0.0  seconds)
    1948 56473825                    ( 0.001978635787963867  seconds)
    1949 564449                      ( 0.0  seconds)
    1950 518                         ( 0.0  seconds)
    1951 564326551                   ( 0.003988504409790039  seconds)
    1952 5128125                     ( 0.000997304916381836  seconds)
    1953 51720487                    ( 0.001995086669921875  seconds)
    1954 5174565                     ( 0.0009963512420654297  seconds)
    1955 563227622                   ( 0.004987955093383789  seconds)
    1956 5112475                     ( 0.0009965896606445312  seconds)
    1957 5115023                     ( 0.0009968280792236328  seconds)
    1958 5618545                     ( 0.000997781753540039  seconds)
    1959 5109750439                  ( 0.007979154586791992  seconds)
    1960 561225                      ( 0.0009970664978027344  seconds)
    1961 51                          ( 0.0  seconds)
    1962 5663155                     ( 0.002992868423461914  seconds)
    1963 566027                      ( 0.0009958744049072266  seconds)
    1964 5096792775                  ( 0.005982875823974609  seconds)
    1965 514                         ( 0.0  seconds)
    1966 55951735                    ( 0.002998828887939453  seconds)
    1967 51398633                    ( 0.0019958019256591797  seconds)
    1968 50818598125                 ( 0.012966632843017578  seconds)
    1969 5078776029                  ( 0.005980730056762695  seconds)
    1970 5588833                     ( 0.0019936561584472656  seconds)
    1971 51299341                    ( 0.001991748809814453  seconds)
    1972 5633925                     ( 0.0019931793212890625  seconds)
    1973 563107                      ( 0.0  seconds)
    1974 507097265                   ( 0.002991199493408203  seconds)
    1975 506836                      ( 0.0  seconds)
    1976 506072875                   ( 0.0029947757720947266  seconds)
    1977 506378913                   ( 0.003989696502685547  seconds)
    1978 56172958094545              ( 0.1775679588317871  seconds)
    1979 5053062709                  ( 0.005988359451293945  seconds)
    1980 5611672278338945            ( 0.6978261470794678  seconds)
    1981 5047960631                  ( 0.004985332489013672  seconds)
    1982 555                         ( 0.0  seconds)
    1983 55981347                    ( 0.001996278762817383  seconds)
    1984 55953125                    ( 0.0029494762420654297  seconds)
    1985 55466                       ( 0.0  seconds)
    1986 5090635                     ( 0.001035928726196289  seconds)
    1987 50830453                    ( 0.0009982585906982422  seconds)
    1988 553325                      ( 0.0009989738464355469  seconds)
    1989 503318809                   ( 0.0029921531677246094  seconds)
    1990 557789                      ( 0.0  seconds)
    1991 5580111                     ( 0.001001596450805664  seconds)
    1992 557781125                   ( 0.0011982917785644531  seconds)
    1993 557                         ( 0.0  seconds)
    1994 557221665                   ( 0.0  seconds)
    1995 5568978                     ( 0.0  seconds)
    1996 556112225                   ( 0.0  seconds)
    1997 500751633                   ( 0.015662431716918945  seconds)
    1998 556111667222778333889445    ( 192.48510360717773  seconds)
    1999 555777889                   ( 0.004984855651855469  seconds)
    2000 5                           ( 0.0  seconds)
    2001 4998001                     ( 0.000997781753540039  seconds)
    2002 5                           ( 0.0  seconds)
    2003 49925667                    ( 0.0009937286376953125  seconds)
    2004 4990025                     ( 0.0009982585906982422  seconds)
    2005 5541202                     ( 0.0009999275207519531  seconds)
    2006 553888335                   ( 0.004943370819091797  seconds)
    2007 5037873                     ( 0.0010373592376708984  seconds)
    2008 54781375                    ( 0.0019979476928710938  seconds)
    2009 49825789                    ( 0.002006053924560547  seconds)
    2010 49801                       ( 0.0  seconds)
    2011 4977628101                  ( 0.0052754878997802734  seconds)
    2012 54721675                    ( 0.0  seconds)
    2013 547                         ( 0.0  seconds)
    2014 55169365                    ( 0.0  seconds)
    2015 550874                      ( 0.0  seconds)
    
    Total solution time =  193.51430010795593  seconds


Preserved for posterity but actually used Python to get solution.

    
```{r, eval=FALSE}
library(gmp)

for (year in 1945:2015) {
    n1=as.bigz(year)
    n2=as.bigz(1)
    while(T) {
        n3=n1*n2
        if (all(strsplit(as.character(n3),"")[[1]] %in% c("0","1"))) {
            print(as.numeric(c(n1,n2)))
            break
        }
        n2=n2+1
    }
}

```

```{r}
library(reticulate)

```

```{python}
import time

tstart=time.time()
for n1 in range(1945,2016):
    t1=time.time()
    n2=1
    while 1:
        j=int(bin(n2)[2:])
        if j%n1 == 0:
            t2=time.time()
            print(n1,j//n1,"   (",t2-t1," seconds)")
            break
        n2=n2+1
tfinish=time.time()
print("Total solution time = ",tfinish-tstart," seconds")
```

The Bent has the same solution.  One of the authors used REXX and
another used Java.  They claimed run-times in the range of minutes
to hours depending on the language and the hardware.  The program
here runs in a little over 3 minutes.
