---
title: "Bent Winter 2001"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Winter 2001


## Problem 1

    Some space travelers landing on a
    distant planet find a thriving civilization
    similar to our own. However, those on
    the planet have eight fingers on each
    hand, and thus they use a hexadecimal
    system. By coincidence, they use coins
    with values of 1, 5, 10, 25, and 50, just as
    we do, but these numbers are
    hexadecimal. Furthermore, they love to
    gamble as evidenced by their toll bridge,
    which instead of being a fixed toll,
    flashes the toll as a random number from
    1 through 100, inclusive, and hexadecimal,
    of course. To use the exact change lane,
    how many coins of each denomination
    must a traveler carry in order to have the
    least total value of coins and still be able
    to exactly pay any toll?
    
    -- William W. Verkuilen, WI Beta'92

Solution:

We can convert these to base 10 and look at the equivalents.
This should be similar to the same problem in base 10 using
our conventional coins:

    0x01 = 1
    0x05 = 5
    0x10 = 16
    0x25 = 37
    0x50 = 80
    
The toll varies from 0x01 to 0x100 = 1 to 256 decimal.

For the smallest tolls we must carry 4 coins with value = 1.

If we add a coin with value = 0x05 we can cover all tolls from 1-9.
If we add a second coin with value = 0x05 we can cover 1-14

We need one more coin with value = 1 to cover a toll of 15.

At this point we can cover the tolls 1-15:

    0x01 Coins = 5
    0x05 Coins = 2

Adding a coin of value 0x10 gives all tolls 16-31.
Adding a second coin of value 0x10 gives all tolls up to 47.

We have overlapped with the coin of value = 0x25 (decimal 37) so we can
add one of those.  This allows us to cover tolls up to 47+37 = 84.

We have overlapped with the coin of value = 0x50 (decimal 80) so we can
add one of those.  This allows us to cover tools up to 84+80 = 164.

This leaves a gap of 256-164 = 92.  The most efficient way to cover this
is to add one more coin of value = 0x50 (decimal 80), two coins of value 0x05,
and two coins of value 0x01.

Final count:

    0x01 Coins  = 7
    0x05 Coins  = 4
    0x10 Coins  = 2
    0x25 Coins  = 1
    0x50 Coins  = 2
    ----------------
    Total Coins = 16
    Total Value = 7+4*5+2*16+37+2*80 = 256

Some questions:

1. Does this really cover all the totals?

2. Can I improve on this and still cover all possible totals?

Answer the second question first:  

We can't get rid of any coins because the total is just precisely enough to cover
the highest toll.  The only way to reduce the coin count is to trade multiple coins
of a lower denomination for ones of a higher value.

Consider the smaller coins: need at least 4 coins of value = 0x01 to cover the
lowest toll values.  This means that only 3 are negotiable.  This is not enough to
exchange for any other coin so we will leave this alone.

Consider the next coin (0x05): could trade three of these and a coin of value = 0x01
for a coin of value 0x10.  This would reduce the coin count by 3 but it would leave
a gap in the toll coverage.  The modified counts would be:

    0x01 Coins = 6
    0x05 Coins = 1
    0x10 Coins = 3
    
These can not cover a toll of 12, 13, 14, or 15.  In order to cover these cases we
would need to add 4 coins of value = 0x01 which is worse than where we started.
Conclude -- no change in the coins with value = 0x05.

Consider coins with value = 0x010: could trade two of these and a coin of value = 0x05
for a coin of value 0x25.  This would reduce the coin count by 2 but it would leave
a gap in the toll coverage.  The modified counts would be:

    0x01 Coins = 7
    0x05 Coins = 3
    0x10 Coins = 0
    0x25 Coins = 2

It is not possible to pay a toll of 23, 24, ..., 36 (and many more).
Conclude -- no change in the coins with value = 0x10.

There is only one coin of value 0x25 so an exchange is not possible.

Coins of value 0x50 are the largest coins to so no exchange is possible.

Final conclusion: this is a minimum set of coins.

We can verify the coverage of the totals with exhaustive search:

```{r}
coin_value=7+4*5+2*16+1*37+2*80
value=rep(0,coin_value)
for (c01 in 0:7) {
  for (c05 in 0:4) {
    for (c10 in 0:2) {
      for (c25 in 0:1) {
        for (c50 in 0:2) {
          total = c01+5*c05+16*c10+37*c25+80*c50
          value[total]=value[total]+1
        }
      }
    }
  }
}
all(value[1:256]>=1)
```

So we conclude that all possible tolls are covered.  In fact, some totals may be
achieved many different ways which some of the toll values are unique.  This
is shown below:

```{r}
table(value)
```

There are 29 toll values that can only be achieved with one combination of the coins.


## Problem 2

    What minimal number of knights is
    required so every square on a standard
    8 x 8 chessboard is either occupied by a
    knight or threatened by a knight? A
    square is threatened if a knight can
    move to that square. Recall that a
    knight moves two squares in any
    direction and then one square
    perpendicular to that direction to land
    on a square of opposite color. The move
    can occur even if the intervening
    squares are occupied.
    
    -- Daryl Cooper

TODO

## Problem 3

    An integer N consisting of five all-different
    nonzero digits has the curious
    property that it is equal to the sum of all
    the different three-digit integers formed
    by the three-digit permutations of its five
    digits. Find N.
    
    -- Thomas R. Diaz-Davilla, PR Alpha'96

Solution: seems like exhaustive search.  Use Python itertools.

```{r}
library(reticulate)
```

```{python}
import itertools

def make_num(a,nbase=10):
  """ Given digits in a
  Calculate value is given base.
  """
  n=0
  for i in a:
      n=nbase*n+i
  return(n)

def perm_sum(a,n=3):
  """ Given set of digits in a
  Calculate the sum of all n digit permutations
  """
  s=0
  for i in itertools.permutations(a,n):
      s=s+make_num(i)
  return(s)

for a in itertools.permutations([1,2,3,4,5,6,7,8,9],5):
  n1=make_num(a)
  n2=perm_sum(a)
  if (n1==n2):
    print(a)
```

The solution is 35964.

I was shocked at how fast this runs!  Less than a second!


Just for grins -- let's try the same thing with R gtools:

```{r}
library(gtools)

t1=Sys.time()
a=permutations(9,5)
for (i in 1:nrow(a)) {
  n1=as.integer(paste0(a[i,],collapse=''))
  b=permutations(ncol(a),3,v=a[i,])
  n2=sum(apply(b,1,function(x) {as.integer(paste0(x,collapse=''))}))
  if (n1==n2) {
    print(a[i,])
  }
}
t2=Sys.time()
print('Execution time:')
t2-t1
```

It yields the same answer as Python but the run-time is much slower.


## Problem 4

    In the tiny kingdom of Podunk, there
    are exactly 1,000 inhabitants, and they
    have their own vehicles with three-digit
    license-plate numbers from 000 through
    999. Plate 000 belongs to comedian Zero
    Mousetell. One day, someone noticed
    that the king's flag had been flown
    upside down, a considerable insult. A
    surveillance camera caught the
    departing car and showed that the first
    two digits of its license plate were 00
    but the third digit was blocked by a
    pole. Zero was the prime suspect. Most
    citizens were out of town attending a
    fair, but those at the fair did not see
    Zero there. On a random basis, how
    many citizens, not counting Zero, would
    have had to remain in town for Zero to
    have just less than a 50% chance of
    being the culprit?
    
    -- Kurt F. Hafner Jr.

Solution: we can try Monte-Carlo or an analytical solution.
Let us number all the other residents 1, 2, ..., 999.
We need to randomly sample N out of 999 and see how many low
license plate numbers (resident numbers) were present.
If this number is zero then Zero Mousetel is the only
suspect.  If the number is 1 then Zero is a 50% chance.
If the number is 9 then ZM has a 10% chance.

In general, if the number is m then the chance that ZM is
the cuplrit is 1/(1+m)


```{r}
n_trial=100000

n=rep(0,10)
a=c()
for (i in 1:n_trial) {
  s=sum(sample(999,168)<10)
  a=c(a,s)
  n[s+1]=n[s+1]+1
}

print(n)
sum(1/(1+0:9)*n)/n_trial

```

A quick run with trial and error suggests you need about 168 people in town.
This is hard to be precise due to Monte-Carlo variation.


## Problem 5

    To produce this odd-looking product
    table, I wrote four digits in ascending
    order from left to right across the top
    and repeated them down the left side
    and then filled in the product matrix in
    the usual way. Then I transliterated the
    numerals, replacing each digit with a
    different one consistently throughout.
    Finally, I rubbed out all the entries in
    the table except 1 x 1 = 0 and 7 x 9 = 54.
    What are the four digits I started with?
    
    +----+----+----+----+----+
    |  x |  1 |  3 |  7 |  9 |
    +----+----+----+----+----+
    |  1 |  0 |    |    |    |
    +----+----+----+----+----+
    |  3 |    |    |    |    |
    +----+----+----+----+----+
    |  7 |    |    |    | 54 |
    +----+----+----+----+----+
    |  9 |    |    | 54 |    |
    +----+----+----+----+----+
    
    -- Martin Hollis

TODO

## Bonus

    Dissect a square into the least
    number of right triangles with legs in the
    ratio of 2 to 1 such that the area of each
    triangle is different. What is this least
    number?
    
    -- Karl Scherer

TODO

## Double Bonus

    What is the largest 11-digit prime that
    contains each of the digits 0 through 9 at
    least once.
    
    -- Richard I. Hess, CA Beta'62

TODO

Need to learn more about primes and how to determine approximate locations.
I don't think I want to exhaustively search all the 11 digit numbers looking
for primes and then checking to see if they have the right pattern.

We might be able to eliminate some ranges looking at the sum of digits.
Consider just the first and last digits.  The central 9 digits are 012345678
in any order.  The last digit must be odd (otherwise not prime).  The last
digit can not be 0 or 5 (otherwise divisible by 5):

    98765432109 --> SOD = 54 --> divisible by 3 --> not feasible
    98765432107 --> SOD = 52 --> 
    98765432103 --> SOD = 48 --> divisible by 3 --> not feasible
    98765432101 --> SOD = 46 --> 
    
    89765432109 --> SOD = 54 --> divisible by 3
    89765432107 --> SOD = 54
    89765432103 --> SOD = 48 --> divisible by 3
    89765432101 --> SOD = 46

    etc.
    
    I looked for additional SOD rules to eliminate some of the other combinations
    but I couldn't really find anything useful.  Wikipedia has a page on division rules
    that might be useful but ...
    
    Need to think about this one some more.
    
