---
title: "Bent Fall 2011"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Fall 2011


## Problem 1

    Joe the butcher received a bill in the mail.
    Unfortunately it had gotten wet, and the first and
    last digits were unreadable. What was legible was:

    72 turkeys $_67.9_ 

    How much was the bill, and what
    is the cost of one turkey assuming they all cost
    the same?
    
    -- The Stanford Mathematics Problem Book,
    by G. Polya and J. Kilpatrick

Solution:

    Simple and quick search
    
```{r}
pmax=96799   # Maximum possible price of 72 turkeys, in cents
pmin=16790   # Minimum possible price of 72 turkeys, in cents

min_price=trunc(pmin/72)
max_price=trunc(pmax/72)

s=(min_price:max_price)*72

print(paste0("Price per turkey = ",s[grepl(".679.",s)]/7200))

```

The price per turkey is 5.11 with a total bill of 367.92


## Problem 2

    Determine the equation of a parabola that passes
    through the four points (xi, yi): (0, -1), (-1,
    0), (1, 0), and (0, 3).

    -- Craig K. Galer, MI A'77

Solution:

    The general equation for a parabola is given by:
    
    a*x^2 + b*x*y + c*y^2 + d*x + e*y + f = 0
    
    with b^2-4*a*c = 0
    
    We should be able to substitute in the four points, apply
    the constraint, and then solve for the 5 independent coeffs.
    
    (1) For (x,y) = (0,-1)   c-e+f=0
    (2) For (x,y) = (-1,0)   a-d+f=0
    (3) For (x,y) = (1,0)    a+d+f=0
    (4) For (x,y) = (0,3)    9*c+3*e+f=0
    (5) b^2-4*a*c=0

    From (2) and (3) we have: a = -f, d = 0
    
    This yields:
    
    a*x^2 + b*x*y + c*y^2 + e*y - a = 0
    
    c-e-a=0     \ 8*c+4*e = 0 --> e = -2*c
    9*c+3*e-a=0 /
    
    This yields:
    
    a*x^2 + b*x*y + c*y^2 - 2*c*y - a = 0
    
    For (x,y) = (0,-1)     c+2*c-a=0 --> a = 3*c
    
    So:
    
    3*c*x^2 + b*x*y + c*y^2 - 2*c*y - 3*c = 0
    
    but:
    
    b^2-4*a*c=0 -> b^2 = 4*a*c = 12*c^2 -> b= +/- 2*sqrt(3)*c
    
    Finally:
    
    3*c*x^2 +/- 2*sqrt(3)*c*x*y + c*y^2 - 2*c*y - 3*c = 0
    
    Dividing out the c:
    
    3*x^2 +/- 2*sqrt(3)*x*y + y^2 - 2*y - 3 = 0

    This is actually two different parabolas -- probably mirror
    images.  If we pick the + sign:

    Solving:
    
    y = +/- sqrt(2)*sqrt(2-sqrt(3)*x) - sqrt(3)*x+1
    
    or,
    
    y = 1 - sqrt(3)*x +/- sqrt(2)*sqrt(2-sqrt(3)*x)

    We can see the two parabolic solutions below:
    
```{r}
x=seq(-2,2,length=200)
y1a = suppressWarnings((-sqrt(2)*sqrt(2-sqrt(3)*x))-sqrt(3)*x+1)
y1b = suppressWarnings(sqrt(2)*sqrt(2-sqrt(3)*x)-sqrt(3)*x+1)
y2a = suppressWarnings((-sqrt(2)*sqrt(sqrt(3)*x+2))+sqrt(3)*x+1)
y2b = suppressWarnings(sqrt(2)*sqrt(sqrt(3)*x+2)+sqrt(3)*x+1)

xx=c(x,x)
y1=c(y1a,y1b)
y2=c(y2a,y2b)

plot(c(0,0),c(1,1),xlim=c(-2,2),ylim=c(-2,4),
     xlab='x',
     ylab='y',
     main="Parabola given 4 points: 2 Solutions")
points(xx,y1,col='blue')
points(xx,y2,col='red')
grid()

# Add the specified points
dotcolor='black'
points( 0,-1,pch=19,cex=2,col=dotcolor)
points( 1, 0,pch=19,cex=2,col=dotcolor)
points(-1, 0,pch=19,cex=2,col=dotcolor)
points( 0, 3,pch=19,cex=2,col=dotcolor)

```

## Problem 3

    Two perfect logicians, Ann and Beth, have each
    been assigned a number by a moderator. Neither
    woman knows either of the numbers, but they do
    know that the numbers consist of two positive odd
    integers that are either the same or differ by 2
    and that, if the two integers are different, then
    the larger one is assigned to Beth. Ann and Beth
    sit in front of computer monitors, arranged so
    that each can see only her own monitor. They take
    turns typing in questions of the form, "Is Y's
    number X?" about the other's number, where X is a
    positive odd integer. For example, Ann could ask,
    "Is Beth's number 7?", and Beth could ask, "Is
    Ann's number 11?" After receiving an honest yes or
    no answer on the monitor from the moderator, the
    questioner states aloud, either "I know my number"
    or "I don't know my number." The following
    conversation ensues:

    Ann: "I don't know my number."
    Beth: "I don't know my number."
    Ann: "I don't know my number."
    Beth: "I don't know my number."
    Ann: "I don't know my number."
    Beth: "I don't know my number."
    Ann: "I don't know my number."
    Beth: "I don't know my number."
    Ann: "I don't know my number."
    Beth: "I don't know my number."
    Ann: "I don't know my number."
    Beth: "I don't know my number."
    Ann: "I don't know my number."
    Beth: "I don't know my number."
    Ann: "I don't know my number."
    Beth: "I know my number."
    What are Ann's and Beth's numbers?

    -- adapted from
    All-Star Mathlete Puzzles by
    Dr. Richard I. Hess, CA B'62

TODO

    Don't care


## Problem 4

    Numbers b and c are chosen at random from the
    range of real numbers bounded by -q and q. Express
    the probability, as a function of q, that the
    quadratic equation, x^2 + bx + c = 0, will have
    complex roots.

    -- William S. Alderson, MI E'43

Solution:

    Complex roots will occur iff b^2-4*c<0 or c > 1/4 * b^2.
    
    If b and c are chosen from random [-q,q] then this is
    a square in b-c axis and the other curve is a parabola.
    
    We need the area inside the square but above the parabola.

    This is illustrated below:
    
```{r}
q=2

x=seq(-q-1,q+1,length=1000)
y=1/4*x^2

xs=c(-q,q,q,-q)
ys=c(-q,-q,q,q)

plot(c(x,xs),c(y,ys),type='n',xlab="b",ylab="c",
     main="Complex Quadratic Roots",
     sub="Case 1: q<4")
polygon(xs,ys,border='gray')
lines(x,y,lwd=3)

xp=seq(-q,q,length=100)
yp=1/4*xp^2
xp=c(xp,q,-q)
yp=c(yp,q,q)
polygon(xp,yp,col='pink')


###

q=8

x=seq(-q-1,q+1,length=1000)
y=1/4*x^2

xs=c(-q,q,q,-q)
ys=c(-q,-q,q,q)

plot(c(x,xs),c(y,ys),type='n',xlab="b",ylab="c",
     main="Complex Quadratic Roots",
     sub="Case 2: q>4")
polygon(xs,ys,border='gray')
lines(x,y,lwd=3)

# Calculate xmin/xma
xmin=-sqrt(4*q)
xmax=sqrt(4*q)
xp=seq(xmin,xmax,length=100)
yp=1/4*xp^2
xp=c(xp,xmax,xmin)
yp=c(yp,q,q)
polygon(xp,yp,col='pink')

```

    There are two cases to consider (depending on the value of q).
    If q is very large then the parabola will intersect the top
    of the square.  If q is very small then the parabola will
    intersect the sides of the square.
    
    The critical value of q that makes the parabola intersect the
    corner of the square is when q = c = 1/4*b^2 = 1/4*q^2
    
    or:
    
    q^2 - 4q = 0 --> q = 0, 4
    
    So: for q<4 the parabola intersects the the sides of the square.
    
        The area is then:
    
        integrate(q-1/4*x^2,x,-q,q) = 2*q^2 - q^3/6
    
        The total area for b,c is (2*q)^2 = 4*q^2
    
        The ratio of the area (e.g. the probability is):
    
        Pr = 1/2 - q/24
        
    For q>4, the parabola intersects the top of the square.
    
        The limits of the integration are:
    
            1/4*x^2 = q --> x = sqrt(4*q) = +/- 2*sqrt(q)
    
        Area between parabola and top of square:
        
            integrate(q-1/4*x^2,x,-2*sqrt(q),+2*sqrt(q)) = 8/3*q^(3/2)
            
        The total parameter space is: 4*q^2
        
        Pr = 8/3*q^(3/2) / (4*q^2) = 2/(3*sqrt(q))

    This might be all well and good but I still don't get a good feel for
    the actual probability.  Let's visualize for q = [0,8]
    
```{r}
q=seq(0,8,length=100)
pr=ifelse(q<4,1/2-q/24,2/(3*sqrt(q)))

plot(q,pr,type='b')
abline(v=4,col='red')
grid()
```

As q->0 the equation becomes:

    x^2 + b*x + c = 0
    
    but b,c are very small so x is very small.
    In this case b*x << c so:
    
    X^2 + c ~ 0
    
    and whether or not the equation has complex roots depends on the
    sign of c.  This is a 50% probability as shown in the plot.
    
As q->infinity the equation becomes:

    x^2 + b*x + c = 0
    
    If b,c are very large then b*x >> c so:
    
    x^2 + b*x ~ 0
    
    x*(x+b) ~ 0
    
    and the equation never has complex roots.  This also matches
    the general trend of the plot above.
    
    This will have complex roots is when b^2<4*c which
    becomes vanishingly unlikely as q->infinity.


## Problem 5

    A uniform billiard ball of diameter D rolls
    without slipping along a pool table toward a
    cushion. What height of cushion creates a rebound
    that maintains rolling-without-slipping motion?

    -- Jeffrey R. Stribling, CA A'92

Solution:

    Need a cushion high enough that the angular momentum is not
    enough to lift the ball off the table.
    
    The Bent solution uses angular acceleration and moment of
    inertia for a sphere.  Can we do this from first principles?
    
    The part of the ball that is above the cushion has momentum
    that will try to raise the ball off the table when it hits
    the cushion.
    
    Let h = height of rail
        r = D/2 = radius of ball
        
    At the point of impact, the mass above the rail will exert a
    moment at the point where the ball hits the rail.  The lower
    part of the ball exerts a countering moment due to gravity.
    
    ... okay ... I gave up and looked at Wikipedia for moment
    of inertia.
    
    The moment of inertia for a sphere is:
    
        I = 2/5*m*R^2
        
    Where this is about an axis through its center of gravity.
        
    Torque is related to moment of inertia as:
    
        T = I*alpha
        
    Where alpha is angular acceleration.
    
    The Torque from the upper half of the ball against the cushion
    is:
    
        F*(h-r) = I * alpha
        
    Where:
    
        F = m*a, a=alpha*r if there is no slipping
        
    Substituting:
    
        m*alpha*r*(h-r) = 2/5*m*r^2*alpha
        
    Solving for h:
    
        h = 7/5*r = 7/10*D

    I still don't get a warm and fuzzy feeling about this.
    

## Bonus

    A geosynchronous orbit is a circular orbit with an
    orbital period of 24 hours. These orbits,
    typically used by the equatorial stationary class
    of communications satellites, have a radius Rs
    equal to about 26,300 miles. Another class of
    satellites, the Russian Molniya class, employ
    inclined, highly eccentric orbits characterized by
    a 12-hour orbital period. Assuming that the
    closest approach to the earth's surface of a
    Molniya orbit is 100 miles to avoid burning in the
    atmosphere, for what range of eccentricities is
    the apogee greater than Rs?

    -- Technology Review

Solution:

    According to the Bent we need to look to Kepler:
    
        T^2 = 4*pi^2/(GM)*a^3
        
    where a is the semi-major axis of the orbit.  I would need
    to investigate orbital mechanics again to figure out what
    any of this means.
    
    The final solution has:
    
        a = 2.66e7 m

    Two solutions:
    
        (1) perigee has R = 100 miles plus radius of earth
                Rp = 6.531e6 m
                Ra = 4.67e7 m
        
        (2) apogee has R = Rs (radius of geosynchronous orbit)
                Ra = 4.23e7 m
                Rp = 1.09e7 m
                
    The range of e is [0.591, 0.755]


## Double Bonus

    What are the last seven digits of (7,777,777)^(7,777,777)?

    -- Adapted from Colin Singleton in New Scientist

Solution:

    I tried brute force with Maxima but I gave up - taking too long.
    I might try with sympy.
    
    I tried with gmp and it was pretty easy:
    
```{r}
library(gmp)

a=as.bigz(7777777)^7777777
b=as.character(a)
n=nchar(b)
print(paste0("Total number of digits = ",n))
print(paste0("Last 20 digits = ",substring(b,nchar(b)-20,nchar(b))))
```

    The last 20 digits are: 951673149961995347697
    The Bent gives the last 7 digits:     5347697

Clearly they did not intend brute force but it was quite simple.
