---
title: "Bent Spring 2010"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Spring 2010


## Problem 1

    Find a 3x3 magic square, consisting
    of nine different, although not necessarily
    consecutive, integers less than
    100, that has the following property.
    When the numbers in the magic square
    are replaced by the number of letters
    in their English names, the square is
    still magic with nine different integers.
    For example, 10 would be replaced by
    3 (ten has 3 letters) and 23 would be
    replaced by 11 (twenty-three has 11
    letters). A magic square is a square
    array of all different integers in which
    each row, column, and major diagonal
    sum to the same constant.

    -- 50 Mathematical IdeasYou
    Really Need to Know by Tony Crilly

TODO

    Some day maybe I should look into magic squares.
    

## Problem 2

    Ann and Beth live on Elm Street,
    which has houses numbered 1 through
    99, but neither knows the other's house
    number. Zack lives in another house
    on Elm Street, and the women are
    eager to discover his number. Ann
    asks Zack two questions: (1) is your
    number a perfect square, and (2) is it
    greater than 50? Having heard Zack's
    answers, she declares she knows his
    number, but she is wrong because only
    Zack's second answer is true. Beth, not
    having heard any of this, asks Zack two
    different questions: (1) is your number
    a perfect cube, and (2) is it greater than
    25? Like Ann, after hearing Zack's
    answers, she declares she knows his
    number, but she is wrong because
    again only Zack's second answer is
    true. Given the further information
    that Zack's number is less than both
    Ann's and Beth's numbers and that the
    sum of their three numbers is twice a
    perfect square, what are Ann's, Beth's,
    and Zack's house numbers?

    -- Brain Puzzlers Delight
    by E.R. Emmet

TODO

    I am sure I have solved this in the past
    but not very interested right now.
    

## Problem 3

    In a certain gambling institution, the
    game of Lucky-10 is played. The house
    provides a pot of $10. The player then
    tosses two coins simultaneously. If two
    heads appear, the player takes $1 from
    the pot; if two tails appear, the player
    adds $1 to the pot; and if a head and a
    tail appear, the player neither adds nor
    takes from the pot. After ten tosses
    (including draws), if there is exactly
    $10 in the pot, the player takes it. Otherwise,
    the house gets whatever is in
    the pot. How much should the player
    initially pay the house (to the nearest
    penny) to make this a fair game, that
    is, with expected winnings of 0?

    -- Joseph-Louis Lagrange
    (1736-1813)

Solution:

    I think this is a duplicate of a different
    problem (or a very minor variation).

    Spring 2005, Problem 2
    
    The result there was $1.76

[Duplicate problem](solution_2005_1.html#problem-2)


## Problem 4

    How many digits long is 100,000!?

    -- Prime Numbers, the Most
    Mysterious Figures in Math
    by David Wells

Solution:

    The product of integers in the factorial can be calculated
    by summing the logs.  In base 10, the integer part of the
    log indicates the number of digits (actually 1 less).
    
    SO: num digits = trunc(sum(log10(1:100000)))+1 = 456574
    
    BTW: using sympy we can actually calculate this:
    
    factorial(100000) = 2824229407... ending in a string of 24999 zeros


## Problem 5

    Solve the following cryptarithm, in
    which none of the letters is zero.

    AB/CD + EF/GHJ = A

    Each of the letters represents a different
    digit, and each digit is a different
    letter.

    -- Joseph Nabutovsky

Solution: solved in cryptic arithmetic notebook

    Rearrange to eliminate fractions:

      AB * GHJ + EF * CD = A * CD * GHJ

    source('rgc_lib.R')
    a=make_crypt_df('abcdefghj')
    x_ab=make_num(a,'ab')
    x_cd=make_num(a,'cd')
    x_ef=make_num(a,'ef')
    x_ghj=make_num(a,'ghj')
    x_a=make_num(a,'a')
    k=(x_ab*x_ghj + x_ef*x_cd == x_a*x_cd*x_ghj)
    a[k,][apply(a[k,],1,function(x) {all(x != 0)}),]
            a b c d e f g h j
    2062764 5 7 1 2 9 6 3 8 4

    m=2062764
    x_ab[m,] = 57
    x_cd[m,] = 12
    x_ef[m,] = 96
    x_ghj[m,] = 384
    x_a[m,] = 5

    ab/cd + ef/ghj = 57/12 + 96/384
                   = 19/4 + 1/4
                   = 20/4
                   = 5


## Bonus

    A hula-hoop is supported on a
    knife edge so that it hangs in a vertical
    plane. It is given a small push and set
    into back and forth oscillatory motion
    in the vertical plane. Later, the bottom
    half of the hula-hoop is cut off so that
    exactly one quarter of the hoop is on
    each side of the knife edge, and the
    half hoop is set into oscillatory motion
    in the vertical plane. What is the ratio
    of the period of oscillation of the half
    hoop to that of the whole hoop?

    -- Mad About Physics by Christopher
    P. Jargodzki and Franklin Potter

Solution:

    I am going to guess that this is just a pendulum
    problem where we need to calculate the center of
    mass for the full hoop vs. the half hoop.
    
    The ratio of these two lengths should define the
    ratio of the periods -- possibly including a sqrt.


## Computer Bonus

    Sequences of four
    consecutive primes whose last digits
    are 1, 3, 7, and 9, but whose other digits
    are the same, are not common. The
    first two such sequence are (11, 13, 17,
    19) and (101, 103, 107, 109). What is the
    thousandth such sequence?

    -- Don A. Dechman, TX A'57

Solution:

    We can try to enumerate sequential primes and then
    see if they have correct final digits.  If so, then
    check the leading digits.  If they match, then
    count the sequence.
    
```{r}
library(gmp)

p1=2
p2=nextprime(p1)
p3=nextprime(p2)
p4=nextprime(p3)
done=FALSE

n=0
while(! done) {
    if (p1%%10 == 1) {
        if (p2%%10 == 3) {
            if (p3%%10 == 7) {
                if (p4%%10 == 9) {
                    if ( ((p1-1)==(p2-3)) & ((p1-1)==(p3-7)) & ((p1-1)==(p4-9)) ) {
                        n=n+1
                        if (n %in% c(1,2,1000)) {
                            print(paste0("Sequence number = ",n))
                            print(c(p1,p2,p3,p4))
                        }
                        if (n==1000) {
                            done=TRUE
                            break
                        }
                    }
                }
            }
        }
    }
    p1=p2
    p2=p3
    p3=p4
    p4=nextprime(p4)
}
```

The 1000th such sequence is:

    11740031 11740033 11740037 11740039

