---
title: "Bent Spring 2008"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Spring 2008


## Problem 1

    A photographer has been hired to
    take a group picture of a high-school
    reunion that is attended by 100
    graduates. How many pictures should
    he take to have at least a 95 percent
    probability that he has at least one
    picture in which no one has his or her
    eyes closed? Assume that the average
    person blinks six times a minute
    and that their eyes are closed for
    one tenth of a second on each blink.
    Assume the shutter clicks instantaneously.

    -- H.G. McIlvried III, PA Gamma'53

Solution:

    Pr(blink) = 0.6/60 = 0.01
    Pr(noblink) = 0.99
    
    For 100 students: Pr(noblink)=(0.99)^100
       Pr(noblink) = 0.3660323
       
    For a given photo: Pr(failure) = 1-Pr(noblink)
                                   = 0.6339677
                                   
    For n photos we want Pr(failure)<0.05 so:
    
    0.6339677^n < 0.05
    
    n >= 6.573
    
    So we need at least 7 photos.


## Problem 2

    Six unit squares can be joined
    along their edges to form 35 different
    hexominos, the simplest being a one
    by six rectangle. How many of these
    hexominos can be folded along edges
    joining the squares to form a unit
    cube? A rotation or reflection is not
    considered to be a different hexomino.

    -- The Colossal Book of Mathematics
    by Martin Gardner

TODO


## Problem 3

    Many of you puzzle fans are familiar
    with the famous cryptic, SEND
    + MORE = MONEY. Here is the
    response: ALAS + LASS + NO +
    MORE = CASH. We want the solution
    to this cryptic addition with the
    biggest NO. Each different letter represents
    a different digit, but the same
    letter always represents the same
    digit. There are no leading zeros.

    -- Madachy's Mathematical
    Recreations by Joseph S. Madachy

Solution: solved in cryptic arithmetic notebook

    > a[k,][4,]
    
           a l s n o m r e c h
    423046 1 2 5 9 6 3 8 4 7 0  

    alas = 1215
    lass = 2155
    no   = 96
    more = 3684
    cash = 7150
    

## Problem 4

    The game of snooker is played
    with 22 balls (15 red balls, six colored
    balls, and a cue ball) on a six pocket
    table that resembles a standard pool
    table. The red balls are not numbered,
    but the colored balls are numbered
    2, 3, 4, 5, 6, and 7.
    A player scores one point for each
    red ball pocketed and the indicated
    number of points for pocketing a
    colored ball. A player must pocket
    a red ball, then any colored ball,
    then another red ball, then another
    colored ball, etc. The red balls are
    not respotted, but as long as any red
    balls are on the table, colored balls
    are respotted.
    After all 15 red balls are pocketed,
    the six colored balls are sunk in
    numerical order without being respotted.
    Al recently ran the table, sinking
    one ball at a time. His cumulative
    score after pocketing each of the first
    14 colored balls and the 15th red ball
    was always a semiprime (the product
    of two primes), but his score after
    pocketing each of the final six colored
    balls was never a semiprime.
    What was his final score?

    -- Richard England in New Scientist

TODO


## Problem 5

    Triangular numbers are numbers
    of the form Tn = n(n + 1)/2. They are
    called triangular numbers because Tn
    points can be arranged, bowling pin
    style, to form a triangle.
    The first few triangular numbers
    are 1, 3, 6, 10, and 15. What is the
    sum of the reciprocals of the triangular
    numbers from 1 to infinity?

    -- Mathematical Journeys
    by Peter D. Schumer

Solution:

    Sympy to the rescue: the sum is 2
    
    In [1097]: summation(1/(n*(n+1)/2),(n,1,oo))
    Out[1097]: 2

    Attempting to do it by hand?  Try partial fraction expansion
    
    S = summation(2/(n*(n+1))) = summation(2/n - 2/(n+1))
    
    As n=1..oo
    
        S1 = 2/1 + 2/2 + 2/3 + ...
        S2 =       2/2 + 2/3 + ...
        
        S = S1-S2 = 2

    Actually not that hard!


## Bonus

    Fred has five dice and decides
    to see how long it will take him
    to throw five 6s using the following
    procedure. He tosses all five dice.
    He then picks up and re-tosses those
    that are not 6s. He repeats this process
    until he has all 6s. What is the
    expected number of tosses to get five
    6s? Give the answer exactly in terms
    of a fraction.

    -- F.J. Tydeman, CA Delta'73

Solution:

    Start with Monte-Carlo and then do the analytics
    
```{r}
ntrials=1000000

set.seed(42)
n=0
for (trial in 1:ntrials) {
   m=0
   nd=5
   while (nd != 0) {
       s=sample(1:6,nd,replace=T)
       # print(s)
       m=m+1
       nd=length(s[s!=6])
   }
   n=n+m
}
print(n/ntrials)
```

For 1e5 trials

    Seed   N
    ===============
    42     13.01873
    123    13.03723
    2020   13.04215
    
For 1e6 trials

    Seed   N
    ===============
    42     13.01179

The Bent solution using combinatorics:

    3,698,650,986/283,994,711 = 13.024

    They work up to this by starting with one die.
    if N1 = expected number of tosses then the
    expression is:
    
        N1 = 1/6 + 5/6*(1+N1)
        
    e.g. there is a 1/6 chance of getting it in one
    toss and a 5/6 chance that it will take more
    than one toss.  The subtlety is the recursion
    in the term (1+N1).
    
    This did not occur to me.  I would express it like
    this:
    
        N1 = 1*1/6 + 2*5/6*1/6 + 3*5/6*5/6*1/6 + ... + n*(5/6)**(n-1)*1/6 + ...
           = summation(1/6*n*(5/6)**(n-1),(n,1,oo))
           = 6
           
    But we note:

        5/6*N1 = 1*5/6*1/6 + 2*5/6*5/6*1/6 + 3*5/6*5/6*5/6*1/6 + ...

        1+5/6*N1 = 1 + 1*5/6*1/6 + 2*5/6*5/6*1/6 + 3*5/6*5/6*5/6*1/6 + ...

        How do we prove this is N1?

        N1 = 1*1/6 + 1*5/6*1/6 + 2*5/6*5/6*1/6 + ... + (n-1)*(5/6)**(n-1)*1/6 + ...
                     1*5/6*1/6 + 1*5/6*5/6*1/6 + ... + 1*    (5/6)**(n-1)*1/6 + ...
        
           = 1/6 + 5/6*(1*1/6 + 2*5/6*1/6 + ... + (n-1)*(5/6)**(n-2)*1/6 + ...) +
                   5/6*1/6*(1 + 5/6 + ...             (5/6)**(n-2) + ...)
                   
           = 1 + 5/6*(1*1/6 + 2*5/6*1/6 + ... + (n-1)*(5/6)**(n-2)*1/6 + ...)
           = 1 + 5/6*N1
           = 1/6 + 5/6 + 5/6*N1
           
        N1 = 1/6 + 5/6*(1+N1)
        
    This just seems way too hard.  Why isn't it obvious from first principles?


    Solving for N1:
    
        N1*(1-5/6) = 1
        
        N1 = 6
        
    So, with only one die we expect to roll 6 times to
    get a 6.
    
    For two dice it gets a bit more complicated.
    They just straight to the combinatorics but we
    will try to derive it here:
    
    On first toss the odds of getting both 6s is:
    
        1/6*1/6
        
    This can only happen one way: C(2,0)
        
    The odds of getting one 6 and one non-6 is:
    
        1/6*5/6
        
    This can happen two ways: C(2,1)
        
    The ods of getting two non 6s is:
    
        5/6*5/6
        
    This can only happen one way: C(2,2)
    
    Summing up the probabilities:
    
        N2 = C(2,0)*1/6*1/6 + 
             C(2,1)*1/6*5/6*(1+N1) +
             C(2,2)*5/6*5/6*(1+N2)
             
    Solving for N2:
    
        N2 = 1/36 + 2*5/36*(1+N1) + 25/36*(1+N2)
        N1 = 6 (from above)
        
        N2 = 1/36 + 10/36*7 + 25/36*(1+N2)
        
        N2*(1-25/36) = 1/36 + 70/36 + 25/36 = 96/36
        N2*11/36 = 96/36
        N2 = 96/11 ~ 8.727273
        
    To get 2 6s with 2 dice requirs ~ 8.7273 tosses
        
    Continuing for N3:
    
        N3 = C(3,0)*1/6*1/6*1/6+
             C(3,1)*1/6*1/6*5/6*(1+N1)+
             C(3,2)*1/6*5/6*5/6*(1+N2)+
             C(3,3)*5/6*5/6*5/6*(1+N3)
             
    Solving for N3:
    
        N3 = 10566/1001

    All five solutions are:
    
    N   Expected Tosses         Numerical
    --------------------------------------------
    1   6                       6.00000000000000
    2   96/11                   8.72727272727273
    3   10566/1001              10.5554445554446
    4   728256/61061            11.9266962545651
    5   3698650986/283994711    13.0236615075553


## Computer Bonus

    A weak prime is a
    prime such that, when any of its digits
    is replaced (one at a time) by any
    different digit, the resulting integer
    is no longer prime. The two smallest
    weak primes are 294,001 and
    505,447. What is the smallest weak
    prime such that, when its digits are
    reversed, the result is also a weak
    prime? This one may take some time.
    As an exception to our usual rule, we
    will acknowledge correct solutions in
    the Fall column.

    -- R. Dudley White, VA Alpha'76

Solution:

    Not sure if this will really be a solution if they
    are admonishing a long solution time.  I will use
    the same sympy routine I used for 2004_2 to find
    weak primes and then check the palindrome.
    
    This is brute force but I don't have any other simple
    solutions.
    
    I could try to re-implement in gmp under R
    
    My sympy search is fairly brain-damaged and does a lot
    of checking of numbers that can't possibly be solutions.
    Need to skip numbers where the first digit is even or 5
    since they can not be prime.
    
    No solutions found up to 400e6.
    or 1e9
    or 2e9
    
    Had to run several hours but solution found:
    
        n =   3376225859
        n2 =  9585226733    
