---
title: "Bent Winter 2005"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Winter 2005


## Problem 1

    The latest release of the Prime Players
    consists of three CDs, each with a
    different playing time. Each CD has
    four tracks with a total playing time of
    not more than one hour. The four tracks
    on any of these CDs each last for a different
    prime number of minutes. Furthermore,
    any combination of three different
    tracks on any of these CDs play
    for a prime number of minutes. What
    are the lengths of the tracks on these
    three CDs?

    -- Andrew Gibbons in New Scientist

Solution: Fairy simple exhaustive search.  The primes are 2, 3, ...  Since the sum of
the four primes must not exceed 60 then the largest is 60-(2+3+5) = 50.  We can enumerate
all the primes up to 50 and look for combinations that meet the requirements above.

```{r}
library(gmp)
library(gtools)
```


```{r}
p=(1:50)[isprime(1:50)>0]

# Get all combinations of 4 primes that sum to 60 (or less)
a=combinations(length(p),4,p)
s=apply(a,1,sum)
a=a[s<=60,]

# Find the cases where any combination of three sums to a prime number
s=apply(a,1,function(x) { 
    m=combinations(4,3,x);
    all(isprime(apply(m,1,sum)))
    }
    )
a=a[s,]

# Candidate solutions
print(a)

```

The only three sets of primes that meet these criteria are:

    [1,]    5    7   17   19
    [2,]    7   11   13   23
    [3,]    7   13   17   23


## Problem 2

    The coordinates, in clockwise order,
    of the vertices of a heptagon H1 are (0,0),
    (2,2), (1,4), (3,6), (3,5), (7,3), and (5,1).
    Form a second heptagon H2 by joining
    the midpoints of adjacent sides of H1.
    Next, form heptagon H3 by joining the
    midpoints of the sides of H2. It should
    be clear that, because these heptagons
    decrease in area, if this process is continued
    indefinitely, the heptagon will
    ultimately shrink to a point. What are
    the coordinates of this point?

    -- You Are a Mathematician by David Wells

Solution: Let us look at a figure and then make some
calculations.

```{r}
h1x=c(0,2,1,3,3,7,5)
h1y=c(0,2,4,6,5,3,1)
plot(h1x,h1y,type='n',asp=1)
polygon(h1x,h1y,lwd=3)

x=h1x
y=h1y
# p=c('red','blue','green','cyan','black','yellow','orange')
p=palette()
for (i in 1:30) {
    x=c(x,x[1]); hx=(x[1:7]+x[2:8])/2
    y=c(y,y[1]); hy=(y[1:7]+y[2:8])/2
    polygon(hx,hy,border=p[i%%length(p)+1])
    x=hx; y=hy
}
grid()

```

In the limit this shold converge to a point.  We can do this
numerically or symbolically.  Since we are lazy the first attempt
will be numerically.

```{r}
# Given the starting value from H1
h1x=c(0,2,1,3,3,7,5,0)
h1y=c(0,2,4,6,5,3,1,0)

while (T) {
    h2x=(h1x[1:7]+h1x[2:8])/2
    h2y=(h1y[1:7]+h1y[2:8])/2
    xxx = (diff(range(h2x))+diff(range(h2y)))
    if (xxx < 1e-10) { break }
    h1x=c(h2x,h2x[1])
    h1y=c(h2y,h2y[1])
}
print(paste(h2x[1],h2y[1]))

```

Numerically the limit appears to by x=3, y=3.

These are the means of the original X,Y coordinates.

The solution calls this the centroid of the figure and demonstrates
that the centroid remains unchanged after a round of coordinate
conversions so it will eventually be the remaining point.


## Problem 3

    Two five-person tennis teams, the
    Aces and the Bulls, play each other in
    the finals of a tournament. The ranking
    of the 10 players from most skillful to
    least skillful is A1, B1, A2, B2, A3, B3,
    A4, B4, A5, and B5. Any player always
    beats a player of lesser skill. The finals
    consists of five one-on-one matches,
    with the winner being the team that
    wins three or more matches. If the pair-
    ings are determined by pulling names
    out of a hat, what is the probability that
    the Bulls will win?

    --Howard G. McIlvried III, PA Gamma'53

Solution:  we could try this analytically or Monte-Carlo.
Since I have no energy today--let's try Monte-Carlo.

Let us encode the values as odd and even integers.

    B5, B4, B3, B2, B1 = 1, 3, 5, 7, 9
    A5, A4, A3, A2, A1 = 2, 4, 6, 8, 10
    
For each trial we will pick a set of 5 from A and 5 from B.
Tabulate the results.

```{r}
ateam=c(2,4,6,8,10)
bteam=c(1,3,5,7,9)

num_trials=100000
num_b_wins=0

for (i in 1:num_trials) {
    a=sample(ateam,5)
    b=sample(bteam,5)
    if (sum(b>a)>2) { num_b_wins=num_b_wins+1 }
}

print(paste("Prob B wins = ",num_b_wins/num_trials,collapse=""))
```

B has a probability of ~ 0.225 of winning even with 1e7 trials.

We can try this again analytically.  It could get a bit messy
since the probability are not independent.  Let's just enumerate
all the permutations of the bteam and find out how many times it
beats the ateam:

```{r}
b=permutations(5,5,bteam)
s=apply(b,1,function(x) { sum(x>ateam) })

print(paste0('Permutations of B team = ',nrow(b),collapse=''))
print(paste0('Number of times B wins = ',sum(s>2),collapse=''))
print(paste0('Probability of B win = ',sum(s>2)/nrow(b),collapse=''))
```

So - enumerating all possibilities yields

    Permutations of B team = 120
    Number of times B wins = 27
    Probability of B win = 0.225


## Problem 4

    Joe was playing with his seven-digit
    calculator. It is a calculator
    where each digit is represented by
    some combination of seven line segments,
    i.e., an "8" uses all seven line
    segments. He displayed a positive integer,
    and his daughter Joan looked
    at it upside down. She declared, "I see
    a number that is a perfect square, and
    I don't see any leading zeroes." Joe
    then multiplied his number by 2. Joan
    looked at it upside down and declared,
    "I see another number, and again it
    is a perfect square." All digits except
    3, 4, and 7 are seen as digits when
    viewed upside down. What number
    did Joe originally display?

    -- Susan Denham in New Scientist

Solution: this is a cryptic arithmetic problem in disguise.
Upside down we get:

    0 --> 0
    1 --> 1
    2 --> 5
    3 --> Illegal
    4 --> Illegal
    5 --> 2
    6 --> 9
    7 --> Illegal
    8 --> 8
    9 --> 6

So the digits 1, 8, and 0 are unchanged.  When you turn the
calculator upside down, of course, the order of the digits is
reversed.

So, from the given problem:

    - Original number does not end in zero
    - The original number only contains 0, 1, 2, 5, 6, 8, 9
    - Original number is perfect square when upside down
    - The number * 2 also contains 0,1,2,5,6,8,9 - no trailing zero
    - The number*2 upside down is perfect square
    - Both numbers have <= 7 digits
    
Find by exhaustive search?

Unfortunately my exhaustive search fails to return any feasible solution!!!

Is it possible they are using "3" inverted to "E" for exponent?
Alternately, maybe "4" inverted to "h" for hexadecimal?

OOOOOps -- they say 2->2 and 5->5 --- LOL

I will fix it and try again.

Solution now exists.

```{r}

flip_digit = function(x) {
    # if (x=="2") return("5")
    # if (x=="5") return("2")
    if (x=="6") return("9")
    if (x=="9") return("6")
    return(x)
}

invert_display = function(n) {
    s=strsplit(sprintf("%d",n),"")[[1]]
    # Change 2->5, 5->2, 6->9, 9->6
    # Reverse order of digits
    for (i in seq_along(s)) { s[i]=flip_digit(s[i]) }
    s=rev(s)
    return(as.numeric(paste0(s,collapse="")))
}

is_flippable = function(n) {
    s=strsplit(sprintf("%d",n),"")[[1]]
    # Check to see if the number contains 3, 4, or 7
    # If so then the display will not be flippable
    return(!any(s %in% c("3","4","7")))
}

is_a_square = function(n) {
    return(round(sqrt(n))^2 == n)
}

zzz=c()
# First number from father
for (i in 1:9999999) {
    if ((i %% 10) == 0) { next }
    if (is_flippable(i)) {
        # First number for daughter
        m=invert_display(i)
        if (is_a_square(m)) {
            # Second number from father
            n=( (2*i) %% 1000000 )
            if (is_flippable(n)) {
                # Second number for daughter
                m2=invert_display(n)
                zzz=c(zzz,i)
                if (is_a_square(m2)) {
                    # This is a solution
                    print(c(i,n,m,m2))
                }
            }
        }
    }
}
```

After correcting for the fact that 2 and 5 remain flip invariant ...

The only solution is an original value of:

    n = 9126      Upside down = 9216 = 96^2
    2*n = 18252   Upside down = 25281 = 159^2


## Problem 5

    Three-dimensional tic-tac-toe is
    played on a 4 x 4 x 4 array of cells.
    Two players take turns by placing
    their symbol in any unoccupied cell.
    The winning player is the first one
    whose symbols occupy four cells in a
    row, either horizontally, vertically, or
    any of the 28 diagonals. What is the
    minimum number of cells that must
    be occupied to ensure that the other
    player cannot win? Submit a sketch
    (in pdf or jpeg format only, if your
    sketch is electronic) of each of the
    four horizontal layers that notes the
    positions of the occupied cells.

    -- Peter E. Rentz, IN A'55

TODO

Do I care?

Solution shows how to block the opponent.


## Bonus

    The staff at a bank includes a
    Director, two Assistant Directors, and
    five Department Heads. The Director
    wants several different padlocks on the
    strongroom door with keys distributed
    so that he can open the door alone, either
    AD can open the door with the
    other AD or with any two DH's (but
    not with only one DH), or else any four
    DH's can open the door (but not only
    three DH's). All padlocks must be open
    to open the door. No key opens more
    than one lock. No lock requires more
    than one key to open. Keys may be duplicated
    to more than one person. What
    is the minimum number of padlocks
    needed, and how should the keys be distributed?

    -- 100 Games of Logic by Pierre Berloquin

TODO

I think I have seen this before.  The *key* to this problem is to
realize that some people have more than one key and others don't.

According to the solution you need 20 locks where the Dir has all 20.
The others get split up as needed.  Solution is simply trial and error
until you figure it out.


## Computer Bonus

    A factorian is a
    positive integer, which is equal to the
    sum of the factorials of its digits. The
    first three factorians are 1 = 1!, 2 =
    2!, and 145 = 1! + 4! +5!. What is the
    next factorian? Note that 0! = 1 by
    definition.

    -- Keys to Infinity by Clifford Pickover

Solution: let us try exhaustive search

```{r}
is_factorian = function(n) {
    s=strsplit(sprintf("%d",n),"")[[1]]
    m=0
    m=sum(sapply(s,function(x) {factorial(as.integer(x))}))
    return(m == n)
}

t1=Sys.time()
n=146
while(T) {
    if (is_factorian(n)) {
        print(paste0("Solution = ",n,collapse=""))
    }
    n=n+1
    if (n>2540160) { break }
    # if ((n%%1000) == 0) { print(n) }
}
t2=Sys.time()
print(t2-t1)
```

The solution is 40585.

According to the solution this is the only factorian larger than 145.
You can not have 8 digit factorian.  Even if all the digits are 9 this
only yields 8*(9!) < 1e7 so the sum of the factorials can never be
large enough.

The largest feasible value is 7*9! = 2540160.

So - a search of 146 to 2540160 is sufficient to ensure all factorians
are found.

Searching the full range takes about 1.5 minutes

