---
title: "Bent Winter 2014"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Winter 2014


## Problem 1

    I have found six different six-
    digit positive integers, whose sum
    also has six digits, with all 42 digits
    having one of only two different
    values. If I told you the sum, you
    would be able to identify all six
    numbers (no leading zeros). What is
    this sum?

    -- An Enigma by Jan Kay in New Scientist

Solution:

    FWIW - this problem was very messy and not a lot
    of fun!  The Bent solution was virtually identical to
    what I did below but they made a table and hand
    calculated the sums to identify the unique one.

    Consider the MS Digit: it must be a 1.
    
    It can not be 0 (by problem statement) and no other digit than
    1 will allow for 6 values that sum to less than 10.  It could be
    something other than 1 mixed with some 1 digits but all of those
    cases would result in the MS digit of the sum that is not one
    of the two allowed digits.
    
    If we look at the least significant digit we know that the sum must
    end in one of the two digits.  We can search for candidate pairs of
    digits that can work.


```{r}
library(gtools)
```

```{r, eval=FALSE}
m=0
for (i in 1) {
    for (j in 0:9) {
        if (j==i) { next }
        a=combinations(2,6,c(i,j),repeats.allowed = T)
        n=apply(a,1,sum)%%10
        if (any(n==i) | any(n==j)) {
            print(c(i,j))
            m=m+1
        }
    }
}
print(paste0("Number of possible solutions = ",m))
```

    So the other digit must be 0, 2, 4, 6, or 8.

    The MS digit must sum to (at least) 6.  This eliminates
    all possibilities except:
    
        1-6
        1-8

    In both cases, all MS digits must be 1.  If there is no
    carry into the MS digit sum then the sum is 6.  If the
    carry into the MS digit sum is 2 then the sum is 8.
    
    The only was the carry into the MS digit is if the next
    column sum is less than 10.  There is no way for this to
    happen unless all the values are 1 (6+5>10 and 8+5>10).
    Conclude: the carry into the MS digit can not be 0 so the
    MS sum can not be 6.
    
    The sum must be 8 so the digits must be 1,8.

    After a lot of messing around I got candidates for each
    column but other than trial and error I don't know how
    to fill these in as the individual numbers.  Only two
    of the columns are unique and the other 4 columns have
    two possible sets of digits.

    (After saying this I looked at the problem statement -- the sum must
    be unique -- maybe that would limit it further.)

    I think I will punt and just look for all the combinations:
    
```{r, eval=FALSE}
library(gtools)

a=permutations(2,6,c(1,8),repeats.allowed=T)
n=apply(a,1,function(x) {as.numeric(paste0(x,collapse=""))})

# I hardcoded the loops since there are ~75e6 combinations of
# the 64 candidates taken 6 at a time.

thesum=c()
for (i1 in 1:59) {
    for (i2 in (i1+1):60) {
        for (i3 in (i2+1):61) {
            for (i4 in (i3+1):62) {
                for (i5 in (i4+1):63) {
                    for (i6 in (i5+1):64) {
                        n2=n[i1]+n[i2]+n[i3]+n[i4]+n[i5]+n[i6]
                        if (n2<1000000) {
                            d=unique(strsplit(as.character(n2),"")[[1]])
                            if (all(d %in% c("1","8"))) {
                                m=c(n[i1],n[i2],n[i3],n[i4],n[i5],n[i6],n2)
                                m=sort(m)
                                print(m)
                                thesum=append(thesum,n2)
                            }
                        }
                    }
                }
            }
        }
    }
}
```

The solution sets are:

        <------ First six numbers --------------> < Sum >
    [1] 111811 111818 111888 118888 181818 181888 818111
    [1] 111811 111818 111888 181818 181888 188888 888111
    [1] 111818 111881 111888 118818 181818 181888 818111
    [1] 111818 111881 111888 118888 181818 181888 818181 <-- Solution
    [1] 111818 111881 111888 181818 181888 188818 888111
    [1] 111818 111881 111888 181818 181888 188888 888181
    [1] 111818 111888 118818 181818 181881 181888 888111
    [1] 111818 111888 118888 181811 181818 181888 888111
    [1] 111818 111888 118888 181818 181881 181888 888181

The only unique sum is 818181 so the numbers are as seen above


## Problem 2

    You are to paint a totem pole,
    consisting of twelve animal images
    stacked on top of each other, using
    only two colors, blue and yellow. No
    two adjacent animals can be painted
    blue, although consecutive yellow
    colors are allowed. In how many
    different ways can you paint the
    totem pole?

    -- BrainMatics Logic Puzzles
    by Ivan Moscovich

Solution:

    No Blue: 12*Y
    1 Blue:  12 cases
    2 Blue:  66 cases but some have 2 blue in a row
        We can have BY+(10 cases of where to put blue)
        We can have YBY + (9 cases of where to put blue)
        Total of 19 cases
    3 Blue
    4 Blue
    5 Blue
    6 Blue
    >6 Blue is not possible without adjacent blue sections.
    
    This gets ugly in a hurry.  I cheated and peeked at the
    solution.  Let us try some shorter totem poles
    and see if we can get a pattern:
    
    One animal: 2 possibilities
    Y
    B
    
    Two animals:  3 possibilities
    B + (any of the 1 animal possibilities, don't start with B) = 1
    Y + (any of the 1 animal possibilities) = 2
    
    Three animals: 5 possibilities
    B + (any two animal possibilities, don't start with B) = 2
    Y + (any two animal possibilities) = 3

    Four animals: 8 possibilities
    B + (any of the three animal solutions, don't start with B) = 3
    Y + (any of the three animal solutions) = 5

    Five animals:
    B + (any four animal not starting with B) = 5
    Y + (any four animal solution) = 8

    etc.
    
    Apparently a Fibonacci sequence: 2,3,5,8,13,21,34,55,89,144,233,377
    Number of sections in the pole:  1,2,3,4,5, 6, 7, 8, 9, 10, 11, 12
    
    For n=12 the value is 377
    
    Generally: F(n) = F(n-2)+F(n-1)
    F(-1)=F(0)=1
    
    We can solve the difference equation:
    
    F(n) = (1/2 + 3*sqrt(5)/10)*(1/2 + sqrt(5)/2)**n +
           (1/2 - 3*sqrt(5)/10)*(1/2 - sqrt(5)/2)**n
 
    I used sympy to to this since I can no longer do anything
    remotely resembling algebra without making huge errors.
    I did rearrange the terms to make them prettier.
    
    I also used maxima to get:
    
                                  n         n                n
                     (sqrt(5) - 1)  c2 (- 1)    (sqrt(5) + 1)  c1
    (%o12)           ------------------------ + -----------------
                                 n                      n
                                2                      2

    With:
    
                           3 sqrt(5) + 5         3 sqrt(5) - 5
    (%o11)           [c1 = -------------, c2 = - -------------]
                                 10                    10


    This appears to be the same thing.  At least they generate the same
    sequence of numbers.
    
    Neither of these is as simple as I recall the solution to the Fibonnaci
    sequence being -- but maybe my memory is fading or my idea of complexity
    has changed.
    
    
## Problem 3

    Solve the following two cryptic
    addition problems simultaneously,
    that is, letters have the same values
    in both cryptics. Each different
    letter stands for a different digit.
    The same letter always stands for
    the same digit. There are no leading
    zeros. An * can stand for any digit.

    * * * * * *    * S E V E N * *
    * * * * * *    * S E V E N * *
    F O U R * *    ---------------
    S E V E N *    F O U R T E E N
    -----------
    E L E V E N

    What is the value of FOURTEEN?

    -- An Enigma by Albert Haddad in New Scientist

TODO

    Never was able to solve this in the cryptic
    arithmetic notebook.


## Problem 4

    "CRYPTOQUIP" is a puzzle in
    a lot of newspapers. It consists of
    a coded sentence or two with each
    letter of the message represented
    by a different unique letter of the
    alphabet. If all 26 letters are used,
    how many codes are possible?
    Either a closed form equation or an
    approximate numerical value is a
    permissible answer.

    -- Don A. Dechman, TX A A'57

Solution:

    I think the possible codes are just all the
    permutations of the letters.  In the case all
    of the letters are used: 26!.
    
        26! = 403291461126605635584000000

    If only a subset of the 26 codes are used then
    the value will be:
    
        C(26,n)*n!
    
    This is (at most) equal to 26! and may be far
    less.  For example, with a single character
    message we only have 26 difference choices.
    For 2 characters we have 26*25 = 250, etc.

    The Bent does not accept this solution.  They argue
    that any arrangement with the letters in the "correct"
    position is not a valid "derangement".  They give
    a result that is ~ 26! / e.
    
    They look at the maximum case of 26! and note that
    there are some codes in the "correct" location so you
    need to subtract those out: C(n,1)*(n-1)! but this is
    an over-correction so you need to add back in C(n,2)*(n-2)!
    etc. in an alternating sequence.
    
        Finally: n! * (1 - 1/1! + 1/2! - 1/3! + ... (-1)^n/n!)
    
    As n->infinity, the quantity in brackets goes to 1/e.
    

## Problem 5

    Consider a five-by-five array of
    25 points, one cm apart horizontally
    and vertically. Select five different
    points, none of which lie on either
    main diagonal of the array, such
    that the distances between pairs of
    points are all different. There are
    multiple solutions, but you only need
    to furnish one. Consider the array
    as consisting of five rows by five
    columns, each numbered 1 to 5, and
    express your answer as five pairs of
    numbers giving the row and column
    of the selected points.

    -- Puzzles 101: A PuzzleMaster's
    Challenge by Nobuyuki Yoshigahara

Solution:

    The puzzle grid is shown below where
    X is not permissible and . is allowed:
    
        1  2  3  4  5
       --------------
    1 | X  .  .  .  X
    2 | .  X  .  X  .
    3 | .  .  X  .  .
    4 | .  X  .  X  .
    5 | X  .  .  .  X

    One proposed solution is shown below where
    the selected points are shown with o:
    
        1  2  3  4  5
       --------------
    1 | X  o  .  o  X
    2 | o  X  .  X  .
    3 | .  .  X  .  o
    4 | .  X  .  X  .
    5 | X  .  o  .  X

    The coordinates are:
    
        2,1
        1,2
        1,4
        3,5
        5,3
        
    This fails to work since the distance from the last
    point to the second and third is identical.  Turns out
    this is hard to solve by trial and error.  I think
    exhaustive search might be easiest.
    
    There are only 16 permissible locations for the 5 dots.
    This is only 4368 combinations to check
    
```{r}
a=data.frame(row=c(1,1,1,2,2,2,3,3,3,3,4,4,4,5,5,5),
             col=c(2,3,4,1,3,5,1,2,4,5,1,3,5,2,3,4))

b=combinations(16,5)

for (i in 1:nrow(b)) {
# for (i in c(135)) {
    d=c()
    for (j in 1:4) {
        for (k in (j+1):5) {
            x1=a[b[i,j],1]; y1=a[b[i,j],2]
            x2=a[b[i,k],1]; y2=a[b[i,k],2]
            d=append(d,((x1-x2)^2+(y1-y2)^2))
        }
    }
    if (length(unique(d))==10) {
        print(i)
    }
}
```

Yields many, many solutions.  One is shown below:

    > a[b[135,],]
       row col
    1    1   2
    2    1   3
    4    2   1
    12   4   3
    13   4   5

        1  2  3  4  5
       --------------
    1 | X  o  o  .  X
    2 | o  X  .  X  .
    3 | .  .  X  .  .
    4 | .  X  o  X  o
    5 | X  .  .  .  X

    The (squared) distances are:
    
        2, 5, 8, 18, 1, 10, 18, 9, 13, 4


## Bonus

    Given a three-by-three grid
    of pigeonholes, fastened to a wall
    with the middle hole covered, place
    envelopes in any or all
    of the outside holes so
    that the sum of the
    number of envelopes
    on each of the four
    sides has the same value, N. For
    example, if N is 10, one solution is
    shown in the accompanying figure.
    How many ways, including reversals,
    rotations, and reflections as different
    arrangements, can this be done?
    Express your answer as a function
    of N.

        1 9 0
        7 * 8
        2 6 2


    -- The Canterbury Puzzles by H.E. Dudeney

TODO

    meh

## Computer Bonus

    Find two different ten-digit numbers,
    each using the digits 0 through 9
    exactly once, such that their square
    roots are the reverses of each other.

    -- You'd Better Be Really
    Smart Brain Bafflers by
    Tim Sole and Rod Marshall

Solution:

    Let's try exhaustive search for $200 Alex.
    
```{r}
library(gmp)

for (i in 31622:100000) {
    n1=as.bigz(i)
    n2=as.bigz(paste0(rev(strsplit(as.character(n1),"")[[1]]),collapse=""))
    n1str=as.character(n1^2)
    n2str=as.character(n2^2)
    if ( (nchar(n1str)==10) && (nchar(n2str)==10) ) {
        # Check for 10 distinct digits
        if (length(unique(strsplit(n1str,"")[[1]]))==10) {
            if (length(unique(strsplit(n2str,"")[[1]]))==10) {
                if (n1str != n2str) {
                    # Solution found
                    print(paste0("n1 = ",n1))
                    print(paste0("   n1^2 = ",n1str))
                    print(paste0("n2 = ",n2))
                    print(paste0("   n2^2 = ",n2str))
                }
            }
        }
    }
}
```

Solution set is:

    n1 = 54918  -->  n1^2 = 3015986724
    n2 = 81945  -->  n2^2 = 6714983025

    Just FYI: this is sooooo much easier in Python.
