---
title: "Bent Spring 2003"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Spring 2003


## Problem 1

    The ratio of the area of a right
    triangle with integral sides (each less
    than 100 units) to its perimeter is a
    single-digit integer. The value of this
    integer is written on a folded piece of
    paper. If you opened the paper and
    saw the integer, you could determine
    the dimensions of the triangle. Given
    that hint, what are the lengths of its
    sides?
    
    -- The Platinic Corner

Solution:

    The only solution is a=48, b=64, h=80, area/perimeter = 8
    
    In my initial analysis I missed the constraint of 100 units on
    the hypotenuse.

    Not sure what the "standard" integer triangles are.
    See the search below -- there are a lot more than I thought.
    

```{r}
zzz=data.frame(a=NA,b=NA,h=NA,p=NA,k=NA)
num_triangles=0
for (a in 1:99) {
    for (b in a:99) {
        h=sqrt(a*a+b*b)
        if (h==round(h)) {
            num_triangles = num_triangles+1
            p=a+b+h
            k=(a*b/2)/p
            if (k==round(k) & k<10) {
                zzz=rbind(zzz,c(a,b,h,p,k))
            }
        }
    }
}
zzz=zzz[!is.na(zzz$a),]
print(paste0('Number of integer sided triangles = ',num_triangles))
```

    I don't understand how you are supposed to be able to solve this?
    There are no unique solutions.
    
    1 2 3 4 5 6 7 8 9  <-- ratio of area to perimeter
    2 3 6 3 4 6 3 2 3  <-- number of solutions

    Oops -- missed one of the constraints.  All sides (including hypotenuse) must be
    less than 100 units.  Repeating the analysis yields:
    
```{r}
zzz=data.frame(a=NA,b=NA,h=NA,p=NA,k=NA)
num_triangles=0
for (a in 1:99) {
    for (b in a:99) {
        h=sqrt(a*a+b*b)
        if (h==round(h) & h<100) {
            num_triangles = num_triangles+1
            p=a+b+h
            k=(a*b/2)/p
            if (k==round(k) & k<10) {
                zzz=rbind(zzz,c(a,b,h,p,k))
            }
        }
    }
}
zzz=zzz[!is.na(zzz$a),]
print(paste0('Number of integer sided triangles = ',num_triangles))
table(zzz$k)
```
 
From above -- the solution for k=8 is unique.
 
```{r}
zzz[zzz$k==8,]
```

So - the only solution is a=48, b=64, h=80, area/perimeter = 8


## Problem 2

    Assume that Venus, Earth, and
    Mars are in circular, coplanar orbits
    around the Sun, with orbital times of
    225, 365, and 687 Earth days, respectively.
    If all three planets are aligned
    along a diameter of the orbit of Mars,
    what is the minimum time (in Earth
    years) before they are again aligned
    on a diameter of Mars' orbit?
    
    -- D.A. Dechman, TX A'57

Solution:

    Need to figure out how to do the modulo arithmetic and solve
    these equations
    
    T = time
    
    mars  = angle of mars  = 360*T/687 %% 360
    earth = angle of earth = 360*T/365 %% 360
    venus = angle of venus = 360*T/225 %% 360

    Need: mars = earth = venus

    venus overtakes earth every 586.6071 days (1.607143 earth years)
    earth overtakes mars  every 778.7422 days (2.13354 earth years)
    
    Need them both to occur at the same time.
    
    586.6071*m1 = 778.7422*m2 where m1,m2 = integer.

    Approximately: occurs at m1=839, m2=632 -- 1189.545 earth years

    As an upper bound:
    
    Prime factors:  687=3*229, 365=5*73, 225=3*3*5*5
    
    LCM = 3*3*5*5*73*229 = 3761325 days
    
    They will all be back in the same position in 10305 earth years.


    For their solution -- they claim t=368+1/28 earth years.
    Turns out this puts venus and mars on the same side of the sun
    and the earth on the other side of the sun.
    
    I admit -- I was not thinking about the other side of the sun
    
    I was only considering all planets on the same side of the sun!!!

    Concidentally (???) 10305 shows up in their solution as earth days.

## Problem 3

    Find a 10-digit integer containing
    the digits 0 through 9 each once such
    that, for n equal to 1 through 10, the
    integer formed by the first n digits is
    divisible by n.
    
    -- Nearly Impossible Brain Bafflers
    by Tim Sole and Rod Marshall

Solution:

    Try exhaustive search using all permutations of the 10 digits.
    
    Only solution = 3 8 1 6 5 4 7 2 9 0

Test:
```{r}
library(gtools)


make_int=function(a,m) {
    n=a[,1]
    for (i in 2:m) { n=10*n+a[,i] }
    return(n)
}

a=permutations(10,10,0:9)

# Start eliminating rows
# First column can not be a zero
# Last column must be a zero
# other columns require specific divisibility rule

k1=a[,1]!=0; a=a[k1,]
k2=make_int(a,2); a=a[k2%%2==0,]
k3=make_int(a,3); a=a[k3%%3==0,]
k4=make_int(a,4); a=a[k4%%4==0,]
k5=make_int(a,5); a=a[k5%%5==0,]
k6=make_int(a,6); a=a[k6%%6==0,]
k7=make_int(a,7); a=a[k7%%7==0,]
k8=make_int(a,8); a=a[k8%%8==0,]
k9=make_int(a,9); a=a[k9%%9==0,]
k10=a[,10]==0;    a=a[k10,]

a
```


## Problem 4

    This cryptic will be more interesting
    if you have a knowledge of
    German. VIER and NEUN are both
    perfect squares. If you were told the
    number represented by VIER, you
    could deduce the number represented
    by NEUN. Alternatively, if
    you were told the number represented
    by NEUN, you could deduce
    the number represented by VIER.
    What are the values of VIER and
    NEUN? (The same letter has the
    same value in both words, and there
    are no leading zeros.)
    
    -- Richard England in New Scientist

Solution: in the cryptic arithmetic notebook

    VIER = 6241
    NEUN = 9409


## Problem 5

    From a large supply of black and
    white marbles, 100 black and 100
    white marbles are placed in an urn.
    Three marbles are drawn at random
    from the urn and, depending on the
    marbles drawn, certain marbles are
    put back in the urn:

    Marbles Drawn       Marbles Put Back
    3 black             1 black
    2 black, 1 white    1 black, 1 white
    1 black, 2 white    2 white
    3 white             1 black, 1 white

    If this procedure is followed until
    fewer than three marbles remain in
    the urn, what is the probability that
    they are all white?
    
    -- Mathematics Teacher

Solution:

    I will try Monte-Carlo first and then try
    it analytically.
    
    From the preliminary results it seems there are always
    two white marbles remaining!
    
```{r}
max_white=0
max_black=0
num_black=100
num_white=100
total=num_black+num_white

while (total>=3) {
    pool=c(rep(0,num_black),rep(1,num_white))
    s=sample(pool,3)
    draw_white=sum(s==1)
    draw_black=sum(s==0)
    num_black=num_black-draw_black
    num_white=num_white-draw_white
    if (draw_black==3) { num_black=num_black+1 }
    if (draw_black==2) { num_black=num_black+1; num_white=num_white+1 }
    if (draw_black==1) { num_white=num_white+2 }
    if (draw_black==0) { num_black=num_black+1; num_white=num_white+1 }
    total=num_black+num_white
    if (num_white>max_white) {max_white=num_white}
    if (num_black>max_black) {max_black=num_black}
}

print(paste(num_black,num_white))
print(paste(max_black,max_white))
```

The analysis is actually fairly easy -- but I did peek at their solution.

    There are four cases.  Consider what happens to the number
    of white/black marbles in each case:
    
    Marbles Drawn       Marbles Put Back    White Marbles    Black Marbles
    3 black             1 black             No change        Decrease by 2
    2 black, 1 white    1 black, 1 white    No change        Decrease by 1
    1 black, 2 white    2 white             No change        Decrease by 1
    3 white             1 black, 1 white    Decrease by 2    Increase by 1
    
    The number of white marbles in mono-decreasing and always an even number.
    if you play the game long enough you must be left with 2 white marbles.
    You can never get lower than this so when the total number of marbles drops
    to less than three they must both be white.
    
    The number of black marbles mostly decreases (although it can go up when
    you draw three white marbles).


## Bonus

    Consider the parabola y = x^2,
    and let P be a point (other than the
    origin) on this parabola. Construct a
    perpendicular to the parabola at P
    (i.e., a perpendicular to the tangent at
    P). This perpendicular, when extended,
    will intersect the parabola at
    Q. Find the coordinates of P (with x
    positive) such that the area bounded
    by the perpendicular and the parabola
    is a minimum.
    
    -- Southwest Missouri State University
    Advanced Problem Archive

Solution:

    The final solution is x0=1/2 to minimize the area.
    Point P = (1/2, 1/4)
    Area = 4/3

    We know the location of the point P, the
    slope of the tangent = dy/dx @ P so we can
    construct the perpindicular line to the tangent.
    
    From there we can calculate the point of intersection
    with the parabola for x<0, call it P'.
    
    Finally, we can integrate from P' to P to calculate the
    area and then differentiate to minimize the area.
    
    Assume P=(x0,y0)
    Then:  dy/dx = 2*x = 2*x0 @ point P.
    Slope of perpindicular = -1/(2*x0)
    
    Equation of line:
    
    At point P: y0 = -1/(2*x0)*x0+b --> b=y0+1/2
    so:
        y = -1/(2*x0) * x + y0 + 1/2
        
    This must intersect y=x^2 at point P' with x<0.
    
        y = -1/(2*x0)*x + y0 + 1/2 = x^2
        Solve this for x at P' --> x0'
        
    Area = integrate(-1/(2*x0)*x+y0+1/2-x^2,(x,x0',x0))

    dA/dx0 = 0 --> solve for x0, y0
    
    I tried this and it really didn't work.
    
    I substituted in x0**2 in place of y0 and it was much
    better.
    
    The final solution is x0=1/2 to minimize the area.

    Area = 4/3
    
    The calculus was done in a separate ipython terminal.

## Computer Bonus

    A palindrome is a
    number that reads the same forwards
    and backwards. What is the smallest,
    and reported to be the only,
    nonpalindromic integer whose cube is
    a palindrome?
    
    -- The Ambidextrous Universe
    by Martin Gardner

Solution:

    I guess we use brute force???
    
    Very quick result = 2201 with 2201^3 = 10662526601
    
    Can do in base-R or gmp.  Also easy in sympy.
