---
title: "2000_1"
author: "R. G. Cronce"
date: "12/18/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(reticulate)
```

# Bent Brain Ticklers, 2000_1

Brain ticklers from Spring 2000


## Problem 1

A farmer has a hemispherical tub of radius 1 m that he uses to water his cattle.  What is the
depth of water when it is half full?

-- C. K. Galer, MI A'77

Need symbolic math so we will use sympy:

This works -- but need to fix this so it produces a real solution.  As shown below I found the
correct root numerically and then examined the real/imaginary parts separately.

```{python}
import sympy
from sympy import *

R=S(1)            # Radius of Tub
x=Symbol('x', real=True)    # X-axis
y=Symbol('y', real=True)    # Y-axis (height of water)
r=Symbol('r', real=True)    # Radius of circular disk used in calculating volume
y0=Symbol('y0', real=True)  # Height of water when tank is half full

r=sqrt(R**2-(R-y)**2)           # radius of circular disk as a function of height
dv=pi*r**2                      # differential volume of circular disk with radius=r
V_full=integrate(dv,(y,0,R))    # Volume of a full hemisphere
V_half=integrate(dv,(y,0,y0))   # Volume of bottom part (up to height of y0)
V_top=integrate(dv,(y,y0,R))    # Volume of upper part (above height of y0)
print('V_full = ',V_full)
print('V_half = ',V_half)
print('V_top = ',V_top)
# We could solve for V_half = V_full/2, or
# We could solve for V_top  = V_full/2, or
# We could solve for V_top  = V_half
#
# Arbitrarily pick the last one
solution=simplify(solve(V_half-V_top,y0)[0])
print('Solution: y0 = ',solution)
print('   re:       = ',trigsimp(re(solution)))
print('   im:       = ',im(solution))
print('Numerically  = ',N(re(solution)))
```

This makes intuitive sense -- need more than half the height of the hemisphere filled with
water to get half the volume.  The upper half has a larger radius so doesn't need as much
height.


## Problem 2

Let f(n) be the number of times the digit 1 is used in printing all the numbers from 1 to n inclusive.
Thus, for example, f(3) = 1 and f(12) = 5 (1,2,3,4,5,6,7,8,9,10,11,12 - 11 has two 1s).  Now, f(1)=1.
What is the next value of n such that f(n)=n?

-- Technology Review

By brute force and some trial and error:

```{r}

num_ones = function(m) {
  n=0
  for (i in 1:m) {
    b=sum(unlist(strsplit(as.character(i),''))=='1')
    n=n+b
  }
  return(n)
}

for (i in c(1,19,199,1999,19999,199990)) {
  n=num_ones(i)
  print(paste(i,n,i-n))
}
```

The smallest value of n I could find was 199990 with trial and error.

I tried LuaJIT to see if it was blazingly faster and it is -- but it still takes a long
time for exhaustive search.  After thinking about the algorithm I realized I was making
it run ridiculously slowly by repeating all of the work for all previous calculations!!!

After fixing the algorithm the Lua code is very fast.

Using the Lua program I found 199981 as the smallest.

The code below uses the modified algorithm where we just add the number of ones in the next number
to the previously calculated value (c0).  This runs much slower than the Lua code but at least
it finishes in a reasonable amount ot time.

FYI: LuaJIT code = 0.1 seconds, R code = 1.2 seconds without storage of results.
Addition of storing the results yields 1.28 seconds.  If we use dynamically sized
storage (e.g. x=c(x,new value)) the run-time goes through the roof!!!


```{r}
t1=Sys.time()
c0=0
x=rep(0,200000)
y=rep(0,200000)
for (i in 1:200000) {
    b=sum(unlist(strsplit(as.character(i),''))=='1')
    c0=c0+b
    x[i]=i
    y[i]=c0
    if (i==c0) {
      print(i)
    }
}
t2=Sys.time()
print(paste0('Elapsed time = ',t2-t1))

```

## Problem 3

The MGGC (Mathematicians Goofy Golf Club) holds banquet in July.  Trying to find the date from
some fellow members who were being obtuse.  One told me the date was an odd number, another
that it was greater than 13, a third that it was not a perfect square, and a fourth that it was
a perfect cube.  Finally, W. W. Webb, the president of the club, told me the date was less than
my highest single hold score last season (which was 17).  Later, when I finally learned the date,
I found that only one of the statements was true.  What is the date of the banquet?

-- Adapted from Brain Puzzlers Delight by E. R. Emmet

TODO


## Problem 4

Cryptic multiplication:  SINK * THEM = DEEPDEEP

TODO -- already done in the cryptic notebook


## Problem 5

An urn contains 80 balls, 72 green and 8 red.  The balls are drawn one at a time without replacement
until all the red balls are drawn.  What is the expected value of the number of balls remaining in
the urn at that point?

-- William Allen Whitworth (1901)

We can solve with Monte Carlo:

```{r}
t1=Sys.time()
num_trials=100000

n=c()
for (i in 1:num_trials) {
  s=sample(80)                # Randomly sample the 80 balls (red=1-8, greep=9-80)
  n=c(n,80-max(which(s<9)))   # Find out how many are left when last red one is picked
}
summary(n)
t2=Sys.time()
print(paste0("Exection time = ",t2-t1," seconds"))
```

So, we conclude the expected number of balls remaining is 8.

Another approach is to consider the probability that the last ball is green, then the second last,
etc. and keep adding balls until we have them all accounted for.  This is like the birthday
problem in reverse.

Let:

    n = number of balls remaining after picking last red ball
    
    Then - before picking the last red ball:

    Number of balls remaining    = n+1
    Number of balls chosen       = 80-(n+1)
    Number of red balls chosen   = 7
    Number of green balls chosen = 80-(n+1)-7 = 72-n
    
    Pr(7 red balls chosen with n+1 remaining) = choose(8,7)*choose(72,72-n)/choose(80,80-(n+1))
    Pr(next ball is red) = 1/(n+1)
    
    Expected value = sum(Pr(7 red)*Pr(next red)*n, n=0 to 79)
    
    Note n = number of balls remaining after picking the last red ball so this is the
    multiplier in calculating the weighted probability (e.g. mean value).


```{r}
xbar=0
for (n in 0:79) {
  p_7_red = choose(8,7)*choose(72,72-n)/choose(80,80-(n+1))
  p_next_red = 1/(n+1)
  xbar = xbar + p_7_red*p_next_red*n
}
print(paste0('Expected number of balls remaining = ',xbar))
```

OK -- numerically this works.

Can I do this symbolically?

```{python}
def choose(n,m):
    return(factorial(n)/(factorial(n-m)*factorial(m)))

n=Symbol('n',integer=True)

pr_7_red = choose(8,7)*choose(72,72-n)/choose(80,80-(n+1))
pr_next_red = 1/(n+1)

xbar = summation(pr_7_red*pr_next_red*n,(n,0,79))

print(xbar)
```

Apparently so -- even symbolically this reduces to precisely 8 balls remaining.


## Bonus

A boy is standing on a dock looking at the water in a smooth pond, when the ball he is holding
slips from his hand. As he watches it disappear below the surface, he fears that it is gone forever.
However, he is relieved a short time later to see it reappear.

The ball, which has a diameter of 6 cm and a mass of 112 gm, falls from rest
at a height of 2 m from the center of
the ball to the surface of the water. Although
not completely accurate, for
the purposes of this problem assume
that the drag force on the ball in the
water equals `k*v**2` where v is the ball's
velocity in m/s, A is the cross sectional
area of the ball in `m**2`, and k has the
value of 353.7 `Newton s**2/m**4`. Ignore
air resistance, and assume that the ball
encounters no drag or buoyancy forces
in the water until the ball is half submerged,
at which point these forces
come into full play. Use a density of
1,000 `kg/m**3` for the water and a value
of 9.8 `m/sec**2` for the acceleration due
to gravity. To three significant figures,
how many seconds elapse between the
time the ball leaves the boy's hand
until it just breaks the surface upon
reappearing?

-- John W. Langhaar, PA A'33

TODO

## Computer Bonus

Find the smallest integer greater than one which has the property that the sum of all the integral
divisors of its square equals a perfect cube. One and N are both divisors of N.

-- Pierre de Fermat (circa 1650)

Solution:

Using Python (see `fermat.py`) yields a solution of n=43098 as the smallest
value (other than 1) with the desired property.  This is also the only solution
below 1e7.  Finding the solution is rather quick (runtime ~4.4 seconds).
Checking all the way to 1e6 takes quite a bit longer (runtime ~210 seconds).
A further check to 1e7 also failed to find angthing (runtime ~2738 seconds).
A further check to 2e7 with LuaJIT failed to find anything (runtime for the
range 1e7-2e7 was 3343 seconds, runtime for 1-1e7 was 1574 seconds).  I
don't have the patience to run this any longer.

Needed Python.sympy to get the prime factor function.

I also rewrote this in Lua (see `fermat.lua`, `ferma2.lua`, and `primes.lua`)

FYI: here is a transcript of the ipython session showing the final results

    In [386]: sympy.factorint(43098*43098)
    Out[386]: {2: 2, 3: 2, 11: 2, 653: 2}

    In [388]: s=(1+2+4)*(1+3+9)*(1+11+121)*(1+653+653*653)

    In [389]: s
    Out[389]: 5168743489

    In [390]: cbrt(s)
    Out[390]: 1729.0

    In [391]: sympy.factorint(1729)
    Out[391]: {7: 1, 13: 1, 19: 1}

I don't understand how to find this algorithmically other than search.
This all begs the question -- what is the next larger value?


I did some research on the sum of the integral divisors and there is a
well known theorem relating to the prime factors of a number and the sume
of the divisors.  Essentially, if you sum all the possible products of
all the powers of the prime factors and then add 1 to the result you get
the sum of the divisors.  Assuming you can do the prime factorization
in a reasonable amount of time the calculation of the sum of divisors
goes very quickly.

The products are relatively easy to calculate if we generate terms of
the form:

    1 + p + p^2 + ... + p^n

Where p = prime factor and n is the multiplicity.

If we form one of these terms for each prime factor and then calculate the
product this will be the same as the sum of the factors.  As an example:

    Assume the number is 360 = 2^3 * 3^2 * 5

    The terms are:
        1+2+2^2+2^3 = 15
        1+3+3^2     = 13
        1+5         = 6

    Sum of all the factors = 15*13*6 = 1170

    Directly listing the factors:
        
          1,   2,   3,   4,   5,   6,
          8,   9,  10,  12,  15,  18,
         20,  24,  30,  36,  40,  45,
         60,  72,  90, 120, 180, 360

For small numbers the difference in calculation is not large.  For big numbers, however,
calculating all the factors becomes difficult.

This is particularly useful in this problem since the prime factors of the square of a number
are just the same as the original number (with double the multiplicity) so we don't even need
to find the prime factors of the larger number.

Python has sympy.factorint() that returns a dictionary with keys of the prime
factors with the values being the multiplicity.

Using Python yields a solution of n=43098 as the smallest value (other than 1) with the
desired property.

```{r}
calc_factors = function(n) {
  fff=c()
  for (i in 1:n) {
    if (n%%i == 0) { fff=c(fff,i) }
  }
  return(fff)
}

```

Turns out that brute force is going to take a really long time.  A scan up to 5745 does
not reveal any solutions (e.g. checking for cube root of numbers up to the sum of the factors
of 5745*5745).  There must be some more systematic way of attacking this -- but it is called
a computer bonus so maybe I just need to search further?

Note that this runs very slowly.  The scan noted above took ~2 hours or so.

