---
title: "Bent Winter 2003"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Winter 2003


## Problem 1

    A circle is divided into 13 sectors.
    Two players alternate by placing a
    coin either on a vacant sector or on
    two adjacent vacant sectors. The
    player who places the last coin wins.
    Which player has the advantage, and
    how should he/she proceed?
    
    -- Sam Loyd, circa 1880

Solution:

    I didn't actually solve the right problem.  I was playing Nim and
    this was talking about ADJACENT vacant sectors.  This game can
    be won by the second person if they pick such that there are
    two groups of 5 adjacent pie slices.  Then - on subsequent moves
    just mirror the move of the other person.
    
    If this was 13 stick Nim:

    First person guarantees a win if they always leave the
    number of remaining sectors equal to a multiple of three.

    The guarantee a win want to leave other person 3 sectors.
    We can do this if we leave the other person 6 sectors.
    If they pick 1 we take 2, if they pick 2 we pick 1.
    
    In a similar way, always leave the other person a multiple
    of 3 and you guarantee a win.  The first player can guarantee
    a win if they pick 1 (leaving 12).  Then the remaining
    sectors will be 9, 6, and 3 for subsequent moves.


## Problem 2

    A committee of four is to be chosen
    from a club which has a membership
    of 10 men and 12 women. In how
    many different ways can the committee
    be selected if it is to include at
    least two women, keeping in mind
    that Mr. and Mrs. Bickerson refuse
    to serve at the same time?
    
    -- Introductory Combinatorics
    by Richard A. Brualdi

Solution:

    Grand total = 495+2145+2871 = 5511 different committees

    We can have committees of:
    
        2 women, 2 men
        3 women, 1 man
        4 women
        
    Need to figure out how to handle Mr/Mrs Bickerson exclusion.
    
    For 4 women: n = choose(12,4) = 495
    
    For 3 women, 1 man: n = 1650+495 = 2145
    
        Without Mrs. Bickerson: n=choose(11,3)=165
        Can have any of the men so n = 165*10 = 1650
        
        With Mrs. Bickerson: n=choose(11,2)=55
        Can have any man except Mr. Bickerson so n=55*9=495

    For 2 women, 2 men: n = 2475+396 = 2871
    
        Without Mrs. Bickerson: choose(11,2)=55
        Number of men = choose(10,2) = 45
        Total = 45*55 = 2475
        
        With Mrs. Bickerson: 11 other women
        Number of men = choose(9,2) = 36
        Total = 11*36 = 396

    Grand total = 495+2145+2871 = 5511 different committees
    

## Problem 3

    Mary cashed a check, and the
    teller made a mistake and paid her
    the amount written in cents in dollars
    and the amount written in dollars in
    cents. Later, after spending $3.50,
    Mary realized that she still had twice
    the amount the check called for. What
    was the amount of the check?
    
    -- Technology Review

Solution:  Check was for $14.32

    Let d = correct amount of dollars, integer > 0
        c = correct amount of cents, integer in range[0-99]
        
    check value = d + c/100
    
    mistake = c + d/100
    
    know: (c+d/100 - 3.50) = 2*(d+c/100)

    c*(1-2/100)+d*(1/100-2) = 3.50
    98*c-199*d = 350
    d = (98*c-350)/199
    
Searching for integer solutions:

    Yields d = 14, c=32 for original check value of $14.32
    Erroneous value = $32.14
    Sanity check:
        (32.14-3.50)/2 = 14.32

Is there a clever way to find the solution other than searching?

```{r}
for (c in 0:99) {
    d=(98*c-350)/199
    if (d==round(d)) {print(paste0(d,"   ",c))}
}
```

## Problem 4

    Five schoolgirls use their only
    nickel to weigh themselves on a scale
    in pairs with one staying on as the
    other gets off and another gets on.
    The 10 readings are 183, 186, 187,
    190, 191, 192, 193, 194, 196, and 200
    pounds. What are the weights of
    each of the five schoolgirls?
    
    -- Discover

Solution:

    Weights: g1=91, g2=92, g3=95, g4=99 g5=101

    The totals are apparently the weights
    of girl 1 + girl2, g1+g3, etc. but we
    don't know the correspondence of girls/weights.
    
    Observations:

    There are 10 distinct totals so no two girls
    have the same weight.  If they did, there would
    be 4 repeated totals.  Let us assume the weights
    are:
    
        g1 < g2 < g3 < g4 < g5

    Then we know:
    
        g1+g2 = 183 --> g1=183-g2
        g1+g3 = 186 --> g3=g2+3  ---+
        g4+g5 = 200                 |- g5-g1=10
        g3+g5 = 196 --> g3=g4-4  ---+
        
        so g4 = g3+4 = g2+7
        
    At this point - sort of stuck.  Have reduced to two
    unknowns but need one more weight.  Consider the third
    lowest weight.  Can either be sum:
    
    g1+g4 = 187 or g2+g3 = 187
    
        if g1+g4 = 187 then g1+g2+7 = 187
        or g1+g2 = 180
        
    but: g1+g2 = 183 --> contradiction so this is incorrect.
    
    if g2+g3 = 187 then g2+g2+3=187
    
        solving --> g1=91, g2=92, g3=95, g4=99 g5=101
    
        sanity checking against the given totals:
        
            g1+g2=183
            g1+g3=186
                      <-+
            g1+g4=190   |
        +->             |
        |   g1+g5=192   |
        |   g2+g3=187 --+
        +-- g2+g4=191
            g2+g5=193
            g3+g4=194
            g3+g5=196
            g4+g5=200
    
    Conclude: this is the solution

    
## Problem 5

    A postal-service company will ship
    a package of up to 70 pounds with a
    length plus girth of up to 108 inches.
    A customer wishes to wrap and mail a
    uniform density right circular cone
    with the maximum volume. What is
    this cone's height, and what is its
    maximum allowable density, assuming
    the wrapping paper has negligible
    weight? Assume the girth is the circumference
    at the base of the cone and the
    length is the height of the cone.
    
    -- Robert A. Grimm, IN A'50

Solution:

    Height = 36 inches
    Max density = 0.01414 pounds/in^3 (24.43 lb/ft^3 < water)

    Details:
    
    R = radius of the base of the cone
    H = height of the cone
    d = density of the cone
    V = Volume of the cone (to be maximized)
    W = Weight of the cone
    
    W = d*V <= 70 pounds
    2*pi*R + H <= 108 inches
    
    Need to calculate V for right circular cone.
    
    V = 1/3 * H * pi * R^2
    
    Solving the girth/height for H = 108-2*pi*R
    
    Substituting in V:
    
        V = 1/3 * (108-2*pi*R) * pi * R^2
    
    Maximize V wrt R:
    
        V = 36 * pi * R^2 - 2*pi/3 * pi * R^3
        
        dV/dR = 2*36*pi*R - 2*pi/3 * pi * 3 * R^2 = 0
        
        2*36 - 2*pi/3 * 3 * R = 0
        
        36 - pi*R = 0
        
        R = 36/pi
    
    Then:

        H = 108-2*pi*R = 108-2*pi*36/pi
          = 108-72
          = 36
          
        V = 1/3 * H * pi * R^2
          = 1/3 * 36 * pi * (36/pi)^2
          ~ 4950 in^3
          
        d ~ 70/4950 = 0.01414 lb/in^3 = 24.43 lb/ft^3

## Bonus

    The island of Bongo is a rectangle
    whose area is an integral number
    of square kilometers. When it was
    first colonized, the inhabitants decided
    to divide it into five counties
    with each county in the shape of a
    right triangle. Abongo, Ebongo, and
    Ibongo each had the same area.
    Obongo was larger, and Ubongo was
    the largest. Abongo, Ebongo, and
    Ibongo later merged into a single
    county that was still triangular.
    Bongo's shorter side is 45 kilometers.
    What is Bongo's area?
    
    -- Roger Schofield in New Scientist

TODO

    Need to draw a picture
    

## Computer Bonus

    Using each integer
    exactly once, separate the integers 2
    through 25 inclusive into two groups,
    such that the difference between the
    product of the integers in the first
    group and the product of the integers
    in the second group is as small as
    possible. What is this difference?
    
    -- John S. Stanczak, MI Delta'70

Solution:

    > best difference = 326592000 (~ 0.327 billion)
    There are many different partitions that achieve this result.
    The optimal results all share the same prime factorization.

    I am pretty sure this is the solution but I don't really
    understand why.  I think it has to do with the prime
    numbers that are singletons -- but I am not sure.
    
    The prime factors in the full product 2*3*...*25 are:
    
    Prime Number
     2    22
     3    10
     5    6
     7    3
    11    2
    13    1
    17    1
    19    1
    23    1

    Note that the prime factors that can not be partitioned
    evenly between the two sets are: 7,13,17,19,23.  We could
    split the other factors: 2,3,5,11 between the sets.
    I tried figuring this out but this is hard.
    
    Comparing a couple of "optimal" solutions" reveals that the
    singleton primes are ditributed as:
    
    set 1: common terms * 11*11*13*23 = common terms * 36179
    set 2: common terms * 16*7*17*19  = common terms * 36176
    
        common terms = 2^9 * 3^5 * 5^3 * 7

    Clearly the relative values of 36179 and 36176 define the best
    case difference -- but I still don't have a good feel for this.

    I think I will let random sampling look for me:
    
    I have found multiple solutions with the same result.
    Pick corresponding rows of best_s1 or best_s2 for
    equivalent partitions.
    
    > sort(best_s1)
        [1]  3  5  8  9 10 11 13 15 16 21 22 23
        [1]  4  5  6  9 10 11 12 13 20 21 22 23
        [1]  5  8  9 11 13 15 20 21 22 23 24
    > sort(best_s2)
        [1]  2  4  6  7 12 14 17 18 19 20 24 25
        [1]  2  3  7  8 14 15 16 17 18 19 24 25
        [1]  2  3  4  6  7 10 12 14 16 17 18 19 25
        
    > best difference = 326592000 (~ 0.327 billion)

```{r}
s=2:25
best_s1=c()
best_s2=c()
best_diff=prod(s)

num_trials=100000

for (i in 1:num_trials) {
    # Random size for the partition
    m=sample(seq(9,16),1)
    s1=sample(s,m)
    s2=setdiff(s,s1)
    p=prod(s1)-prod(s2)
    if (abs(p)<best_diff) {
        best_s1=s1
        best_s2=s2
        best_diff=abs(p)
        print(best_diff)
    }
}
```
