---
title: "Bent Fall 2004"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Fall 2004


*EDITORIAL COMMENT:* I did not really enjoy this set of problems.
About half the problems were simply not the kind of problems I enjoy.
Problems 1 and 2 were ok and problem 5 was a standard cryptic.  The
other four problems were just not my cup of tea.

## Problem 1

    At the end of the soccer season, it
    was found that the total number of
    goals scored by each of the 11
    players on the team was a prime
    number and that the average was
    also a prime number. Furthermore,
    no two players had scored the same
    number of goals, nor had anyone
    scored the same as the average. If
    no one scored more than 45 goals,
    how many goals did each player
    score?

    -- Nearly Impossible Brain Bafflers
    by Tim Sole and Rod Marshall

Solution:  need a list of the primes up to 45.

```{r}
library(gmp)

p=c()
for (i in 2:45) {
    if (isprime(i)) { p=c(p,i) }
}
print(p)
print(paste0("Number of primes = ",length(p),collapse=""))

n=length(p)
m=choose(n,11)
print(paste0("Number of 11 player combinations = ",m,collapse=""))
```

We can just enumerate the combinations since there aren't too many of them.
We can then check to verify the mean is a prime and not included in the
player scores

```{r}
library(gtools)

a=combinations(14,11,p)

s=apply(a,1,mean)
k=(s %in% p)

print(paste0("Number of prime means = ",sum(k),collapse=""))

print(cbind(a[k,],s[k]))
```

Only the final case has a mean that is different from all the player scores.

    Players = 5    7   11   13   17   19   29   31   37    41    43
    Mean = 23


## Problem 2

    What is the exact probability that
    a randomly selected month (in the
    Gregorian calendar) will have five
    Sundays?

    -- Howard G. McIlvried, PA Gamma'53

Solution: I wonder what the Gregorian calendar means for the leap
year?  I know the rules but can this affect the calculation?
You can have 5 Sundays in a month of 29 days -- but not in 28.

    For months with 31 days there are three possibilities (out of 7)
    For months with 30 days there are two possibilities (out of 7)
    For months with 29 days there is one possibility (out of 7)
    For months with 28 days there is no possibility
    
For a year:

    7 months have 31 days
    4 months have 30 days
    1 month has 28 or 29 days

If I ignore interaction between months (since they ask for a random month):

    Pr = (7/12)*(3/7)+(4/12)*(2/7)+(1/12)*(0/7) for non-leap years
       = 29/84
       ~ 0.3452381
       
    Pr = (7/12)*(3/7)+(4/12)*(2/7)+(1/12)*(1/7) for leap years
       = 5/14
       ~ 0.3571429

For the most part leap years occur every 4 years except if divisible by 100
except if divisible by 400.  In a 400 year span we "miss" 3 leap years so we
have 97 leap years and 303 non-leap years.  Using these to weight the Pr from
above we get:

    Pr = (303*(29/84) + 97*(5/14))/400
       = 557/1600
       ~ 0.348125
       
Another way of calculating this: in a 400 year span we have the following:

    Months of 31 days = 400*7, p=3/7
    Months of 30 days = 400*4, p=2/7
    Months of 29 days = 97,    p=1/7
    Months of 28 days = 303,   p=0
    
    P = ((400*7*3 + 400*4*2 + 97)/7)/(400*12)
      = 1671/4800
      = 557/1600


## Problem 3

    Four men and their wives are
    scheduled for a mixed-doubles
    tennis tournament on two adjacent
    courts for three consecutive time
    periods. Everyone plays three
    matches, each with a different partner
    of the opposite sex and with no
    repeated opponent. No one ever
    plays either with or against his or
    her own spouse. How many possible
    tournaments are there, ignoring
    permutations of courts and time
    slots?

    -- Byron R. Adams, TX A'58

Solution:  I think this is a standard scheduling problem.
Unfortunately, I never paid attention on how to make
tournament schedules.  I am actually not sure I care.

* We have 8 participants

* Two adjacent courts

* Three consecutive time periods

* Everyone plays three matches

* No repeated opponents

* Not with or against spouse




## Problem 4

    Marrieds are complete liars in the
    village of Loose Chippins, whereas
    singles are wholly truthful. Take
    that group of three men and three
    women over there, for instance.
    They are two couples and two
    singles. If you asked Charlie
    whether Bert is married to Queenie,
    he would say "Urp." If you asked
    Pauline if Pauline is married to
    Charlie, she would say "Urp." If you
    asked Alf if Alf is married to Rhona,
    he would say "Ong." And if you
    knew enough of the local dialect to
    know whether "Urp" means "Yes"
    and "Ong" means "No," or vice
    versa, you could work out who is
    married to whom. Who is married
    to whom?

    -- New Scientist

Solution:  Let us summarize the facts

    Marrieds are complete Liars
    Singles always tell the truth
    Two couples and two singles

    1. Charlie: Bert married to Queenie?  Reply=Urp
    2. Pauline: Pauline married to Charlie?  Reply=Urp
    3. Alf: Alf married to Rhona?  Reply=Ong

    People: Charlie, Bert, Queenie, Pauline, Alf, Rhona

    Potential Status: M=1, S=0  with exactly 4 M and 2 S
    
    Let us eliminate Gay marriage so need exactly two men and two women marrieds

The following combinations are possible:

       Charlie Bert Queenie Pauline Alf Rhona
    2        0    1       0       1   1     1
    3        0    1       1       0   1     1
    5        0    1       1       1   1     0
    6        1    0       0       1   1     1
    7        1    0       1       0   1     1
    9        1    0       1       1   1     0
    11       1    1       0       1   0     1
    13       1    1       1       0   0     1
    15       1    1       1       1   0     0


    Assume Urp=Yes, Ong=No

    1. If Charlie is single then Bert is married to Queenie
       If Charlie is married then Bert is not married to Queenie
       
    2. If Pauline is single then Pauline is married to Charlie --> Contradiction
       If Pauline is married then she is married to someone other than Charlie

    3. If Alf is single then he is not married to Rhona
       If Alf is married then he is married to Rhona

        Conclusions:
        
        Pauline is married to someone other than Charlie
       
    After eliminating everything I can I get the following where 1=married, 0=single

       Charlie Bert Queenie Pauline Alf Rhona
    11       1    1       0       1   0     1 --> Bert+Pauline, Charlie+Rhona
    15       1    1       1       1   0     0 --> Bert+Pauline, Charlie+Queenie

    I don't know how to reduce this any further

    
    Assume: Urp=No, Ong=Yes
    
    1. If Charlie is single then Bert is not married to Queenie
       If Charlie is married then Bert is married to Queenie
       
    2. If Pauline is single then she is not married to Charlie
       If Pauline is married then she is married to someone other than Charlie
       
    3. If Alf is single then he is married to Rhona --> Contradiction
       If Alf is married then he is not married to Rhona
        
        Conclusions:
        
        Alf is married to someone other than Rhona

    After eliminating what I can

       Charlie Bert Queenie Pauline Alf Rhona
    3        0    1       1       0   1     1 --> Bert+Rhona, Alf+Queenie
    5        0    1       1       1   1     0 --> Bert+Pauline, Alf+Queenie

    I do not know how to reduce this any further.


## Problem 5

    TWO + TWENTY = TWELVE +
    TEN, with the first three divisible by
    their namesakes. Note: different
    letters stand for different digits, and
    same letter is always the same digit.

    -- Technology Review

Solution: solved in the cryptic arithmetic notebook

    TWO    = 876
    TWENTY = 872480
    TWELVE = 872532
    TEN    = 824


## Bonus

    On what day of the year is
    twilight shortest at a place of given
    latitude at sea level? Express your
    answer in the form: March 21st +/-
    f(latitude). Assume the inclination of
    the ecliptic is 23 degrees 27'. Assume that the
    angular distance of the sun from the
    equinox (assumed to be noon March
    21 for this problem) changes at an
    average daily rate of 360 deg/365.25.
    Assume twilight starts the moment
    at which the midpoint of the sun is
    intersected by the horizon. Assume
    twilight ends when the midpoint of
    the sun stands 6.5 degrees below the
    horizon; this is civil twilight. Ignore
    effects of atmospheric refraction. As a
    check of your math, the shortest
    twilight in Leipzig, Germany, (51 deg 20'
    6" N latitude) is when?

    -- 100 Great Problems of Elementary
    Mathematics by Heinrich Dorrie

TODO

I have no idea how to do this.

## Double Bonus

    Given the base and
    altitude of a triangle, with ruler and
    compasses, find the length of a side
    of the inscribed square.

    -- Alsop, 1848

TODO

I have no idea how to do this.
