---
title: "Bent Spring 2019"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Spring 2019


## Problem 1

    Solve the following cryptic addition in base
    11. All the usual rules apply; each different
    letter stands for a different digit, and each
    different digit is always represented by the
    same letter; there are no leading zeros. Use
    a to represent the digit 10.

        TICKLER = STRAINS + BRAINS

    -- Howard G. McIlvried, III, PA G'53

Solution: from cryptic arithmetic notebook

    Cryptic addition in base 11:

      TICKLER = STRAINS + BRAINS

    source('rgc_lib.r')
    a=make_crypt_df('ticklerstrainsbrains',11)
    x_tickler=make_num(a,'tickler',11)
    x_strains=make_num(a,'strains',11)
    x_brains=make_num(a,'brains',11)
    k=(x_tickler == x_strains+x_brains)
    k=k&(a$t!=0)&(a$s!=0)&(a$b!=0)
    sum(k)
    [1] 1
    a[k,]
             t i  c k l e r s a n b
    33007270 9 0 10 6 1 4 5 8 3 7 2
    x_tickler[k,]
    [1] 16098615
    x_strains[k,]
    [1] 15699230
    x_brains[k,]
    [1] 399385

    Note the digit 10 = 'a' in conventional base-11 notation.

    The values for tickler, strains, and brains given in base 10!!!


## Problem 2

    What is the maximum number of knights that
    can be placed on a standard 8x8 chess board,
    so that each knight threatens (i.e., can
    capture) exactly one other knight, and how
    are they arranged? A knight moves two squares
    in one direction and one square in another
    direction to end up on a square of opposite
    color. This move can occur even if
    intervening squares are occupied. In this
    puzzle, each knight can move to exactly one
    square occupied by another knight, but not to
    two or more occupied squares. Present your
    answer as an 8x8 grid (representing a
    chessboard) with the locations of the knights
    represented by Ns and the unoccupied squares
    represented by dashes.

    -- A Passion for Mathematics by Clifford A.
       Pickover

Solution:

    I think I remember this problem.  Pack them into
    rectangles in each corner of the board.  The
    rectangles are 2x4.


## Problem 3

    I am getting a new sign for my shop, and
    instead of showing the shop number as numeric
    digits, it will be spelled out in capital
    letters, one word per digit. For example, if
    my shop number were 103, it would be painted
    as ONE ZERO THREE. The sign painter is a
    little eccentric, and instead of charging by
    the hour, he charges by the brush stroke,
    which can be any shape and can touch, but
    must not cover, an already painted area
    (except where changing direction). He paints
    capital letters in a simple style (no serifs)
    and does not use two strokes where a single
    stroke will do. For example, E, F, G, H, I,
    N, O, R, S, T, U, V, W, X, Z would require 2,
    2, 2, 3, 1, 1, 1, 1, 1, 2, 1, 1, 1, 2, and 1
    strokes, respectively. My shop number, which
    I had spelled out for the painter when he
    arrived, is a prime number, consisting of
    three different digits. He charges $1 per
    stroke, regardless of the length of the
    stroke. By coincidence, the cost of my number
    in dollars was equal to the sum of the three
    digits in the shop number and was a prime.
    What is my shop number?

    -- Enigma in New Scientist by Susan Denham

Solution:

    I remember doing this problem.  I am sure it was
    a brute force search.
    
    Need three digit prime, all digits distinct, with SOD also prime.
    From this set, convert to alpha and assess the cost
    
```{r}
library(gmp)

# Letters used in the sign and the associated cost
letter = strsplit(c('EFGHINORSTUVWXZ'),"")[[1]]
cost   = c(2, 2, 2, 3, 1, 1, 1, 1, 1, 2, 1, 1, 1, 2, 1)

# Digits
digits=c("ZERO","ONE","TWO","THREE","FOUR","FIVE","SIX","SEVEN","EIGHT","NINE")
digit_cost=list()
for (d in digits) {
    digit_cost[[d]] = sum(cost[sapply(strsplit(d,"")[[1]],function(x) {which(letter==x)})])
}

# All three digit primes
p=(100:999)[isprime(100:999)==2]
# Filter to remove duplicate digits
k=sapply(p,function(x) {length(unique(strsplit(as.character(x),"")[[1]]))==3})
p=p[k]
# Calculate the sum of digits
sod=sapply(p,function(x) {sum(as.numeric(strsplit(as.character(x),"")[[1]]))})

# Filter list to only include if SOD is prime
p=p[isprime(sod)==2]
sod=sod[isprime(sod)==2]

# Calculste the cost of the sign
# Convert number to digits to words to cost - sum the result
# My goal was to make this as incomprehensible as possible
# by jamming it all into one line
sign_cost=sapply(p,function(n){sum(unlist(digit_cost[digits[as.numeric(strsplit(as.character(n),"")[[1]])+1]]))})

# Final filter, cost of sign = sum of digits
p[sign_cost==sod]
sign_cost[sign_cost==sod]
```

    The shop number is 829 with a sign cost of 19.


## Problem 4

    Joe is playing five-card draw poker. He is
    dealt the 4, 5, and 6 of hearts, the 3 of
    diamonds, and the 10 of spades. He decides to
    discard the 3 and 10 and draw two new cards.
    What is the probability that he will improve
    his hand by getting one pair or better?
    Assume that Joe sees only the cards dealt to
    him and knows nothing about the other
    players' cards. In five-card draw, each
    player is dealt five cards face down. After
    looking at the cards, if he is not satisfied,
    he can turn in up to three cards (face down)
    and get replacements.

    -- Howard G. McIlvried, III, PA G'53

Solution:

    I am sure they want combinatorics but I will use
    Monte Carlo.
    
```{r}

pip  = trunc((0:51)/4)+1   # 1, 2, ..., 13
suit = (0:51)%%4           # 0=diamonds, 1=hearts, 2=spades, 3=clubs

discard = c(which(pip==3 & suit==0),which(pip==10 & suit==2))
keep    = which((pip %in% c(4,5,6)) & suit==1)

available=(1:52)[-c(discard,keep)]

num_trials=100000

n=0
np=0
nf=0
ns=0

set.seed(1)
for (trial in 1:num_trials) {
    s=sample(available,2)
    hand=c(keep,s)
    hand=sort(hand)
    if (length(unique(pip[hand]))<5) {           # 2, 3, or 4 of a kind
        n=n+1
        np=np+1
    } else if (length(unique(suit[hand]))==1) {  # flush
        n=n+1
        nf=nf+1
    } else if (max(diff(sort(pip[hand])))==1) {  # straight
        n=n+1
        ns=ns+1
    }
}

print(n/num_trials)
print(c(np,nf,ns))
```

    Monte-Carlo summary with 100K trials

    Seed   P(improvement)   P/3/4   Flush  Straight
    -----------------------------------------------
    1      0.47721          40077   4237   3407
    42     0.47618          40066   4167   3385
    123    0.47519          39841   4197   3481
    2020   0.47493          40037   4098   3358

    The likelihood of improvement is about 0.476.  Most of the time
    the improvement is in getting a pair, three or four of a kind.
    This occurs about 80% of the time when there is improvement.
    About 10% of each improvement is a flush or a straight.
    Did not check for straight-flush.

    The analytical result is really not that bad (see Bent).
    The probability of improvement is 514/1081 ~ 0.4754857.


## Problem 5

    We are all familiar with the story of the
    1001 Arabian Nights. Find N such that N! (N
    factorial) has one digit for each of the
    Arabian Nights.

    -- H. S. Uhler

Solution:

    The result is obvious with base-10 logs.
    
    log10(n) = x.yyyyy
    
    where number of digits in n = x+1
    
    We can calculate a factorial by repeated adding log10() values
    until we reach a sum of 1000.
    
```{r}

s=0
n=1
while (TRUE) {
    s=s+log10(n)
    if (s>1000) {
        break
    }
    n=n+1
}
print(n)
```

    The solution is 450! has 1001 digits.  BTW - here they are:
    
    > factorial(as.bigz(450))
    Big Integer ('bigz') :
        17333687331126326593447131461045793996778112652090
        51015569207509555333001683436750604675088290438710  (100)
        61458112845184240978586185838063016502083472961813
        51667570171918700422280962237272230663528084038062  (200)
        31236934267413503661010150883822049497092973901163
        67937661650237308538964039015908361441495944326842  (300)
        04513784716402303182604094683993315061302563918385
        30334151060676146242020582000693635209596741718319  (400)
        15387256175095213805567813091954298002292738033425
        53558164591996298912368598547771179158461351340068  (500)
        90564712765816483637712630377492336007807230746200
        85543550683614481266062811457609604991878134283979  (600)
        24840592504537849487425060488481036571447957046788
        63574293671461517621914846974310297994974071448510  (700)
        47161696640523973926028484086940074089989011274929
        05171514473431386633392492040661522692303043813960  (800)
        54196609322424380922513726885171790430321405823844
        79361116785682369730362384046265078906880000000000  (900)
        00000000000000000000000000000000000000000000000000
        00000000000000000000000000000000000000000000000000  (1000)
        0


## Bonus

    Find three consecutive triangular numbers
    whose product is a perfect square. The first
    such set is 6*10*15 = 900 = 30^2, and the
    second such set is 300*325*351 = 34,222,500 =
    5850^2? What is the third such set, and what
    is the square root of their product? A
    triangular number is a number of the form n(n
    + 1)/2, where n is a positive integer.

    -- Elementary Number Theory by David M.  Burton

Solution:

    We need a value of n,m such that:
    
    T(n-1)*T(n)*T(n+1) = m^2
    
    Where T(n) = n*(n+1)/2
    
    or: n^6 + 3*n^5 + n^4 - 3*n^3 - 2*n^2 = 8*m^2
    
    I don't know how to solve this analytically.
    Can try to solve numerically.
    
    If something is a perfect square then the multiplicity
    of the prime factors must all be even.  We will use this
    as the test:
    
```{r}
library(gmp)

T = function(n) { n*(n+1)/2 }

m=0
n=1

while (TRUE) {
    if (all(rle(sort(as.numeric(c(factorize(T(n)),factorize(T(n+1)),factorize(T(n+2))))))$lengths%%2 == 0)) {
        m=m+1
        print(c(T(n),T(n+1),T(n+2)))
        if (m>4) { break }
    }
    n=n+1
}

```

    The first five solutions are:
    
    T(n)      T(n+1)   T(n+2)
    ===========================
           6       10       15
         300      325      351
       10878    11026    11175 <-- requested solution
      373680   374545   375411
    12718446 12723490 12728535


Analytical solution:

    The Bent does this algebraically.
    
    They get another form of the Pell equation.
    
    Consider the product: T(n-1)*T(n)*T(n+1)

        = (n^2*(n+1)^2/4) * (n-1)*(n+2)/2

    Clearly the first term is always a square = n*(n+1)/2
    
    We need to find n such that the second term is a square.
    
    Let:
    
        (n-1)*(n+2)/2 = p^2
        
        n^2 + n - 2 = 2*p^2
        
        4*n^2 + 4*n - 8 = 8*p^2
        
        (2*n+1)^2 - 9 = 8*p^2
        
        (2*n+1)^2 - 8*p^2 = 9
        
        (2*n+1)^2 - 2*(2*p)^2 = 9
        
    Let:
        r=2*n+1, q=2*p then
        
        r^2 - 2*q^2 = 9
        
    Which is a form of the Pell equation.
    
    The primitive form of the Pell equation is:
    
        x^2 - n*y^2 = 1
        
    If we can find any solution to the original equation
    and to the primitive equation we can generate an
    infinite number of solutions to the original equation.
    If the solution to the original equation is minimal then
    we will find all the solutions.
    
    Going back to Wikipedia for the Pell equation:
    
[Wikipedia](https://en.wikipedia.org/wiki/Pell%27s_equation)

    The primitive form:
    
        x^2 - 2*y^2 = 1 has solution of x=3, y=2
        
    By inspection/search we find a solution of:
    
        r^2 - 2*q^2 = 9  --> r=9, q=6
        
    From the Bent, successive solutions may be found as:
    
        r_i - q_i * sqrt(2) = (9-6*sqrt(2))*(3-2*sqrt(2))^x
        
    expanding albebraically and equating terms

    (I don't really understand this but I used maxima and
    looked at multipls of sqrt(2) for the q term):
        
    for x=0: r_i=  9, q_i=  6 --> n=4    T(3)  =    6
    for x=1: r_i= 51, q_i= 36 --> n=25   T(24) =  300
    for x=2: r_i=297, q_i=210 --> n=148  T(147)=10878

    The Wikipedia page contains a recurrence equation
    where x(1),y(1) are the solution to the primitive equation.
    
    If x(k),y(k) are solutions to the desired problem then
    this generates the next solution.  I think it is just the
    same logic used above in a different form.  The powers
    of x(1),y(1) are hidden in x(k),y(k):
    
    x(k+1) = x(1)*x(k) + n*y(1)*y(k)
    y(k+1) = x(1)*y(k) + y(1)*x(k)
    
    if n = 2, x(k)=9, y(k)=6, x(1)=3, y(1)=2 we get
    
    3* 9+2*2* 6 = 51,  3* 6+2* 9 = 3 6
    3*51+2*2*36 = 297, 3*36+2*51 = 210 

    Which is the same as above.


## Computer Bonus

    What is the largest base-10 prime, such that
    when the leading digit is successively
    removed until only one digit remains, the
    resulting series of integers, including the
    final digit, are all primes? Leading zeros
    are not allowed, and 1 is not a prime number.

    -- Penguin Dictionary of Curious and
    Interesting Numbers by David Wells

TODO

    Not sure why there even is a largest but I
    would have to resort to brute force and I
    don't think it would be pretty.

    The solution is given in the Bent without any
    explanation as give below:
    
        357,686,312,646,216,567,629,137 is
        the largest decimal prime integer
        with the property that as the leading
        digit is successively removed, the
        resulting integers are still all primes.
