---
title: "Bent Summer 2013"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Summer 2013


## Problem 1

    What conclusions (about shingles, happiness,
    Klingon, Ph.D. candidate, mom, and mixed
    dominance) can be drawn from these eight
    premises?

    1. Everyone who works in the Reliable
       Data Dump has shingles.
    2. All hippies are unhappy.
    3. No one whose mother is, or has been a
       shaman suffers simultaneously from
       shingles and mixed dominance.
    4. Christopf L. Biggleswade works in the Reliable Data Dump.
    5. All hippies are fluent in Klingon.
    6. Christopf L. Biggleswade is a hippy.
    7. Everyone fluent in Klingon is a
       candidate for the Ph.D. degree, or else
       his mother is, or has been, a shaman.
    8. No one who is unhappy can be a candidate for the Ph.D.

-- Professor H. Webb

TODO

    meh
    

## Problem 2

    You have an eight-pint jug full of wine and
    want to give exactly half to your friend.
    However, all you have is a three-pint jug and
    a five-pint jug, both empty. How do you
    arrive at four pints in each of the five-pint
    and eight-pint jugs by just pouring wine from
    one jug to another (in the minimum number of
    pours)? Show your answer as the state of each
    jug after each pour.

    -- Master Book of Mathematical Recreations by Fred Schuh

Solution:

    Start with pour into 5
    
    8   5   3
    =========
    8   0   0
    3   5   0
    3   2   3
    6   2   0
    6   0   2
    1   5   2
    1   4   3
    4   4   0


## Problem 3

    A, B, C and D all played each other once at
    soccer. Some of the figures in the table of
    results (in which they are not necessarily
    arranged in the order of points scored) are
    given below.

                               Goals  Goals
    Played  Won  Lost  Drawn   For    Against
    -----------------------------------------
    A 3                                5
    B 3      3                  4
    C 3
    D 3                  1      0      3

    With the additional information that the
    match between A and C was a draw (3-3) and
    that A scored more than 5 goals altogether,
    please fill in the table, and provide the
    scores of each game.

    -- Brain Puzzler's Delight by E.R. Emmet

TODO

    meh
    

## Problem 4

    Find 10 primes in arithmetic progression,
    that is find an expression of the form a +
    n*b, which generates primes for values of n
    from 0 through 9 inclusive. We want the
    expression where a + 9*b is the smallest.

    -- Penguin Dictionary of Curious and
    Interesting Numbers by David Wells

Solution:

    I did not understand this problem.  I thought
    the primes had to be sequential.  That is not
    what they were looking for.
    
    The solution is 199+210*N
    
    From the Bent:
    
    Number theory tells us that the difference
    between consecutive terms in an arithmetic
    sequence of primes is the product of the 
    prime numbers that are less than the required 
    number of terms in the sequence.
    
    In this case, for 10 terms: 2*3*5*7 = 210
    
    So we know we are looking for values that are
    
    a + 210*b for b=1,2,...,9 and a is prime
    
    We should be able to find a fairly easily:
    
```{r}
library(gmp)

n=2
while(T) {
    s=n+210*(1:9)
    if (all(isprime(s)==2)) {
        print(n)
        break
    }
    n=nextprime(n)
}
```
   
This runs almost instantly.

The first value in the sequence is 199.  The full
sequence is:

    199  409  619  829 1039 1249 1459 1669 1879 2089
    
Original - misguided attempt

    Exhaustive search looking at the diff between
    adjacent primes.  If the diff is constant for
    9 diffs then we have a solution.
    
    NOTE - THIS CODE IS FOR REFERENCE ONLY -- NOT THE SOLUTION
    FOR THE CORRECT PROBLEM!!!
    
```{r, eval=FALSE}
library(gmp)

n1=2
ndiff=0
currdiff=-1
while(T) {
    n2=nextprime(n1)
    if ((n2-n1) == currdiff) {
        ndiff=ndiff+1
        if (ndiff == 9) {
            # Done
            print(as.character(n2))
            print(as.character(n1))
            print(as.character(n1-currdiff))
            print(as.character(n1-2*currdiff))
            break
        }
    } else {
        currdiff = (n2-n1)
        ndiff=1
    }
    n1=n2
}
```


## Problem 5

    Solve the following cryptic addition problem
    with the largest TIGER.  Usual rules: Base
    ten, no leading zeros, different letters are
    different digits, same letter is same digit
    throughout.

        B E A V E R
     +    T I G E R
        -----------
        R A B B I T

    -- 150 Puzzles in Crypt-Arithmetic by Maxey
    Brooke

Solution: from the cryptic arithmetic notebook

    Cryptic addition with constraint

      BEAVER + TIGER = RABBIT

      TIGER should have largest possible value

    a=make_crypt_df('beavertigerrabbit')
    x_beaver=make_num(a,'beaver')
    x_tiger=make_num(a,'tiger')
    x_rabbit=make_num(a,'rabbit')
    k=(x_beaver+x_tiger == x_rabbit)
    k=k & (a$b != 0) & (a$t != 0) & (a$r != 0)
    > a[k,]
           b e a v r t i g
    446833 2 5 1 4 3 6 0 7
    447565 2 5 1 7 3 6 0 4

    First solution has maximum TIGER = 60753


## Bonus

    There are three pegs located on the
    circumference of a circle: peg A is at 12
    o'clock; peg B is at 4 o'clock; and peg C is
    at 8 o'clock. Initially, there are N disks
    (all different sizes) on peg A, stacked in
    size from largest at the bottom to smallest
    at the top. Find the minimum number of moves
    to transfer all of the disks to peg B, one at
    a time, never having a larger disk on top of
    a smaller one. Only clockwise moves (to
    adjacent peg) are permitted. Also, what is
    the minimum number of moves to transfer all
    the disks to peg C (from peg A) with the same
    rules? We want direct (not recursive)
    formulas. As a check, fill in the following:

        N 0 1 2 3 4 5 6 7 8 9
        B 0 1 5
        C 0 2

    where, N is the number of disks on the A peg,
    B is the number of moves to move those N
    disks to the B peg, and C is the number of
    moves to move those N disks to the C peg
    (from the A peg).

    -- Allan Gottlieb's Puzzle Corner
    in Technology Review

TODO

    Isn't this just the Tower of Hanoi?
    
    Yes - with the cyclic move constraint.  They even
    mention this in the Bent and refer people to the
    OEIS for the formula.


## Computer Bonus

    A bingo card consists of a 5 x 5 grid with
    numbers in each cell, except the center,
    which is marked FREE. The five columns are
    labeled B, I, N, G, and O. Column B consists
    of five different numbers in the range 1
    through 15, column I of five different
    numbers from 16 through 30, N of four numbers
    between 31 and 45, G of five numbers between
    46 and 60, and O of five numbers between 61
    and 75. The caller has 75 balls, numbered 1
    to 75, and randomly calls numbers (without
    replacement) until someone calls "Bingo."
    Bingo is scored by getting five numbers (or
    four numbers and the FREE cell) in a row,
    vertically, horizontally, or two main
    diagonals. What is the expected number of
    calls to get Bingo on an arbitrarily selected
    Bingo card?

    -- H.G. McIlvried III PA Gamma'53

Solution:

    Monte-Carlo?
    
    We can make random cards and then play games until
    we win.
    
```{r}
ntrials=10000

set.seed(42)
total_moves=0
for (trial in 1:ntrials) {
    # Make a random bingo card
    b = sample(1:15,5)
    i = sample(16:30,5)
    n = sample(31:45,5); n[3]=-1
    g = sample(46:60,5)
    o = sample(61:75,5)
    # print(b); print(i); print(n); print(g); print(o)
    # Check the numbers until winner found
    s=sample(1:75)
    for (j in seq_along(s)) {
        m=s[j]
        num_moves=j
        # Change the value in the card to -1
        b[b==m]=-1
        i[i==m]=-1
        n[n==m]=-1
        g[g==m]=-1
        o[o==m]=-1
        # Check for a win
        # Down the columns
        if (all(b==-1) | all(i==-1) | all(n==-1) | all(g==-1) | all(o==-1)) {
            break
        }
        r1 = c(b[1],i[1],n[1],g[1],o[1]); if (all(r1 == -1)) { break }
        r2 = c(b[2],i[2],n[2],g[2],o[2]); if (all(r2 == -1)) { break }
        r3 = c(b[3],i[3],n[3],g[3],o[3]); if (all(r3 == -1)) { break }
        r4 = c(b[4],i[4],n[4],g[4],o[4]); if (all(r4 == -1)) { break }
        r5 = c(b[5],i[5],n[5],g[5],o[5]); if (all(r5 == -1)) { break }
        d1 = c(b[1],i[2],n[3],g[4],o[5]); if (all(d1 == -1)) { break }
        d2 = c(b[5],i[4],n[3],g[2],o[1]); if (all(d2 == -1)) { break }
    }
    # print(b); print(i); print(n); print(g); print(o)
    total_moves=total_moves+num_moves
}

print(paste0("Total number of moves = ",total_moves))
print(paste0("Mean  number of moves = ",total_moves/ntrials))

```

Monte-Carlo summary:

    Seed   Trials   Result
    42     10K      41.4080
    123    10K      41.4165
    2020   10K      41.4490
    42     100K     41.39153
    42     1000K    41.373945
    
The Bent, unbelievably, also uses Monte-Carlo.  Their
estimate is 41.37 but they do not list the number of
trials or any other measure of variability.  One
interesting note:

    Since the card and draws are random there is no
    need to generate a random card every time.
       
I am wondering if this observation could be used to
speed up the solution.

