---
title: "Bent Summer 2005"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Summer 2005


## Problem 1

    Five chess fanatics always manage
    to fit in a round-robin before their train
    reaches Waterloo. Yesterday there
    were no ties in the final order (2 points
    for win, 1 for draw, 0 for loss). Alapin
    won his game against the only person
    who took a game off Bird. Catalan was
    the only person to lose a game against
    the financier, who was the only person
    to lose a game to Dunst. Alapin finished
    below the ghost-writer. The interpreter
    scored only 1 point total. The
    journalist fared worse than the hairdresser.
    As for Evans, you can work
    that out for yourself. What is each
    person's score and occupation?

    -- New Scientist

Solution:

    People: Alapin, Bird, Catalan, Dunst, Evans
    Jobs:   financier, ghost-writer, interpreter, journalist, hairdresser
    
    Each match is Win (2), Draw (1), or Loss (0)
    
    Round-robin --> each person plays all other people
    Possible scores for individual = 0-8
    In each match there are exactly 2 points awarded
    Must be an even number of draws
    Number of wins = number of losses
    
    Bird has 1 Loss.  Whoever beat him lost to Alapin.
        Bird score >= 3
        Alapin score >= 2
    financier has 1 Win, he beat Catalan
        Catalan is not the financier
        financier score <= 5
    Dunst has 1 Win, he beat financier
        Dunst is not the financier
        Dunst score <= 5
    Alapin < ghost-writer
        Alapin is not the ghost-writer
    interpreter = 1 (e.g. no wins)
        Dunst is not the interpreter
        Alapin is not the interpreter
    journalist < hairdresser

    Yuck -- this is supposed to be fun?

The Bent provides the answer but no explanations.


## Problem 2

    Consider a set of N distinct positive
    integers of your choice, the sum of any
    K of which is prime. What is the maximum
    possible value of N for K = 2, 3,
    4, and 5?

    -- Technology Review

TODO

I have no idea.  Even after reading the solution in the Bent
this is not much fun and they even admit they don't know the
anwer for 5 but 7 is the most they can find.


## Problem 3

    Frances was trying to remember
    how to arrange the numbers 1 to 9 in a
    magic square. She tried:

        7 8 9
        6 1 2
        5 4 3

    This was her first shot, and, as you can
    see, rather than getting the three columns,
    three rows, and two diagonals
    to all add to the same total, they all
    summed to different totals. "Look,"
    she said, "I have invented an antimagic
    square." What is the smallest
    anti-magic square, consisting of nine
    positive, not necessarily all different,
    integers? In judging smallest, the first
    criterion is to minimize the largest integer
    used, the second criterion is to
    minimize the total of all nine integers,
    and the third criterion is to minimize
    the sum of the first row.

    -- Anti-Magic Square
    by Stephen Ainley

TODO

I have no idea.

No real explanation in the Bent -- jut a proposed anti-magic square.


## Problem 4

    If N is a positive integer, what rational
    values may be assumed by R =
    sqrt((N/7)^2 + 5,036)?

    -- American Mathematical Monthly

Solution:

    R = sqrt( (N/7)^2 + 5036)
      = sqrt( N^2 + 7^2*5036)/7
      
    The only was this is rational is if N^2 + 49*5036 is a square.
    
    This equation can not have a solution when N is "big enough"
    since the separation between adjacent integer squares will
    exceed 49*5036 = 246764:
    
    (n+1)^2 - n^2 > 246764
    2*n+1 > 246764
    n > 123381
    
    So we only need to consider values of n in the range [1,123381].
    
    We can find this with exhaustive search:

```{r}
for (n in 1:123381) {
    m=n^2+49*5036
    k=round(sqrt(m))
    if (k*k == m) { print(n) }
}
```

The only solutions are for n = 1210, 8806, 61690

The values for R = 1308/7, 8820/7=1260, and 61692/2


## Problem 5

    The Parliament of Puevigi plans to
    divide the population into 10 income
    groups and to average the wealth between
    each pair of neighboring groups.
    The averaging will be done as nine distinct
    steps done in order: starting with
    the two lowest (poorest) groups, then
    groups two and three, ending with the
    top two groups. An amendment to this
    plan proposes to work from the top
    down. Which plan, if either, should the
    richest groups prefer? The poorest
    groups prefer?

    -- Litton Industries
    "Problematical Recreations"

Solution: I'm not sure I understand the problem but I will
try it anyhow.  Lets assume the groups have the following
income levels: v1, v2, v3, ..., v10 where group 1 is
the poorest and group 10 is the richest.

Going from low to high:

    After one round: (v1+v2)/2, (v1+v2)/2, v3, ..., v10
    After two: (v1+v2)/2, ((v1+v2)/2+v3)/2, ((v1+v2)/2+v3)/2, ... v10
    at the end, the final (hightest) group will get:
    
    (v1+v2)/2^9 + v3/2^8 + v4/2^7 + ... + v10/2
    
Going from high to low:

    After one round: v1, v2, ..., v8, (v9+v10)/2, (v9+v10)/2
    After two: v1,v2, ..., v8/2+(v9+v10)/2^2, (v9+v10)/2
    and the end, the final (lowest) group will get:
    
    v1/2 + v2/2^2 + ... , (v9+v10)/2^9

For the poorest group:

    Low to High --> (v1+v2)/2
    High to Low --> (v1+v2)/2 - v2/4 + ....
    
For the richest group:

    Low to High --> (v9+v10)/2 - v9/4 + ....
    High to Low --> (v9+v10)/2

The relative merit will depend on the magnitude of group 2 or group 9 versus the
other groups (not including group 1 or group 10).  If there is a large gap in
income near the top/bottom of the range the second group from the end will
suffer the precise opposite of all the other income groups.

If we assume integer values for the groups as they do in the Bent:

```{r}
n=1:10
for (i in 1:9) {
    t=(n[i]+n[i+1])/2
    n[i]=t
    n[i+1]=t
}
print('Bottom-Up')
n
```

```{r}
n=1:10
for (i in 9:1) {
    t=(n[i]+n[i+1])/2
    n[i+1]=t
    n[i]=t
}
print("Top-Down")
n
```

Under this distribution the top and bottom groups benefit from top-down.  The other groups,
however, have a more complex behavior.  In the calculations above, 8 of the 10 income
groups do BETTER WITH BOTTOM UP.  Only the top/bottom groups benefit from top-down.


What if we assume something other than simple integers -- e.g. a big gap 1-2, 8-9, or 9-10:

```{r}
n=c(1,1002,1003,1004,1005,1006,1007,1008,1009,1010)
for (i in 1:9) {
    t=(n[i]+n[i+1])/2
    n[i]=t
    n[i+1]=t
}
print('Bottom-Up')
n
```

```{r}
n=c(1,1002,1003,1004,1005,1006,1007,1008,1009,1010)
for (i in 9:1) {
    t=(n[i]+n[i+1])/2
    n[i]=t
    n[i+1]=t
}
print('Top-Down')
n
```

Playing with various scenarios moves the pain around in the central 8 groups.

The Bent solution only considers the extreme two groups.


## Bonus

    At a summer camp, three campers
    show up at the mess hall for their
    share of the lemonade. One camper has
    a 10 Liter container, the second has a 6
    L container, and the third has a pail of
    unknown capacity, but greater than 2 L.
    The mess-hall attendant has a 15 L container
    full of lemonade, and she needs
    to distribute exactly 8 L to the 10 L container,
    exactly 5 L to the 6 L container,
    and exactly 2 L to the pail. If no other
    containers are available, how does she
    accomplish this? Present your answer
    as a table with four columns labeled 15L,
    10L, 6L, and Pail, with each row giving
    the contents of the containers after each
    transfer. For example, the first row is
    (15, 0, 0, 0); now if the 10 L container is
    filled, the second row would be (5, 10, 0,
    0), etc. Containers cannot be tipped at
    an angle to provide half or any other
    fraction of their volume. We want the
    shortest sequence of exchanges that will
    work.

    -- Dr. Richard I. Hess, CA B'62

Solution:

This is harder than you might think!

    0        8    5   2  <-- Target

    Kitchen  10L  6L  Pail
    ----------------------
    15       0    0   0
    5        10   0   0

What if we built a program to make random transfers until
the proper totals were achieved?

Even with 1e6 random pours this does not find the solution!  Multiple runs
do not even find any two containers with 8 and 5.

The solution from the Bent is given here:

    15L 10L 6L Pail
    15  0   0   0 => START
    9   0   6   0
    9   6   0   0
    3   6   6   0
    3   10  2   0
    13  0   2   0 also: (3, 10, 0, 2)
    13  0   0   2 also: (3, 4, 6, 2)
    7   0   6   2
    7   6   0   2
    1   6   6   2
    1   10  2   2
    11  0   2   2
    11  2   0   2
    5   2   6   2
    5   8   0   2
    0   8   5   2 => END


```{r}

q = c(15,0,0,0)     # Quantity in each container
cap = c(0,10,6,2)   # maximum capacity of each container, cap[1]=X

max_pours=1000000
num_pours=0
num_8=0
num_5=0
num_2=0
while(T) {
    source=sample(1:4,1)   # Source container
    dest=sample(1:4,1)     # Destination container
    
    if (source != dest) {      # Transfer might take place
        if (q[source] != 0) {  # There is fluid to transfer
            if (dest == 1) {   # Put full amount back in the kitchen
                q[dest] = q[dest]+q[source]
                q[source]=0
            } else if ((dest == 2) || (dest==3)) {
                space=cap[dest]-q[dest]
                if (space>0) {  # There is room for some fluid
                    if (q[source]>space) { # Partial trasfer from source to dest
                        q[dest]=q[dest]+space
                        q[source]=q[source]-space
                    } else {               # Complete transfer from source to dest
                        q[dest]=q[dest]+q[source]
                        q[source]=0
                    }
                }
            } else {
                if (q[source]<=cap[dest]) {  # Container can't measure - just accept
                    q[dest]=q[source]
                    q[source]=0
                }
            }
            s=sprintf("(%d,%d): %2d %2d %2d %2d",source,dest,q[1],q[2],q[3],q[4])
            num_pours=num_pours+1
            if (num_pours>max_pours) {
                print("Target not achieved")
                break
            }
            if (any(q==8) && any(q==5)) {
                print(s)
            }
            if (any(q==8)) { num_8 = num_8 + 1 }
            if (any(q==5)) { num_5 = num_5 + 1 }
            if (any(q==2)) { num_2 = num_2 + 1 }
            if (all(q == c(0,8,5,2))) {
                print(s)
                break
            }
        }
    }
}
print(paste0("Number of times 2 achieved = ",num_2,collapse=""))
print(paste0("Number of times 5 achieved = ",num_5,collapse=""))
print(paste0("Number of times 8 achieved = ",num_8,collapse=""))
```



## Double Bonus

    Given three circles,
    each with a different radius, externally
    tangent to each other. For each pair of
    circles, construct the two lines that are
    tangent to both circles, but not the tangent
    line where the two circles meet.
    Each of these pairs of lines will meet
    at a point. Prove that the three meeting
    points lie on a straight line.

    -- Daryl Cooper

TODO

This seems like fun.  Need paper/pen for this.
