---
title: "Bent Fall 2018"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Fall 2018


## Problem 1

    Solve this cryptic addition with
    the usual rules: different letters
    are different digits, same letter is
    same digit, no leading zeros, base
    10:

        SIXTEEN + TWENTY +
        TWENTY + TEN + TWO + TWO =
        SEVENTY.

    -- Journal of Recreational
    Mathematics

Solution: from the cryptic arithmetic notebook

    Cryptic addition:

      SIXTEEN + TWENTY + TWENTY + TEN + TWO + TWO = SEVENTY

    a=make_crypt_df('sixteentwentytentwoseventy')
    x1=make_num(a,'sixteen')+
       make_num(a,'twenty')*2+
       make_num(a,'ten')+
       make_num(a,'two')
    x2=make_num(a,'seventy')
    k=(x1==x2)
    k=k&(a$s!=0)&(a$t!=0)
    sum(k)
    [1] 1
    a[k,]
            s i x t e n w y o v
    1125065 3 0 9 2 6 5 7 8 1 4


## Problem 2

    Hook, Line, and Sinker returned from a day's
    fishing, and each reported his catch
    secretly, but accurately, to George Gaff,
    landlord of The Complete Idiot. "Well,
    Gents," George announced later.  "Hook caught
    the most fish, and Sinker caught the fewest
    (no ties). If you divide Hook's catch by
    Sinker's catch, you get Line's catch." Upon
    reflection, Hook remarked, "I know how many
    fish each of us caught." Line then chimed in
    with, "I also know how many each of us
    caught." "But I don't know," Sinker
    complained after a short pause.  "Never mind,
    Old Chap," said George. "I'll give you a
    clue. I've been fishing, too, and I caught
    fewer fish than Hook. If you know how many I
    caught, you can figure out how many fish Line
    caught." Without further help, Sinker managed
    to work out Line's catch. Sinker was proud of
    himself and remarked that it had been a good
    day, since his catch was twice what he had
    caught the previous week. How many fish did
    Hook, Line, and Sinker each catch?

    -- A Tantalizer by Martin Hollis in New
       Scientist

TODO

    meh


## Problem 3

    In how many different ways (order does not
    matter) can one change a $100 bill using $1,
    $2, $5, $10, $20, and $50 bills?

    -- Challenging Mathematical Problems with
    Elementary Solutions by A.M. & I.M. Yaglom

Solution:

    Brute force?
    
```{r}
n=0
solution=c()
for (n1 in 0:100) {
    for (n2 in 0:50) {
        for (n5 in 0:20) {
            for (n10 in 0:10) {
                for (n20 in 0:5) {
                    for (n50 in 0:2) {
                        t=n1+2*n2+5*n5+10*n10+20*n20+50*n50
                        if (t==100) { 
                            n=n+1
                            solution=append(solution,c(n1,n2,n5,n10,n20,n50))
                        }
                    }
                }
            }
        }
    }
}
solution=matrix(solution,ncol=6,byrow=T)
colnames(solution)=c("1","2","5","10","20","50")

print(n)

```

    There are only 4562 ways to do it.
    
    If you really want -- see the solution matrix for all
    the combinations.


## Problem 4

    What is the maximum number of queens that can
    be placed on an 8x8 chessboard so that each
    queen threatens (that is, could capture on
    the next move) exactly two other queens? A
    queen can move horizontally, vertically, or
    diagonally as many squares as desired or
    until another piece or the edge of the board
    is encountered. Using Q's and -'s, provide an
    8x8 grid that represents the chess board.

    -- The Colossal Book of Mathematics by Martin
       Gardner

TODO

    meh


## Problem 5

    The There'sAPillForEverything Pharmaceutical
    Company received a shipment of 12 bottles of
    pills, each bottle containing 1,000 pills.
    The labels on the bottles indicated that each
    pill weighed 100 mg. However, before any
    pills had been sold, the pharmacist received
    a notice from the supplier that the
    manufacturer had some problems, and some of
    the pills might weigh 110 mg. However, if a
    110 mg pill was found in a bottle, all the
    pills in that bottle would weigh 110 mg. The
    pharmacist has an ordinary pan scale that is
    accurate to 1 mg. What is the minimum number
    of weighings of pills from the various
    bottles required to determine which bottles,
    if any, contain pills weighing 110 mg?
    Explain how the weighings are to be carried
    out. The pan on the scale can hold at most 24
    pills at the same time. Pills on the scale
    can be identified as to which bottle they
    came from.

    -- The Crucible

TODO

    Use binary encoding to weight 1+2+4+8
    pills from each of four bottles.  The
    result should be 1500 mg.  If it is not,
    the size of the error will tell you which
    bottle was wrong.
    
    Repeat for two more sets of four bottles.
    
    In total, maximum of 3 weighings required.


## Bonus

    Two positive integers with no common integer
    factor (other than one) are selected. Their
    sum is written on the forehead of logician A
    and the sum of their squares is written on
    the forehead of logician B.  The two
    logicians are allowed to see the numbers on
    each-other's foreheads.  The following
    conversation takes place.

    A (sees sum of squares): I don't
    know my number.
    B (sees sum): I don't know my
    number.

    A: I don't know my number.
    B: I don't know my number.
    A: I don't know my number.
    B: I don't know my number.
    A: I don't know my number.
    B: I don't know my number.
    A: I know my number.

    What are the numbers on A's and B's
    foreheads?

    -- Richard I. Hess, CA B'62

TODO

    meh


## Computer Bonus

    Find five different positive integers such
    that the sum of any two of them is a perfect
    square.

    -- Allan Gottlieb's Puzzle Corner in
    Technology Review

Solution:

    Hmmmm ...
    
    Not even sure how to brute force this unless the
    integers are pretty small.
    
    Here is the Bent solution without any explanation:
    
        7442, 28658, 148583, 177458, and 763442
        are five different positive integers such
        that the sum of any two of them is a
        perfect square.

    I can confirm this is true but I don't know how
    they would have found them.
