---
title: "Bent Fall 2012"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Fall 2012


## Problem 1

    Two small rockets, one with a mass of 2 kg
    and the other with a mass of 26 kg, are
    traveling through space with no forces acting
    on them. The velocity of each is a positive
    integral number of meters per second such
    that the difference in their kinetic energies
    is 1 J. Each rocket then fires a short burst
    of its ion engine, after which each again has
    a velocity of an integral number of meters,
    and the difference in their kinetic energies
    is still 1 J. What are the smallest positive
    velocities which the two rockets can have
    before and after they fire their engines?
    Assume that the masses of the rockets do not
    change when firing the engines.

    -- Adapted from Theodore J. Stadnik, OH A'05

Solution:

    Initial state:
    
    E1 = 1/2*m1*v1^2, m1=2 kg
    E2 = 1/2*m2*v2^2, m2=26 kg
    
    E1-E2 = 1 or E2-E1 = 1
    
    v1=integer>0, v2=integer>0

    After engine burn -- essentially the same but need new solution set.

    I don't know how to find just integer solutions so I will make an assumption
    about v2 and then calculate required v1 and see if it is an integer.
    
```{r}
vmax=1000
m1=2
m2=26

for (v2 in 1:vmax) {
    e2=1/2*m2*v2^2
    v1a=sqrt(2*(e2+1)/m1)
    v1b=sqrt(2*(e2-1)/m1)
    if (v1a == round(v1a)) {
        print(paste0("v2 = ",v2," v1 = ",v1a))
    }
    if (v1b == round(v1b)) {
        print(paste0("v2 = ",v2," v1 = ",v1b))
    }
}

```

Initial state:

    v2 = 5, v1 = 18

Final state:

    v2 = 180, v1 = 649

The Bent notes that this is something called a Pell equation
which has the propery that if you know one solution the others
can be generated from it.  I just did the exhaustive search and
it ran very quickly.


## Problem 2

    Find a six-digit number with the following
    properties: a) the first (leftmost) digit is
    four more than the second digit, b) the third
    digit is one less than the second digit, c)
    the product of the first two digits equals
    the two-digit number formed by the third and
    fourth digits, d) the fifth digit equals the
    sum the first and third digits, and e) the
    sixth digit equals the sum of the fourth and
    fifth digits.

    -- Mark A. Noblett, P.E., NY N'70

Solution:

    We will use exhaustive search with baseOf() from gtools.
    
```{r}
library(gtools)

a=baseOf(100000:999999)
k1=(a[,1] == (a[,2]+4))
k2=(a[,3] == (a[,2]-1))
k3=((a[,1]*a[,2])==(10*a[,3]+a[,4]))
k4=(a[,5] == (a[,1]+a[,3]))
k5=(a[,6] == (a[,4]+a[,5]))
k=(k1 & k2 & k3 & k4 & k5)

print(paste0("Number of solutions = ",sum(k)))
a[k,]
```

The only solution is: 621279


## Problem 3

    A band of pirates has stolen a small chest
    containing nine identical bottles of gold
    dust that each weigh exactly the same. They
    decide to wait until the next day to divide
    the loot. During the night, one pirate
    surreptitiously opens one of the bottles and
    removes a small amount of gold. Later he
    starts to worry that, if he gets caught, he
    will have to walk the plank, so he sneaks
    back, grabs a bottle at random and replaces
    the gold he took. Using only a simple two-pan
    balance scale, what is the minimum number of
    weighings required to assure either that the
    bottles all have the same weight or to
    determine the light and heavy bottles?

    -- You'd Better Be Really Smart Brain Bafflers by
    Tim Sole and Rod Marshall

TODO

    meh.  My head hurts.
    

## Problem 4

    Alice and Beth each toss n fair coins. What
    is the probability that they throw the same
    number of heads?

    -- adapted from Problems in
    Probability Theory, Mathematical Statistics
    and Theory of Random Functions by A.A.
    Sveshnikov

Solution:

    Let us try some induction:
    
    Case 1: n = 1
    
    Alice   Beth   Result
    ---------------------
    T       T      1
    T       H      0
    H       T      0
    H       H      1
    
    Pr = 1/2
    
    Case 2: n = 2
    
    Alice   Beth   Result
    ---------------------
    T T     T T    1
    T T     T H    0
    T T     H T    0
    T T     H H    0
    T H     T T    0
    T H     T H    1
    T H     H T    1
    T H     H H    0
    H T     T T    0
    H T     T H    1
    H T     H T    1
    H T     H H    0
    H H     T T    0
    H H     T H    0
    H H     H T    0
    H H     H H    1
    
    Pr = 6/16 = 3/8
    

    After playing with this a bit -- it generalizes to:
    
    sum((choose(n,0:n)*0.5^n)^2)

    This can probably be algebraically reduced to something
    a lot prettier.  They do this in the Bent.  Apparently
    the summation can be replaced with choose(2*n,n) and then
    they factored out the 1/4^n.

    This function is nice and linear on log-log scale:
    
```{r}

x=1:100
y=sapply(x,function(n) {sum((choose(n,0:n)*0.5^n)^2)})

plot(x,y,type='b',log="xy",xlab="Number of Coins",ylab="Pr(Match Heads)")
grid()

```

## Problem 5

    A Tau Bate is standing 5 meters away from a
    long, tall, vertical wall. She holds a garden
    hose in her hand. The nozzle is 1 meter above
    the ground and water emerges from the nozzle
    at a velocity of 10 m/s. Using only her wrist
    to vary the horizontal and vertical angles at
    which water leaves the nozzle, what is the
    maximum area of the wall that she can wet
    down? Use 9.81 m/s2 for the acceleration due
    to gravity and report the answer to 3
    significant digits.

    -- John L. Bradshaw, PA A'82

Solution:

    This was fun.  Spent a lot of time on this and
    got a good chance to play with Maxima for the
    symbolic calculations.

    We will assume simple ballistic motion for the
    droplets of water.  We need to consider two
    parts of the geometry - the angle of elevation
    (aka the vertical angle) and the angle of the
    trajectory wrt the wall (aka the horizontal
    angle).
    
    The given data:
    
        V0 = 10 m/s
        x0 = 5 m
        g  = 9.81 m/s^2
        
    Let:
    
        theta = vertical angle
        phi   = horizontal angle
        
    Then - the water droplet trajectory is defined
    by:
    
        y = y0 + v0*sin(theta)*t - 1/2*g*t^2
        x = x0 + v0*cos(theta)*t

    Where:

        x0 = 0, y0 = 1, v0=10

    Solving and maximimizing this value symbolically, however, gets
    very ugly.  BTW: The Bent did this symbolically so I worked it
    through at the end of this problem.
    
    For now, let us try a numerical approach:
    
    Consider a plan view.  If we measure phi wrt the perpindicular
    to the wall where phi=0 then we can calculate the distance from
    the hose to the wall as:
    
        distance = 5/cos(phi)
        
    The time for the water to cover this distance in the x-direction
    depends on the launch angle and is given by:
    
        x = 5/cos(phi) = 10*cos(theta)*t
        
        t = (5/cos(phi))/(10*cos(theta))
        
    The height of the water at this time is calculated from:
    
        y = y0 + v0*sin(theta)*t - 1/2*g*t^2
        
    If this height is <0 then the water failed to reach the wall
    since it hit the ground first.
    
    For each value of phi we want to calculate the max value of y.
    We can do this with numerical optimization.
    
    So the strategy is this:
    
    1. Sweep phi over a range of angles from 0 to phi_max.  Recognize
       the there is a max value of phi where the water will never
       reach the wall.  A quick calculation shows this is ~ 1.1 radians.
    2. For each value of phi, calculate required x distance
    3. For each x distance calculate maximum y
    4. Integrate this mess to get an area.  Note the integration
       delta along the wall will change as sin(phi) so we will have
       numerical vectors for x,y with uneven spacing in x that needs
       a numerical summation.  If we take enough steps then rectangles
       should be good enough.
    
    
```{r}

# Calculate intersection of wall and water stream
# Given the distance the water flows in the x-plane
# Our control variable is the launch angle (theta) so
# that is the input parameter to the optimization.
# The important result, however, is the height of
# the water on the wall -- this is called $objective
# in the syntax of the optimize result.
max_y = function(x_distance) {
    ym=optimize(
        function(theta) {
            tx=x_distance/(10*cos(theta))
            y=1+10*sin(theta)*tx-1/2*9.81*tx^2
            return(y)
        },
        c(0,pi/2),
        maximum=TRUE)
    return(ym$objective)
}

phi=seq(0,1.2,length=10000)   # Horizontal angle
distance=5/cos(phi)           # Distance to wall
y=sapply(distance,max_y)      # Find max y for each distance
y=ifelse(y<0,0,y)             # if y<0 then can not reach wall
x=5*tan(phi)                  # distance along wall where stream intersects
dx=diff(x)                    # width of differential rectangles
area=sum(y[1:length(y)-1]*dx)*2    # Area (double for +/- phi)

print(paste0("Area = ",area))
plot(x,y,type='b',xlab="Distance along wall",ylab="Height on Wall")

```

Solution with many divisions:

    Num div   Area
    1e6       64.7130947940944
    100K      64.7138801451422
    10K       64.7217361967214
    1K        64.8005330493871

I guess I will take 64.713 as the answer.

Guess I didn't need all those extra divisions to get the digits since
they only requested three significant digits.

They actually have a closed form solution in the Bent:

    Area = (2^2.5*v / ((3*g^0.5 )*(y0 + v^2/(2*g) - g*xo^2/(2*v^2)))
    
    They only solve as 64.7
    
    I can not get the answer from their closed form solution.
    
    Need to look at this some more...

    Took a couple of hours but I found it -- they have a type setting error.
    They dropped a 3/2 power on the final parenthesized expression.  When
    shown properly it should look like this:
    
                                  5/2                  2     2       3/2
                             v0  2   ( y0 - g/2 (x0/v0)  + v0 /(2g) )
    (%o64)                 ----------------------------------------------
                                                  1/2
                                               3 g
    
    They have the same expression without the 3/2 power!!!  All of their
    other results and equations seem to be correct.

    Before I found their error I worked through their equations and came up
    with a closed form solution that yields:
    
    Area = 64.71300758305074
    
    My closed form solution was:
    
                    2       2   2     4 3/2
           2 (2 g v0  y0 - g  x0  + v0 )
    Area = --------------------------------
                       2   2
                    3 g  v0

    Note that in these solutions the value of x0=5 (x when phi=0).
    In the Bent they use z as the variable of distance along the
    wall.  The zmax value and formula seem correct but the Area
    formula is wrong as noted above.

    Some interesting results:
    
    You can express ymax=y(x) instead of y(t) if x0=0:
    
        ymax = y0 - g*x^2/(2*v0^2) + v0^2/(2*g)
        
    If we make x0=0 in my original equation then we can solve for
    t = f(theta) and substitute this into the equation for y:
    
                                     2         2
                                g sec (theta) x
    (%o108)            y = y0 - ---------------- + tan(theta) x
                                         2
                                     2 v0

    We can diff wrt theta and set = 0 and find optimal angle:
    
                                                  2
                                                v0
    (%o110)                       [tan(theta) = ---]
                                                g x

    If we let sec^2 = 1+tan^2 and sub tan(theta) back into equation
    above we get ymax = f(x).
    
    Finally, need to switch to coordinates along the wall using:
    
        z^2 + x0^2 = x^2
        
    Where x0 = shortest value of x to wall (e.g. for phi=0).
    
    Substituing back in yields:
    
                                       2     2            2
                                   g (z  + x0 )         v0
    (%o133)                 y = (- ------------) + y0 + ---
                                          2             2 g
                                      2 v0

    Finally, need to calculate max limits on z and integrate this along
    the z-wall to get total area.
    
    Note that xmax/zmax occurs when y finally drops to zero.  We can solve
    above for zmax with y=0 as:
    
                                         2       2   2     4
                              sqrt(2 g v0  y0 - g  x0  + v0 )
    (%o163)            zmax = -------------------------------
                                             g
    
    With: g=9.81, v0=10, x0=5 we get z = 9.964861771079583.
    
    This means the water can never reach more than +/-9.965m along the wall.
    
    So we need the following integral:
    
                          zmax
                         /             2     2            2
                         [         g (z  + x0 )         v0
    (%o172)            2 I     ((- ------------) + y0 + ---) dz
                         ]                2             2 g
                         /            2 v0
                          0

    Evaluating yields:
    
                   g  zmax  + ((- 6 g v0  y0) + 3 g  x0  - 3 v0 ) zmax
    (%o173)      - ---------------------------------------------------
                                               2
                                         3 g v0

    Numerically with g=9.81, zmax=9.964861771079583, v0=10, y0=1, x0=5:
    
        Area = 64.71300758305074

    An alternate formula for area without using zmax was given above in
    this derivation.


## Bonus

    A plane takes off from an airport on the
    equator and heads due northeast. If it always
    maintains a true northeast heading, what is
    the exact latitude at which it will cross its
    starting longitude for the first time? Assume
    that the earth is a perfect sphere and that
    the plane can be refueled while in flight.

    -- Technology Review

TODO

    Huh?

    Seems like the plane will spiral into the north pole but I don't
    know about calculating this on a globe.  Seems hard.  I would
    only guess that it is nearly to the north pole.


## Double Bonus

    This cryptogram is in honor of Jim Froula and
    his many years of service to Tau Beta Pi; he
    was named Secretary-Treasurer in 1982 and
    served as Executive Director until his
    retirement in 2011.

    J*FROULA = RETIRE*S

    Of course, we want Jim to enjoy maximum REST.
    The symbol * stands for multiplication. The
    problem is in base 12, and all the usual
    rules apply.

    -- Howard G. McIlvried, III, PA Gamma'53

Solution: in the cryptic arithmetic notebook

    Cryptic in Base 12 with constraint

      J * FROULA = RETIRE * S

      REST should have maximum value


    I switched to Python to use itertools for this since I didn't want to try
    creating a dataframe with 480e6 rows.  I am just dumping the solutions and
    then I will maximize REST by inspection.
    
    The run-time is not fast (~ 1500-2000 seconds) but it should at least finish.

      j   f   r   o   u   l   a   e   t   i   s      rest (base-12)
    ( 2, 10,  7,  1,  5,  4,  6,  0,  8, 11,  3) --> 7038
    ( 2, 10,  7,  5,  8,  4,  6,  0, 11,  9,  3) --> 703b
    ( 3,  2,  0,  4,  1,  5, 10,  6,  7,  8, 11) --> 06b7, also r=0
    ( 3,  2,  0,  4,  5,  1, 10,  6,  7,  9, 11) --> 06b7, also r=0
    ( 3,  2,  0,  5,  7,  9, 10,  6,  8,  1, 11) --> 06b8, also r=0
    ( 5,  1,  0,  6, 10,  3,  4,  8, 11,  9,  7) --> 087b, also r=0
    ( 5,  3,  9, 10,  1, 11,  0,  6,  7,  4,  2) --> 9627
    ( 6,  2,  1,  3,  7,  8,  0,  4, 10,  5,  9) --> 149a
    ( 7,  2,  5,  6,  1, 11,  0,  8, 10,  4,  3) --> 583a
    (11,  0,  2,  5, 10,  6,  9,  3,  4,  8,  1) --> 2xxx, also f=0
    (11,  0,  2,  6, 10,  5,  8,  4,  3,  7,  1) --> 2xxx, also f=0
    (11,  0,  2,  8,  6,  4,  7,  5,  9, 10,  1) --> 2xxx, also f=0
    (11,  0,  4, 10,  8,  2,  7,  5,  9,  6,  1) --> 4xxx, also f=0
    (11,  1, 10,  4,  5,  9,  6,  3,  0,  7,  2) --> a320 (maximum REST)
    
    Solution:
    
    Maximum REST = a320 (base-12) = 17736 (base-10)

      j  f  r  o  u  l  a  e  t  i  s
      2 10  7  1  5  4  6  0  8 11  3
      2 10  7  5  8  4  6  0 11  9  3
      5  3  9 10  1 11  0  6  7  4  2
      6  2  1  3  7  8  0  4 10  5  9
      7  2  5  6  1 11  0  8 10  4  3
     11  1 10  4  5  9  6  3  0  7  2  --> REST = a320 (base-12)
