---
title: "Bent Spring 2009"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Spring 2009


## Problem 1

    Scott, the astronaut, has landed on
    Planet X, which has the same density
    as Earth but only half the mass.
    Finding no atmosphere, he decides
    to leave; he fires his hyperdrive (that
    has the ability to generate thrust
    while consuming very little fuel) and
    lifts off. Unfortunately, just as he
    reaches half the escape velocity, his
    engine shuts off prematurely. How
    much time does Scott have to eject
    in his escape pod before his ship is
    pulled back to the planet's surface
    by gravity and crashes? Assume
    that the space ship has a constant
    mass of 1,000 kg and that its engine
    generates a constant thrust of 35,000
    N as long as it is firing. Ignore the
    effect of the planet's rotation and the
    change in gravity with distance from
    the surface of Planet X.

    --Scott L. Bilker, NJ Gamma'90

Solution:

    t = 1217 seconds (about 20 minutes)

    I don't think they really mean to ignore the change in gravity
    with distance from the surface of Planet X or else I don't
    understand escape velocity.  I think that admonition is just
    referring to the last part of the calculation.

    From Wikipedia the escape velocity for stationary object is:
    
        ve = sqrt(2*G*M/r) = sqrt(2*g*r)
        
        G = 6.67e-11 m^3 kg^-1 s^-2
        M(earth) = 5.9736e24 kg

    Note that all of this changes for rotating bodies.

    When fired vertically at velocity, v, the maximum height is:
    
        h = R*x^2/(1-x^2), where x=v/ve

    Assuming the planet is spherical with the same density as earth
    we have:
    
        r(planet X) = r(earth)*(1/2^(1/3))
        
    So:
    
        ve = sqrt(2*G*1/2*M(earth)/(r(earth)*(1/2^(1/3)))
           = sqrt(2*G*M(earth)/r(earth)*(1/2)/(1/2^(1/3)))
           = ve(earth) * sqrt((1/2)/(1/2^(1/3)))
           = ve(earth) / 2^(1/3)
           ~ 0.7937 * ve(earth)
           
    ve(earth) = 11.186 km/sec --> ve(planet X) ~ 8.878 km/sec
    
    Question: how long does it take to accelerate to this speed?
    
    We have the rocket pushing up and gravity pulling down.
    
    g(planet X) = g(earth)*(1/2)/((1/2)^(2/3)) = g(earth)/2^(1/3)
    
    During lift-off:
    
        F = ma = +35000 N - 1000 * 9.80665 / (2^(1/3))
            a = 27216.46 N / 1000 kg = 27.21646 m/s^2
            
    To reach 1/2 of the escape velocity:
    
        t = 1/2 * 8.878 * 1000 / 27.21646 = 163.1 seconds
        
    At this point the height will be:
    
        h = 1/2*a*t^2 = 362000 meters
        
    Velocity is upwards at:
    
        v0 = 1/2*8878 = 4439 m/s
        
    Falling from this height only under the acceleration of gravity:
    
        h = h0 + v0*t + 1/2*a*t^2
        
    To find h=0 we need to solve the quadratic with h0, v0, a given above.
    
        1/2*a = - 1/2 * (9.80665 / (2^(1/3))) = -3.891772 m/s^2
        v0 = 1/2*8878 = 4439
        h0 = 362000
    
        -3.891772*t^2 + 4439*t + 362000 = 0
    
        t = 1217 seconds (about 20 minutes)

    
## Problem 2

    Students Al, Bill, Carl, Don, and
    Ed live in rooms A, B, C, D, and E in
    the dorm, although not necessarily
    in that order. They have just taken
    a test in which they each got a different
    score. When questioned, they
    made the following remarks:

    Al   a1: Carl scored lower than Bill.
         a2: Don is in Room D.

    Bill b1: Ed scored lower than Carl.
         b2: Don is in Room C.

    Carl c1: Al is in Room C.
         c2: The student in Room B did
             not have the second-best score.

    Don  d1: Carl is in Room E.
         d2: Ed is not in Room D.

    Ed   e1: I scored lower than Bill.
         e2: Al got the highest score.

    Remarks in which the first person
    mentioned scored lower than the
    speaker are true, while remarks in
    which the first person mentioned
    scored higher than the speaker are
    false. Also, when making a remark
    about himself, only those who scored
    highest or next highest tell the truth.
    How did the students rank in the
    test, and what rooms are they in?

    --Brain Puzzler's Delight by
    E.R. Emmet

TODO

    Oh Joy - another logic problem ...


## Problem 3

    Solve the following cryptic division,
    which results in a repeating
    decimal.

    TBPI/AAAAA = .ONANDONANDON...

    We want the solution with the biggest
    PI. The usual rules for cryptics
    apply.

    -- H.G. McIlvried III, PA Gamma'53

Solution:  solved in the crypic arithmetic notebook

    Cryptic division with repeating decimal:

      TBPI / AAAAA = .ONANDONANDON ...
      Want solution with largest value of PI

    rewrite this to avoid the repeating decimal

      TBPI / AAAAA = ONAND / 99999
    or

      TBPI * 99999 = ONAND * AAAAA

```{r}
source('rgc_lib.R')

a=make_crypt_df('tbpiaaaaaonand')
x_tbpi=make_num(a,'tbpi')
x_aaaaa=make_num(a,'aaaaa')
x_onand=make_num(a,'onand')
k=(x_tbpi*99999 == x_onand*x_aaaaa)
k=k & (a$t != 0) & (a$a != 0)
print(paste0("Number of soutions = ",sum(k)))
print(paste0("TBPI  = ",x_tbpi[k,]))
print(paste0("AAAAA = ",x_aaaaa[k,]))
print(paste0("ONAND = ",x_onand[k,]))
print(paste(colnames(a),collapse=" "))
print(paste(a[k,],collapse=" "))

# x_tbpi[k]/x_aaaaa[k]
# x_tbpi[k]*99999
# x_aaaaa[k]*x_onand[k]
```

I originally overspecified the problem and required ONAND
to start with a nonzero digit.  With that constraint there
is only one solution  If you allow this to be zero
then there are two solutions.  The one with the larger value of
PI is:

    TBPI = 1784
    AAAAA = 33333
    ONAND = 05352
    
    So: 1784/33333 = 0.05352...


## Problem 4

    Ignoring colons, a 24-hour digital watch
    shows many palindromes during the day, e.g.,
    1:01:01 and 23:55:32.  How many palindromes
    show in 24 hours, what are the longest and
    shortest times between palindromes, and what
    two palindromes have a difference that is
    closest to 12 hours?  A 24-hour clock goes
    from 23:59:59 to 0:00:00.

    -- Technology Review

Solution:  We can make exhaustive search

```{r}

h=rep(0:23,each=3600)
m=rep(0:59,each=60,times=24)
s=rep(0:59,times=24*60)
a=data.frame(h=h,m=m,s=s)
a$t=apply(a,1,function(x) {paste0(x[1],sprintf("%02d",x[2]),sprintf("%02d",x[3]),collapse="")})
a$t2=apply(a,1,function(x) {paste0(rev(strsplit(x[4],"")[[1]]),collapse="")})
k=(a$t==a$t2)
b=a[k,]
print(paste0("Number of palindromes = ",nrow(b)))
rownames(b)=seq(nrow(b))
b$x=b$h*3600+b$m*60+b$s

n=which.max(diff(b$x))
print(paste0('Max difference (seconds) = ',b$x[n+1]-b$x[n]))
print(sprintf("%d:%02d:%02d",b[n,'h'],b[n,'m'],b[n,'s']))
print(sprintf("%d:%02d:%02d",b[n+1,'h'],b[n+1,'m'],b[n+1,'s']))
n=which.min(diff(b$x))
print(paste0('Min difference (seconds) = ',b$x[n+1]-b$x[n]))
print(sprintf("%d:%02d:%02d",b[n,'h'],b[n,'m'],b[n,'s']))
print(sprintf("%d:%02d:%02d",b[n+1,'h'],b[n+1,'m'],b[n+1,'s']))

d=outer(b$x,b$x,"-")
n=which(d==(12*3600))
colnum=trunc(n/nrow(b))+1
rownum=n%%nrow(b)
print("12 hour time difference:")
print(sprintf("%d:%02d:%02d",b[colnum,'h'],b[colnum,'m'],b[colnum,'s']))
print(sprintf("%d:%02d:%02d",b[rownum,'h'],b[rownum,'m'],b[rownum,'s']))

```

There are 660 palindromes.

The maximum time between palindromes is from 15:55:51 to 20:00:02, 14651 seconds
The minimum time between palindromes is from  9:59:59 to 10:00:01, 2 seconds

Palindromes at 1:33:31 and 13:33:31 are precisely 12 hours apart


## Problem 5

    What is the probability of getting a royal flush
    (A-K-QJ- 10 in the same suit in any order)
    in seven-card stud poker played with a 53-card
    deck (standard 52-card deck plus a joker), if
    the joker is wild and can be used to represent
    any card?

    -- F.J. Tydeman, CA Delta'73

Solution:  I wonder if Monte-Carlo is up to the task or if
this takes too many trials to be practical?  We shall see:

```{r}
# Deck of 53 cards
# 1 = Joker
# 1xx = A-K of diamonds
# 2xx = A-K of hearts
# 3xx = A-K of spades
# 4xx = A-K of clubs
cards=c(1,seq(101,113),seq(201,213),seq(301,313),seq(401,413))

ntrials=1000000

zzz=matrix(ncol=7)
set.seed(42)
win=0
winj=0
for (trial in 1:ntrials) {
    s=sample(cards,7)
    # Check for a possible royal flush
    if (1 %in% s) {
        # evaluate with the joker
        for (j in 1:4) {
            hand=s[trunc(s/100)==j] %% 100
            if (sum(c(10,11,12,13,1) %in% hand)>3) {
                winj=winj+1
                zzz=rbind(zzz,s)
            }
        }
    } else {
        for (j in 1:4) {
            hand=s[trunc(s/100)==j] %% 100
            if (all(c(10,11,12,13,1) %in% hand)) {
                win=win+1
                zzz=rbind(zzz,s)
            }
        }
    }
}

print(paste0("Winning hands = ",win+winj))
print(paste0("   Without joker = ",win))
print(paste0("   With    joker = ",winj))
print(paste0("Total   hands = ",ntrials))
print(paste0("Pr(winning)   = ",(win+winj)/ntrials))
```

Results seem reasonable.  Adding the joker
increases the likelihood of a royal flush
by about a factor of 5.

The Monte-Carlo results with 1e6 trials are:

    Seed   Wins   No_Joker  Joker   Pr
    ----------------------------------------
    42     160    23        137     0.000160
    123    176    29        147     0.000176
    2020   186    28        158     0.000186
    1      163    24        139     0.000163

The Monte-Carlo results with 1e7 trials are:

    Seed   Wins   No_Joker  Joker   Pr
    ----------------------------------------
    42     1743   296       1447    0.0001743


The Bent solution is actually not too bad:

    To get a royal flush without the joker:
    
        4 suits
        choose(5,5) ways to get royal flush
        choose(48,2) ways to get other two cards
    
        Total = 4512
    
    To get a royal flush with the joker
    
        4 suits
        choose(5,4) ways to get 4 of the cards in 5 
        1 way to get the joker
        choose(47,2) ways to get the other two cards
        
        Total = 21620
        
        note this is choose(47,2) since you can't draw
        the 5th card of the royal flush?
    
    The total number of hands
    
        choose(53,7) = 154143080
        
    Pr = (4512+21620)/154143080 = 139/819910
       ~ 0.0001695308


    Still not convinced about the choose(47,2).  If you
    use choose(48,2) you get Pr = 0.000175629
    
    This is more in line with the Monte-Carlo result
    with 1e7 trials.
    

## Bonus

    Alice's doctor has prescribed that every
    morning she is to take one each of k different
    pills. The pills all look identical except
    that each medicine type has a unique color.
    In preparation for an n-day trip, Alice put n
    pills from each of her k bottles into one larger
    bottle to save space. As a result Alice, who is
    totally color blind but forgot that fact while
    packing, was forced each morning during the trip
    to withdraw randomly k pills from the bottle and
    take them.  What is the overall probability that
    every day of the trip Alice indeed took exactly
    what her doctor prescribed? Present your answer
    as a function of n and k.

    -- Toby Berger, CT Alpha'62

Solution:

    At the start of the trip there are n*k pills.
    
    On the first day the probability of getting
    exactly the correct k pills is:
    
    p1 = n/(n*k) * n/(n*k-1) * ... * n/(n*k-(k-1))
    
       = n^k /( (n*k)*(n*k-1)*...*(n*k-(k-1)))
       
       = n^k / (choose(n*k,k)*factorial(k))
       
    For the second day we have:
    
    p2 = (n-k)^k / (choose(n*k-k,k)*factorial(k))
    
    etc.
    
    The product of the n terms is the overall probability:
    
    product((n-m)**k/(binomial(n*k-m*k,k)*factorial(k)),(m,0,n))
    
    which sympy does not simplify very much.
    
    
    The Bent is able to simplify this as:
    
    factorial(n)^k * factorial(k)^n / factorial(n*k)
    
    They don't have a real clear explanation of how you get there.
    
    What is this numerically?
    
    What if we have a 10 day trip and 5 different pills:
    
    n=3, k=2  --> Pr = 0.4
    n=3, k=6  --> Pr = 0.00272
    n=4, k=10 --> Pr = 1.35e-8
    ...
    n=10, k=5 --> Pr = 1.281e-11

    Generally Pr->0 really fast!
    

## Double Bonus

    The sequence 29, 149, 269, 389, 509 contains five
    consecutive primes having a common difference of
    120; however, the next member of the sequence
    629 is composite.  In a sequence of positive
    integers having a common difference of 210, what
    is the maximum possible number of consecutive
    terms that are all primes? Give an example of
    such a sequence.

    -- Technology Review

TODO


