---
title: "Bent Winter 2009"
author: "R. G. Cronce"
date: "12/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bent Brain Ticklers, Winter 2009


## Problem 1

    What is the smallest right triangle
    that will fit completely inside another
    right triangle, such that all six sides
    of the two triangles have integral
    values, and the larger triangle has
    one side that is shorter than all three
    sides of the smaller triangle?

    -- Almost Impossible Brain Bafflers
    by Tim Sole and Rod Marshall

Solution:

    May need to generate some right triangles and
    try them.  I seems like the most likely candidate
    for the small triangle would lay on its hypotenuse
    along the lower side of the larger triangle with the
    shortest leg of the smaller triangle nearest the
    shortest leg of the larger triangle.
    
    Assume smaller triangle has sides a<b<c and larger
    triangle has sides A<B<C.
    
    We require that the small triangle lay on side c on
    leg B of the larger triangle.  The height of the
    smaller triangle must fit under the hypotenuse of the
    larger triangle.  This requires:
    
        H = a*sin(theta), theta = acos(a/c)
        
    The height of the larger triangle at this point is:
    
        H2 = A*(B-a*cos(theta))/B
        
    So we need: H2>H and A<a

    Not really sure where to go from here.        

```{r}
# Generate some triangles and make a matrix
zzz=matrix(ncol=3)

for (a in 1:100) {
    for (b in a:1000) {
        for (c in b:1000) {
            if (a^2+b^2 == c^2) {
                zzz=rbind(zzz,c(a,b,c))
            }
        }
    }
}
zzz=zzz[2:nrow(zzz),]
```

We can then look at these lists to find some candidates.  The smallest two
triangles that we can consider are:

    smaller = 6,8,10
    larger  = 5,12,13
    
    If we put the smaller triangle with hypotenuse along the x-axis
    theta = acos(6/10) --> height = 6*sin(acos(6/10)) = 4.8
    
    In theory this might fit inside the 5,12,13 triangle but it actually does not.
    
    Need to look deeper.
    
    Let's try this programmatically
    
```{r}
# arrange the triangles in ascending area
# The length of the sides is increasing across the columns of zzz
# So the columns of zzz are a,b,c or A,B,C
zzz=zzz[order(1/2*zzz[,1]*zzz[,2]),]
# For each triangle search through all the larger triangles for one with a side
# smaller than any of the sides in the current triangle and then check to see
# if the current triangle can fit inside:
done=0
for (i in 1:(nrow(zzz)-1)) {
    for (j in i:nrow(zzz)) {
        if (min(zzz[j,])<min(zzz[i,])) {
            # Check the height of the current triangle
            theta=acos(zzz[i,1]/zzz[i,3])
            h=zzz[i,1]*sin(theta)
            # Check the height under the hypotenuse of triangle j
            # H2 = A*(B-a*cos(theta))/B
            h2=zzz[j,1]*(zzz[j,2]-zzz[i,1]*cos(theta))/zzz[j,2]
            if (h2>h) {
                # We have a winner
                print("Winner")
                print(zzz[i,])
                print("inside")
                print(zzz[j,])
                done=1
                break
            }
        }
    }
    if (done==1) { break }
}

```

A triangle of 12x16x20 will fit inside a triangle of 11x60x61



## Problem 2

    Find the smallest positive integer
    such that, if you place a 4 in front of
    it, you get a number that is exactly
    four times as large as you get if you
    place a 4 at the end of the number.

    -- The Numerology of Dr. Matrix by
    Martin Gardner

Solution:

    Can I do this with exhaustive search or is it
    trickier than that?
    
```{r}
    for (i in 1:1000000) {
        n1=as.numeric(paste0(4,i))
        n2=as.numeric(paste0(i,4))
        if (n1==4*n2) {
            print(i)
            break
        }
    }
```

Apparently exhaustive search works:

    n  = 10256
    
    n1 = 410256  \ n1 = 4*n2
    n2 = 102564  /


## Problem 3

    Al has two 12-hour clocks that,
    when fully wound, will run for nearly
    eight days. Both clocks were keeping
    different times, with each being
    wrong by a different exact number of
    minutes per day, although less than
    one hour each. Al took his clocks to
    the local clock mender, who works
    only from 9:30 a.m. to 5 p.m. Monday
    through Friday. He immediately
    wound both clocks fully and set them
    to the correct time, a whole number
    of minutes after the hour, and put
    them on the shelf for observation.
    The following Monday, as he went
    to take down the clocks to start
    working on them, they started to
    strike eight o'clock simultaneously.
    This occurred some hours and minutes
    past the correct time. What day
    and exact time did the clock mender
    set them originally?

    -- Puzzles, Mathematical Diversions,
    and Brainteasers by Erwin Brecher

TODO


## Problem 4

    Two dominoes are chosen at random
    from a 28-domino, double-six
    set. What is the probability that an
    end-to-end chain of all the dominoes
    can be formed with the two chosen
    dominoes at the ends, with one in the
    first position, and the other in the
    last position? The usual rule applies
    that the numbers on the ends of
    adjoining dominoes must match.

    -- Classical Mathemagic by R. Blum, A.
    Hart-Davis, B. Longe, and D. Niederman

TODO

    Somehow I just don't get excited about
    dominoes.


## Problem 5

    A number of identical planes, each
    with a tank that will hold just enough
    fuel to travel exactly halfway around
    the world, are based on a small island
    on the equator. If the planes can refuel
    only from the island or from another
    plane, what is the smallest number of
    planes that are required for one plane
    to complete an equatorial trip around
    the world, with each plane involved
    returning safely to the island? Assume
    that a plane can refuel and transfer
    fuel instantaneously and that all
    planes travel at the same speed and
    use fuel at the same uniform rate. Ignore
    rotation of the earth and winds.

    -- Almost Impossible Brain Bafflers
    by Tim Sole and Rod Marshall

TODO

    I looked at the solution in the Bent.
    Only three planes are required.  They
    start out together, go 1/8 of the way and
    plane A transfers fuel to B.  B and C
    continue until B has to turn back and
    gives C the excess fuel.  By this time
    A can take off the other direction and
    meet C and get it around the world.
    
    I might have missed one more refueling
    in there -- but you get the idea.


## Bonus

    In the game of Yahtzee, players
    try to achieve specific combinations
    of five dice. A player first tosses
    all five dice and then retosses those
    that don't have the desired values
    and then finally repeats this process
    for a third toss. Suppose you are
    trying to maximize the number of 6s.
    After your tosses, your score is the
    sum of all the 6s thrown. What is the
    expected value of your score? Your
    maximum score is 30 since you stop at
    any time that you have tossed five 6s.

    -- Howard G. McIlvried III, PA Gamma'53

Solution: sounds like Monte-Carlo time

```{r}
ntrials=100000

set.seed(2020)
score=0
for (trial in 1:ntrials) {
    s=sample(6,5,replace=T)
    score=score+6*sum(s==6)
    n=sum(s!=6)
    if (n!=0) {
        s=sample(6,n,replace=T)
        score=score+6*sum(s==6)
        n=sum(s!=6)
        if (n!=0) {
            s=sample(6,n,replace=T)
            score=score+6*sum(s==6)
        }
    }
}

score/ntrials
```

For 1e5 trials the mean score is ~ 12.6

    Seed   Result
    --------------
    42     12.65988
    123    12.62184
    2020   12.61902
    
    The closed form solution from the Bent is
    
    455/36 = 12.6389

    They originally derived it with combinatorics and it was ugly.
    
    They also had a solution from a reader that is quite elegant.
    
    Consider one die.  The probability that you do not get a 6 in
    three rolls is (5/6)^3.  Thus, the probability that you get (at least)
    one 6 is 1-(5/6)^3.  The score for this die is 6*(1-(5/6)^3).
    
    The same logic applies to all 5 dice so the result is:
    
    5*6*(1-(5/6)^3) = 30*91/216 = 455/36 = 12.63889


## Computer Bonus

    Fermat, in about
    1650, asked for all solutions in positive
    integers for x**2 + 2 = y**3. It turns
    out that x = 5, y = 3 is the only solution.
    Now consider x**2 - 15 = y**3. One
    solution in positive integers is x = 4,
    y = 1. Find all other solutions.

    -- Don A. Dechman, TX A'57

Solution:

    Performing exhaustive search with sympy.
    Up to y=1e9 the solutions are:
    
    x        y
    ----------------
    4        1
    1138     109
    
    The Bent notes these as the only solutions but
    provides no explanation as to how these are the ONLY
    solutions.  Yes - it is easy to generate these numbers
    but how do we know there are no other solutions?
