---
title: 'Le Monde 1120: Token passing game'
author: "R. G. Cronce"
date: "10/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Le Monde 1120

[R-Bloggers Website](https://www.r-bloggers.com/2020/01/le-monde-puzzle-1120)

[Xi'an Blog Website](https://xianblog.wordpress.com/2020/01/14/le-monde-puzzle-1120/)

Another Xi'an special.

    A board game as Le weekly Monde current mathematical puzzle:

    11 players in a circle and 365 tokens first owned by a single player.
    Players with at least two tokens can either remove one token and give
    another one left or move two right and one left. How quickly does the
    game stall, how many tokens are left, and where are they?

As usual, Xi'an only provides a partial solution:

```{r}

od=function(i)(i-1)%%11+1

muv<-function(bob){
  if (max(bob)>1){
    i=sample(rep((1:11)[bob>1],2),1)
    dud=c(0,-2,1)
    if((runif(1)<.5)&(bob[i]>2))dud=c(2,-3,1)
    bob[c(od(i+10),i,od(i+1))]=bob[c(od(i+10),i,od(i+1))]+dud
  }
  bob}

```

Expanding this to actually solve the problem and investigate the number of
moves in a typical game:

```{r}
library(dplyr)

num_games=1000
start_config=rep(0,11); start_config[1]=365

num_steps=c()
solution=c()
for (i in 1:num_games) {
  bob=start_config
  n=0
  while(max(bob)>1) { n=n+1; bob=muv(bob) }
  num_steps=append(num_steps,n)
  solution=rbind(solution,bob)
}
solution=distinct(as.data.frame(solution))

print("Starting Configuration:")
print(start_config)
print(c("Number of distinct solutions:",nrow(solution)))
print("Distinct solutions:")
print(as.matrix(solution))
print("Number of steps to find solution:")
summary(num_steps)

```

To duplicate the histogram from the article:

```{r}
hist(num_steps,freq=FALSE)
```