---
title: "Buckets"
author: "R. G. Cronce"
date: "10/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

Originally seen on R-bloggers from [Xian's Blog](https://xianblog.wordpress.com/2020/07/01/le-monde-puzzle-1149/)

[R-Bloggers Website](https://www.r-bloggers.com/2020/06/le-monde-puzzle-1149)

The code he provided, and the example he cited, do not make sense as written
but I have tweaked them to sort of solve the problem.

This is a simulation of the classic water bucket transfer problem with a little twist.

> Three buckets have capacities of 8, 5 and 3 litres, respectively.
> At the start of the game, the 8 litre bucket is full and both others are empty.
> Aiming at reaching exactly 4 litres in one bucket, water is transferred between buckets.
> However, a fraction 1/k is lost with each transfer.
> If k=9, it is possible to reach 4 litres in three operations?
> If k=7, is it at all possible to reach 4 litres?

Note that he uses k=1/n instead of k=n below.  The only thing he provided was the k, z, and
m (function) definitions.  I don't know how he suggests these alone solve the problem.  The
function provided makes a single transfer between two randomly selected buckets.  You
need to add some additional logic to itereate this and look for optimal solutions.

In addition, he gives a sample solution for k=9 (I think this means 1/9):

    8.000    0    0
    2.375    5    0
    0.000    5    2.11
    0.000    4    3

I tried several hundred simulations and I never got anything this "good".  In addition,
the code for k=9 (1/9) generates 2.444 as the total in line 2.  If I change the value
to k=8 (1/8) then I get 2.375 in line 2 but the rest of the totals are wrong.  I need
to figure out if his code is any good.

If k=9 (e.g. you lose 1/9 of the fluid during the transfer) then pouring from the 8 liter
to the 5 liter bucket should spill an amount x given by:

    (5+x)*8/9 = 5
    (5+x)*8 = 5*9
    8*x = 5
    x = 5/8
    
So - this means the first pour should put 5 liters in the second bucket and take 5.625 liters
out of the first bucket which should leave 2.375 (as given in his table).

The code, however, makes a calculation of:

    s[i]=s[i]+ifelse(
      rep( ( a <- z[i[1]]-s[i[1]] ) < (b <- s[i[2]])*(1-k),2),
      a*c(1,-1-k),b*c(1-k,-1))

which is utterly incomprehensible!!!!!  and I think it is wrong.  Here is what I think
it says:

    Given i = c(n1,n2) we want to transfer liquid from bucket n2 to n1
    a = space remaining in bucket n1
    b = amount of liquid coming to bucket n1 if all of bucket n2 is transferred
    
    if (space remaining < space required) then
        transfer enough to fill remaining space in n (a)
        remove liquid with spillage (a*(1+k))  <-- this is the error here *******
    else
        transfer remaining liquid (with spillage) = b*(1-k)
        remove all of liquid from the other bucket
    end if
    
I think there is an arithmetic error.  If you want to transfer an amount, a, to a bucket
you need to actually remove a/(1-k) such than a/(1-k)*(1-k) == a.  The equation used above
is a*(1+k) instead of a/(1-k).

I modified the code and now it works properly (and reproduces the data in the table).
Unfortunately, the code no longer seems to find solutions for some other values of k.

I don't know if these are really no solution cases or if this is a problem with floating
point arithmetic.  Note that the *solver* looks for equality of a floating point number
and should probably be rewritten to allow for a nonzero error.

Looks like it was round-off error.  I modified the solver to look for error of 1e-3 and it
now returns solutions.  I later made the error 0.0001 to eliminate spurious solution for k=1/17.
I suppose I could use rational arithmetic and get rid of all the round off errors but this should
be good enough for now.  Here are the shortest solution strings for k=1/2 through k=1/20.
The only value without any solution appears to be k=3 (1/3).

    k=1/20 - solution in 13 steps
    k=1/19 - solution in 8 steps
    k=1/18 - solution in 12 steps
    k=1/17 - solution in 11 steps
    k=1/16 - solution in 6 steps
    k=1/15 - solution in 15 steps
    k=1/14 - solution in 9 steps
    k=1/13 - solution in 4 steps
    k=1/12 - solution in 5 steps
    k=1/11 - solution in 4 steps
    k=1/10 - solution in 9 steps
    k=1/9  - solution in 3 steps
    k=1/8  - solution in 2 steps
    k=1/7  - solution in 6 steps
    k=1/6  - solution in 8 steps
    k=1/5  - solution in 3 steps
    k=1/4  - solution in 1 step
    k=1/3  - no solution
    k=1/2  - solution in 1 step


```{r}
k=1/9
z=c(8,5,3)

# This was the code from the web site
# This has a logic/arithmetic error -- see above for details
#
### m<-function(s){
###   i=sample(1:3,2)
###   s[i]=s[i]+ifelse(
###     rep((a<-z[i[1]]-s[i[1]])<(b<-s[i[2]])*(1-k),2),
###     a*c(1,-1-k),b*c(1-k,-1))
###   s}

# Make one bucket exchange.  It is not smart, however, and may
# attempt to transfer from an empty bucket - thus the returned
# configuration may be the same as on entry.  Need to check for
# that in the calling function.
m<-function(s){
  i=sample(1:3,2)
  s[i]=s[i]+ifelse(
    rep((a<-z[i[1]]-s[i[1]])<(b<-s[i[2]])*(1-k),2),
    a*c(1,-1/(1-k)),b*c(1-k,-1))
  s}

# This will find a valid solution.  If the bucket total falls
# below 4 liters it will restart.
solve_it <- function() {
  s=c(8,0,0)
  zzz=c()
  zzz=rbind(zzz,s)
  while (all(abs(s-4)>0.0001)) {
    if (sum(s)<4) {
      s=c(8,0,0)
      zzz=c()
      zzz=rbind(zzz,s)
    }
    s2=m(s)
    while(all(s2==s)) { s2=m(s2) }
    s=s2
    zzz=rbind(zzz,s)
  }
  zzz
}

# This does not actually find a min
# It simply runs 100 loops of the solve_it() function and returns
# the best result
find_min <- function() {
  best=solve_it()
  zzz=solve_it()
  for (i in 1:100) {
    zzz=solve_it()
    if (nrow(zzz)<nrow(best)) { best=zzz; print(nrow(best)) }
  }
  best
}

b=find_min()
print("Solution for k=1/9:")
print(nrow(b))
print(b)
print("Set k=1/n and run b=find_min() to look for other solutions")
print("Note that k=1/3 has no solutions")

```

