---
title: "Cryptic Arithmetic"
author: "R. G. Cronce"
date: "10/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Le Monde 1133

Another one of Xi'an's solutions to a Le Monde puzzle.  The statement on R-bloggers
is wrong (it might a spell check issue) but it is correct on Xi'an blog site.

[R-Bloggers (erroneous)](https://www.r-bloggers.com/2020/03/le-monde-puzzle-1133)

[Xi'an Blog](https://xianblog.wordpress.com/2020/03/28/le-monde-puzzle-1133/)

This is his approach to solving a cryptic arithmetic problem:

    ADULE - ELUDA = POINT
    
    With each character a different digit.  No value starts with a zero.
    
    What is the largest possible value of POINT?

The problem also looks for a solution to:

    ADULE + ELUDA = POINT

The code he provides dumps all the feasible solutions -- it does not find the
largest value of POINT.  I added the code to just dump the final solution.

Note -- this code does not explore all the permutations -- it is done with sampling.

    
```{r}
library(dplyr)

num_trials=1e6

dify<-function(adule) 
  (sum(adule*10^(4:0))-sum(rev(adule)*10^(4:0)))

num2dig<-function(dif) (dif%/%10^(0:4))%%10

sl=NULL
for (t in 1:num_trials){
  adule=sample(0:9,5)
  while((dify(adule)<=0)||(!prod(adule[c(1,5)])))
     adule=sample(0:9,5)
point=rev(num2dig(dify(adule)))
if ((!sum(duplicated(point)))&(prod(point%in%(0:9)[-adule-1])))
  sl=rbind(sl,c(adule,point))}
sl=as.matrix(distinct(as.data.frame(sl),.keep_all = TRUE))

# Added by RGC
k=which.max(apply(sl,1,function(x) {sum(x[6:10]*(10^(4:0)))} ))
print(sl[k,])
adule = sum(sl[k,1:5]*(10^(4:0)))
eluda = sum(sl[k,5:1]*(10^(4:0)))
point = sum(sl[k,6:10]*(10^(4:0)))
print(sprintf("%d - %d = %d",adule,eluda,point))

```

Solving the second half of the problem involves changing dify() to calculate the
sum instead of the difference.  This has been done below:


```{r}
library(dplyr)

dify<-function(adule) 
  (sum(adule*10^(4:0))+sum(rev(adule)*10^(4:0)))

num2dig<-function(dif) (dif%/%10^(0:4))%%10

sl=NULL
for (t in 1:num_trials){
  adule=sample(0:9,5)
  while((dify(adule)<=0)||(!prod(adule[c(1,5)])))
     adule=sample(0:9,5)
point=rev(num2dig(dify(adule)))
if ((!sum(duplicated(point)))&(prod(point%in%(0:9)[-adule-1])))
  sl=rbind(sl,c(adule,point))}
sl=as.matrix(distinct(as.data.frame(sl),.keep_all = TRUE))

# Added by RGC
k=which.max(apply(sl,1,function(x) {sum(x[6:10]*(10^(4:0)))} ))
print(sl[k,])
adule = sum(sl[k,1:5]*(10^(4:0)))
eluda = sum(sl[k,5:1]*(10^(4:0)))
point = sum(sl[k,6:10]*(10^(4:0)))
print(sprintf("%d + %d = %d",adule,eluda,point))

```

Note that if you don't try enough random samples you may not get the solution(s).
I tried num_trials = 1e2, 1e3, 1e4 and I get progressively more correct
solutions but even with 1e4 the solution is not always correct.

I left num_trials at 1e6 for the values above.

I have confirmed the solutions above with permutations.

