---
title: "LeMonde #1053: Coin Denominations"
output: html_notebook
---

[Website](http://www.statsblogs.com/2018/06/20/le-monde-puzzle-1053/ )


# Overview

1. If coins come in units of 1, x, and y, what is the optimal value of
   (x,y) that minimises the number of coins representing an arbitrary
   price between 1 and 149? 

2. If the number of units is now four, what is the optimal choice?

```{r}
library(tidyverse)
```


# Problem 1

From the original author:

    The first question is fairly easy to code

```{r}
coinz <- function(x,y){
  z=(1:149)
  if (y<x){xx=x;x=y;y=xx}
  ny=z%/%y
  nx=(z%%y)%/%x
  no=z-ny*y-nx*x
  return(max(no+nx+ny))
}

```

    and returns M=12 as the maximal number of coins, corresponding to x=4       and y=22. And a price tag of 129.

----

While the statement from the author is true (based on the algorithm he
coded here), there are a large number of combinations that generate the
same *minimal* number of coins.  In addition, the price that requires
12 coins (129) is not unique.  In fact, the patterns make for rather
interesting plots.

Having said all that, the author notes down below that the algorithm
above is not optimal.  There are ways to use fewer coins than using
the remainder division above. (See the cos() function near the end).

* Why is this not optimal?

I can easily demonstrate that it is not optimal.  Consider a case of
attempting to reach a price of 8 using coins of values 1, 4, and 5.
Using the algorithm above would yield:

    1 coin = 5
    0 coins = 4
    3 coins = 1
    4 coins total
    
Versus using 2 of the coins with a value of 4.  In general, how do we
ensure we are getting a minimal set?  More of that at the end of this
notebook.

----

Going back to the algorithm used above:

I think I have seen this problem before where the additional constraint
was to minimize the value of x+y.  This results in a unique coin
combination but still has two purchase prices with the same number
of coins.  In the case of x=4, y=22 the prices are 129 and 147.

There are 39 combinations of coin values that result in the minimum
coin count of 12.  Some of these coin combinations have a single
maximum, 2 maxima, and 3 maxima.  These combinations are enumerated
below.  First, all results.  Second, the best worst-case performance.
Finally, the number of maxima for each of the best results.

```{r}
coinz2 <- function(x,y){
  z=(1:149)
  if (y<x){xx=x;x=y;y=xx}
  ny=z%/%y
  nx=(z%%y)%/%x
  no=z-ny*y-nx*x
  return(data.frame(z=z,count=no+nx+ny))
}

m=c()
for (x in 2:148) {
  for (y in x:149) {
    s=coinz2(x,y)
    m=rbind(m,c(x,y,max(s$count)))
  }
}
m=data.frame(m)
names(m)=c("x","y","worst")
m
```

```{r}
mbest <- m[m$worst==min(m$worst),]
mbest
```

Even for a given combination of coins, there may be multiple different
purchase totals that result in the maximum number of coins.

For example, with x=4, y=22 we have:

```{r}
s=coinz2(4,22)
which(s$count==max(s$count))
```

Both of these totals will require 12 coins.  The number of coins
versus price is shown below:

```{r}
plot(s$z,s$count)
```

We can determine the number of maxima for each coin configuration
as shown below:

```{r}
nmax=c()
for (i in 1:nrow(mbest)) {
  x=mbest[i,]$x
  y=mbest[i,]$y
  s=coinz2(x,y)
  nmax=c(nmax,length(which(s$count==max(s$count))))
}
mbest=cbind(mbest,nmax)
mbest %>% group_by(nmax) %>% summarize(n=n())
```
 
 Combinations with 3 maxima:
 
```{r}
mbest[mbest$nmax==3,]
```

```{r}
mm <- mbest[mbest$nmax==3,]
for (i in seq(nrow(mm))) {
  s=coinz2(mm[i,'x'],mm[i,'y'])
  plot(s$z,s$count)
}
```

Which prices are the hardest to reach?

```{r}
print('Prices requiring most coins:')
for (i in seq(nrow(mm))) {
  s <- coinz2(mm[i,'x'],mm[i,'y'])
  z <- s$z[s$count==12]
  print(c("Coin Values:",1,mm[i,'x'],mm[i,'y']))
  print(z)
}
```


# Problem 2

From the original author:

    For the second question, one unit is necessarily 1 (!) and there is
    just an extra loop to the above, which returns M=8, with other units
    taking several possible values:

    [1] 40 11  3
    [1] 41 11  3
    [1] 55 15  4
    [1] 56 15  4

----

Similar to the code above (using the non-optimal algorithm) we can make
the nested loops and find the *optimal* combinations:

```{r}
# From above with three coins
# Note: we know this isn't optimal - but want to reproduce
# the results as presented in the Web posting.
coins <- function(w,x,y) {
  # We require 1<w<x<y
  s <- c(w,x,y)
  s <- sort(s)
  w=s[1]; x=s[2]; y=s[3]
  z=1:149
  ny <- z %/% y
  nx <- (z%%y) %/% x
  nw <- ((z%%y) %% x) %/% w
  no <- z-ny*y-nx*x-nw*w
  return(max(no+nw+nx+ny))
}

mbest=999
mworst=0
for (w in 2:147) {
  for (x in w:148) {
    for (y in x:149) {
      m=coins(w,x,y)
      if (m<mbest) { mbest = m }
      if (m>mworst) { mworst = m }
    }
  }
}
c(mbest,mworst)
```

Let's find the coin combinations that yield the optimal result:

```{r}
moptimal=c()
for (w in 2:147) {
  for (x in w:148) {
    for (y in x:149) {
      m=coins(w,x,y)
      if (m==mbest) {
        moptimal=rbind(moptimal,c(w,x,y))
      }
    }
  }
}
moptimal=data.frame(moptimal)
names(moptimal)=c('w','x','y')
moptimal
```

The author adds:

A quick search revealed that this problem (or a variant) is solved in
many places, from stackexchange (for an average-why average?, as it does
not make sense when looking at real prices-number of coins, rather than
maximal), to a paper by Shalit calling for the 18¢ coin, to Freakonomics,
to Wikipedia, although this is about finding the minimum number of
coins summing up to a given value, using fixed currency denominations
(a knapsack problem). This Wikipedia page made me realise that my
solution is not necessarily optimal, as I use the remainders from the
larger denominations in my code, while there may be more efficient
divisions. For instance, running the following dynamic programming code


```{r}
coz=function(x,y){
  minco=1:149
  if (x<y){ xx=x;x=y;y=xx}
  for (i in 2:149){
    if (i%%x==0)
      minco[i]=i%/%x
    if (i%%y==0)
      minco[i]=min(minco[i],i%/%y)
    for (j in 1:max(1,trunc((i+1)/2)))
          minco[i]=min(minco[i],minco[j]+minco[i-j])
      }
  return(max(minco))}
```

returns the lower value of M=11 (with x=7,y=23) in the first case and M=7
in the second one.

Test this out below:

```{r}
mbest=999
for (x in 2:148) {
  for (y in x:149) {
    m<-coz(x,y)
    if (m<mbest) { mbest=m }
  }
}
mbest
```

So this code is more efficient than the original strategy of removing
the maximum number of maximum denomination coins.  The first strategy
resulted in a minimum of 12 coins, here we achieve the result with 11
coins.

```{r}
moptimal=c()
for (x in 2:148) {
  for (y in x:149) {
    m<-coz(x,y)
    if (m==mbest) { moptimal<-rbind(moptimal,c(x,y)) }
  }
}
moptimal<-data.frame(moptimal)
names(moptimal)<-c('x','y')
moptimal
```

There are 28 combinations that have an optimal coverage.

* How do we prove this works?

- If the price is an exact multiple of the largest denomination coin
  then the min coin count is the number of these coins
- If the price is an exact multiple of the second largest coin then
  the min coin count might be the number of these coins
- Find the smallest number of coins based on the sum of each pair
  of prices that sum up to the target price.

Here is a commented version of the code with storage of the actual
coin combination that creates each total.

```{r}
# Determine the minimum number of coins required to achieve any
# amount in the range [1,149].  The coins are defined to have the
# following denominations:
#
#    1 < y < x
#
# The return value is a data frame consisting of the following
# columns:
#
#    index = total value
#    minco = minimum number of coins required
#
#    One combination of minco coins with the correct total
#    is represented by the final three columns
#
#    no = number of coins with value = 1
#    ny = number of coins with value = y
#    nx = number of coins with value = x
#
# Note that there is no guarantee that the no,ny,nx combination
# is unique.  The only guarantee is that minco gives the smallest
# number of coins.
coz2=function(x,y) {
  minco=1:149
  no=1:149
  nx=rep(0,149)
  ny=rep(0,149)
  no[1]=1
  if (x<y){ xx=x;x=y;y=xx }
  for (i in 2:149) {
    if (i%%x==0) {
      minco[i]=i%/%x
      nx[i]=i %/% x
      ny[i]=0
      no[i]=0
    }
    if (i%%y==0) {
      if ( (i%/%y)<minco[i] ) {
        minco[i] = i %/% y
        nx[i] = 0
        ny[i] = i %/% y
        no[i] = 0
      }
    }
    for (j in 1:max(1,trunc((i+1)/2))) {
      if ( (minco[j]+minco[i-j]) < minco[i]) {
        minco[i]=minco[j]+minco[i-j]
        nx[i]=nx[j]+nx[i-j]
        ny[i]=ny[j]+ny[i-j]
        no[i]=no[j]+no[i-j]
      }
    }
  }
  d=data.frame(minco=minco,no=no,ny=ny,nx=nx)
  return(d)
}

nmax=c()
for (i in seq(nrow(moptimal))) {
  d=coz2(moptimal[i,'x'],moptimal[i,'y'])
  nmax=c(nmax,sum(d$minco==max(d$minco)))
}

which(nmax==max(nmax))
```

