---
title: "Fun with Primes"
author: "R. G. Cronce"
date: "10/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## LeMonde 1139

Another one of Xi'an blog gems with a very terse solution.

[R-bloggers website](https://www.r-bloggers.com/2020/04/le-monde-puzzle-1139)

[Xi'an blog](https://xianblog.wordpress.com/2020/04/24/le-monde-puzzle-1139/)

The problem involves finding numbers where the difference in divisors are also
divisors of the given number.  Specifically:

    The integer n=36 enjoys the property that all the differences between its ordered
    divisors are also divisors of 36. Find the only 18≤m≤100 that enjoys this property
    such that all its prime dividers are of multiplicity one. Are there other such m’s?

The code is given below:

```{r}
library(primes)

y=z=18:2000
for(x in y)z[y==x]=!sum(x%%diff((1:x)[!x%%(1:x)]))
y=y[z==1]
for(k in generate_primes(2,max(y)))y=y[!!y%%k^2]

y
```

At least up to 100K, the only solutions are 42 and 1806.  The problem only asked to check the
range 18 to 100 which only has one solution (42) but this has been checked 1 to 100000.  I modified
the upper limit to 2K for the saved file since it runs much faster.

