---
title: "Virus Simulation"
author: "R. G. Cronce"
date: "10/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Simulation of a Virus

LeMonde #1141

Another concoction from Xi'an - in honor of Conway's passing and the Game of Life.
This is a simulation of a virus on an 8x8 board.  Assuming two or more sick neighbors
will infect a cell - what is the minimum number of infected cells that results in
the entire board being infected?

[R-Blogger's Website](https://www.r-bloggers.com/2020/05/le-monde-puzzle-1141)

[Xi'ans Blog](https://xianblog.wordpress.com/2020/05/04/le-monde-puzzle-1141/)

Code was taken from tio.run (Try It Online) that was linked from his article.  The code is
actually in the URL and (apparently) gets loaded into tio when you visit the site.

[Code](https://tio.run/#%23dZLtboMgFIb/cxX0hwtMycqyD2NLb6QxS7diRqJoKK5LjNfuDvgxdZ2Jyjnv6@ObA6brivIL7xnOav1hVanJe2nOtEEYK30WhCcvTxQKLZXY3kNrEOA@bpw17cWjb2zgGQRxmgonLXshX1oJZ71CwX67HcIH/TXDzQxsjYQX4zR48BH@6Ydz1kxg8Qr2ymbyKuRaW0BXYujBk2GgpAcOSI4xaEba2gyTh7JFCGUqz29vi5bXN1cI2Ljpk@unyiUm5FIXZDTQPWzd3bJ1cJVfUdpgNI5WjIbdX3r7G3DJEgL4Tka2rGCNstJgYuF4YJ5w@ezT9qyTNeqbbKM4iuk4i8upqHLpD1g0rB8TIEWcUhiMO2cZJm4OQ@DG/WbKz/jOOrZPXRmlLQEd4rSo634A)

I copy/pasted the code below so I can *try* to understand it.

I infer from the results that 7 cells are required - but the code does not dump the
actual board configuration.  That is available in `tord`.

After looking at the code it appears that the number printed is one less than the
number of infected cells.  So - I guess the answer is really 8 cells.

After a cursory examination of the code I must admit that the arithmetic in the mov() function
is an amazing tour-de-force of index manipulation.  The `!!` essentially changes the scope
so `!!ind%8` just generates a vector that can be used to index `ind`.  If you try to use
`ind[ind%%8]` you don't get what you expect.  If you use `ind[!!ind%%8]` it generates the
values from 1:64 skipping all the multiples of 8.  This is done to generate the index
values of the left (top?) neighbor.  In a similar way he adds together the values from
the bord in the left/right/top/bottom positions and accumulates them in a matrix nei[].
He then updates any cells where the nei[] value exceeds 1.

This is definitely not readable - but it seems to work.  Note that `%/%` is integer
division.  It also appears that any index values of 0 just disappear from the returned vectors
but I didn't work it all out in detail.

My first thought on solving this was to use 2D convolution but it appears that is
not built into R so I ended up installing a bunch of stuff just to get 2D convolution.
The filter2() function from the EBImage package seems to work.  I also looked at the
magick package - but it doesn't work for a matrix input - only a matrix kernel.  I
also looked at the Imagine package - which works - but the convolution routine does
not handle edge conditions and marks all the edges as NA.  You need to manually pad
the input matrix with the boundary value, run the convolution, then extract the
center portion of the matrix.  Ugly!  Might want to uninstall magick and Imagine but
they are installed for now.

In any case, once you get a 2D convolution routine you can just run a kernel to add
up the four neighbors and use some ifelse logic.  The result is fairly clean.
Assume zzz=board, k=convolution matrix, then:

    k=matrix(c(0,1,0,1,0,1,0,1,0),3,3)                 # kernel for the four neighbors
    ...
    zzz <- ifelse(filter2(zzz,k,boundary=0)>1,1,zzz)   # New virus value
    ...

Will simulate the next generation of the virus (e.g. the mov() function).  Keep iterating
until the number of cells reaches 64 or it stops increasing -- same as the logic below.

Note that this would be trivial to update if the threshold was three neighbors but I don't
know if the logic below would be so simple.

```{r}

mov <- function(bord){
  ind=(1:64)
  nei=0*ind
  ind=ind[!bord]
  nei[ind[!!ind%%8]]=bord[ind[!!ind%%8]+1]
  nei[ind[!!(1-ind%%8)]]=nei[ind[!!(1-ind%%8)]]+
        bord[ind[!!(1-ind%%8)]-1]
  nei[ind[!!(ind-1)%/%8]]=nei[ind[!!(ind-1)%/%8]]+bord[ind[!!(ind-1)%/%8]-8]
  nei[ind[!!(7-(ind-1)%/%8)]]=nei[ind[!!(7-(ind-1)%/%8)]]+bord[ind[!!(7-(ind-1)%/%8)]+8]
  bord[ind[nei[ind]>1]]=1  
  return(bord)
  }


fill <- function(bord){
  new_bord=mov(bord)
  while ((sum(new_bord)<64)&(sum(new_bord)>sum(bord))){ 
    bord=new_bord;new_bord=mov(bord)}
  return((sum(new_bord)==64))}
 
top=64
for (t in 1:1e5){
  bord=matrix(0,8,8)
  bord[sample(1:64,sample(2:top,1))]=1
  if (fill(bord)){top=sum(bord)-1;tord=bord;print(top)}
}



```

