---
title: "Lemonde 1020: Liars and Truthtellers"
output: 
  html_notebook:
    toc: TRUE
---


# Overview

Three logic problems are presented that employ a circle of liars and
truth-tellers.  For each problem, the responses of the members of the circle
may be used to determine the maximum number of liars in the group.  The
solutions presented here attempt to enumerate all possible sequences of
members.

For more details see the [original Web posting](http://www.statsblogs.com/2017/09/14/le-monde-puzzle-1020/).


## Problem Statements

1. A circle of 16 liars and truth-tellers is such that everyone states that
their immediate neighbours are both liars. How many liars can there be? 

2. A circle of 12 liars and truth-tellers is such that everyone state that
their immediate neighbours are one liar plus one truth-teller. How many liars
can there be?

3. A circle of 8 liars and truth-tellers is such that four state that their
immediate neighbours are one liar plus one truth-teller and four state that
their immediate neighbours are both liars . How many liars can there be?

## Purpose of this Notebook

This notebook is as much about trying to understand the author's solutions as
it is about solving the puzzle(s).  I find the code to be cryptic and there is
no explanation of the algorithm or comments in the code to help.  On the other
hand, there are some very interesting coding techniques used by the author and
the general approach of using Monte Carlo to solve problems rather than
exhaustive enumeration is intriguing.

## Observations and Conclusions

After spending some time with the code and these problems I make the
following conclusions:

* Using Monte Carlo simulations to solve deterministic problems is
  interesting.  I would never have thought of doing that.  I am still
  struggling with understanding how many trials are required to ensure
  the solution space is fully explored - but there are times when the
  randomized solution is far more compact than exhaustive search.

* The indexing and vectorization of the calculations is actually very
  impressive (but terse).  I really liked the ability of indexing with embedded
  logic conditions to nearly eliminate the need for if-else testing.  The tour
  de force is the author's solution to problem 3.

  Of particular note are the following:

  1. Wrap-around logic for (1-based) vector indexing.  Given an item
     as vec[i], the previous/next item in the vector are found as shown
     below.  This allows vectorization of expressions and extremely
     compact code.

    * Predecessor logic: `vec[i-1+veclen*(i==1)]`

    * Successor logic: `vec[i%%veclen+1]`

  2. Multiplication/Addition of logic vectors in place of if-then-else.
     See the original author's code for Problem 3 for an example of
     this.  Combined with the wrap-around logic this yields single
     line vectorized expressions in place of multiple, nested if tests.
  
* I found mistakes in the code for two of the author's solutions.  The
  stated solutions were correct in all cases, but the code did not generate
  the correct results in problem 3 and in problem 2 it rarely found one
  of the solutions.  I don't know if this was a publication/typo issue
  or something else.  (To be fair, I also had a mistake in my exhaustive
  search code for problem 3 but I quickly found/fixed it.)
  
  Deeper question: the author's code is very terse.  Does this yield
  higher defect rates?  Or am I over-analyzing this?


# Problem #1

Repeating the problem statement

    1. A circle of 16 liars and truth-tellers is such that everyone states that
       their immediate neighbours are both liars. How many liars can there be? 

Per the original author:

    These questions can easily be solved by brute force simulation. For the
    first setting, using 1 to code truth-tellers and -1 liars, I simulate
    acceptable configurations as (see code below)

With a cryptic conclusion:

    which produces 8, 9, and 10 as possible (and obvious) values.

## Author's Code for Problem #1

The author's original code is given here:

```{r}
# Original code
tabz=rep(0,16)
tabz[1]=1 #at least one
tabz[2]=tabz[16]=-1
for (i in 3:15){
  if (tabz[i-1]==1){
   tabz[i]=-1}else{
   if (tabz[i+1]==-1){
    tabz[i]=1}else{
    if (tabz[i+1]==1){
     tabz[i]=-1}else{
     if (tabz[i-2]==-1){
      tabz[i]=1}else{
       tabz[i]=sample(c(-1,1),1)
}}}}}
# Added by RGC
print(c('Num Liars =',sum(tabz==-1)))
tabz
```

I don't understand the *conclusion* from above.  Why is the solution set
considered as *obvious*?

The original code has no output and only runs one trial to generate a single
solution in tabz[].  The code is not deterministic so you have to run it
multiple times to cover the solution space.  Running some trials manually, I
could only get it to produce 10 and 9 as the possible number of liars.  It is
possible I didn't try hard enough since the case of 8 liars is rare.

* How does this code cover the solution space and verify the results?

I added a loop and I can get it to generate 8, 9, or 10 with varying
probabilities.  The comments and format changes were added during an analysis
of the code but the source itself should be identical to the code from the
author.

```{r}
# Author solution wrapped in a loop to try many
# iterations.  Also changed the indentation to
# make the structure more obvious.
#
# Call
#    tabz[i-1] = left neighbor - always known
#    tabz[i+1] = right neighbor - might be unknown

num_liars <- rep(0,16)
num_trials <- 100000
for (trial in 1:num_trials) {
  tabz=rep(0,16)
  tabz[1]=1 #at least one
  tabz[2]=tabz[16]=-1
  for (i in 3:15){
    if (tabz[i-1]==1) {         # Left neighbor tells truth
      tabz[i]=-1
    } else {                    # Left neighbor is liar
      if (tabz[i+1]==-1) {      # Right neighbor is liar
        tabz[i]=1
      } else {                  # Right neighbor tells truth or undefined
        if (tabz[i+1]==1) {     # Right neighbor tells truth
          tabz[i]=-1
          print("Should never print")
          print(tabz)
          break
        } else {                # Right neighbor undefined
          if (tabz[i-2]==-1) {  # This prevents three liars in a row
            tabz[i]=1
          } else {              # No rule applies - randomly pick veracity
            tabz[i]=sample(c(-1,1),1)
          }
        }
      }
    }
  }
  m <- sum(tabz==-1)   # Added by RGC
  num_liars[m] <- num_liars[m]+1
}
print(c('Total trials =',num_trials))
cbind(num_liars=1:16,solution_count=num_liars,probability=num_liars/num_trials)
```

I don't know how many trials to run.  If I use 10 trials I rarely get
the 8 liar solution.  If I use 100 trials I get the 8 liar solution most
of the time.  If I use 100K trials I see all three solutions all the
time.

I don't know how to prove that this is the complete solution set.

----

I can get a lower and upper bound on the possible solutions from the
constraints (more details below):  (1) Every truth-teller has two liars for
neighbors, and (2) can never have three liars in a row since that would make
the middle liar a truth-teller.

Using these we can find the maximum number of truth-tellers as:

    L-T-L-T-L-T-L-T-L-T-L-T-L-T-L-T-wrap back around to start

This is the lower bound on number of liars (8).

With the constraint on liars we get and maximum of liars with:

    L-T-L-L-T-L-L-T-L-L-T-L-L-T-L-L-wrap back around to start

But this has (circularly) too many liars in sequence.  We can't move any
of the L or T locations without getting three liars in a row so the actual
upper bound must be one less: the maximum number of liars is 10.

**Conclusion:** Number of liars is between 8 and 10 (inclusive).

As a note: this hasn't demonstrated 9 as a feasible solution but it is
trivial to modify the maximum liar sequence to add one more truth
teller to achieve 9 liars.

After thinking about the problem for a while - it is apparent that any
feasible solution needs to satisfy the following two conditions:

* Rule 1: Truth-teller must be surrounded by two liars

* Rule 2: Can't have three liars in a row

Consider traversing the circle.  These rules imply that only certain
sequences are valid.  If we look at two-member groups, the next
member in sequence is defined as follows:

* L-L: next member is T (followed by L)
* L-T: next member is L
* T-L: next member is random

Note that we only need consider three cases.  T-T is illegal since it
would make a truth-teller a liar.

We can further simplify this by looking two members in the past.  Given
the current member is i, then:

* if tabz[i-2] is Liar then tabz[i] is opposite of tabz[i-1]
* if tabz[i-2] is Truth-teller then tabz[i] is random

The final condition is continuity of the circle at the end:

* L-?-L: center member must be T
* T-?-L: center member must be L

In other words, when we reach the last member, it must be the opposite
of the previous member of the group.

So it seems that the sequence can be constructed in a partially
deterministic way.

* Starting with L-T-L for index 16, 1, and 2
* Generate members for index 3-14 using tabz[i-2]
* Generate member for index 15 using tabz[i-1]

## RGC rewrite of Author's Code for Problem #1

```{r}
# Use Booleans instead of +1/-1
gen_sequence <- function() {
  liar <- FALSE
  truther <- TRUE
  tabz <- rep(liar,16)
  tabz[1] <- truther
  for (i in 3:14) {
    if (tabz[i-2]) {
      tabz[i] <- sample(c(liar,truther),1)
    } else {
      tabz[i] <- !tabz[i-1]
    }
  }
  tabz[15] <- !tabz[14]
  return(tabz)
}

run_simulation <- function(num_trials=100000) {
  num_liars <- rep(0,16)
  for (i in 1:num_trials) {
    s <- 16-sum(gen_sequence())
    num_liars[s] <- num_liars[s]+1
  }
  return(cbind(num_liars=1:16,solutions=num_liars,p=num_liars/num_trials))
}

run_simulation(100000)
```

The author's solution is coded in a slightly different way.  The code
starts with L-T-L and then builds the members of the circle sequentially
by following these rules:

1. If the previous member is a truth-teller then the current member must be a
   liar. (Rule 1)

2. If the previous member is a liar and the following member is a liar then the
   current member must be a truth-teller.  This condition ensures continuity
   around the circle at member 15-16.  Except for member 15, however, the
   veracity of the following member is unknown. (Rule 1)

3. If the following member is a truth-teller then the current member must be
   a liar.  (Rule 1) Except for member 15 the next member is not known.
   In the case of member 15, however, the following member is known to be a
   liar so this rule can never be true.  I confirmed this by putting
   in a diagnostic print/break that did not trigger in 200000 trials
   so I am reasonably confident in my conclusion.

4. If the previous two members are liars then the current member must
   be truth-teller. (Rule 2)

5. If no rule applies, generate a random choice

If you do this *enough* times you should cover the solution space.
I don't know how to quantify, a priori, how many trials are required.
That is the problem with Monte Carlo sims - when do you have enough samples?


## Probabilities of the Monte Carlo Sim

I think we should be able to determine the odds of finding the 8 liar
solution (since it is unique) starting as follows:

   L-T-L
   
The next element is T (50%) which then defines the next element as L.

  L-T-L-T-L
  
Each time we add T-L to the list is a 50% probability.  To get to the full
length sequence we do this 6 times:

  L-T-L-T-L-T-L-T-L-T-L-T-L-T-L-?
  
The final element is forced to be T by condition 2 in the code above.  In
this case then, the odds of finding this solution must be `r 1/2^6`.

So, the odds of not finding this in N trials is:

$$P_{miss} = \left( 1-\frac{1}{2^6} \right)^N$$

The odds of not finding/finding the 8 liar solution can be tabulated as:

```{r}
num_trials <- c(10,20,30,40,44,50,100,1000,10000)
p_miss <- unlist(lapply(num_trials,function(N) {(1-2^-6)^N}))
p_hit <- 1-p_miss
cbind(num_trials=num_trials,p_miss=p_miss,p_hit=p_hit)

```

So with 10 trials we have 85% chance of missing the 8 liar case.  With 100
trials we have 21% chance of a miss.  With 1000 trials we have less than
a 1 in a million chance of missing.  For 10K trials the probability is
vanishingly small.

Calculating the odds for 9 or 10 liars is much more complicated since there
are many combinations that can yield these cases and each path will have
a slightly different probability.  Here are two difference sequences that
each have 10 liars.  Each 50% probability is marked with a caret (^).

    L-T-L-L-T-L-L-T-L-L-T-L-L-T-L-T
          ^     ^     ^     ^        
          +-----+-----+-----+--------- Four random chances
    
    L-T-L-T-L-T-L-L-T-L-L-T-L-L-T-L
          ^   ^   ^     ^     ^      
          +---+---+-----+-----+------- Five random chances

So the probabilistic simulation is twice as likely to generate the first
sequence as the second.

The previous simulations generate feasible solutions of length 9 and 10
about equally with a much smaller number of 8 liar solutions.

The exhaustive enumeration (see below) has a total of 37 feasible
solutions with the following distribution:

    10 liars: 15 cases
    9  liars: 21 cases
    8  liars:  1 case
    
Since the probabilities of randomly generating a particular solution
depends on the number of random choices it is difficult to directly compare
this distribution with that obtained in the Monte Carlo sim.

One way to calculate the actual probabilities is the following:

1. Start with: seq=L-T-L, prob=1.0

2. Generate the next item according to the rules above.  In the event
   of a random choice - bifurcate the solution and create both new
   sequences with the new prob = prob/2.
   
3. Repeat step 2 until a sequence reaches 15 elements.  The final element
   is deterministic so it will not change the probability.
   
The coding may be recursive or it might be a repeating loop.  In any event,
the solution is reached when all sequences reach the full length of 16
members.

```{r}
# RGC Solution: enumeration of feasible solutions

last <- function(x) {
  return(tail(x,1))
}
second_last <- function(x) {
  return(head(tail(x,2),1))
}

starting_list <- list(s=c(0,1,0),p=1.0)

# Mast list of all feasible solutions and the probability
ll <- list(starting_list)

done <- FALSE
while (!done) {
  # Assume we are done and then go through the main list
  done <- TRUE
  idx <- 1
  while (idx <= length(ll)) {
    # Examine the sequence and figure out what to do
    s <- ll[[idx]]$s
    if (length(s)>15) {
      # Sequence is full length - skip it
      idx <- idx+1
    } else if (length(s)==15) {
      # Finish this one off
      ll[[idx]]$s <- append(s,1-last(s))
      idx <- idx+1
    } else {
      done <- FALSE
      if (second_last(s)==0) {
        # Deterministic next element so add it
        ll[[idx]]$s <- append(s,1-last(s))
        idx <- idx+1
      } else {
        # Clone the current vector and drop p by factor of 2
        # Since these each need more processing do not increment
        # the index
        new_p <- ll[[idx]]$p/2
        s1 <- append(s,0)
        s2 <- append(s,1)
        # Modify the current solution
        ll[[idx]] <- list(s=s1,p=new_p)
        # Add new candidate at end of the list
        m <- length(ll)
        ll[[m+1]] <- list(s=s2,p=new_p)
        # print(c('Total candidates = ',length(ll)))
      }
    }
  }
}

# Now total up the probabilities
p_total <- rep(0,16)
n_cases <- rep(0,16)
for (i in ll) {
  m <- sum(i$s == 0)
  n_cases[m] <- n_cases[m]+1
  p_total[m] <- p_total[m]+i$p
}
cbind(num_liars=1:16,p_total=p_total,n_cases=n_cases)
```

Algebraically the probabilities are:

    8 liars  ->  1/64
    9 liars  -> 31/64
    10 liars -> 32/64

This matches well with the Monte Carlo simulations for a large number
of trials above.


## Exhaustive Search for Problem #1

I would prefer an enumeration of the possible cases or an exhaustive search.

The code above enumerates all possible cases.

The code below performs exhaustive search.  It starts with positions 1,2,
and 16 defined as the original author does and then enumerates all possible
combinations of the remaining 13 positions.  This is only 2^13 cases and
seems like a reasonable thing to try.

In the end - I agree with the solution presented by the author but I
still don't like probabilistic solutions to problems with deterministic
answers.  I grudgingly admit, however, that I am impressed with the
ability of this code to cover the solution space - given enough random
trials.

```{r}
# RGC Solution: exhaustive search

truth_teller <- 1
liar <- 0

to_binary <- function(x,nbits=13) {
  n=rep(0,nbits)
  for (i in 1:nbits) {
    n[i] <- x %% 2
    x <- x %/% 2
  }
  return(n)
}

valid_solution <- function(x) {
  valid <- TRUE
  for (i in 3:(length(x)-1)) {
    if (x[i]==truth_teller) {
      if ( (x[i-1]==truth_teller) | (x[i+1]==truth_teller) ) {
        # Bad solution - truth_teller lied
        valid <- FALSE
        break
      }
    } else {
      if ( (x[i-1]==liar) & (x[i+1]==liar)) {
        # Bad solution - liar told truth
        valid <- FALSE
        break
      }
    }
  }
  return(valid)
}

tabz=rep(0,16)
tabz[1]=truth_teller #at least one
tabz[2]=tabz[16]=liar
solution=c()
for (i in 0:8191) {
  tabz[3:15] <- to_binary(i)
  if (valid_solution(tabz)) {
    solution <- c(solution,tabz)
  }
}
# Each row is a valid solution
solution <- t(matrix(solution,nrow=16))
nnn <- c()
for (i in 1:nrow(solution)) {
  nnn <- c(nnn,sum(solution[i,]==0))
}
print('Number of liars:')
sort(unique(nnn))
```

So - exhaustive search yields all three solutions.  In addition, the various
solutions can be enumerated:

```{r}
for (i in sort(unique(nnn))) {
  print(c('Number of liars =',i))
  print(c("Num cases =",sum(nnn==i)))
  for (m in which(nnn==i)) {
    print(paste0(solution[m,],collapse=""))
  }
  print("")
}
```


# Problem #2

Repeating the problem:

    2. A circle of 12 liars and truth-tellers is such that everyone state that
       their immediate neighbours are one liar plus one truth-teller. How many
       liars can there be?

Per the original author:

    The second puzzle is associated with the similar R code

```{r}
tabz=sample(c(-1,1),12,rep=TRUE)
rong=FALSE
while (!rong){
 for (i in sample(12)){
  if (tabz[i-1+12*(i==1)]*tabz[i%%12+1]==-1){
   tabz[i]=1}else{ 
   tabz[i]=sample(c(-1,1),1)}
  }
  rong=TRUE
  for (i in (1:12)[tabz==1])
    rong=rong&(tabz[i-1+12*(i==1)]*tabz[i%%12+1]==-1)
  if (rong){
   for (i in (1:12)[tabz==-1])
     rong=rong&(tabz[i-1+12*(i==1)]*tabz[i%%12+1]!=-1)
  }}
sum(tabz==-1)
```

With the conclusion:

    with numbers of liars (-1) either 12 (obvious) or 4.

I have executed the code and it seems that it *always* yields 4 liars as the
solution.  I have wrapped the code in a loop below to see if more trials
finds both solutions.  I have changed the format and added comments but
it should be identical to the code above:

```{r}
num_trials <- 100
num_liars <- rep(0,12)

for (i in 1:num_trials) {
  # Random candidate solution
  tabz=sample(c(-1,1),12,rep=TRUE)
  rong=FALSE
  # Keep adjusting the solution until a feasible solution is found
  while (!rong){
    # Take the 12 members in a random order
    for (i in sample(12)) {
      # If the two neighbors are opposite veracity
      if (tabz[i-1+12*(i==1)]*tabz[i%%12+1]==-1) {
        # then this member must be a truth-teller
        tabz[i]=1
      } else {
        # I don't understand the original code.
        #
        # The two neighbors are the same veracity so this
        # must be a liar -- for some unknown reason the
        # original author uses random veracity.  Is this
        # to break out of an infinite loop?  I changed this
        # to force the current member to be a liar and it
        # generates the all liar solution much more often
        # and requires far fewer trials.
        #
        # With random logic required 10000 trials to find
        # the 12 liars solution only 3 times.
        #
        # With correction in place the 12 liars solution is
        # found about 1/4 of the time which matches the expected
        # result from exhaustive enumeration.
        #
        # Original code:
        # tabz[i]=sample(c(-1,1),1)
        #
        # Correct code:
        tabz[i]=-1
      }
    }
    # Check validity of the candidate solution
    # Assume it is correct
    rong=TRUE
    # For every truth-teller verify the neighbors are opposites
    # Note that (1:12)[tabz==1] is the same as which(tabz==1)
    for (i in (1:12)[tabz==1])
      rong=rong&(tabz[i-1+12*(i==1)]*tabz[i%%12+1]==-1)
    # If still a valid seqence, check the liars
    # You could always check the liars at a slight performance penalty
    # The logic would still work.
    if (rong) {
        # For every liar verify the neighbors are the same veracity
        for (i in (1:12)[tabz==-1])
          rong=rong&(tabz[i-1+12*(i==1)]*tabz[i%%12+1]!=-1)
    }
    # If we get here and rong is still TRUE then we are done
    #
    # If rong is FALSE then we repeat the process taking a new random
    # permutation of the 12 members and *tweaking* the veracity.
    #
    # Since the sequencing is done with random permutations it is
    # difficult to know a priori how long this will execute.
  }
  m=sum(tabz==-1)
  num_liars[m] <- num_liars[m]+1
}
cbind(1:12,num_liars)
```

Well -- if I run enough trials I find both solutions.  Running this code
for 10000 trials takes a while so I really don't like the original code
approach to using random veracity.  I don't know the theoretical distribution
of the results but this does not give me much confidence that all the
possible solutions have been found.

With the correction in place the number of cases required to find both
solutions has been reduced by 2-3 orders of magnitude.

Exhaustive search is only 2^12 = 4096 cases and should run much faster.
Let us try that:

```{r}
# RGC Exhaustive search
# Liar = 0
# Truth-teller = 1

# to_binary() defined above

# Return element before x[i] with wrap-around
# Coded as tabz[i-1+12*(i==1)] by original author
pred <- function(x,i) {
  if (i==1) {
    return(tail(x,1))
  }
  return(x[i-1])
}

# Return element after x[i] with wrap-around
# Coded as tabz[i%%12+1] by original author
succ <- function(x,i) {
  if (i==length(x)) {
    return(x[1])
  }
  return(x[i+1])
}

# Check for valid solution
valid_solution <- function(x) {
  for (i in 1:12) {
    if (x[i]==1) {
      if (pred(x,i)==succ(x,i)) { # Truth-teller lied
        return(FALSE)
      }
    } else {
      if (pred(x,i)!=succ(x,i)) { # Liar told the truth
        return(FALSE)
      }
    }
  }
  return(TRUE)
}

# Exhaustive search
num_liars=rep(0,12)
for (i in 1:2^12) {
  s <- to_binary(i,12)
  if (valid_solution(s)) {
    m=sum(s==0)
    num_liars[m] <- num_liars[m]+1
    print(s)
  }
}
cbind(1:12,num_liars)
```

Only 4 of the 8192 possible combinations are valid.

Note that the 4 liar solutions are just circular shifts of the same pattern.

This can be derived if we consider the center member of a three member group.  

The are four infeasible patterns involving three members:

* 0-0-1: liar told the truth
* 1-0-0: liar told the truth
* 0-1-0: truth-teller lied
* 1-1-1: truth-teller lied

This leaves four feasible patterns involving three members:

* 0-0-0
* 1-0-1
* 0-1-1
* 1-1-0

If we look at this, we can see that the third member is defined by the
first two members.  In fact, the third member is always the exclusive-OR
of the previous two members.

This means we can extend any group of two members to three members.
If we then look at the last two members we can extend to four.
This may be continued until the entire sequence is completed and we can
assess the feasibility by continuity with the first member of the circle.

Applying the feasible patterns above to the possible starting cases yields:

* 0-0-0-0-0-0-0-0-0-0-0-0-wrap around to start
* 1-0-1-1-0-1-1-0-1-1-0-1-wrap around to start
* 0-1-1-0-1-1-0-1-1-0-1-1-wrap around to start
* 1-1-0-1-1-0-1-1-0-1-1-0-wrap around to start

Which generates the four patterns found above.

----

# Problem #3

Repeating the problem:

    3. A circle of 8 liars and truth-tellers is such that four state that
       their immediate neighbours are one liar plus one truth-teller and
       four state that their immediate neighbours are both liars . How 
       many liars can there be?

Per the original author:

    The final puzzle is more puzzling in that figuring out the validating
    function (is an allocation correct?) took me a while, the ride back
    home plus some. I ended up with the following code that samples liars
    (-1) and thruth-seekers (1) at random, plus forces wrong and right
    answers (in 0,1,2) on these, and check for the number of answers
    of both types:

```{r}
rong=FALSE
while (!rong){
 tabz=sample(c(-1,1),8,rep=TRUE) #truth
 tabz[1]=1;tabz[sample(2:8,1)]=-1
 tt=(1:8)[tabz==1];lr=(1:8)[tabz==-1]
 statz=rep(0,8) #stmt
 statz[tt]=(tabz[tt-1+8*(tt==1)]*tabz[tt%%8+1]==-1)+
           2*(tabz[tt-1+8*(tt==1)]+tabz[tt%%8+1]==-2)
 #answering 0 never works
 statz[lr]=2*(tabz[lr-1+8*(lr==1)]*tabz[lr%%8+1]==-1)+
          (tabz[lr-1+8*(lr==1)]+tabz[lr%%8+1]==-1)+
           sample(c(1,2),8,rep=TRUE)[lr]*
           (tabz[lr-1+8*(lr==1)]+tabz[lr%%8+1]==1)
 rong=(sum(statz==1)==4)&(sum(statz==2)==4)}

c(sum(tabz==-1),tabz)
```

Conclusion:

    with solutions 3, 4, 5 and 6.

My comment: **Huh?**  I have tried running this manually and I always get
the same solution with 4 liars.  I will try that code in a loop and see
if we get anything with more trials.

```{r}

# Original author solution wrapped in a function
# I think the solution is supposed to be in tabz (?)
#
# lr    = index values of liars
# tt    = index values of truth-tellers
# statz = replies from members of tabz
# For truth-tellers:
#    1 = L-T or T-L as neighbors
#    2 = L-L as neighbors
#    0 = T-T as neighbors (Invalid solution)
# For liars:
#    2 = L-T or T-L as neighbors (Liar lies)
#    1 = L-L as neighbors (Liar lies)
#    Random = T-T as neighbors (Liar makes up answer)
# tabz  = liar/truth-teller solution

gen_solution <- function() {
rong=FALSE
while (!rong){
 tabz=sample(c(-1,1),8,rep=TRUE) #truth
 tabz[1]=1;tabz[sample(2:8,1)]=-1
 tt=(1:8)[tabz==1];lr=(1:8)[tabz==-1]
 statz=rep(0,8) #stmt
 statz[tt]=(tabz[tt-1+8*(tt==1)]*tabz[tt%%8+1]==-1)+
           2*(tabz[tt-1+8*(tt==1)]+tabz[tt%%8+1]==-2)
 #answering 0 never works
 # Original code doesn't work
 # The sum of two tabz members can never equal -1 or +1
 # so the random replies never gets activated. I think
 # he meant for the sums to -2 or +2 to find neighbors
 # that are both Liars (-2) or truth-tellers (+2).
 if (FALSE) {
 statz[lr]=2*(tabz[lr-1+8*(lr==1)]*tabz[lr%%8+1]==-1)+
             (tabz[lr-1+8*(lr==1)]+tabz[lr%%8+1]==-1)+  # RGC: never true
             sample(c(1,2),8,rep=TRUE)[lr]*
             (tabz[lr-1+8*(lr==1)]+tabz[lr%%8+1]==1)    # RGC: never true
 } else {
 statz[lr]=2*(tabz[lr-1+8*(lr==1)]*tabz[lr%%8+1]==-1)+  # RGC: Liar lies
             (tabz[lr-1+8*(lr==1)]+tabz[lr%%8+1]==-2)+  # RGC: Liar lies
             sample(c(1,2),8,rep=TRUE)[lr]*
             (tabz[lr-1+8*(lr==1)]+tabz[lr%%8+1]==2)    # RGC: Liar random
 }
 rong=(sum(statz==1)==4)&(sum(statz==2)==4)}
 return(tabz)
}

num_liars <- rep(0,8)
num_trials <- 100
l <- list()
for (i in 1:num_trials) {
  s <- gen_solution()
  m <- sum(s==-1)
  num_liars[m] <- num_liars[m]+1
  found <- FALSE
  if (length(l)==0) {
    l[[1]] <- s
  }
  for (j in 1:length(l)) {
    if (all(l[[j]]==s)) {
      found <- TRUE
    }
  }
  if (!found) {
    l[[length(l)+1]] <- s
  }
}

print("Feasible number of liars:")
print(which(num_liars!=0))
print("Unique solutions saved in list = l")
print("")
print("Simulated solutions:")
cbind(num_liars=1:8,num_solutions=num_liars)

```

OK - the original code always generates one of the two same sequences.

I fixed the problem and it works now.

----

Let us try exhaustive search on 2^8 cases:

```{r}
# to_binary() from above
# pred() from above
# succ() from above

valid_solution <- function(x) {
  n1 <- 0   # Answer: 1 Liar, 1 Truth-teller
  n2 <- 0   # Answer: 2 Liars
  n3 <- 0   # Answer: Unknown - Liar with TT,TL, or LT neighbors
  n4 <- 0   # Answer: 2 Truth-tellers -> invalid sequence
  for (i in 1:8) {
    if (x[i]==1) {
      if (pred(x,i) != succ(x,i)) {
        n1 <- n1+1   # Truther with L-T/T-L neighbors
      } else if ((pred(x,i)==0)&(succ(x,i)==0)) {
        n2 <- n2+1   # Truther with L-L neighbors
      } else {
        n4 <- n4+1   # Truther with T-T neighbors
                     # Invalid sequence - could force return here
      }
    } else {
      if ((pred(x,i)==0)&(succ(x,i)==0)) {
        n1 <- n1+1   # Liar with LL neighbors
      } else {
        n3 <- n3+1   # Liar can lie
      }
    }
  }
  return( list(n1=n1,n2=n2,n3=n3,n4=n4) )
}

num_liars <- rep(0,8)
for (i in 0:255) {
  s <- to_binary(i,8)
  # Check if sequence is valid
  #    - No answers of T-T, and
  #    - Known L-T/T-L no more than 4, and
  #    - Known L-L no more than 4
  # Remainder will be liars telling lies and they can answer
  # however they need to make the sequence *valid*
  l <- valid_solution(s)
  if ((l$n4==0) & (l$n1<=4) & (l$n2<=4)) {
    m=sum(s==0)
    # print(c(m,":",s))
    num_liars[m] <- num_liars[m]+1
  }
}
cbind(num_liars=1:8,solutions=num_liars)
```

First try exhaustive search only finds solutions of 3, 4, and 5 liars.
Did I miss something?

Yes -- liar surrounded by anything other than L-L can reply
however they want to make the *correct* total.  Modified the logic and I
now get solutions of 3,4,5, or 6 liars.

The original author claims there are solutions of 3, 4, 5, and 6 liars.

He appears to be correct.

Need to think about this.

Answers are always:

1. One Liar + One Truth-teller
2. Both Liars

Let's consider all three member combinations.  What is the reply of the
member in the middle of the sequence?  The reply is either #1 or #2 or a lie
or an invalid response (two truth-tellers).

    Combo  Case
    ---------------
    L-L-L  n1
    L-L-T  n2
    L-T-L  n2
    L-T-T  n1
    T-L-L  n2
    T-L-T  Either
    T-T-L  n1
    T-T-T  Invalid
    
Seven of the cases are deterministic.  Only T-L-T allows the liar to lie
in a way we can't predict.

Not sure this is going anywhere.

How about trying to generate the prime implicants for all the valid
sequences.  I had an old program called pigen.c that I might be able to
find.  If not - I wil have to recreate it in R (that does not sound like
a lot of fun).  Whew!  Found it in my old cd repo.  Compilation and a
simple test case seem to work properly.  Now I need to figure out how
to write output from R!

```{r}
# Generate ASCII output file with the valid sequences.
# Each line of the output file should consist of precisely
# 8 digits (1 or 0) that represent the minterms.

# Using the logic from above:
# valid_sequence() needs to be checked and the sequences put into
# a matrix.

sss=c()
for (i in 0:255) {
  s <- to_binary(i,8)
  # Check if sequence is valid
  #    - No answers of T-T, and
  #    - Known L-T/T-L no more than 4, and
  #    - Known L-L no more than 4
  # Remainder will be liars telling lies and they can answer
  # however they need to make the sequence *valid*
  l <- valid_solution(s)
  if ((l$n4==0) & (l$n1<=4) & (l$n2<=4)) {
    sss=c(sss,s)
  }
}
sss=matrix(sss,ncol=8,byrow=TRUE)
for (i in 1:nrow(sss)) {
  write(paste0(sss[i,],collapse=""),"problem3.txt",append=TRUE)
}

```

The output of this works fine to feed into pigen.  Unfortunately, the
results are somewhat disappointing.  Given the 106 different patterns of
valid sequences the pigen process only reduces this to 32.  The prime
implicants are:

    101-0--0
    1010--0-
    1-010--0
    10-10--0
    -1010--0
    0101-0--
    01-010--
    010-10--
    -0101-0-
    0-1010--
    10--01-0
    10--010-
    -01-010-
    -010-10-
    --0101-0
    -0-1010-
    1-0--010
    10--0-10
    -10--010
    010--01-
    --01-010
    --010-10
    --0-1010
    0--0101-
    01-0--01
    010--0-1
    -010--01
    0-10--01
    0--01-01
    0--010-1
    -0--0101
    0--0-101

I suspect there is a XOR pattern buried in here but I'm not sure.

Maybe I got greedy -- let us try each class of solutions.

Wow - that is even more disappointing than the previous result.  There
is **no** simplification possible which is blindingly obvious after a bit
of thought.  If two sequences have the same number of liars they must
differ in two positions - hence they can never be combined.

Looking at the cases of 3 liars, 4 liars, etc. individually shows that there
are a few basic patterns that occur multiple times as circular shifts.
Let's see what this yields:

```{r}
# Check if two vectors are just circular shifts
circ_shift <- function(x1,xcompare) {
  x2 <- xcompare
  # Return TRUE if x2 is a circular shift of x1
  m <- length(x2)
  if (length(x1) != m) return(FALSE)
  for (i in 1:m) {
    if (identical(x1,x2)) {
      return(TRUE)
    } else {
      temp <- x2[1]
      x2[1:(m-1)] <- x2[2:m]
      x2[m] <- temp
    }
  }
  return(FALSE)
}
unique_set <- function(x) {
  # Return the unique rows from matrix x
  # A row is considred unique if it is not a circular shift of any
  # other row.
  zzz <- x
  nc <- ncol(x)
  output=c()
  while (nrow(zzz)>0) {
    t <- zzz[1,]
    m <- c()
    for (i in 1:nrow(zzz)) {
      m <- c(m,circ_shift(t,zzz[i,]))
    }
    output=c(output,zzz[which(m)[1],])
    zzz <- zzz[!m,]
  }
  output <- matrix(output,ncol=nc,byrow=TRUE)
  return(output)
}

zzz <- unique_set(sss)
zzz
```

So - the original set of 106 solutions has been reduced to 15 after removing
solutions that are simply circular shifts of other solutions.

The unique solutions are given below:

```{r}
for (num_liar in c(3,4,5,6)) {
  m <- apply(zzz,1,function(x) {sum(x==0)==num_liar})
  print(c("Num liars = ",num_liar))
  print(zzz[m,])
  print("")
}
```

You could consider these as *basis solutions*.

