---
title: "Pick Some Boxes"
author: "R. G. Cronce"
date: "10/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Le Monde 1130: Pick some boxes

[R-Bloggers Website](https://www.r-bloggers.com/2020/02/le-monde-puzzle-1130)

[Xi'an blog Website](https://xianblog.wordpress.com/2020/02/07/le-monde-puzzle-1130/)


    Abishag and Caleb fill in alternance a row of N boxes in a row by picking
    one then two then three etc. consecutive boxes. When a player is unable
    to find enough consecutive boxes, the player has lost.
    
    Who is winning when N=29? When N=30?

Solution (per Xi'an):

    returns Abishag as the winner for N=29 (as well as for N=1,2,7,…,13,19,…,29)
    and Caleb as the winner for N=30 (as well as for N=3,…,6,14,…,18). I am actually
    surprised that the recursion operates that deep, even though this means a sqrt(N)
    depth for the recursion. While the code took too long to complete, the function
    operates for N=100. A side phenomenon is the apparent special case of N=47,
    which makes Abishag the looser, while N=46 and N=48 do not.
    
    This is an unusual pattern as otherwise (up to N=59), there are longer and longer
    stretches of adjacent wins and looses as N increases.

There is a recursive function provided by Xi'an to solve this problem.  The
reuturn value is (apparently) TRUE for Abishag as the winner.

When I run this code I do not get the same results as Xi'an.  The code does not run
with N=1.  I do not see the anomaly for N=47.  I agree that the code is very slow
once N is more than (about) 30.

```{r}

f<-function(b=!1:N,r=0){
  for(i in 1:(N-r)){
    if(p<-!max(b[j<-i+r:0])){
      q=b;q[j]=1
      if(p<-!f(q,r+1))break}}
  p}

```

using this to tabulate some results:

```{r}
t1=Sys.time()
abishag_wins=c()
for (i in 2:50) {
  N=i
  r=f()
  abishag_wins=append(abishag_wins,r)
  if (r) { print(c(N,"Abishag"))} else { print(c(N,"Caleb")) }
}
t2=Sys.time()

print("Abishag wins:")
print(which(abishag_wins)+1)

print("Caleb wins:")
print(which(!abishag_wins)+1)

print("Run lengths:")
print(rle(abishag_wins))

print(c("Run time",t2-t1))
```

Note that the lengths of alternating runs appear to be intermixed sequences of primes
with run lengths:

    for Ashibag: 2, 7, 11, 13
    for Caleb:   3, 5, 7,  ???
    
With the final sequence for Caleb terminated by the upper limit of this simulation.
Even this modest upper limit of 50 takes a substantial amount of run time.