---
title: 'Le Monde 1124: Fun with Primes'
author: "R. G. Cronce"
date: "10/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Le Monde 1124: Fun with Primes


Wow - I don't even pretend to understand this one.


Courtesy of Xi'an via R-Bloggers:

[R-Blogger's Website](https://www.r-bloggers.com/2019/12/le-monde-puzzle-1124)

[Xi'an Blog Website](https://xianblog.wordpress.com/2019/12/29/le-monde-puzzle-1124/)

The puzzle:

    A prime number challenge [or rather two!] as Le weekly Monde current mathematical puzzle:

    When considering the first two integers, 1 and 2, their sum is 3, a prime number.
    For the first four integers, 1,2,3,4, it is again possible to sum them pairwise to
    obtain two prime numbers, eg 3 and 7. Up to which limit is this operation feasible?
    And how many primes below 30,000 can write as n^p+p^n?


Here is his one-line solution:

```{r}
library(primes)

n=10

print(c(n,max(apply(apply(b<-replicate(1e6,(1:n)+sample(n)),2,is_prime)[,b[1,]>2],2,prod))))

n=15

print(c(n,max(apply(apply(b<-replicate(1e6,(1:n)+sample(n)),2,is_prime)[,b[1,]>2],2,prod))))


```

This code is amazingly incomprehensible.  I don't know if it works.

I get inconsistent results for n=14.


I'm not sure I even understand Xi'an's conclusion:

    provides a solution for the first question until n=14 when it stops.
    A direct explanation is that the number of prime numbers grows too slowly
    for all sums to be prime. And the second question gets solved by direct
    enumeration using again the is_prime R function from the primes package:

    [1] 1 1
    [1] 1 2
    [1] 1 4
    [1] 2 3
    [1] 3 4

But he does not provide any code or other explanation.
